# Regional & Geographic Deep-Dives
*Country profiles and detailed breakdowns*

```{r setup}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)

has_reactable <- requireNamespace("reactable", quietly = TRUE)
if (has_reactable) {
  library(reactable)
  .or_filter <- JS("function(rows, columnId, filterValue) {
    const terms = filterValue.split('|').map(t => t.trim().toLowerCase()).filter(t => t.length > 0);
    if (terms.length === 0) return rows;
    return rows.filter(row => {
      const val = String(row.values[columnId] || '').toLowerCase();
      return terms.some(t => val.includes(t));
    });
  }")
}

# Top-N country plot helper (used in nested service tabs)
make_country_top_plot <- function(df, svc, n = 20) {
  d <- df %>%
    filter(service == svc) %>%
    slice_max(relative_intensity, n = n) %>%
    arrange(relative_intensity) %>%
    mutate(group = factor(group, levels = group))
  ggplot(d, aes(x = relative_intensity, y = group, fill = relative_intensity > 1)) +
    geom_col() +
    geom_vline(xintercept = 1, linetype = "dashed", color = "black", linewidth = 0.5) +
    scale_fill_manual(values = c("TRUE" = "#E83737", "FALSE" = "gray70"), guide = "none") +
    labs(x = "Relative Intensity", y = NULL) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
}

source("../../../R/paths.R")
source("../../../R/plotting_functions.R")

area_stats <- read_csv(
  file.path(data_dir(), "processed", "tables", "hotspot_area_stats.csv"),
  show_col_types = FALSE
)

multiservice_stats <- read_csv(
  file.path(data_dir(), "processed", "tables", "hotspot_multiservice_stats.csv"),
  show_col_types = FALSE
)

# Country subset: exclude micro-states (<100 grid cells ≈ <1,000 km²)
country_area_stats <- area_stats %>%
  filter(grouping_var == "nev_name", !is.na(group), as.numeric(n_total) >= 100)
```

## Hotspot Intensity by Geography

Relative intensity: each group's share of global hotspot cells divided by its share of global land area. Dashed line = equal distribution (1.0).

:::: {.panel-tabset}

### World Bank Region

```{r intensity-region}
#| fig-height: 10
#| fig-width: 12
make_intensity_plot(area_stats, "region_wb")
```

### Income Group

```{r intensity-income}
#| fig-height: 8
#| fig-width: 12
make_intensity_plot(area_stats, "income_grp")
```

### Biome

```{r intensity-biome}
#| fig-height: 12
#| fig-width: 12
make_intensity_plot(area_stats, "WWF_biome")
```

### Country

Top 20 countries per service. Select a service tab to explore.

::: {.panel-tabset}

#### Coastal Risk
```{r intensity-country-crisk}
#| fig-height: 6
#| fig-width: 9
make_country_top_plot(country_area_stats, "C_Risk")
```

#### Coastal Risk Reduction
```{r intensity-country-criskredratio}
#| fig-height: 6
#| fig-width: 9
make_country_top_plot(country_area_stats, "C_Risk_Red_Ratio")
```

#### Nitrogen Export
```{r intensity-country-nexport}
#| fig-height: 6
#| fig-width: 9
make_country_top_plot(country_area_stats, "N_export")
```

#### Nitrogen Retention
```{r intensity-country-nret}
#| fig-height: 6
#| fig-width: 9
make_country_top_plot(country_area_stats, "N_Ret_Ratio")
```

#### Nature Access
```{r intensity-country-nature}
#| fig-height: 6
#| fig-width: 9
make_country_top_plot(country_area_stats, "Nature_Access")
```

#### Pollination
```{r intensity-country-poll}
#| fig-height: 6
#| fig-width: 9
make_country_top_plot(country_area_stats, "Pollination")
```

#### Sediment Export
```{r intensity-country-sedexp}
#| fig-height: 6
#| fig-width: 9
make_country_top_plot(country_area_stats, "Sed_export")
```

#### Sediment Retention
```{r intensity-country-sedret}
#| fig-height: 6
#| fig-width: 9
make_country_top_plot(country_area_stats, "Sed_Ret_Ratio")
```

:::

::::

## Compound Risk by Geography

Average number of overlapping service hotspots per 10 km grid cell.

:::: {.panel-tabset}

### World Bank Region

```{r compound-region}
#| fig-height: 6
#| fig-width: 10
make_multi_plot(multiservice_stats, "region_wb")
```

### Income Group

```{r compound-income}
#| fig-height: 6
#| fig-width: 10
make_multi_plot(multiservice_stats, "income_grp")
```

### Biome

```{r compound-biome}
#| fig-height: 10
#| fig-width: 10
make_multi_plot(multiservice_stats, "WWF_biome")
```

### Country

```{r compound-country}
#| fig-height: 10
#| fig-width: 9
#| fig-cap: "Top 40 countries by average overlapping hotspot services per cell."
multiservice_stats %>%
  filter(grouping == "nev_name", !is.na(group)) %>%
  slice_max(mean_hotspots, n = 40) %>%
  arrange(mean_hotspots) %>%
  mutate(group = factor(group, levels = group)) %>%
  ggplot(aes(x = mean_hotspots, y = group, fill = mean_hotspots)) +
  geom_col() +
  scale_fill_distiller(palette = "Reds", direction = 1, guide = "none") +
  labs(x = "Mean overlapping hotspot services per cell", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

::::

## Interactive Country Table

```{r country-table}
#| echo: false
country_data <- country_area_stats %>%
  select(service, group, n_hot, pct_area, relative_intensity) %>%
  arrange(desc(relative_intensity)) %>%
  rename(
    Service             = service,
    Country             = group,
    "Hotspot Cells"     = n_hot,
    "% of Country Area" = pct_area,
    "Relative Intensity" = relative_intensity
  ) %>%
  mutate(
    "% of Country Area"  = round(`% of Country Area`, 2),
    "Relative Intensity" = round(`Relative Intensity`, 2)
  )

if (has_reactable) {
  reactable(
    country_data,
    filterable  = TRUE,
    searchable  = TRUE,
    pagination  = TRUE,
    defaultPageSize = 25,
    striped     = TRUE,
    highlight   = TRUE,
    compact     = TRUE,
    columns = list(
      Service             = colDef(minWidth = 140, filterMethod = .or_filter),
      Country             = colDef(minWidth = 160, filterMethod = .or_filter),
      `Hotspot Cells`     = colDef(format = colFormat(separators = TRUE), minWidth = 120),
      `% of Country Area` = colDef(format = colFormat(suffix = "%", digits = 2), minWidth = 140),
      `Relative Intensity` = colDef(format = colFormat(digits = 2), minWidth = 140)
    )
  )
} else {
  knitr::kable(head(country_data, 20), caption = "Country-Level Hotspot Statistics — install `reactable` for interactive filtering")
}
```

## Regional Hotspot Profiles

Source: `hotspot_area_stats.csv`, grouping_var == "region_wb", 10 km² grid. Per-region population exposure in Chapter 6.

::: {.callout-note}
**How to read "relative intensity" below:** it is a region's (or biome's, or income group's) share of global hotspot cells for a service, divided by its share of global land area. **1.0** means its hotspot burden exactly matches its size; **1.5×** means it holds 50% more of that service's hotspots than its land area would predict; **0.7×** means 30% less. It is a measure of disproportionate burden, not an absolute count — a small region can have a very high relative intensity while contributing few hotspot cells in absolute terms, and vice versa.
:::

### Sub-Saharan Africa
- **Top services by relative intensity**: Nature Access (1.17×), Sediment Retention (1.10×), Pollination (1.09×)
- **Geographic footprint**: ~14,700 Nature Access cells; ~13,700 Pollination and Sediment Retention cells (out of 241,000 regional cells)
- **Regional burden**: All three top services exceed the 1.0 equal-distribution benchmark — the region hosts a disproportionate share of global hotspot area across terrestrial and freshwater services

### Latin America & Caribbean
- **Top services by relative intensity**: Sediment Export (2.51×), Pollination (1.91×), Nitrogen Export (1.65×), Sediment Retention (1.61×)
- **Geographic footprint**: ~26,600 Sediment Export cells; ~20,300 Pollination cells (out of 204,000 regional cells)
- **Regional burden**: Sediment Export at 2.51× is the highest region-service relative intensity in the entire dataset; all four top services substantially exceed the 1.0 benchmark; coastal risk is well below expected levels (0.51×)

### East Asia & Pacific
- **Top services by relative intensity**: Nitrogen Retention (1.80×), Nitrogen Export (1.57×), Sediment Retention (1.56×)
- **Geographic footprint**: ~23,000 Nitrogen Retention cells; ~20,000 Nitrogen Export cells (out of 246,000 regional cells)
- **Regional burden**: Water quality services (nitrogen and sediment) are strongly disproportionate; coastal risk is at approximately expected levels (0.95×)

### South Asia
- **Top services by relative intensity**: Nitrogen Retention (1.37×), Nature Access (1.08×), Nitrogen Export (1.08×)
- **Geographic footprint**: ~3,600 Nitrogen Retention cells; ~2,800 Nature Access cells (out of 51,000 regional cells)
- **Regional burden**: Moderate relative intensities but the region's very high population density means geographic footprint substantially understates human exposure

### Middle East & North Africa
- **Top services by relative intensity**: Nature Access (1.31×), Nitrogen Retention (1.13×)
- **Geographic footprint**: ~5,800 Nature Access cells; ~5,000 Nitrogen Retention cells (out of 86,000 regional cells)
- **Regional burden**: Nature Access (1.31×) and Nitrogen Retention (1.13×) both exceed the benchmark; remaining services at or below expected levels

### Europe & Central Asia
- **Top services by relative intensity**: All services below 1.0; highest is Coastal Risk Reduction (0.88×)
- **Geographic footprint**: ~9,300 Nature Access cells; ~8,700 Pollination cells (out of 280,000 regional cells)
- **Regional burden**: No service exceeds its expected share

## Summary

::: {.callout-note}
**Key findings from regional analysis:**

1. **Disproportionate burden in low- and middle-income regions.** Sub-Saharan Africa, Latin America & Caribbean, East Asia & Pacific, and South Asia all show relative intensities above 1.0 for their top services — they host more hotspot area than their land share would predict.

2. **Compound risk concentrates in tropical regions.** Countries with high average overlapping hotspot counts tend to be in tropical biomes.

3. **East Asia & Pacific burden is concentrated in freshwater services.** Nitrogen Retention hotspots run 80% above what the region's land area would predict, Nitrogen Export 57% above, and Sediment Retention 56% above (relative intensity 1.80×, 1.57×, 1.56×); coastal risk sits at approximately the expected level (0.95×).

4. **Europe & Central Asia is the clear outlier.** No service exceeds 1.0 — the region hosts fewer global hotspot cells than its land area would predict across all services.
:::
