# Regional & Geographic Deep-Dives
*Country profiles and detailed breakdowns*

```{r setup}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)

has_dt <- requireNamespace("DT", quietly = TRUE)
if(has_dt) library(DT)

source("../../../R/paths.R")
source("../../../R/plotting_functions.R")

# Load data
area_stats <- read_csv(
  file.path(data_dir(), "processed", "tables", "hotspot_area_stats.csv"),
  show_col_types = FALSE
)

pop_exposure <- read_csv(
  file.path(data_dir(), "processed", "tables", "hotspot_pop_exposure.csv"),
  show_col_types = FALSE
)
```

## Overview

This chapter provides detailed geographic deep-dives. You can explore patterns by:

- **World Bank Region** – Continental/regional hotspot patterns
- **Income Group** – Socioeconomic clustering of hotspots
- **Biome** – Ecosystem-specific service change patterns
- **Country** – Specific country-level priorities

Use the tabs below to drill into specific geographies of interest.

## Hotspot Intensity by Region

```{r intensity-region-detail}
#| fig-cap: "Detailed regional hotspot intensity. Explore which services are most affected in each region."
#| fig-height: 8
#| fig-width: 12

make_intensity_plot(area_stats, "region_wb")
```

## Country-Level Rankings

```{r country-rank}
#| fig-cap: "Top countries by hotspot relative intensity, faceted by service. Countries above the dashed line have disproportionately high hotspot concentration."
#| fig-height: 12
#| fig-width: 14

# Filter out micro-states (<10,000 sq km) to prevent skewed relative intensities
filtered_area_stats <- area_stats %>%
  filter(!(grepl("countr|nev_name|admin", grouping_var, ignore.case = TRUE) & as.numeric(n_total) < 100))

make_country_intensity_plot(filtered_area_stats)
```

## Interactive Country Profiles

Explore country-specific hotspot statistics. Filter by service, country, or intensity to find regions of interest:

```{r country-table}
#| fig-cap: "Interactive table of country-level hotspot statistics. Click column headers to sort; use the search box to filter."

country_data <- area_stats %>%
  filter(grouping_var == "country") %>%
  select(service, group, n_hot, pct_area, relative_intensity) %>%
  arrange(desc(relative_intensity)) %>%
  rename(
    Service = service,
    Country = group,
    "Hotspot Cells" = n_hot,
    "% of Country Area" = pct_area,
    "Relative Intensity" = relative_intensity
  ) %>%
  mutate(
    "% of Country Area" = round(`% of Country Area`, 2),
    "Relative Intensity" = round(`Relative Intensity`, 2)
  )

if(has_dt) {
  DT::datatable(
    country_data,
    filter = "top",
    options = list(
      pageLength = 30,
      columnDefs = list(
        list(targets = 2, render = JS("function(data) {return Math.round(data);}"))
      )
    ),
    rownames = FALSE,
    caption = "Country-Level Hotspot Statistics"
  )
} else {
  knitr::kable(country_data, caption = "Country-Level Hotspot Statistics")
}
```

## Compound Risk by Geography

```{r compound-by-region}
#| fig-cap: "Countries ranked by compound risk (average overlapping hotspots per cell). High values indicate areas where multiple services are failing simultaneously."
#| fig-height: 8
#| fig-width: 10

multiservice_stats <- read_csv(
  file.path(data_dir(), "processed", "tables", "hotspot_multiservice_stats.csv"),
  show_col_types = FALSE
)

if (nrow(multiservice_stats) > 0 && "grouping" %in% names(multiservice_stats)) {
  make_country_multi_plot(multiservice_stats)
} else {
  ggplot() + theme_void() + ggtitle("Multi-service data not available")
}
```

## Population Exposure by Country

Below is an interactive breakdown of population exposure to hotspots at the country level:

```{r pop-exposure-country}
#| fig-cap: "Population exposure to hotspots by country. Shows absolute numbers of people affected by ecosystem service decline."

if (nrow(pop_exposure) > 0) {
  # Identify the geographic column dynamically
  geo_col <- if("country" %in% names(pop_exposure)) "country" else if("group" %in% names(pop_exposure)) "group" else "service"
  
  pop_filtered <- pop_exposure
  if("grouping_var" %in% names(pop_exposure)) {
    pop_filtered <- pop_filtered %>% filter(grouping_var == "country")
  }
  
  pop_by_geo <- pop_filtered %>%
    filter(!is.na(.data[[geo_col]])) %>%
    group_by(.data[[geo_col]]) %>%
    summarise(total_exposed = sum(exposed_population, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(total_exposed)) %>%
    rename(
      Geography = !!geo_col,
      "Exposed Population" = total_exposed
    ) %>%
    mutate("Exposed Population (Millions)" = `Exposed Population` / 1e6)

  if(has_dt) {
    DT::datatable(
      pop_by_geo %>% select(Geography, "Exposed Population (Millions)"),
      rownames = FALSE,
      options = list(pageLength = 10),
      caption = "Total Population Exposed to Ecosystem Service Declines"
    )
  } else {
    knitr::kable(head(pop_by_geo %>% select(Geography, "Exposed Population (Millions)"), 15), caption = "Total Population Exposed to Ecosystem Service Declines (Top 15)")
  }
} else {
  cat("Population exposure data not available")
}
```

## Regional Policy Priorities

Regional hotspot burden is quantified by **relative intensity** (observed hotspot share ÷ expected share given regional land area): values above 1.0 indicate a region hosts a disproportionate share of global hotspot cells. All figures below are derived from `hotspot_area_stats.csv` (grouping_var == "region_wb") at 10 km² grid resolution. Per-region population counts are not directly available in current outputs; see the multiplier effect dumbbell plots in Chapter 6 for regional exposure scale.

### Sub-Saharan Africa
- **Top Hotspot Services** (by relative intensity): Nature Access (1.17×), Sediment Retention (1.10×), Pollination (1.09×)
- **Geographic footprint**: ~14,700 Nature Access cells; ~13,700 Pollination cells (out of 241,000 regional cells)
- **Key Countries**: Nigeria, DRC, Ethiopia, Tanzania
- **Priority**: Ecosystem-based agriculture; soil and water conservation; access to natural areas for rural communities
- **Regional burden**: All three services exceed the 1.0 benchmark — disproportionate risk given the region's low historical contribution to drivers

### Latin America & Caribbean
- **Top Hotspot Services** (by relative intensity): Sediment Export (2.51×), Pollination (1.91×), Nitrogen Export (1.65×)
- **Geographic footprint**: ~26,600 Sediment Export cells; ~20,300 Pollination cells (out of 204,000 regional cells)
- **Key Countries**: Brazil, Colombia, Mexico, Bolivia
- **Priority**: Upstream watershed protection; riparian buffer restoration; sustainable agricultural intensification to reduce sediment and nutrient loads
- **Regional burden**: Sediment Export at 2.51× is the highest region-service relative intensity in the entire dataset — a global priority signal

### East Asia & Pacific
- **Top Hotspot Services** (by relative intensity): Nitrogen Retention (1.80×), Nitrogen Export (1.57×), Sediment Retention (1.56×)
- **Geographic footprint**: ~23,000 N_Ret_Ratio cells; ~20,000 N_export cells (out of 246,000 regional cells)
- **Key Countries**: Indonesia, Philippines, Vietnam, China
- **Priority**: Nutrient management; watershed restoration; wetland conservation for water purification
- **Regional burden**: Water quality services are strongly disproportionate here; coastal risk is at approximately expected levels (relative intensity ≈ 0.95)

### South Asia
- **Top Hotspot Services** (by relative intensity): Nitrogen Retention (1.37×), Nature Access (1.08×), Nitrogen Export (1.08×)
- **Geographic footprint**: ~3,600 N_Ret_Ratio cells; ~2,800 Nature Access cells (out of 51,000 regional cells)
- **Key Countries**: India, Bangladesh, Pakistan
- **Priority**: Fertilizer management; groundwater and riparian protection; urban green access
- **Regional burden**: Moderate relative intensities but very high absolute population density — geographic footprint understates human exposure

### Middle East & North Africa
- **Top Hotspot Services** (by relative intensity): Nature Access (1.31×), Nitrogen Retention (1.13×)
- **Geographic footprint**: ~5,800 Nature Access cells; ~5,000 N_Ret_Ratio cells (out of 86,000 regional cells)
- **Key Countries**: Morocco, Egypt, Iran, Algeria
- **Priority**: Access to nature for urban and semi-arid populations; water retention in dryland watersheds
- **Regional burden**: Nature access disproportionate (1.31×); most other services near or below expected levels

### Europe & Central Asia
- **Top Hotspot Services**: All services below the 1.0 equal-distribution benchmark (highest: Coastal Risk Reduction ≈ 0.88×)
- **Notable Pattern**: No service exceeds its expected share — region hosts fewer global hotspot cells than its land area would predict
- **Mechanisms**: Protected area systems; ecosystem recovery in post-industrial landscapes; agri-environment policy schemes
- **Lesson for Others**: Policy-driven land management can yield measurable ecosystem recovery at regional scale

## Key Takeaways

::: {.callout-note}
**Key findings from regional analysis:**

1. **Disproportionate burden in the Global South.** Sub-Saharan Africa, Latin America & the Caribbean, East Asia & Pacific, and South Asia all show hotspot relative intensities above 1.0 for at least their top services, meaning they host more hotspot area than their land share would predict — while contributing least to the historical land-use pressures driving those changes.

2. **Compound risk concentrates in tropical regions.** The country-level compound risk chart above shows that countries with high *average* overlapping hotspot counts tend to be in tropical biomes — where multiple services co-decline in the same cells, leaving communities with no substitute service buffer.

3. **East Asia & Pacific faces concentrated water quality risk.** The region's top signals are Nitrogen Retention (1.80×), Nitrogen Export (1.57×), and Sediment Retention (1.56×) — all substantially above the equal-distribution benchmark. Coastal risk in EAP is at approximately expected levels (0.95×); the dominant burden is freshwater ecosystem degradation.

4. **Europe & Central Asia diverges.** No service in this region exceeds the 1.0 equal-distribution benchmark — it hosts *fewer* global hotspot cells than its land area would predict. This likely reflects a combination of ecosystem recovery post-industrial decline and stronger protected-area systems, though low SPC values can reflect either ecological health *or* already-depleted baselines.

5. **Country-level variation within regions is substantial** (see interactive table above). Regional averages mask heterogeneity: within Sub-Saharan Africa, for instance, small land-locked states often show higher relative intensity than large forested nations.
:::

## How to Use This Chapter

**For policymakers**: Locate your country/region and check which services are most at-risk; cross-reference with Chapter 6 for driver-specific interventions.

**For conservation organizations**: Use country rankings to allocate resources to highest-burden regions; use population exposure data to identify vulnerable communities.

**For researchers**: Detailed geographic data enables hypothesis testing; compound risk patterns may reveal ecological principles about multi-service failure.

## Data Caveats

- Country-level data aggregated from 10km grid cells; within-country variation may be substantial
- Population exposure estimates based on grid-cell-level socioeconomic data; local variations not captured
- Historical data (1992-2020); recent trends may differ
- Services modeled with varying data quality (see Technical Appendix)

