---
title: "DaTa Preparation NBS"
output: html_notebook
---

I will use this project to peprare and extract insights from the raster data for the polygons in the 5 prioritized areas.


1. Prepare Environemnt (Load necessary libraries)
```{r load libraries}
packs <- c('terra', 'purrr', 'landscapemetrics', 'sf','dplyr',
           'here', 'gdalUtilities', 'jsonlite', 'devtools', 'stringr',
           'parallel', 'dplyr', 'tidyr', 'ggplot2')
sapply(packs, require, character.only = TRUE, quietly=TRUE)
rm(packs)
```

2. Load The polygons with the NBS areas
```{r}

poly <- st_read('/Volumes/Jero_HDD/POSDOC/NBS_OPS/vector/areas_nbs.geojson') 
#Split the vector into a list of individual polygons
#create vector with names
nam <- poly$COUNTR
nam <- sort(nam)
poly <- poly%>%split(.$COUNTR)
```

3. Prepare Inventory to organize data 

I am having to many issues with the data ialready have and the one i don't and it gests confusing, so I am workin on a table that shhould make this easier

```{r organize inventory, eval=FALSE, include=FALSE}

tab <- as_tibble(read.csv('/Volumes/Jero_HDD/POSDOC/Downloaded_data_ES/inventory_global_products.csv'))
library(dplyr)

# Assuming your tibble is called 'data'
tab <- tab %>%
  select(-matches("^X(\\.\\d+)?$"))

# Print the cleaned tibble
print(cleaned_data)

tab <- tab %>% slice(1:218)
tab <- tab %>% mutate(Filename= basename(Link))

write.csv(tab, file= '/Volumes/Jero_HDD/POSDOC/Downloaded_data_ES/inventory_global_products.csv')

```
Test with one raster
I already did this, it is quite annoying but anyway, no need to do again
```{r load rst, eval=FALSE, include=FALSE}
#Land cover
esamod <- rast('/Volumes/Jero_HDD/WWF/global_outputs/ESAmodVCFv2_md5_05407ed305c24604eb5a38551cddb031.tif')

esamod <- map(1:length(poly), function(x) crop(esamod, poly[[x]]))
esamod <- map(1:length(esamod), function(x) mask(esamod[[x]], poly[[x]]))

map(1:length(esamod), function(x) writeRaster(esamod[[x]], paste0('/Volumes/Jero_HDD/WWF/NBS_OPS/raster/', 'esaLC_', nam[x], '.tif'), overwrite=TRUE))

# Access
access_nature<- rast('/Volumes/Jero_HDD/WWF/global_outputs/nature_access_lspop2019_esa2020modVCFhab_md5_a6519ebd8b941444921e749da2e645bb_WGS84.tif')

access_nature <- map(1:length(poly), function(x) crop(access_nature, poly[[x]]))
access_nature <- map(1:length(access_nature), function(x) mask(access_nature[[x]], poly[[x]]))

map(1:length(access_nature), function(x) writeRaster(access_nature[[x]], paste0('/Volumes/Jero_HDD/WWF/NBS_OPS/raster/', 'nature_access_ESA_2020', nam[x], '.tif'), overwrite=TRUE))

# Access restoration
acces_rest <- rast('/Volumes/Jero_HDD/WWF/global_outputs/nature_access_diff_Sc3v1_PNVnoag-esa2020.tif')
acces_rest <- project(acces_rest, acc2, method='bilinear')

acces_rest <- map(1:length(poly), function(x) crop(acces_rest, poly[[x]]))
acces_rest <- map(1:length(acces_rest), function(x) mask(acces_rest[[x]], poly[[x]]))
map(1:length(access_nature), function(x) writeRaster(acces_rest[[x]], paste0('/Volumes/Jero_HDD/WWF/NBS_OPS/raster/', 'nature_access_diff_Sc3v1_PNVnoag-esa2020_', nam[x], '.tif'), overwrite=TRUE))


acces_rest <- rast('/Volumes/Jero_HDD/WWF/global_outputs/global_people_access_population_2019_60.0m_md5_d264d371bd0d0a750b002a673abbb383.tif')
acces_rest <- project(acces_rest, acc2, method='bilinear')

acces_rest <- map(1:length(poly), function(x) crop(acces_rest, poly[[x]]))
acces_rest <- map(1:length(acces_rest), function(x) mask(acces_rest[[x]], poly[[x]]))

map(1:length(access_nature), function(x) writeRaster(acces_rest[[x]], paste0('/Volumes/Jero_HDD/WWF/NBS_OPS/raster/', 'global_people_access_population_2019_60.0m_', nam[x], '.tif'), overwrite=TRUE))

access_nature<- rast('/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/DataOBS_op/global_products/ESAmodVCFv2_cv_habitat_value_md5_c01e9b17aee323ead79573d66fa4020d.tif')

access_nature <- map(1:length(poly), function(x) crop(access_nature, poly[[x]]))
access_nature <- map(1:length(access_nature), function(x) mask(access_nature[[x]], poly[[x]]))

map(1:length(access_nature), function(x) writeRaster(access_nature[[x]], paste0('/Users/sputnik/Documents/Natural_capital/NBS_OP/data/raster/', 'coastal_risk_reduction', nam[x], '.tif'), overwrite=TRUE))

```


# Iterate over all rasters (this has been annoying)

```{r iterate crop}
wd <- '/Volumes/Jero_HDD/POSDOC/Downloaded_data_ES'
setwd(wd)

tiffes <- list.files(wd, pattern= '2020-1992')
tiffes <- tiffes[-2]
paths <- file.path(wd, tiffes)
n <- 3

t <- rast(paths[[n]])
clean_filename <- function(filename) {
  sub("_md5.*", "", filename)
}

output_path <- '/Volumes/Jero_HDD/POSDOC/NBS_OPS/cropped_raster_data/'

 cropped <- map(1:length(poly), function(x) crop(t, poly[[x]]))
  masked <- map(1:length(cropped), function(x) mask(cropped[[x]], poly[[x]]))
  
  # Extract the base filename before 'md5'
  base_filename <- clean_filename(basename(paths[[n]]))
  
  # Write the cropped and masked rasters to files
  map(1:length(masked), function(x) (writeRaster(masked[[x]], paste0(output_path, base_filename, "_", nam[x], ".tif"), overwrite = TRUE)))

# Remove faulty data
#tiffes <- tiffes[-1]

# Loop through each geotiff file
for (i in seq_along(tiffes)) {
  
  # Get the current tiff file
  tiff_file <- paths[i]
  
  # Load the current geotiff (don't load all at once)
  access <- rast(tiff_file)
  
  # Apply crop and mask operations to each polygon
  cropped <- map(1:length(poly), function(x) crop(access, poly[[x]]))
  masked <- map(1:length(cropped), function(x) mask(cropped[[x]], poly[[x]]))
  
  # Extract the base filename before 'md5'
  base_filename <- clean_filename(basename(tiff_file))
  
  # Write the cropped and masked rasters to files
  map(1:length(masked), function(x) {
    writeRaster(masked[[x]], paste0(output_path, base_filename, "_", nam[x], ".tif"), overwrite = TRUE)
  })
  # Optionally, clear the variable to free up memory
  rm(access, cropped, masked)
  gc()  # Garbage collection to clear memory
}

```


# Reproj Accesibility Raster

```{r reproj access}
acc <- rast('/Volumes/Jero_HDD/WWF/global_outputs/nature_access_lspop2019_esa2020modVCFhab_md5_a6519ebd8b941444921e749da2e645bb.tif')
acc

acc2 <- rast('/Volumes/Jero_HDD/WWF/global_outputs/temp/access_WGS84.tif')

acc <- project(acc,acc2)
writeRaster(acc, '/Volumes/Jero_HDD/WWF/global_outputs/nature_access_lspop2019_esa2020modVCFhab_md5_a6519ebd8b941444921e749da2e645bb_WGS84.tif')
```

# Explore rasters
(max min values, unique, frequencies)
```{r explore rasters}
sediment <- rast('/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/DataOBS_op/raster/sediment_ESAmod2-Sc3v1_MEXICO.tif')
```

#summaries

For summarizing geotiff raster files and comparing values across protected areas or different land covers for an initial assessment, especially for an audience without geospatial expertise, here are some accessible metrics you can use:
1. Basic Statistical Summaries:

# set class names and colors

```{r set symbols}
lb <- data.frame(
  Class <- c("Cultivated", "Forest", "Shrubs and Grasses", "Sparse",
            "Mangroves", "Urban", "Bare", "Water"),
  color=c("#dcf064", "#00b809", "#d29000", "#ffebaf", "#009678", "#c31400", 
            "#fff5d7", "#0046c8"),
  value=c(1:8))
names(lb) <- c("Class", "color", "value")
```

#Summaries Land cover
```{r basic summaries}

path_lc <- '/Volumes/Jero_HDD/POSDOC/ESA_LC'
# load reclassified land cover map
tf <- file.path(path_lc, list.files(path_lc, pattern= "rec"))
lc <- lapply(tf,rast)

#this needs to be converted into projected coordinates before extracting the metrics
props <- lapply(lc, function(r){
  mets <- freq(r)
})
props <- lapply(seq_along(props), function(i) {
  props[[i]]$AREA <- nam[i]
  return(props[[i]])
})

props <- do.call(rbind, props)

props <- as_tibble(props)
lb <- as_tibble(lb)
props <- left_join(props, lb)


# Assume your data is stored in a tibble called `data`
props <- props %>%
  group_by(AREA, layer) %>%
  mutate(percentage = count / sum(count) * 100) %>%  # Calculate percentage
  ungroup()

# Create the bar plot using ggplot2
ggplot(props, aes(x = Class, y = percentage, fill = color)) +
  geom_bar(stat = "identity") +  # Use identity since we are plotting pre-calculated values
  scale_fill_identity() +  # Use the exact colors provided in the `color` column
  facet_grid(layer ~ AREA) +  # Create a grid with rows for `AREA` and columns for `layer`
  labs(title = "Class Distribution by Area and Year",
       x = "Class",
       y = "Percentage") +
  #theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability


color_mapping <- props %>% 
  distinct(Class, color) %>% 
  pull(color, name = Class)
# By Region

t <- ggplot(props, aes(x=Class, y=value, fill=Class)+
              geom_bar()+
              scale_fill_manual(values=color_mapping)+
            facet_grid(AREA~layer)+
              labs(
                x= "Class",
                y= "pix",
                color="Class"
              )
)
              

#this needs to be converted into projected coordinates before extracting the metrics
lsmet <- lapply(lc, function(r){
  mets <- lsm_c_ai(r)
})

lsmet <- lapply(seq_along(lsmet), function(i) {
  lsmet[[i]]$AREA <- nam[i]
  return(lsmet[[i]])
})
```
Get the restoration and send the charts 


Beneficiaries and coastal protection pop< 10 downstreams
I already produced the map.

45 - 19 (calculate the values) for coastal risk reduction

```{r coastal risk red}

wd <- '/Volumes/Jero_HDD/POSDOC/NBS_OPS/vector'

coastal_risk_reduction_MEXICO.geojson

pat <- file.path(wd, list.files(wd, pattern="coastal_risk"))

pat <- pat[-5]

vat <- lapply(pat, st_read)
vat <- lapply(vat, function(v){
  v <- v %>% mutate(rt_fin=Rt_nohab_all-Rt)
})


library(sf)
library(janitor)

# Standardize column names and save each sf object
lapply(seq_along(vat), function(i) {
  # Clean column names to ensure compatibility
  vat[[i]] <- clean_names(vat[[i]])
  
  # Save the sf object to the specified path
  st_write(vat[[i]], pat[i], append = FALSE, delete_layer = TRUE)
})


lapply(seq_along(vat), function(i) {
  st_write(vat[[i]], pat[i], append = TRUE)
})

```

Substract these columnns, and map this thing again 
Add legend (in terms of hig-low)

include access change

then, go through these change raster (priority 4 and take a look at them and think what can be done here, and show differences for land covers), inside outside protected areas (UICN cat National park Ia , Ib, II, III, IV , V, VI)

# Get summaries

Barplots/distributions outisde inside protected areas/land cover classes as a way to see if protected areas are abnormally represented.
## Get summaries per PA.

```{r get summaries}
#load protected areas
poly = st_read('/Volumes/Jero_HDD/POSDOC/NBS_OPS/vector/wdpa_nbs.gpkg')
#remove those without UICN cathegory
filt <- c("Not Reported", "Not Assigned", "Not Applicable")
poly <- poly %>% filter(!(IUCN_CAT %in% filt))
poly <- poly%>%split(.$COUNTR)

#Get baseline values for ecosystem services. Will use the 2020baseline values
# set inpath
wd <- '/Volumes/Jero_HDD/POSDOC/NBS_OPS/cropped_raster_data'
wd <- '/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/DataOBS_op/cropped_rasters'


tiffes <- file.path(wd, list.files(wd, pattern="tif"))
#tiffes <-  tiffes[c(36:40, 46:50, 66:70, 86:90)]
tiffes <-  tiffes[c(1:5, 11:15, 21:25, 41:45)]
lab <- sapply(tiffes, function(x) {
  basename(tools::file_path_sans_ext(x))
})

baseES <- lapply(tiffes, rast)

  # Extract only the file name without the extension
tfs <- sapply(tiffes, function(x) {
  basename(tools::file_path_sans_ext(x))
})
tfs <- as_tibble(tfs)

# Define a function to compute the total sum for each SpatRaster
compute_sum <- function(raster) {
  total_sum <- global(raster, fun = "sum", na.rm = TRUE)[,1]  # Use global() to get   return(total_sum)
}

# This has the total ES per site
totSE <-  mclapply(baseES, compute_sum, mc.cores = detectCores() - 4)
totSE <- as_tibble(unlist(totSE))
totSE <- cbind(tfs,totSE)


ap <- st_read('/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/DataOBS_op/pa_nbs.shp')
filt <- c("Not Reported", "Not Assigned", "Not Applicable")
ap <- ap %>% filter(!(IUCN_CAT %in% filt))
ap <- ap%>%split(.$COUNTR)
# This is very bad work, and i did not even document it correctly. Terrible practices here.
# This was to fix the order of the lists of ecosystem services ws the list of protected areas. I don;'t remeber why this happened and why i decided this was a good way to  fixing 
#baseES <- as.list(as.vector(matrix(baseES, nrow = 5, byrow = FALSE)))

s1 <- list(baseES[[1]],baseES[[6]],baseES[[11]], baseES[[16]])
s2 <- list(baseES[[2]],baseES[[7]],baseES[[12]], baseES[[17]])
s3 <- list(baseES[[3]],baseES[[8]],baseES[[13]], baseES[[18]])
s4 <- list(baseES[[4]],baseES[[9]],baseES[[14]], baseES[[19]])
s5 <- list(baseES[[5]],baseES[[10]],baseES[[15]], baseES[[20]])

#baseEs_ap <- map(baseES, ap, mask)
s1 <- lapply(s1, function(r){
  r <- mask(r,ap[[1]])
})
s2 <- lapply(s2, function(r){
  r <- mask(r,ap[[2]])
})
s3 <- lapply(s3, function(r){
  r <- mask(r,ap[[3]])
})
s4 <- lapply(s4, function(r){
  r <- mask(r,ap[[4]])
})
s5 <- lapply(s5, function(r){
  r <- mask(r,ap[[5]])
})


# this is becasue it is a list, not a multiband raster, but we can fix that
baseES_ap <- list(s1[[1]],s2[[1]],s3[[1]],s4[[1]],s5[[1]],s1[[2]],s2[[2]],s3[[2]],s4[[2]],s5[[2]],s1[[3]],s2[[3]],s3[[3]],s4[[3]],s5[[3]], s1[[4]],s2[[4]],s3[[4]],s4[[4]], s5[[4]])

# Assume you have a list of SpatRasters called raster_list
msk_all <- lapply(baseES, function(r) {
  # Replace all non-NA values with 1 using ifel()
  ifel(!is.na(r), 1, NA)
})

msk_ap <- lapply(baseES_ap, function(r) {
  # Replace all non-NA values with 1 using ifel()
  r <- ifel(!is.na(r), 1, NA)
})


freq_a <- lapply(msk_all, function(r){
  f <- freq(r)
})
freq_a <- do.call(rbind, freq_a)

freq_ap <- lapply(msk_ap, function(r){
  f <- freq(r)
})

freq_ap <- do.call(rbind, freq_ap)

freq_a <- freq_a %>% select(3)
freq_ap <- freq_ap %>% select(3)

#calculate ES output
totSE_ap <-  lapply(baseES_ap, compute_sum)#, mc.cores = detectCores() - 4)
totSE_ap <- as_tibble(unlist(totSE_ap))

totSE <- cbind(totSE, totSE_ap)

totSE <- cbind(totSE, freq_a, freq_ap)
names(totSE) <- c("Service", "WholeArea", "PArea", "pix_count_area", "pix_count_pa")

#Calcualte share protected area/total
totSE <- totSE %>% mutate(share_service= (PArea/WholeArea)*100)
totSE <- totSE %>% mutate(share_area= (pix_count_pa/pix_count_area)*100)

totSE <- totSE%>% mutate(Country = str_extract(Service, "[^_]+$")) %>%  # Extract the text after the last '_'
  # Step 2: Remove the last '_' and country name from the original value column
  mutate(Service = str_remove(Service, "_[^_]+$"))

write.csv(totSE, file=here('data', "totSE.csv"))

totSE_l <- totSE %>%
  # Create a new column with simplified service names
  mutate(Service = case_when(
    Service == "global_n_retention_esamod2_compressed" ~ "Nitrogen Retention",
    Service == "global_sed_retention_esamod2_compressed" ~ "Sediment Retention",
    Service == "pollination_ppl_fed_on_ag_10s_esa2020" ~ "Pollination",
    Service == "nature_access_ESA" ~ "Access",
    TRUE ~ Service  # Keep other names unchanged, if any
  )) %>%
  # Select relevant columns and reshape the data into long format
  select(Service, share_service, share_area, Country) %>%
  pivot_longer(cols = c(share_service, share_area),
               names_to = "Metric",
               values_to = "Value")

##########################################################################
##########################################################################

# Get the histograms
baseES_v <- lapply(baseES, function(r){
  t <- as_tibble(values(r, na.rm=TRUE))
})

baseESap_v <- lapply(baseES_ap, function(r){
  t <- as_tibble(values(r, na.rm=TRUE))
})

# waht the fuck does this freaking line do???/ It adds the id for the service
# There is a better way to do it, check the next part 
baseES_v <- lapply(seq_along(baseES_v), function(i) {
  cbind(baseES_v[[i]], NewColumn = lab[i])
})

baseES_v <- lapply(baseES_v, function(r){
  r <- as_tibble(r)
})

baseESap_v <- lapply(seq_along(baseESap_v), function(i) {
  cbind(baseESap_v[[i]], NewColumn = lab[i])
})


baseESap_v <- lapply(baseESap_v, function(r){
  r <- as_tibble(r)
})
new_col <- "Value"
# Use lapply to rename the first column of each tibble
baseES_v <- lapply(baseES_v, function(tbl) {
  tbl %>%
    rename(!!new_col := 1)  # Rename first column using column position
})

baseESap_v <- lapply(baseESap_v, function(tbl) {
  tbl %>%
    rename(!!new_col := 1)  # Rename first column using column position
})


baseES_v <- do.call(rbind, baseES_v)
baseESap_v <- do.call(rbind, baseESap_v)
##################### do not run inside the Rmd!!!!!!!!
names(baseES_v) <- c("Value", "Service")
names(baseESap_v) <- c("Value", "Service")
###########################################################
###########################################################

baseES_v <- baseES_v%>% mutate(Country = str_extract(Service, "[^_]+$")) %>%  # Extract the text after the last '_'
  # Step 2: Remove the last '_' and country name from the original value column
  mutate(Service = str_remove(Service, "_[^_]+$"))

baseES_v <- baseES_v%>% mutate(set = "all")


baseESap_v <- baseESap_v%>% mutate(Country = str_extract(Service, "[^_]+$")) %>%  # Extract the text after the last '_'
  # Step 2: Remove the last '_' and country name from the original value column
  mutate(Service = str_remove(Service, "_[^_]+$"))

baseESap_v <- baseESap_v%>% mutate(set = "protected_areas")
baseES_vf <- rbind(baseES_v, baseESap_v)

baseES_vf <- baseES_vf %>%
  # Create a new column with simplified service names
  mutate(Service = case_when(
    Service == "global_n_retention_esamod2_compressed" ~ "Nitrogen Retention",
    Service == "global_sed_retention_esamod2_compressed" ~ "Sediment Retention",
    Service == "pollination_ppl_fed_on_ag_10s_esa2020" ~ "Pollination",
    Service == "nature_access_ESA" ~ "Nature Access",
    Service == "nature_access_ESA_2020" ~ "Nature Access",
    TRUE ~ Service  # Keep other names unchanged, if any
  )) 


write.csv(baseES_vf, here('data', "distributionServ.csv"))
# Calculate the 98th percentile for each combination of Service and Country
percentile_98 <- baseES_vf %>%
  group_by(Service, Country) %>%
  summarize(threshold = quantile(Value, 0.98)) %>%
  ungroup()
percentile_95 <- baseES_vf %>%
  group_by(Service, Country) %>%
  summarize(threshold = quantile(Value, 0.95)) %>%
  ungroup()


# Join the threshold back to the original data and filter values up to the 98th percentile
baseES_vf_filtered <- baseES_vf %>%
  left_join(percentile_98, by = c("Service", "Country")) %>%
  filter(Value <= threshold)

baseES_vf_filtered <- baseES_vf %>%
  left_join(percentile_95, by = c("Service", "Country")) %>%
  filter(Value <= threshold)



Count <- c("BRAZIL", "MADAGASCAR", "MEXICO", "PERU","VIETNAM")

C <- 1

df <- baseES_vf_filtered %>% filter(Country==Count[C]) %>% filter(Service!="Nature Access")

df <- baseES_vf_filtered %>% filter(Service=="Nature Access")

t <- ggplot(df, aes(x = Value, fill = set)) +
  geom_density(alpha = 0.4) +  # Use transparency with alpha to see overlapping densities
  #facet_wrap(~Service, scales = "free") +
  facet_wrap(~Country, scales = "free") +
  # Set both x and y axes to be free for each facet
  scale_fill_manual(values = c("blue", "orange")) +  # Set colors for the `set` variable
  labs(title = paste("Density Plot of Services Inside and Outside Protected Areas -", Count[C]),
       x = "Value",
       y = "Density",
       fill = "Set") +
  #scale_y_continuous(trans = 'log10')# Label for the fill legend
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12),  # Increase facet labels' font size
    legend.position = "bottom",  # Place the legend at the bottom
    legend.box = "horizontal"  # Arrange legend items horizontally
  )
t
```
# Plot the shares

```{r ggplot, h=8, w=8}

# Step 2: Create the bar plot
p <- ggplot(totSE_l, aes(x = Metric, y = Value, fill = Service)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +  # Side-by-side bars
  facet_wrap(~ Country, ncol = 5) +  # Create 5 columns (one for each country)
  scale_fill_brewer(palette = "Set3") +  # Use a color palette for services
  labs(title = "Share of the protected areas and of the total ES provided inside them",
       x = "Metric",
       y = "%",
       fill = "Service") +
  #theme_minimal() +
   theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    legend.position = "bottom",  # Place the legend at the bottom
    legend.box = "horizontal",  # Arrange legend items horizontally
    legend.spacing.x = unit(0.1, 'cm'),  # Add horizontal spacing between legend items
    legend.title = element_text(size = 12, face = "bold"),  # Adjust legend title size
    legend.text = element_text(size = 10),  # Adjust legend text size
    legend.box.just = "center"  # Center the legend items
  )



```

# Restorastion Areas Griscom

```{r resto}
wd <- '/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/global_products/Restoration'



tiffes <- file.path(wd,list.files(wd, pattern = "\\.tif$", full.names = FALSE))
tiffes <- tiffes[-5]


wd <- '/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/global_products/Restoration/test'


rest <- lapply(tiffes, rast)
               
               # This function extracts the 1 and 0 for mthe categorical rasters from restoration and corrects the weird values. did this because i had no metadata. Add the function to a new package eventully
process_raster <- function(raster) {
  # Step 1: Extract unique values
  unique_values <- unique(values(raster, na.rm = TRUE))
  unique_values <- as.vector(unique_values)  # Convert to vector

  # Step 2: Get frequency counts of each unique value
  value_frequencies <- freq(raster)
  val_f <- as.numeric(value_frequencies[, 2])  # Get the frequencies

  # Step 3: Convert raster to numeric
  raster_numeric <- as.numeric(raster)

  # Step 4: Substitute the values with the frequencies
  raster_subst <- subst(raster_numeric, from = unique_values, to = val_f)

  # Return the reclassified raster
  return(raster_subst)
}

rest  <- lapply(rest, process_raster)
#reorganize the list by countries, not by products. 

# Step 1: Extract country names from the "source" attribute of each SpatRaster object
get_country <- function(r) {
  # Extract the source file path
  source_name <- sources(r)[1]  # Extract the first source (in case there are multiple)

  # Use a regular expression to extract the country name from the file name
  # Assuming the file name format is "something_COUNTRY.tif"
  country <- sub(".*_([^_]+)\\.tif$", "\\1", source_name)
  return(country)
}

country_names <- unique(sapply(baseES, get_country))  # Get unique country names
country_list <- setNames(vector("list", length(country_names)), country_names)

# Step 3: Group SpatRasters by country and populate the nested list
for (i in seq_along(baseES)) {
  country <- get_country(baseES[[i]])
  country_list[[country]] <- c(country_list[[country]], list(baseES[[i]]))
}

baseES <- country_list
rm(country_list)

baseES_rp <- lapply(seq_along(baseES), function(i) {
  # Get the current sublist of SpatRasters from baseES
  current_sublist <- baseES[[i]]
  # Get the corresponding template SpatRaster for reprojection
  template_raster <- rest[[i]]
  # Use lapply to reproject each SpatRaster in the sublist using the template raster
  reprojected_sublist <- lapply(current_sublist, function(r) {
    # Reproject the SpatRaster to match the CRS and extent of the template raster
    project(r, template_raster, method = "bilinear")
  })
  # Return the reprojected sublist
  return(reprojected_sublist)
})


baseES <- baseES_rp
rm(baseES_rp)

baseES_0 <- lapply(seq_along(baseES), function(i) {
  # Get the current sublist of SpatRasters from baseES
  current_sublist <- baseES[[i]]
  # Get the corresponding template SpatRaster for reprojection
  template_raster <- rest[[i]]
  # Use lapply to reproject each SpatRaster in the sublist using the template raster
  masked_sublist <- lapply(current_sublist, function(r) {
    # Reproject the SpatRaster to match the CRS and extent of the template raster
    mask(r, template_raster, maskvalues=0)
  })
  # Return the reprojected sublist
  return(masked_sublist)
})
# ok so far so good, but of course resolution with freaking access is always an issue
baseES_1 <- lapply(seq_along(baseES), function(i) {
  # Get the current sublist of SpatRasters from baseES
  current_sublist <- baseES[[i]]
  # Get the corresponding template SpatRaster for reprojection
  template_raster <- rest[[i]]
  # Use lapply to reproject each SpatRaster in the sublist using the template raster
  masked_sublist <- lapply(current_sublist, function(r) {
    # Reproject the SpatRaster to match the CRS and extent of the template raster
    mask(r, template_raster, maskvalues=1)
  })
  # Return the reprojected sublist
  return(masked_sublist)
})

 # Step 1: Flatten the nested list into a single list of SpatRaster objects
flattened_list <- unlist(baseES_1, recursive = FALSE)

# Step 2: Extract information for the data frame

# Create a dataframe to store the information
raster_info <- data.frame(
  Country = character(),   # Empty column for country names
  Product = character(),   # Empty column for product names
  stringsAsFactors = FALSE # Prevent automatic conversion to factors
)

# Step 3: Generate the correct country names for each raster
# Calculate the number of rasters per country
rasters_per_country <- length(flattened_list) / length(country_names)

# Repeat each country name for the number of rasters per country
repeated_country_names <- rep(country_names, each = rasters_per_country)

# Step 4: Iterate over the flattened list and extract the relevant information
for (i in seq_along(flattened_list)) {
  # Get the corresponding country name from the repeated country names vector
  country_name <- repeated_country_names[i]
  
  # Get the product name from the `name` slot of the SpatRaster
  product_name <- names(flattened_list[[i]])
  
  # If the product name contains a "~" or "_md5", extract only the part before it
  if (grepl("_md5", product_name)) {
    product_name <- sub("_md5.*$", "", product_name)
  }

  # Step 5: Add the extracted information to the dataframe
  raster_info <- rbind(raster_info, data.frame(Country = country_name, Product = product_name))
}

# Step 5: Optional - If you need to store the raster info back into a named list with country and product
# Create a named list where each raster is named by country and product

# baseES_0 <- setNames(flattened_list, paste(raster_info$Country, raster_info$Product, sep = "_"))
baseES_1 <- setNames(flattened_list, paste(raster_info$Country, raster_info$Product, sep = "_"))

# Get the histograms
baseES_v <- lapply(baseES_0, function(r){
  t <- as_tibble(values(r, na.rm=TRUE))
})

baseESap_v <- lapply(baseES_1, function(r){
  t <- as_tibble(values(r, na.rm=TRUE))
})


# Adding corresponding rows from `df` to each tibble in `list_of_tibbles`
baseES_v <- map2(baseES_v,                  # First input: List of tibbles
  split(raster_info, seq(nrow(raster_info))),         # Second input: Split the dataframe into individual rows
  ~ mutate(.x, Country = .y$Country, Product = .y$Product)  # Add `Country` and `Product` columns from df to each tibble
)

baseESap_v <- map2(baseESap_v,                  # First input: List of tibbles
  split(raster_info, seq(nrow(raster_info))),         # Second input: Split the dataframe into individual rows
  ~ mutate(.x, Country = .y$Country, Product = .y$Product)  # Add `Country` and `Product` columns from df to each tibble
)


new_col <- "Value"
# Use lapply to rename the first column of each tibble
baseES_v <- lapply(baseES_v, function(tbl) {
  tbl %>%
    rename(!!new_col := 1)  # Rename first column using column position
})

baseESap_v <- lapply(baseESap_v, function(tbl) {
  tbl %>%
    rename(!!new_col := 1)  # Rename first column using column position
})


baseES_v <- do.call(rbind, baseES_v)
baseESap_v <- do.call(rbind, baseESap_v)
##################### do not run inside the Rmd!!!!!!!!
baseES_v <- baseES_v%>% mutate(restoration = "0")
baseESap_v <- baseESap_v%>% mutate(restoration = "1")

baseES_v <- rbind(baseES_v, baseESap_v)

baseES_v <- baseES_v %>%
  # Create a new column with simplified service names
  mutate(Service = case_when(
    Product == "global_n_retention_esamod2_compressed" ~ "Nitrogen Retention",
    Product == "global_sed_retention_esamod2_compressed" ~ "Sediment Retention",
    Product == "pollination_ppl_fed_on_ag_10s_esa2020" ~ "Pollination",
    Product == "access_WGS84" ~ "Nature Access",
    TRUE ~ Product  # Keep other names unchanged, if any
  )) 

write.csv(baseES_v, here('data', "distributionServ_restoration.csv"))
# Calculate the 98th percentile for each combination of Service and Country
percentile_98 <- baseES_v %>%
  group_by(Service, Country) %>%
  summarize(threshold = quantile(Value, 0.98)) %>%
  ungroup()
percentile_95 <- baseES_v %>%
  group_by(Service, Country) %>%
  summarize(threshold = quantile(Value, 0.95)) %>%
  ungroup()


# Join the threshold back to the original data and filter values up to the 98th percentile
baseES_filtered <- baseES_v %>%
  left_join(percentile_98, by = c("Service", "Country")) %>%
  filter(Value <= threshold)

baseES_filtered <- baseES_v %>%
  left_join(percentile_95, by = c("Service", "Country")) %>%
  filter(Value <= threshold)



C <- 5

df <- baseES_filtered %>% filter(Country==country_names[C]) %>% filter(Service!="Nature Access")

rm(t)
t <- ggplot(df, aes(x = Value, fill = restoration)) +
  geom_density(alpha = 0.4) +  # Use transparency with alpha to see overlapping densities
  facet_wrap(~Service, scales = "free") +
  #facet_wrap(~Country, scales = "free") +
  # Set both x and y axes to be free for each facet
  scale_fill_manual(values = c("blue", "orange")) +  # Set colors for the `set` variable
  labs(title = paste("Density Plot of Services Inside and Outside Griscom restoration Areas -", country_names[C]),
       x = "Value",
       y = "Density",
       fill = "Restoration") +
  #scale_y_continuous(trans = 'log10')# Label for the fill legend
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12),  # Increase facet labels' font size
    legend.position = "bottom",  # Place the legend at the bottom
    legend.box = "horizontal"  # Arrange legend items horizontally
  )
t
```

```{r get prportions restoration}

rest <- lapply(rest, function(r){
  r <- project(r, baseES[[1]])
  })


s1 <- list(baseES[[1]],baseES[[6]],baseES[[11]], baseES[[16]])
s2 <- list(baseES[[2]],baseES[[7]],baseES[[12]], baseES[[17]])
s3 <- list(baseES[[3]],baseES[[8]],baseES[[13]], baseES[[18]])
s4 <- list(baseES[[4]],baseES[[9]],baseES[[14]], baseES[[19]])
s5 <- list(baseES[[5]],baseES[[10]],baseES[[15]], baseES[[20]])

#baseEs_ap <- map(baseES, ap, mask)
s11 <- lapply(s1, function(r){
  r <- mask(r,rest[[1]],maskvalues=0)
})
s12 <- lapply(s1, function(r){
  r <- mask(r,rest[[1]],maskvalues=1)
})
s2 <- lapply(s2, function(r){
  r <- mask(r,ap[[2]])
})
s3 <- lapply(s3, function(r){
  r <- mask(r,ap[[3]])
})
s4 <- lapply(s4, function(r){
  r <- mask(r,ap[[4]])
})
s5 <- lapply(s5, function(r){
  r <- mask(r,ap[[5]])
})



# Produce land cover change maps
```{r land cover change}

#path to lc classifications
path_lc <- '/Volumes/Jero_HDD/POSDOC/ESA_LC'
# load reclassified land cover map
tf <- file.path(path_lc, list.files(path_lc, pattern= "all"))
lc <- lapply(tf,rast)
lc <- c(lc[[1]],lc[[2]])

#load study areas
poly <- st_read('/Volumes/Jero_HDD/POSDOC/NBS_OPS/vector/areas_nbs.geojson')
nam <- poly$COUNTR
nam <- sort(nam)
poly <- poly%>%split(.$COUNTR)
lc <-map(1:length(poly), function(x) crop(lc, poly[[x]]))
lc <-map(1:length(poly), function(x) mask(lc[[x]], poly[[x]]))
map(1:length(poly), function(x) writeRaster(lc[[x]], paste0(path_lc, '/', 'ESA_LC_rec_', nam[x], '.tif')))

chmaps <- lapply(lc, function(r){
  ch_mapR(r)
})

map(1:length(chmaps), function(x) writeRaster(chmaps[[x]][[1]], paste0(path_lc, '/', 'ESA_LC_change', nam[x], '.tif')), overwite=TRUE)


# pending: set a rule to deal wit those combinations that have to few pixels, not worth the time and effort,

```


#######################################################################

Second part with justin.

# Explore Data. 

The ecorregiosn Raster is not suited, 830 different values. Could be done but does not make a lot of sense
```{r explore data}

ecoregions <- rast('/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/DataOBS_op/global_products/Ecoregions2017_compressed_md5_316061.tif')

ec_uniqe <- unique(ecoregions)
```

Use vector file, just the main biomes. 

# PA 

```{r filter out PAs}

pa <- st_read('/Users/sputnik/Documents/Natural_capital/NBS_OP/data/vector/global-2024-05-08.gpkg')

att <- unique(pa$IUCN_CAT)

filt <- c("Not Reported", "Not Assigned", "Not Applicable")
pa <- pa %>% filter(!(IUCN_CAT %in% filt))

st_write(pa, '/Users/sputnik/Documents/Natural_capital/NBS_OP/data/vector/PA_filt.gpkg')



countries <- st_read('/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/DataOBS_op/global_products/cartographic_ee_ee_r264_correspondence.gpkg')

pa <- st_transform(pa, crs=st_crs(countries))

```

split the rasters by biomes

```{r split countries}
wd. <- '/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/DataOBS_op/global_products'

tf <- list.files(wd., pattern='1992')
tf <- clean_filename(tf)

tf <- tf[-c(2,3,5)]


tiffes <- file.path(wd., list.files(wd., pattern='1992'))
rs <- lapply(tiffes, rast) 

rs <- rs[-3]

biomes <- st_read('/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/global_products/biomes_WWF.gpkg')


nam <- sort(biomes$BIOME)

biomes <- biomes %>% split(.$BIOME)

# Apply cropping and masking recursively using lapply and map
rs_msk <- lapply(rs, function(raster) {
  map(biomes, function(polygon) {
    # Crop the raster to the extent of the polygon
    cropped_raster <- crop(raster, polygon)
    masked_raster <- mask(cropped_raster, polygon)
    return(masked_raster)
  })
})

outpath <- '~/Documents/Natural_capital/Global_mapping/rest_masked_biomes'
# Use lapply to iterate over each list in nested_list
lapply(seq_along(rs_msk), function(i) {
  # For each outer list, iterate through the inner list (rasters)
  lapply(seq_along(rs_msk[[i]]), function(j) {
    
    # Construct the filename using vector_1 and vector_2
    file_name <- paste0(outpath, tf[[i]], '_', nam[[j]], '.tif')
    
    # Write the raster to the file
    writeRaster(rs_msk[[i]][[j]], file_name, overwrite = TRUE)
  })
})

summary_df <- map2_dfr(seq_along(rs_msk), rs_msk, function(i, inner_list) {
  
  # i corresponds to the index of the outer list (e.g., 1, 2, 3...)
  outer_label <- paste0("Outer_", i)  # Label for the outer list
  
  # Loop through each SpatRaster in the inner list
  map_dfr(inner_list, function(raster) {
    extract_raster_summary(raster, outer_label)
  })
})

```


# Split by Countries/continents
# mask by protected areas (all togeter, chategories I-VI)
# Reclassify Land Cover maps. 


#  Change Maps
```{r change_map_calc}
esat <- c('/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/DataOBS_op/global_products/ESACCI-LC-L4-LCCS-Map-300m-P1Y-2020-v2.1.1_md5_2ed6285e6f8ec1e7e0b75309cc6d6f9f.tif','/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/DataOBS_op/global_products/ESACCI-LC-L4-LCCS-Map-300m-P1Y-1992-v2.0.7cds_compressed_md5_60cf30.tif')


esas <- lapply(esat,rast)


nam <- c(2020,1992)

map(1:2, function(x) writeRaster(esas[[x]], paste0('ESA_LC_', nam[x],'.tif')))
cls <- lapply(esas,freq)


wd <- '~/Documents/Natural_capital/NBS_OP/data/ESA_LC'

tf <- file.path(wd, list.files(wd, pattern='tif'))

tf <- tf[c(2,3)]

esas <- lapply(tf,rast)



countries <- st_read('/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/DataOBS_op/global_products/cartographic_ee_ee_r264_correspondence.gpkg')

cont <- countries %>% filter(subregion=='Central America')

# Define the reclassification rules in one operation
reclass_table <- data.frame(
  from = c(10, 11, 12, 20, 30, 40, 50, 60, 61, 62, 70, 71, 72, 80, 81, 82, 90, 100,
           110, 120, 121, 122, 130, 140, 150, 151, 152, 153, 160, 170, 180,
           190, 200, 201, 202, 210, 220),
  to = c(1, 1, 1, 1, 1, 1,   # Cultivated
         2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2,  # Forests
         3, 3, 3, 3, 3,   # Grasses and Shrubs
         4, 4, 4, 4, 4,   # Sparse Vegetation
         5, 5, 5,   # Mangroves
         6,   # Urban
         7, 7, 7,   # Bare
         8,   # Water
         9)   # Ice
)

esas <- lapply(esas, function(r) {
  subst(r, from = reclass_table$from, to = reclass_table$to)
})


map(1:2, function(x) writeRaster(esas[[x]], paste0('~/Documents/Natural_capital/NBS_OP/data/ESA_LC/', "Forests_esa", year[x], '.tif')))

# Extract forests
esas <- lapply(esas, function(r) {
  subst(r, from = 2, to =1, others=NA)
})


chg <- ch_mapR(esas[[1]],esas[[2]])
freq(chg)

mat_1 <- crosstabm(esas[[1]],esas[[2]])

diffs_p <-  diffTablej(mat, digits=3, analysis = "change")

```


# Get Sankey Diagrams
```{r Sankey Diagram}
classes <- c("Cultivated", "Forests", "Grasses/shrubs", "Sparse", "Mangroves", "Urban", "Bare", "Water")
groupColor <- c("#dcf064","#00b809","#8ca000","#ffebaf",'#009678', "#c31400", "#fff5d7", "#0046c8")

years <- c(1992,2020)


nodeInfo <- nodeInfoR(years, classes, groupColor)

NodeCols<- sort(unique(nodeInfo$nodeCol))

#create paths to the rasters/bands
num_bands <- length(NodeCols)
tf <- list.files(wd, pattern= "Rec")
tf <- file.path(wd,tf)
fileInfo <- tibble(
  nodeCol = seq_along(tf),
  rast = tf,
  rasterBand = 1
)

# join path strings to nodeInfo
nodeInfo <- dplyr::left_join(nodeInfo, fileInfo, by='nodeCol') 
linkInfo <- linkInfoR(NodeCols, nodeInfo)


fontSize <- 0.5
nodeWidth <-30
fontFamily <- "sans-serif"
colorScaleJS <- sprintf("d3.scaleOrdinal().domain(%s).range(%s)",
                        jsonlite::toJSON(unique(nodeInfo$nodeGroup)),
                        jsonlite::toJSON(unique(nodeInfo$groupColor)))


sankeyNetwork(Links = linkInfo, Nodes = nodeInfo,
              Source = "source",
              Target = "target",
              Value = "value",
              NodeID = "nodeName",
              NodeGroup = "nodeGroup",
              LinkGroup = "LinkGroup",
              fontSize = fontSize,
              fontFamily = fontFamily,
              nodePadding = 10,
              margin=1,
              nodeWidth = nodeWidth,
              colourScale = JS(colorScaleJS))
```



# Next step
1. Zonal Countries. one per service

mask for the forest 

not sum, average. 

Historic services. 

name(test)


Map 4 maps for each servie for the average change fior that time peiod 


map(1:2, function(x) writeRaster(esas[[x]], paste0('~/Documents/Natural_capital/NBS_OP/data/ESA_LC/', "ESA_LC_Rec_", year[x], '.tif')))



cls
[[1]]
   layer value      count
1      1    10  105009519
2      1    11  102087508
3      1    12    2415511
4      1    20   29796402
5      1    30   48088904
6      1    40   44613941
7      1    50  137316150
8      1    60   84649974
9      1    61   11061590
10     1    62   39391781
11     1    70  113472860
12     1    71   45563958
13     1    72      19882
14     1    80  109865531
15     1    81      55745
16     1    82         21
17     1    90   37417117
18     1   100   58048747
19     1   110   15730987
20     1   120  140408175
21     1   121    3007375
22     1   122   32663372
23     1   130  177774130
24     1   140   42447574
25     1   150  155885143
26     1   151         61
27     1   152    1072086
28     1   153    3664319
29     1   160   11677666
30     1   170    2394625
31     1   180   34499895
32     1   190   10933052
33     1   200  247641877
34     1   201    1572727
35     1   202    1244423
36     1   210 5675137546
37     1   220  871449826


So, now what does this mean?



Add column with the class names with multiple aggregation levels (ESA)

    Mean: The average value within each zone (e.g., inside/outside protected areas or across land cover types). This provides a simple understanding of the central tendency.
    Median: The middle value when data is ordered, useful when your data has outliers that could skew the mean.
    Standard Deviation: Indicates the variation or dispersion from the average, helping show if values are clustered or spread out.
    Min/Max: The range of values, showing the extremes within each zone.

2. Zonal Statistics (Per Category):

    Sum: The total value across a zone, particularly useful for variables like ecosystem services (e.g., total nitrogen retention within protected areas).
    Count of Pixels: The number of pixels in each zone, which could give an idea of the area covered by specific land covers.
    Percentage Area Contribution: For each zone or land cover class, express the variable as a percentage of the total landscape (e.g., “30% of nitrogen retention occurs in protected areas”).

3. Comparisons by Category:

    Inside vs. Outside Protected Areas:
        You can calculate mean, sum, and standard deviation inside and outside protected areas to compare the effectiveness of protection measures in preserving ecosystem services.
    Land Cover Comparison:
        Use zonal means to compare how different land cover types contribute to ecosystem services like sediment retention or distance to beneficiaries.
        Express results as percentages, e.g., “Forests account for 60% of nitrogen retention.”

4. Histogram or Distribution Analysis:

    A frequency distribution showing how the values of each variable are distributed across the entire study area or within protected vs. unprotected zones.
    This can help show if most of the service is concentrated in a few high-value areas or spread out.

5. Change Detection (If Time Series Exists):

    If your rasters cover different time periods, you could show net changes in ecosystem service values (e.g., changes in nitrogen retention) inside and outside protected areas or across different land covers.

6. Normalized or Standardized Scores:

    Convert variables into percentile ranks or Z-scores to facilitate comparisons across variables that might have different units (e.g., how “high” or “low” the ecosystem service is in relation to the mean).

7. Threshold Analysis:

    For variables where a threshold is meaningful (e.g., distance to beneficiaries < 10 km), you can count the number of pixels above or below the threshold in each category (protected areas, land covers).
    For example, "80% of the people dependent on agriculture live within 5 km of high-sediment retention areas."

8. Spatial Autocorrelation (Advanced):

    If you're dealing with spatially explicit relationships, calculating spatial autocorrelation (Moran’s I) might help quantify clustering patterns of ecosystem services inside and outside protected areas.

9. Cumulative Impact Maps:

    Create a composite map where multiple variables are aggregated to show high- vs. low-value areas for ecosystem services. This could simplify complex data into an easily interpretable map for non-specialists.

These summaries will provide enough detail to highlight key insights, while also being interpretable by those without technical expertise. Let me know if you'd like to dive deeper into any specific metric or approach!


Produyces forest loss maps/gain permanences


Country/continent

Biomes. (Köppen)








