---
title: "Clean Mapping Data"
output: html_notebook
---


Here, write down the output of Yesterday meeting and what i need to do. 

1. Organize output Dataframes (make sure that they are complete and correclty named)

2. Make sure all the Vector files have the necessary attributes.

a. Subregion (IPBES) *done*
b. Countries top/less change *running to complete*
c. Biomes 
d. Income level *done*
e. regioin_wb *done*
f. Continent *done*


```{r load polygons, eval=FALSE, include=FALSE}
# Read spatial polygon dataset
# Set attribute column for analysis
#set <- 'Country_id_t.gpkg'
#set <- 'IPBES_subregion.gpkg'
# set <- "Income_grp.gpkg"
# set <- 'Biome.gpkg' 
# set <- "WB_region.gpkg"
# set <- "Continent.gpkg"
# load polygons
poly <- st_read(here('vector', set))#, layer= 'Country_id')) 
#poly <- poly[1]
t <- names(poly)
write.csv(t, here('vector', 'names.csv'))
# select column with the asttribute to iterate through.
col <- "id"
# col <- 'continent'
# #col <- 'subregion'
# col <- "income_grp"
 col <- "WWF_biome"
# col <- 'region_wb'
# col <- "Sub_Region"
cols <- col

```

```{r biomes ghott}
biome_labels <- c(
  "Temperate Grasslands, Savannas & Shrublands" = "Temperate Grasslands",
  "Tropical & Subtropical Moist Broadleaf Forests" = "Moist Broadleaf Forests",
  "Temperate Conifer Forests" = "Temperate Conifers",
  "Tropical & Subtropical Coniferous Forests" = "Tropical Conifers",
  "Rock & Ice" = "Rock & Ice",
  "Boreal Forests/Taiga" = "Boreal Forests",
  "Montane Grasslands & Shrublands" = "Montane Grasslands",
  "Lakes" = "Lakes",
  "Flooded Grasslands & Savannas" = "Flooded Grasslands",
  "Mediterranean Forests, Woodlands & Scrub" = "Mediterranean Forests",
  "Tropical & Subtropical Dry Broadleaf Forests" = "Dry Broadleaf Forests",
  "Temperate Broadleaf & Mixed Forests" = "Temperate Broadleaf",
  "Tundra" = "Tundra",
  "Tropical & Subtropical Grasslands, Savannas & Shrublands" = "Tropical Grasslands",
  "Mangroves" = "Mangroves"
)
```

```{r get amounts of lc chnage}


# get pct as standalone data to add to the table
poly <- st_transform(poly, crs = 5880)
poly$area_ha <- as.numeric(st_area(poly))/10000

plt <- st_drop_geometry(poly)


plt <- as_tibble(plt)
#filter all thins smaller than 150.000 ha (takes away the 54 smallest features) 
plt  <- plt %>% select(1,2, area_ha,contains("pct"))%>% filter(area_ha>400000)


service <- c("Coastal_Protection", "Nitrogen_Export", "Sediment_Export", "Usle", "Nature_Access", "Pollination", "Pot_Sed_ret")
color <- c("#9e9ac8", "#2c944c", "#08306b", "#17c0ff", "#A57C00", "#dd1c77", "#8C510A")
cd <- as_tibble(cbind(service,color))

################## THIS IS VERY IMPPORTAnt. Instead of struggling  with the multiple dataframes, it is easier to load the vector file swith all the column and pivot longer as necessary. Easier to manage, adjust on the fly!
plt <- as_tibble(plt %>%
  pivot_longer(
    cols = c(ends_with("pct_ch")),  # Select all columns ending with "pct_ch"
    names_to = "service",        # New column to store the service names
    values_to = "pct_ch"         # New column to store the percentage change values
  ))

plt <- plt %>%
  mutate(service = str_remove(service, "_pct_ch$"))
plt <- left_join(plt,cd)
plt <- plt%>% filter(!is.na(pct_ch))

plt <- plt %>%
  mutate(biome_short = recode(WWF_biome, !!!biome_labels))
 
```


```{r ggplot 2020, echo=FALSE, fig.height=8, fig.width=14}
#

plot_ecosystem_services <- function(data, var, col) {
  col_sym <- sym(col)  # Convert column name to symbol for dplyr
  
  # Step 1: Prepare Data (Remove NA values and zero mean values, then reorder names)
  data_prepped <- data %>%
    filter(!is.na(mean) & mean > 0 & year == !!year) %>%  # Exclude cases where mean is 0 and filter year
    mutate(temp_col = reorder_within(!!col_sym, -mean, service))  

  # Step 2: Compute min/max values for selected column
  service_range <- data_prepped %>%
    group_by(service) %>%
    summarize(
      min_val = min(pct_chg, na.rm = TRUE),
      max_val = max(pct_chg, na.rm = TRUE),
      .groups = "drop"
    )

  # Step 3: Create "invisible" data for min/max range
  range_data <- service_range %>%
    pivot_longer(cols = c(min_val, max_val), names_to = "range_type", values_to = "mean") %>%
    mutate(year = !!year, temp_col = "dummy")

  # Step 4: Extract top 10 and bottom 10 per selected column
  top_10 <- data_prepped %>%
    group_by(service) %>%
    slice_max(order_by = mean, n = 10, with_ties = TRUE) 

  bottom_10 <- data_prepped %>%
    group_by(service) %>%
    slice_min(order_by = mean, n = 10, with_ties = TRUE)

  # Step 5: Combine only top 10 and bottom 10
  filtered_data <- bind_rows(top_10, bottom_10) %>%
    arrange(service, desc(mean))

  # Step 6: Reorder selected column to appear correctly
  filtered_data <- filtered_data %>%
    mutate(temp_col = reorder_within(!!col_sym, -mean, service))

  # Step 7: Plot the filtered data
  p <- ggplot(filtered_data, aes(x = temp_col, y = mean, fill = color)) +
    geom_bar(stat = "identity", show.legend = FALSE) +
    scale_fill_identity() +
    facet_wrap(~ service, scales = "free") +
    scale_x_reordered() +  
    labs(
      title = paste("Mean Ecosystem Service Values,", year),
      x = col,
      y = "Mean Value"
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(face = "bold", size = 12),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 5)
    )
  
  return(p)
}

# Generate plots for different years using a specified column
p_1992 <- plot_ecosystem_services(zonal_df, 1992, col)
p_2020 <- plot_ecosystem_services(zonal_df, 2020, col)

# Display the plots
print(p_1992)
print(p_2020)


```

## 5. Here, add plots with the differences 

```{r plot chg, fig.height=8, fig.width=14}
library(dplyr)
library(ggplot2)
library(tidytext)
library(rlang)

col <- "biome_short"

plot_es_changes <- function(data, label_col = col, 
                            filter_type = "top_bottom", filter_val = 10) {
  
  label_sym <- sym(label_col)
  
  # Step 1: Filter problematic service and NAs
  filtered_data <- data %>%
    filter(service != "Nature_Access", !is.na(pct_ch))
  
  # Step 2: Apply filtering based on the chosen type
  if (filter_type == "top_bottom") {
    # Top/bottom n observations per service
    top_bottom <- filtered_data %>%
      group_by(service) %>%
      slice_max(pct_ch, n = filter_val, with_ties = FALSE) %>%
      bind_rows(
        filtered_data %>%
          group_by(service) %>%
          slice_min(pct_ch, n = filter_val, with_ties = FALSE)
      ) %>%
      ungroup()
    
  } else if (filter_type == "quantile") {
    # Top/bottom quantile per service (e.g., top/bottom 10%)
    top_bottom <- filtered_data %>%
      group_by(service) %>%
      filter(pct_ch >= quantile(pct_ch, 1 - filter_val, na.rm = TRUE) |
             pct_ch <= quantile(pct_ch, filter_val, na.rm = TRUE)) %>%
      ungroup()
    
  } else if (filter_type == "all") {
    top_bottom <- filtered_data
  } else {
    stop("Invalid `filter_type`. Use 'top_bottom', 'quantile', or 'all'.")
  }
  
  # Step 3: Reorder labels per service
  top_bottom <- top_bottom %>%
    mutate(temp_label = reorder_within(!!label_sym, -pct_ch, service))
  
  # Step 4: Plot
  ggplot(top_bottom, aes(x = temp_label, y = pct_ch, fill = color)) +
    geom_bar(stat = "identity", show.legend = FALSE) +
    scale_fill_identity() +
    scale_x_reordered() +
    facet_wrap(~ service, scales = "free", ncol = 3) +
    labs(
      title = paste("% Change 1992–2020 By", cols, sep= " "),
      x = NULL,
      y = "% Change"
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(face = "bold", size = 10),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
    )
}

# Top/bottom 10 countries per service
#plot_es_changes(plt, label_col = "name_long", filter_type = "top_bottom", filter_val = 10)

# Top/bottom 5% per service
#plot_es_changes(plt, label_col = "iso3", filter_type = "quantile", filter_val = 0.05)

# Show all values
plot_es_changes(plt, filter_type = "all")


```


```{r plot chg, fig.height=8, fig.width=14}


library(dplyr)
library(ggplot2)
library(forcats)



# 1. Filter out "Nature_Access"
filtered_plt <- plt %>%
  filter(service != "Nature_Access" & !is.na(pct_ch))

# 2. Get top 10 and bottom 10 by service
top_bottom <- filtered_plt %>%
  group_by(service) %>%
  slice_max(pct_ch, n = 10, with_ties = FALSE) %>%
  bind_rows(
    filtered_plt %>%
      group_by(service) %>%
      slice_min(pct_ch, n = 10, with_ties = FALSE)
  ) %>%
  ungroup()

# 3. Reorder country names *within each service*
top_bottom <- top_bottom %>%
  mutate(name_long = reorder_within(name_long, -pct_ch, service))  # descending order


ggplot(top_bottom, aes(x = name_long, y = pct_ch, fill = color)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  scale_fill_identity() +
  scale_x_reordered() +  # <<-- THIS enables correct facet ordering
  facet_wrap(~ service, scales = "free", ncol = 3) +
  labs(
    title = "% Change 1992-2020 By Income Group
    x = NULL,
    y = "% Change"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)
  )
```








write.csv(plt, here('output_data', paste0('pct_chg_',cols, '.csv')))



Percentage of Change. This is the key. I almot have it for everything! 


Variables: 

Pot. Sediment Retention for Each Grouping (Initial/ Final)

######################################################################

Time Series!!!

% percentage of change in the charts.... amount.





I extract change stats (stdev + mean) for 

- Subregions (repeat for stdev)
- Biome (running right now, lost connection to the internet, but just adusting to add the new columns to the vector file)
- Country. Here to ee those with the most change (%) It's the same, its a ratio, so we don't need to put too much effort normalizing or stuff
- Income
 
 
Get all the differentials and include % change in service and in export (so retention and export)....make sure i know which inputs am i using.

Paper: Review question to make sure that i

Pending:

Incorporate the Fertilier EASA database to improve Fertilizer (N) mode.

M<ake sure i m using the right dstasert and olumn! Wrtite tomorrow to justin!!!

Connecting LC analysis -> is it possible to identify s relationship betwe them?

How do the amoung of change between nwture tono nature relatw with the amount of change in the provision. This is zero banal

Get emtrics for Retention, export and the calcualtion (in short repeat the whole thing. again)



Re run the export part

```{r create list of columns}

poly <- st_read('/Users/rodriguez/Global_ES_TS/global_NCP/vector/vector_f/Continent.gpkg')

serv <- names(poly)
serv <- serv[-c(1,30)]
serv


#Get filenames (add paths to stored data)





```
