---
title: "Clean Mapping Data"
output: html_notebook
---


Here, write down the output of Yesterday meeting and what i need to do. 

1. Organize output Dataframes (make sure that they are complete and correclty named)

2. Make sure all the Vector files have the necessary attributes.

a. Subregion (IPBES) *done*
b. Countries top/less change *running to complete*
c. Biomes 
d. Income level *done*
e. regioin_wb *done*
f. Continent *done*


```{r load polygons, eval=FALSE, include=FALSE}
# Read spatial polygon dataset
# Set attribute column for analysis
# By individual territory; "Country" 
set <- 'Country_id.gpkg'
# By Continent
#set <- 'Continent.gpkg' 
# By Subregion
set <- 'subregion.gpkg'
## By Income Group
set <- "Income_grp.gpkg"
# By WWF biome
set <- 'Biome.gpkg' 
# by WB Region
set <- "WB_region.gpkg"
# By Continent
set <- "Continent.gpkg"
# load polygons
poly <- st_read(here('vector', set)) 
#poly <- poly[1]
t <- names(poly)
write.csv(t, here('vector', 'names.csv'))
# select column with the asttribute to iterate through.
col <- "id"
col <- 'continent'
#col <- 'subregion'
col <- "income_grp"
col <- "WWF_biome"
#col <- "WWF_biome"
col <- 'region_wb'
cols <- col

```



```{r get amounts of lc chnage}





Percentage of Change. This is the key. I almot have it for everything! 


Variables: 

Pot. Sediment Retention for Each Grouping (Initial/ Final)

######################################################################

Time Series!!!

% percentage of change in the charts.... amount.





I extract change stats (stdev + mean) for 

- Subregions (repeat for stdev)
- Biome (running right now, lost connection to the internet, but just adusting to add the new columns to the vector file)
- Country. Here to ee those with the most change (%) It's the same, its a ratio, so we don't need to put too much effort normalizing or stuff
- Income
 
 
Get all the differentials and include % change in service and in export (so retention and export)....make sure i know which inputs am i using.

Paper: Review question to make sure that i

Pending:

Incorporate the Fertilier EASA database to improve Fertilizer (N) mode.

M<ake sure i m using the right dstasert and olumn! Wrtite tomorrow to justin!!!

Connecting LC analysis -> is it possible to identify s relationship betwe them?

How do the amoung of change between nwture tono nature relatw with the amount of change in the provision. This is zero banal

Get emtrics for Retention, export and the calcualtion (in short repeat the whole thing. again)



Re run the export part

```{r create list of columns}

poly <- st_read('/Users/rodriguez/Global_ES_TS/global_NCP/vector/vector_f/Continent.gpkg')

serv <- names(poly)
serv <- serv[-c(1,30)]
serv


#Get filenames (add paths to stored data)





```
