---
title: "Beneficiary consolidation
output: html_notebook
---
# Zonal Stat Exctraction and exploratory analysis 

```{r setup, message=FALSE, warning=FALSE}
library(terra)
library(sf)
library(dplyr)
library(ggplot2)
library(glue)
library(tidyr)
library(purrr)
library(diffeR)
library(here)
library(stringr)
library(tidytext)
library(rlang)
library(tidyr)
library(forcats)
# source the helper functions
source(here("R", "utils_lcc_metrics.R"))
source(here("R","utils_pct_change.R"))
```


THis cleans the data/consolidates columns that are going to be analyzed keep at hands, the rest is sent to a storage. there should be a database structure here. 
```{r load procesed data and build final gpkgs}

inpath <- here("summary_pipeline_workspace")

gp <- list.files(inpath, pattern = ".gpkg")

gp <- file.path(inpath, gp)

synth <- lapply(gp, st_read)
synth <- lapply(synth, function(s){
  s <- s %>% mutate(fid=row_number()-1)
})

synth <- lapply(synth, function(s){
  s <- st_drop_geometry(s)
})

synth[[1]] <- select(synth[[1]], fid, GHS_BUILT_S_E2020_GLOBE_R2023A_4326_3ss_V1_0_mean)
synth[[2]] <- select(synth[[2]], fid, rast_gdpTot_1990_2020_30arcsec.7_mean)
synth[[3]] <- select(synth[[3]], fid, GHS_POP_E2020_GLOBE_R2023A_4326_3ss_V1_0.1_mean)
synth[[4]] <- select(synth[[4]], fid, GlobPOP_Count_30arc_2020_I32.1_sum)
synth[[5]] <- select(synth[[5]], fid, hdi_raster_predictions_2020.1_mean)
synth[[6]] <- select(synth[[6]], fid,farmsize_mehrabi.1_mean)


synth <- reduce(synth, left_join, by= "fid")
# Load the polygons with the calcualted ES change variables
synth_ES <- st_read(here('vector',"hydrosheds_lv6_synth.gpkg"))
# Keep the columns of interest (original hydrosheds and pct_ch in this case)
synth_ES <- dplyr::select(synth_ES, 1:13, contains("pct_ch"))

synth_ES <- synth_ES %>% mutate(fid = row_number()-1)


synth_ES <- left_join(synth_ES, synth, by = 'fid')
synth_ES$fid <- NULL
############# CURRENT OUTPUT WITH THE FINAL BANDS TO ANALYZE (SCATTERPLOTS, VIOLINS)
st_write(synth_ES, here('vector', "hydrosheds_lv6_synth.gpkg"), layer = "benef_analysis", append = TRUE)
```

WHY is this here? I think is mee trying (again) to get straight what i have already processed and see if sme day it doesn to take me days to retake the work when o finally extract the data i need. 
```{r pivot and preprare to plot the data}

tt <- st_layers('/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_NCP/vector/hydrosheds_lv6.gpkg')
t2 <- st_layers('/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_NCP/vector/hydrosheds_lv_tmp.gpkg')
t3 <- st_layers('/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_NCP/vector/hydrosheds_lv_6_tmp.gpkg')
t <- st_read('/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_NCP/vector/hydrosheds_lv6.gpkg', layer = "ch_92_2020")
tt2 <-st_read('/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_NCP/vector/hydrosheds_lv_tmp.gpkg', layer = "clipped") 

```

Here, another dialogue to input the variable names, plot colors and set the factor order.
```{r, fig.height=8, fig.width=14  }
plt <- as_tibble(read.csv(here('output_data', "metrics_change_hb_lev_6.csv")))
# plt[1] <- NULL Update this with the names of the variables being assessed. There should be a less annoying way to do this.

service <- unique(plt$service)
# Actually, GDP and pop density are not services here, we can in theory dorp them
service <- c("GHS_BUILT","GDP","GHS_POP","HDI", "Farm_size", "GlobPOP")
 color <- c("#9e9ac8", "#dd1c77","#2c944c", "#2c711c","#08306b", "#02606b",)# "#d4ac0d", "#85929e")
 service_levels <- service
 cd <- as_tibble(cbind(service,color))
 
 #check and join. Think about a more intuitive/relatable) variable order to plot. 
 plt <- left_join(plt, cd)

ly <- st_layers('/Users/sputnik/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_NCP/vector/hydrosheds_lv6_synth.gpkg')
synth_ES <- st_read('/Users/sputnik/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_NCP/vector/hydrosheds_lv6_synth.gpkg', layer = "benef_analysis")
spring_ES <- st_read('/Users/sputnik/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_NCP/vector/hydrosheds_lv6_synth.gpkg', layer = "hydrosheds_lv6_synth")
```

```{r chunk1, eval=FALSE, include=FALSE}
# Get the columns that we need (can;t believe i have been struggling with this for more than 2 months)
# 
spring_ES <- spring_ES[c(1:17,36:40,44, 45)]
synth_ES <- synth_ES[c(1,7,21,22,23,24,25,26)]
#  get pct as standalone data to add to the table
# Most recent version of synthesis data:
 
 # calculate POP Density
synth_ES <- synth_ES %>%
  mutate(GlobPOP_sqkm = GlobPOP_Count_30arc_2020_I32.1_sum / SUB_AREA)

 plt_spring <- st_drop_geometry(spring_ES) # if necessary when i load 
 plt_synth <- st_drop_geometry(synth_ES) # if necessary when i load 

```

```{r join tables}

plt_synth$HYBAS_ID <- as.numeric(plt_synth$HYBAS_ID)
plt <- left_join(plt, plt_synth) 
```

 This could be a dialogue box!!!
```{r remove lowestvalues to reduce artifacts}
# Calculate pseudo_log val for display purposes. 
plt <- plt %>%
  mutate(
    service = factor(service, levels = service_levels),
    pct_ch_trans = pseudo_log(pct_ch)
  )

# apply filter (5% or 2%) # Remove the smallest basins. 
area_threshold <- quantile(plt$SUB_AREA, probs = 0.05, na.rm = TRUE)
# Filter the dataset
df <- plt %>% filter(SUB_AREA > area_threshold)

write.csv(plt, here("output_data", "serv_benef_synth_01.csv"))
# get the filtered locations
```



## 5. add plots with the differences 

Convert into function. We don't need this right now!!!
```{r, fig.height=8, fig.width=14}
col <- "HYBAS_ID"
# create thew function. THIS is what i need to get righ (save as a function)
plot_es_changes <- function(data, label_col, #labelr col has the column with the ind IDs.
                            filter_type = "top_bottom", filter_val = 10) {
  label_sym <- sym(label_col)
  # Step 2: Apply filtering based on the chosen type
  if (filter_type == "top_bottom") {
    # Top/bottom n observations per service
    top_bottom <- data %>%
      group_by(service) %>%
      slice_max(pct_ch, n = filter_val, with_ties = FALSE) %>%
      bind_rows(
        data %>%
          group_by(service) %>%
          slice_min(pct_ch, n = filter_val, with_ties = FALSE)
      ) %>%
      ungroup()

  } else if (filter_type == "quantile") {
    # Top/bottom quantile per service (e.g., top/bottom 10%)
    top_bottom <- data %>%
      group_by(service) %>%
      filter(pct_ch >= quantile(pct_ch, 1 - filter_val, na.rm = TRUE) |
             pct_ch <= quantile(pct_ch, filter_val, na.rm = TRUE)) %>%
      ungroup()

  } else if (filter_type == "all") {
    top_bottom <- data
  } else {
    stop("Invalid `filter_type`. Use 'top_bottom', 'quantile', or 'all'.")
  }

  # Step 3: Reorder labels per service ## This is stupid. Better to order the labels outside
  top_bottom <- top_bottom %>%
    mutate(temp_label = reorder_within(!!label_sym, -pct_ch, service))

  # Step 4: Plot
  ggplot(top_bottom, aes(x = temp_label, y = pct_ch)) +#, fill = color)) +
    geom_bar(stat = "identity", show.legend = FALSE) +
    scale_fill_identity() +
    scale_x_reordered() +
    facet_wrap(~ service, scales = "free", ncol = 3) +
    labs(
      #title = paste("% Change 1992–2020 By", cols, sep= " "),
      x = NULL,
      y = "% Change"
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(face = "bold", size = 10),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
    )
}

# Top/bottom 10 countries per service Send to the docuentnbsation, but we don't need to keep this here, just know thqat i is possible
# There are several filter types availalbe, jus add to the docuemtnaiton 
#plot_es_changes(plt_long, label_col = col, filter_type = "top_bottom", filter_val = 10)

# Top/bottom 5% per service
plot_es_changes(df, label_col = col, filter_type = "quantile", filter_val = 0.5)
# 
# # Show all values
# plot_es_changes(plt, filter_type = "all")
```


## Scatterplots

```{r other plot, fig.height=7, fig.width=7, warning=FALSE}

# 1. Filter to top/bottom 10% based on pct_ch # per service not anymore change
df_top_bottom <- df %>%
  group_by(service) %>%  # Do it separately for each service (optional, if needed)
  filter(pct_ch >= quantile(pct_ch, 0.90, na.rm = TRUE) |
         pct_ch <= quantile(pct_ch, 0.10, na.rm = TRUE)) %>%
  ungroup()
#n <- floor(nrow(df) * 0.1)  # 10% of total rows
######



###### HERE IS OUR ISSEU. WE JUST NEED TO KNOW WHICH COLUMN NAME COMES HERE INSTEAD IF DIR CH (FROM LAND COVER ANALYSIS TO DO BENEFICIARIES ANALYSISI)
# df_top_bottom <- df %>%
#   slice_max(dir_ch_2, n = n, with_ties = FALSE) %>% # BY SERVICE AND GROUPED, NOT AROUND LAND COVER.
#   bind_rows(
#     df %>% slice_min(dir_ch_2, n = n, with_ties = FALSE)
#   )
# 
# df_top_bottom <- df %>%
#   group_by(service) %>%
#   group_modify(~ bind_rows(
#     slice_max(.x, dir_ch_2, n = floor(0.1 * nrow(.x)), with_ties = FALSE),
#     slice_min(.x, dir_ch_2, n = floor(0.1 * nrow(.x)), with_ties = FALSE)
#   )) %>%
#   ungroup()



# 2. Optional: relevel services if you want
service <- c("Coastal_Protection","Pollination","N_export","N_Ret_Ratio", "Sed_export", "Sed_Ret_Ratio")
 color <- c("#9e9ac8", "#dd1c77","#2c944c", "#2c711c","#08306b", "#02606b")
 service_levels <- service
df_top_bottom <- df_top_bottom %>%
  mutate(
    service = factor(service, levels = service_levels),
    pct_ch_trans = pseudo_log(pct_ch)  # Apply pseudo-log transformation
  )

# 3. List of Benef data to analyze
lc_metrics <- c("rast_gdpTot_1990_2020_30arcsec.7_mean","GHS_POP_E2020_GLOBE_R2023A_4326_3ss_V1_0.1_mean","GlobPOP_Count_30arc_2020_I32.1_sum","hdi_raster_predictions_2020.1_mean","farmsize_mehrabi.1_mean","GlobPOP_sqkm", "dir_ch_2")

Benef
#df_top_bottom <- filter(df_top_bottom, dir_ch_2 != 0)
# 4. Create the scatterplots
for (metric in lc_metrics) {
  p1 <- df_top_bottom %>%
    ggplot(aes_string(x = metric, y = "pct_ch_trans")) +
    geom_hex(bins = 60) +  # Hexbin layer
    scale_fill_viridis_c(option = "D", direction = 1) +  # Use viridis scale
    facet_wrap(~ service, scales = "free",ncol=2) +
    labs(
      title = paste("Benef ", "vs. ES % Change"),
      x = metric,
      y ="% Change in Ecosystem Service Provision, 1992-2020",
      fill = "Density"
    ) +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5),
      axis.text = element_text(size = 9)
    )

  print(p1)
}

```

```{r scatt_ ver 2, fig.height=9, fig.width = 12}

#df_top_bottom <- filter(df_top_bottom, dir_ch_2 != 0)
# 4. Create the scatterplots
for (metric in lc_metrics) {
  p2 <- df_top_bottom %>%
    ggplot(aes_string(x = "pct_ch_trans", y = metric)) +
    geom_hex(bins = 60) +  # Hexbin layer
    scale_fill_viridis_c(option = "D", direction = 1) +  # Use viridis scale
    facet_wrap(~ service, scales = "free",ncol=2) +
    labs(
      title = paste("Benef ", "vs. ES % Change"),
      x = "% Change in Ecosystem Service Provision, 1992-2020",
      y = metric,
      fill = "Density"
    ) +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5),
      axis.text = element_text(size = 9)
    )

  print(p2)
}
```
```{r export_scatterplots, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(viridis)

# Set PDF output: landscape (width > height), e.g., A4 landscape is ~11.7 x 8.3 inches
pdf(here("output_charts", "scatterplots_by_service5.pdf"), width = 8,3, height = 11.7)

# Loop through each LC metric and generate a faceted plot
for (metric in lc_metrics) {
  p <- df %>%
    ggplot(aes_string(x = "pct_ch", y = metric)) +
    geom_hex(bins = 50) +  # density shading
    scale_fill_viridis_c(option = "D", direction = 1) +
    facet_wrap(~ service, scales = "free", ncol = 2) +
    labs(
      title = paste("Benef", "vs. ES % Change"),
      x = "% Change in Ecosystem Service Provision, 1992–2020",
      y = metric,
      fill = "Density"
    ) +
   # theme_minimal() +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5),
      axis.text = element_text(size = 9)
    )

  print(p)
}

# Close the PDF device
dev.off()
```



```{r fig.height=8, fig.width=10, warning=FALSE}
# 1. Define desired facet order
df <- plt %>% filter(SUB_AREA > area_threshold)
# df <- df %>%
#   filter(id %in% top_bottom_10k$id)
# 2. Apply pseudo-log transformation
pseudo_log <- function(x) sign(x) * log1p(abs(x))

df <- df %>%
  mutate(
    service = factor(service, levels = service_levels),
    pct_ch_trans = pseudo_log(pct_ch)
  )

# 3. Reorder countries based on dir_ch_2 (descending)
df <- df %>%
  mutate(ee_r264_name = reorder(ee_r264_name, -dir_ch_2))
# 4. Create plot
p <- ggplot(df, aes(x = ee_r264_name, y = pct_ch_trans, fill = ee_r264_name)) +
  geom_violin(trim = FALSE, scale = "width") +
  facet_wrap(~ service, scales = "free_y") +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "Pseudo-log Scaled % Change by Service ",
    x = NULL,
    y = "Transformed % Change"
  ) +
  #theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )
# 4. Create plot
p2 <- ggplot(df, aes(x = ee_r264_name, y = pct_ch, fill = ee_r264_name)) +
  geom_violin(trim = FALSE, scale = "width") +
  facet_wrap(~ service, scales = "free_y") +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "% Change by Service ",
    x = NULL,
    y = "Transformed % Change"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )
p
p2

```
######################################################################################
######################################################################################



```{r , fig.height=9, fig.width=7, warning=FALSE}

df <- plt %>% filter(SUB_AREA > area_threshold)
#df <- plt %>% filter(id %in% top_bottom_10k$id)
# List of LC metrics to analyze
lc_metrics <- c("dir_ch_2")
pseudo_log <- function(x) sign(x) * log1p(abs(x))

df <- df %>%
  mutate(
    service = factor(service, levels = service_levels),
    pct_ch_trans = pseudo_log(pct_ch)
  )
# Loop over each LC metric and generate a faceted scatterplot
for (metric in lc_metrics) {
  p <- df %>%
    ggplot(aes_string(x = metric, y = "pct_ch_trans", color = "color")) +
    geom_point(alpha = 0.5, size = 1.2) +
    facet_wrap(~ service, scales = "free") +
    scale_color_identity() +
    labs(
      title = paste("Relationship Between", metric, "and ES % Change"),
      x = metric,
      y = "% Change in Ecosystem Service"
    ) +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5),
      axis.text = element_text(size = 9)
    )

  print(p)
}
```


```{r sctp, fig.height=9, fig.width=7, warning=FALSE}


# List of LC metrics to analyze
lc_metrics <- c("dir_ch_2")

df_1 <- df %>% filter(service != "GDP") %>% filter(service != "Pop_sqkm")



 for (metric in lc_metrics) {
  p1 <- df_1 %>%
    ggplot(aes_string(x = metric, y = "pct_ch_trans")) +
    geom_hex(bins = 60) +  # Hexbin layer
    scale_fill_viridis_c(option = "D", direction = -1) +  # Use viridis scale
    facet_wrap(~ service, scales = "free",ncol=2) +
    labs(
      title = paste("Density of", metric, "vs. ES % Change"),
      x = metric,
      y = "% Change in Ecosystem Service",
      fill = "Bin Count"
    ) +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5),
      axis.text = element_text(size = 9)
    )

  print(p1)
}
 
df_split <- split(df_1, df_1$service)


```


# Scatterplots change 2

```{r unnamedchunk, fig.height=9, fig.width=7, warning=FALSE}

# LC metrics of interest
lc_metrics <- c("dir_ch_2")

# Step 1: Remove top 2% outliers for each service
filtered_plt <- plt %>%
  group_by(service) %>%
  mutate(threshold = quantile(pct_ch, 0.98, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(pct_ch <= threshold)

# Step 2: Apply pseudo-log transformation
filtered_plt <- filtered_plt %>%
  mutate(pct_ch_log = sign(pct_ch) * log1p(abs(pct_ch)))

# Step 3: Loop through each LC metric and create faceted plots
for (metric in lc_metrics) {
  p <- ggplot(filtered_plt, aes_string(x = metric, y = "pct_ch_log", color = "color")) +
    geom_point(alpha = 0.5, size = 1.2) +
    facet_wrap(~ service, scales = "free") +
    scale_color_identity() +
    labs(
      title = paste("% Change vs", metric),
      x = metric,
      y = "Log(% Change in ES)"
    ) +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5),
      axis.text = element_text(size = 9)
    )

  print(p)
}
```

```{r eval=FALSE, fig.height=9, fig.width=7, warning=FALSE}


# LC metrics of interest
lc_metrics <- c("dir_ch_2")

# Step 1: Remove top 2% outliers for each service
filtered_plt <- plt %>%
  group_by(service) %>%
  mutate(threshold = quantile(pct_ch, 0.98, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(pct_ch <= threshold)

# Step 3: Loop through each LC metric and create faceted plots
for (metric in lc_metrics) {
  p3 <- ggplot(filtered_plt, aes_string(x = metric, y = "pct_ch", color = "color")) +
    geom_point(alpha = 0.5, size = 1.2) +
    facet_wrap(~ service, scales = "free") +
    scale_color_identity() +
    labs(
      title = paste(" % Change vs", metric),
      x = metric,
      y = "% Change in ES"
    ) +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5),
      axis.text = element_text(size = 9)
    )

  print(p3)
}
```


```{r eval=FALSE, fig.height=8, fig.width=14, include=FALSE}

# apply filter (5 or 2%) # Remove the smalles basins. That help. Also, it seems that some biomes are more prone to artifacts than others!!!
area_threshold <- quantile(plt$SUB_AREA, probs = 0.05, na.rm = TRUE)
# Filter the dataset
df <- plt %>% filter(SUB_AREA > area_threshold)


t_5 <- ggplot(df, aes(x = service, y = pct_ch, fill = color)) +
  geom_boxplot() +  # or geom_violin()
  scale_fill_identity() +
  facet_wrap(~ service, scales = "free_y") +
  labs(
    title = "Distribution of % Change by Service (Excl. Smallest 5% of Basins)",
    x = NULL,
    y = "Percentage Change"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )

library(ggplot2)
library(dplyr)

# Custom pseudo-log transformation
pseudo_log <- function(x) sign(x) * log1p(abs(x))

# Apply  the filtered dataframe
df <- plt %>%
  mutate(pct_ch_trans = pseudo_log(pct_ch))

# Plot with transformed values
p <- ggplot(df, aes(x = service, y = pct_ch_trans, fill = color)) +
  geom_violin(trim = FALSE, scale = "width") +
  scale_fill_identity() +
  facet_wrap(~ service, scales = "free_y") +
  labs(
    title = "Pseudo-log Scaled % Change by Service",
    x = NULL,
    y = "Transformed % Change"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )
p
```




#######################

## 5. Here, add plots with the differences 

```{r eval=FALSE, fig.height=8, fig.width=14, include=FALSE}
print(col)
print(names(df))
stopifnot(col %in% names(df))

plot_es_changes <- function(data, label_col,
                            filter_type = "top_bottom", filter_val = 10) {
  label_sym <- sym(label_col)

  # Step 2: Apply filtering based on the chosen type
  if (filter_type == "top_bottom") {
    # Top/bottom n observations per service
    top_bottom <- data %>%
      group_by(service) %>%
      slice_max(pct_ch, n = filter_val, with_ties = FALSE) %>%
      bind_rows(
        data %>%
          group_by(service) %>%
          slice_min(pct_ch, n = filter_val, with_ties = FALSE)
      ) %>%
      ungroup()

  } else if (filter_type == "quantile") {
    # Top/bottom quantile per service (e.g., top/bottom 10%)
    top_bottom <- data %>%
      group_by(service) %>%
      filter(pct_ch >= quantile(pct_ch, 1 - filter_val, na.rm = TRUE) |
             pct_ch <= quantile(pct_ch, filter_val, na.rm = TRUE)) %>%
      ungroup()

  } else if (filter_type == "all") {
    top_bottom <- data
  } else {
    stop("Invalid `filter_type`. Use 'top_bottom', 'quantile', or 'all'.")
  }

  # Step 3: Reorder labels per service
  top_bottom <- top_bottom %>%
    mutate(temp_label = reorder_within(!!label_sym, -pct_ch, service))

  # Step 4: Plot
  ggplot(top_bottom, aes(x = temp_label, y = pct_ch, fill = color)) +
    geom_bar(stat = "identity", show.legend = FALSE) +
    scale_fill_identity() +
    scale_x_reordered() +
    facet_wrap(~ service, scales = "free", ncol = 3) +
    labs(
      #title = paste("% Change 1992–2020 By", cols, sep= " "),
      x = NULL,
      y = "% Change"
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(face = "bold", size = 10),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
    )
}

# Top/bottom 10 countries per service
plot_es_changes(df, label_col = col, filter_type = "top_bottom", filter_val = 10)

# Top/bottom 5% per service
#plot_es_changes(df, label_col = col, filter_type = "quantile", filter_val = 0.1)

# Show all values
#plot_es_changes(plt, filter_type = "all")
```


## Scatterplots

```{r}


# List of LC metrics to analyze
lc_metrics <- c("Gain_2", "Persistence_2", "Loss_2", "dir_ch_2")

# Loop over each LC metric and generate a faceted scatterplot
for (metric in lc_metrics) {
  p <- plt %>%
    ggplot(aes_string(x = metric, y = "pct_ch", color = "color")) +
    geom_point(alpha = 0.5, size = 1.2) +
    facet_wrap(~ service, scales = "free") +
    scale_color_identity() +
    labs(
      title = paste("Relationship Between", metric, "and ES % Change"),
      x = metric,
      y = "% Change in Ecosystem Service"
    ) +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5),
      axis.text = element_text(size = 9)
    )

  print(p)
}
```


# Scatterplots change 2

```{r, fig.height=8, fig.width=14}


# LC metrics of interest
lc_metrics <- c("Gain_2", "Persistence_2", "Loss_2", "dir_ch_2")

# Step 1: Remove top 2% outliers for each service
filtered_plt <- plt %>%
  group_by(service) %>%
  mutate(threshold = quantile(pct_ch, 0.98, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(pct_ch <= threshold)

# # Step 2: Apply pseudo-log transformation
# filtered_plt <- filtered_plt %>%
#   mutate(pct_ch_log = sign(pct_ch) * log1p(abs(pct_ch)))

# Step 3: Loop through each LC metric and create faceted plots
for (metric in lc_metrics) {
  p <- ggplot(filtered_plt, aes_string(x = metric, y = "pct_ch", color = "color")) +
    geom_point(alpha = 0.5, size = 1.2) +
    facet_wrap(~ service, scales = "free") +
    scale_color_identity() +
    labs(
      title = paste("Log-Transformed % Change vs", metric),
      x = metric,
      y = "Log(% Change in ES)"
    ) +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5),
      axis.text = element_text(size = 9)
    )

  print(p)
}
```

```{r, fig.height=8, fig.width=14}
# library(dplyr)
# library(ggplot2)
# 
# plt <- as_tibble(read.csv(here('output_data', "metrics_change_hb_lev_6.csv")))
# plt[1] <- NULL 
# # apply filter (5 or 2%) # Remove the smalles basins. That help. Also, it seems that some biomes are more prone to artifacts than others!!!
# area_threshold <- quantile(plt$SUB_AREA, probs = 0.05, na.rm = TRUE)
# # Filter the dataset
# df <- plt %>% filter(SUB_AREA > area_threshold)
# 
# 
# t_5 <- ggplot(df, aes(x = service, y = pct_ch, fill = color)) +
#   geom_boxplot() +  # or geom_violin()
#   scale_fill_identity() +
#   facet_wrap(~ service, scales = "free_y") +
#   labs(
#     title = "Distribution of % Change by Service (Excl. Smallest 5% of Basins)",
#     x = NULL,
#     y = "Percentage Change"
#   ) +
#   theme_minimal() +
#   theme(
#     axis.text.x = element_blank(),
#     strip.text = element_text(face = "bold"),
#     legend.position = "none"
#   )
# 
# library(ggplot2)
# library(dplyr)
# 
# # Custom pseudo-log transformation
# pseudo_log <- function(x) sign(x) * log1p(abs(x))
# 
# # Apply to the filtered dataframe
# df <- plt %>% 
#   mutate(pct_ch_trans = pseudo_log(pct_ch))
# 
# # Plot with transformed values
# p <- ggplot(df, aes(x = service, y = pct_ch_trans, fill = color)) +
#   geom_violin(trim = FALSE, scale = "width") +
#   scale_fill_identity() +
#   facet_wrap(~ service, scales = "free_y") +
#   labs(
#     title = "Pseudo-log Scaled % Change by Service",
#     x = NULL,
#     y = "Transformed % Change"
#   ) +
#   theme_minimal() +
#   theme(
#     axis.text.x = element_blank(),
#     strip.text = element_text(face = "bold"),
#     legend.position = "none"
#   )
# p
```





```{r fig.height=8, fig.width=10, warning=FALSE}
library(dplyr)
library(ggplot2)
library(forcats)

# 1. Filter the dataframe for top and bottom 10 by dir_ch_2
df_top_bottom <- df %>%
  arrange(desc(dir_ch_2)) %>%
  slice_head(n = 10) %>%
  bind_rows(
    df %>%
      arrange(dir_ch_2) %>%
      slice_head(n = 10)
  )

# 2. Reorder countries inside the filtered data
df_top_bottom <- df_top_bottom %>%
  mutate(ee_r264_name = fct_reorder(ee_r264_name, -dir_ch_2))
df_top_bottom <- df_top_bottom %>% filter(service != "GDP") %>% filter(service != "Pop_sqkm")

# 3. Create plot (pseudo-log scale)
p <- ggplot(df_top_bottom, aes(x = ee_r264_name, y = pct_ch_trans, fill = service)) +
  geom_violin(trim = FALSE, scale = "width") +
  facet_wrap(~ service, scales = "free_y") +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "Pseudo-log Scaled % Change by Service (Top and Bottom 10 Countries)",
    x = NULL,
    y = "Transformed % Change"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )

# 4. (Optional) Plot regular % change
p2 <- ggplot(df_top_bottom, aes(x = ee_r264_name, y = pct_ch, fill = service)) +
  geom_violin(trim = FALSE, scale = "width") +
  facet_wrap(~ service, scales = "free_y") +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "% Change by Service (Top and Bottom 10 Countries)",
    x = NULL,
    y = "% Change"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )

# 5. Print
p
# p2  # if you also want the regular % plot

```




