---
title: "R Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
library(sf)             # Spatial vector operations
library(dplyr)          # Data manipulation
library(terra)          # Raster processing
library(exactextractr)  # Zonal statistics
library(ggplot2)        # Visualization
library(forcats)        # Factor reordering
library(tidytext)
library(here)           # Managing paths
library(patchwork)      # Combining plots
library(tidyr)
library(parallel)       # Parallel computing
library(purrr)
```


## **1. Compute Differences**



```{r get differences, eval=FALSE, include=FALSE}

# Define service names and colors
service <- rep(c("Coastal_Protection", "Nitrogen_Export", "Sediment_Export", "Usle", "Nature_Access", "Pollination"), each=2)
year <- rep(c(1992,2020), 6)
filename <- as_tibble(cbind(service, raster_name, year))

# create a table with the content to fill the table
color <- rep(c("#9e9ac8", "#2c944c", "#08306b", "#17c0ff", "#A57C00", "#dd1c77"), each=2)
cd <- as_tibble(cbind(filename, color))
rm(filename)
cd

outpath <- here("input_ES")
m_index <- length(rasters_list) %/% 2
lab <- unique(service)
lab <- gsub(" ", "_", lab)


l1 <- rasters_list[seq(1,length(rasters_list), by=2)]
l2 <- rasters_list[seq(2,length(rasters_list), by=2)]
 # test with future_map2 next time to get this to run faster. 

diffs <- map2(l2, l1, ~ .x - .y)

# Export difference rasters
map(1:length(diffs), function(x) writeRaster(diffs[[x]], paste0(outpath, "/change_calc/", lab[x], '_diff.tif'),overwrite=TRUE))
```



## 2.  Prepare Potential of Sed retention data.

```{r getpotential Sed retention , eval=FALSE, include=FALSE}
inpath <- here("input_ES", "Pot_Sed_retention")
tiffes <- file.path(inpath, list.files(paste0(inpath),pattern= 'tif$'))
filename<- basename(tiffes)
# Use terra::rast() to load each file as a SpatRaster object


rasters_list <- lapply(tiffes, rast)
rcl <- matrix(c(
  -Inf, 0, NA   # Any value from 0 to Infinity becomes 1
), ncol = 3, byrow = TRUE)

rasters_list <- lapply(rasters_list, function(r){
  r <- classify(r,rcl, right=FALSE)
},mc.cores = num.cores))



m_index <- length(rasters_list) %/% 2
#lab <- unique(service)
l1 <- rasters_list[seq(1,length(rasters_list), by=2)]
l2 <- rasters_list[seq(2,length(rasters_list), by=2)]
 # test with future_map2 next time to get this to run faster. 

pot_SedRet1 <- (l1[[2]]-l1[[1]])/l1[[2]]


pot_SedRet2 <- (l2[[2]]-l2[[1]])/l2[[2]]

psr_92 <- classify(pot_SedRet1,rcl, right=FALSE)
psr_20 <- classify(pot_SedRet2,rcl, right=FALSE)

writeRaster(psr_92, paste0(inpath, '/', 'pot_SedRet_92.tif'), overwrite=TRUE)
writeRaster(psr_20, paste0(inpath, '/', 'pot_SedRet_20.tif'), overwrite=TRUE)
diff <- psr_92 - psr_20
writeRaster(diff, paste0(inpath, '/', 'pot_Sed_r_diff.tif'), overwrite)

```



# Multi Temporal Analysis on NDR.


## 3. Calculte N export potential
Here , I need to add the target rasters and organzie them by lists. i think i already did that but need to revisit here
```{r prepare retention potential per year}

inpath <- here('input_ES', 'mod_n_ret')


#load the rasters with the calculated differences 
tiffes <- file.path(inpath, list.files(paste0(inpath),pattern= 'global'))
tiffes <- tiffes[-c(1,3,5,9,11,15)]
tiffes <- tiffes[c(1,6,2,7,3,8,4,9,5,10)]
filename<- basename(tiffes)
 # here, adjust to call the correct ones and buld thje corresponding lists.


# There must be a better way to do this, but basically the datatype in wehich geosharding/gdealtrnaslatre export result in issues when performing map algebra. I think the best way to fix this is someho to specify the data type. Keeep in mind for the next iteratioons, but for some reson it dd not iterate well (lapply/mclapply) so i ended up doing this stupid think i had to finish. 
# The important thing is i did that and we don't need anymore. (mnaybe for map making) 

rasters_list <- lapply(tiffes, rast)
rcl <- matrix(c(
  -Inf, 0, NA,    # Values strictly less than 0 become NA
  5000, Inf, NA   # Values greater than 3000 become NA
), ncol = 3, byrow = TRUE)

r2 <- classify(rasters_list[[2]],rcl)
writeRaster(r2, paste0(inpath, '/', names(r2), '_rec.tif'))
r2 <- classify(rasters_list[[3]],rcl)
writeRaster(r2, paste0(inpath, '/', names(r2), '_rec.tif'))
r2 <- classify(rasters_list[[4]],rcl)
writeRaster(r2, paste0(inpath, '/', names(r2), '_rec.tif'))
r2 <- classify(rasters_list[[5]],rcl)
writeRaster(r2, paste0(inpath, '/', names(r2), '_rec.tif'))
r2 <- classify(rasters_list[[6]],rcl)
writeRaster(r2, paste0(inpath, '/', names(r2), '_rec.tif'))
r2 <- classify(rasters_list[[7]],rcl)
writeRaster(r2, paste0(inpath, '/', names(r2), '_rec.tif'))
r2 <- classify(rasters_list[[8]],rcl)
writeRaster(r2, paste0(inpath, '/', names(r2), '_rec.tif'))
r2 <- classify(rasters_list[[9]],rcl)
writeRaster(r2, paste0(inpath, '/', names(r2), '_rec.tif'))
r2 <- classify(rasters_list[[10]],rcl)
writeRaster(r2, paste0(inpath, '/', names(r2), '_rec.tif'))


# This is waht somehow did not work, but i am sure there is a better way to handle this. 
rasters_list2 <- lapply(rasters_list, function(r){
  r <- classify(r,rcl)
  return(r)
})#, mc.cores = 10)

map(1:length(rasters_list2), function(x) writeRaster(rasters_list[[x]], paste0(inpath, '/', names(rasters_list[[x]]), '_rec.tif')))

```

```{r get retention potential} 


tiffes <- file.path(inpath, list.files(paste0(inpath),pattern= 'rec'))
tiffes <- tiffes[c(1,6,2,7,3,8,4,9,5,10)]
filename<- basename(tiffes)
 # here, adjust to call the correct ones and buld thje corresponding lists.

rasters_list <- lapply(tiffes, rast)


l1 <- rasters_list[seq(1,length(rasters_list), by=2)]
l2 <- rasters_list[seq(2,length(rasters_list), by=2)]
 # test with future_map2 next time to get this to run faster. 
 year <- c(1992,1995,1998,2001,2004)
# 
 pot_s_loss <- Map(function(r1,r2) r2/(r2+r1), l1,l2)
# 
 map(1:length(pot_s_loss), function(x) writeRaster(pot_s_loss[[x]], paste0(inpath, '/', 'pot_n_loss_', year[x],'2.tif')))

 # i think i need to extract some percentages from here. 
# pending to check: max min values, nas and visuialization

```



```{r load poly]}
# Read spatial polygon dataset
# Set attribute column for analysis
 
# By Continent
#set <- 'Continent.gpkg' 
# By Subregion
# set <- 'subregion.gpkg'
# By Income Group
# set <- "Income_grp.gpkg"
# By individual territory; "Country" 
#set <- 'Country.gpkg'
# By WWF biome
set <- 'Biome.gpkg' 
# by WB Region
#set <- "World_Bank_Region.gpkg"
# load polygons
poly <- st_read(here('vector', set)) 
poly <- poly[c(1,2)]

# select column with the asttribute to iterate through.
#col <- "id"
#col <- "WWF_biome"
#col <- 'continent'
col <- "WWF_biome"
#col <- 'continent'
#col <- "income_grp"
#col <- 'subregion'
#col <- 'region_wb'
cols <- col
```
Here, i need to think if iu add the new extracted values to the existing polygons or waht to do wit hthat. Variables: potentiaL 

```{r load raster es, include=FALSE}
inpath <- here('input_ES', 'mod_n_ret')


#load the rasters with the calculated differences 
tiffes <- file.path(inpath, list.files(paste0(inpath),pattern='pot'))

# inpath <- "/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_NCP/input_ES"
# tiffes <- file.path(inpath, list.files(inpath, pattern= 'tif$'))
filename <- basename(tiffes)
raster_name <- gsub(".tif$", "", filename)

# Load raster files
rasters_list <- lapply(tiffes, rast)
# Define service names and colors
year <- c(1992,1995,1998,2001,2004)
filename <- as_tibble(cbind(raster_name, year))

```

```{r compute stats 1, message=TRUE, include=FALSE}
target_stats <- c("mean", "stdev")
num.cores <- length(rasters_list) # set as many cores as rasters to be evaluated.  sequentially, and there is no progress bar available 
# here, set a way to change the input poly and columns externally instead of editing inside of the functions. That is cumbersome and prone to errors. 

results_list <- mclapply(rasters_list, function(r) {
  # Perform exact extraction
  # - append_cols = "country" retains the country identifier with each result
  res <- exact_extract(r, poly,
                       fun = target_stats,
                       append_cols = cols)
  r_name <- names(r)
  # Add a column labeling which raster these results are from
  res$raster_name <- r_name
  return(res)},mc.cores = num.cores)



results_list <- Map(function(df,y)  {
  df$year <- y
  return(df)
}, results_list, year) 

# Combine results from all rasters into one data frame
zonal_df <- do.call(rbind, results_list)
zonal_df <- zonal_df %>% mutate(service = "n_ret")

zonal_wide <- pivot_wider(zonal_df, id_cols=c(col), names_from=c(year, service), values_from = c(mean, stdev))


poly <- left_join(poly,zonal_wide)

st_write(poly, paste0(here('vector'), '/', cols, '_sdr_chg.gpkg'))


zonal_df <- as_tibble(zonal_df)
write.csv(zonal_df, here('output_data', 'TS_Biome.csv'))

```



```{r timeseries plot}

library(ggplot2)

# Create the line plot
ggplot(zonal_df, aes(x = year, y = mean, color = subregion, group = subregion)) +
  geom_line(size = 1) +          # Add lines for each subregion
  geom_point(size = 2) +         # Add points for visibility
  labs(
    title = "Ecosystem Service Trends by Subregion",
    x = "Year",
    y = "Mean Value"
  ) +
  theme_minimal() +
  theme(legend.position = "right")  # Adjust legend position
```
```{r faceted plot fig.height=18, fig.width=18}


ggplot(zonal_df, aes(x = year, y = mean, group = WWF_biome)) +
  geom_line(size = 1, color = "blue") +
  geom_point(size = 2, color = "red") +
  facet_wrap(~ WWF_biome, scales = "free_y", ncol = 4) +
  labs(
    title = "Pot N Loss by Subregion",
    x = "Year",
    y = "Mean Value"
  ) +
  theme(
    legend.position = "right",
    text = element_text(size = 12),  # Reduce overall text size
    axis.text.x = element_text(size = 6, angle = 45, hjust = 1),  # Adjust x-axis labels
    axis.text.y = element_text(size = 6),  # Adjust y-axis labels
    legend.text = element_text(size = 6),  # Reduce legend font size
    legend.title = element_text(size = 6),  # Adjust legend title size
    strip.text = element_text(size = 8)  # Adjust facet labels (if using facets)
  )

  

```





# add color and service columns
zonal_df <- left_join(zonal_df,cd)
# Add the year. this uses the original filenames to extract the year, but might be too hardcoded for generalization 



zonal_wide <- pivot_wider(zonal_df, id_cols=all_of(col), names_from=year, values_from = mean)
# adjust column names #

######### HERE: Include a way/convention to adjust the column names autmatically. Saves work and keeps consistency. I had to do this in Stpe 6, but this is not organized. should +be done herer bedore exporting

poly <- left_join(poly, zonal_wide, by= col)
st_write(poly, here('vector', set), append = FALSE) 
write.csv(zonal_df, here('output_data', paste0('zonal_df_',col, '.csv')))



zonal_df



```{r load data}

inpsath <- '/home/jeronimo/OneDrive/Global_ES_mapping/ESA_LC'
tiffes <- file.path(inpath, list.files(paste0(inpath),pattern= 'tif$'))
filename<- basename(tiffes)
#set <- 'continent.gpkg' 

raster_list <- lapply(rast,fiffes)

poly <- st_read(here('vector', set)) 

poly <- poly %>% filter(continent=='South America')


raster_list <- lapply(raster_list, function(r){
  r <- crop(r,poly)
  r <- mask(r,poly)
}

n_ch_sa <- noch_msk(raster_list, type='nochange')




poly <- 
# Use terra::rast() to load each file as a SpatRaster object





```




### 6.1  Prepare Potential of Sed retention data.

```{r getpotential Sed retention , eval=FALSE, include=FALSE}
inpath <- here("input_ES", "Pot_Sed_retention")
tiffes <- file.path(inpath, list.files(paste0(inpath),pattern= 'tif$'))
filename<- basename(tiffes)
# Use terra::rast() to load each file as a SpatRaster object


rasters_list <- lapply(tiffes, rast) # why am i doingthis? There is a good reasoner, but now i don't know anymore 
rcl <- matrix(c(
  -Inf, 0, NA   # Any value from 0 to Infinity becomes 1
), ncol = 3, byrow = TRUE)

rasters_list <- lapply(rasters_list, function(r){
  r <- classify(r,rcl, right=FALSE)
})

m_index <- length(rasters_list) %/% 2
#lab <- unique(service)
l1 <- rasters_list[seq(1,length(rasters_list), by=2)]
l2 <- rasters_list[seq(2,length(rasters_list), by=2)]
 # test with future_map2 next time to get this to run faster. 

pot_SedRet1 <- (l1[[2]]-l1[[1]])/l1[[2]]


pot_SedRet2 <- (l2[[2]]-l2[[1]])/l2[[2]]

psr_92 <- classify(pot_SedRet1,rcl, right=FALSE)
psr_20 <- classify(pot_SedRet2,rcl, right=FALSE)

writeRaster(psr_92, paste0(inpath, '/', 'pot_SedRet_92.tif'), overwrite=TRUE)
writeRaster(psr_20, paste0(inpath, '/', 'pot_SedRet_20.tif'), overwrite=TRUE)
diff <- psr_92 - psr_20
writeRaster(diff, paste0(inpath, '/', 'pot_Sed_r_diff.tif'), overwrite)

```
