---
title: "Land Cover Change Metrics Workflow"
author: "Jeronimo Rodriguez-Escobar"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(terra)
library(sf)
library(dplyr)
library(ggplot2)
library(glue)
library(tidyr)
library(purrr)
library(diffeR)
# source the helper functions
source(here::here("R", "utils_lcc_metrics.R"))
```


Here i am creating my maps of land cover change. Just from natural to transformed. As usual, this is easier said that n done


```{r load LC maps}
inpath <-  '/home/jeronimo/global_ES_modeling/esos-c/data/ndr'
tiffes <- file.path(inpath, list.files(paste0(inpath),pattern= 'landcover'))
filename<- basename(tiffes)
#set <- 'continent.gpkg' 

raster_list <- lapply(tiffes,rast)
num.cores <- length(raster_list)

# Define the reclassification rules in one operation
reclass_table <- data.frame(
  from = c(10, 11, 12, 20, 30, 40, 50, 60, 61, 62, 70, 71, 72, 80, 81, 82, 90, 100,
           110, 120, 121, 122, 130, 140, 150, 151, 152, 153, 160, 170, 180,
           190, 200, 201, 202, 210, 220),
  to = c(1, 1, 1, 1, 1, 1,   # Cultivated
         2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2,  # Forests
         3, 3, 3, 3, 3,   # Grasses and Shrubs
         4, 4, 4, 4, 4,   # Sparse Vegetation
         5, 5, 5,   # Mangroves
         6,   # Urban
         7, 7, 7,   # Bare
         8,   # Water
         9)   # Ice
)


raster_list <- lapply(raster_list, function(r) {
  subst(r, from = reclass_table$from, to = reclass_table$to)
})#,mc.cores = num.cores)

#save rasters 
lapply(1:length(raster_list), function(x) writeRaster(raster_list[[x]], paste0(here("input_ES", "LandCovers"),'/', filename[x])))

inpath <- here("input_ES", "LandCovers")

tiffes <- file.path(inpath, list.files(paste0(inpath),pattern= 'landcover'))
filename<- basename(tiffes)

raster_list <- lapply(tiffes,rast)


reclass_table <- data.frame(
  from = c(3,4,5,6,7,8,9),
  to = c(2,2,2,1,2,2,2)
)

raster_list <- lapply(raster_list, function(r) {
  subst(r, from = reclass_table$from, to = reclass_table$to)
})#,mc.cores = num.cores)

#save rasters 
lapply(1:length(raster_list), function(x) writeRaster(raster_list[[x]], paste0(here("input_ES", "LandCovers"),'/', "rec_", filename[x])))


# second reclassifciation Natural-not natural 

```

## Load Inputs

```{r load-data}
# Example file paths (replace with actual paths)
landcover_files <- list.files(
  here("input_ES/LandCovers"),
  pattern = "rec.*\\.tif$", 
  full.names = TRUE
)


# Name rasters by year
names(landcover_files) <- stringr::str_extract(basename(landcover_files), "\\d{4}")
landcover_rasters <- lapply(landcover_files, terra::rast)

# Load polygon spatial units (e.g., HydroBASINS level 6)
#prelimnarily test only in South America and on;y fpr 3 years. 
basins <- st_read(here('hydrosheds',"hybas_sa_lev01-12_v1c", "hybas_sa_lev06_v1c.shp"))
#basins <- st_read(here("vector", "hydrosheds_lev6.gpkg"))
```

## Run Land Cover Change Metric Extraction

```{r run-extraction}
lcc_metrics_df <- iterate_lcc_metrics(
  raster_list = landcover_rasters,
  polygons = basins,
  id_col = "HYBAS_ID",
  percent = TRUE,
  verbose = TRUE,
  mc = TRUE,
  ncores = 3
)



# Save to CSV (optional)
readr::write_csv(lcc_metrics_df, "output_data/lcc_ch_metrics_lv6.csv")
```

## Inspect Results

```{r summarize-results}
# Summarize by year and metric
summary_df <- lcc_metrics_df %>%
  group_by(year_step, Category) %>%
  summarize(across(c(Gain, Persistence, Loss, Quantity, Exchange, Shift), 
                   mean, na.rm = TRUE), .groups = "drop")

# Preview
head(summary_df)
```

## Plot Change Metrics Over Time

```{r plot-changes}
ggplot(summary_df, aes(x = year_step, y = Gain, group = Category, color = Category)) +
  geom_line(size = 1) +
  geom_point() +
  labs(title = "Average Gain in Transformed vs Natural Land Over Time",
       x = "Year Step",
       y = "% Area Changed") +
  theme_minimal()
```

---

This RMarkdown file shows how to:

1. Load land cover rasters and polygon units
2. Apply `extract_zonal_lcc_metrics()` and `iterate_zonal_lcc_metrics()`
3. Export the results
4. Summarize and plot key land cover transition metrics


```{r landcover_differ_change, message=TRUE, warning=FALSE, eval=FALSE}

library(terra)
library(sf)
library(diffeR)
library(dplyr)
library(tidyr)
library(purrr)
library(progress)

# 1. Define inputs ----

# Path to input rasters
raster_dir <- here("input_landcover")
raster_files <- list.files(raster_dir, pattern = "\\.tif$", full.names = TRUE)

# Sort chronologically (assuming names contain 4-digit years)
raster_files <- raster_files[order(gsub("\\D", "", basename(raster_files)))]
years <- gsub("\\D", "", basename(raster_files))

# Load reclassified land cover rasters
lc_rasters <- lapply(raster_files, terra::rast)

# Load spatial units
polygons <- vect(here("vector", "hydrosheds_lev6.gpkg"))  # or lev7

# Ensure proper alignment
lc_rasters <- lapply(lc_rasters, function(r) terra::project(r, crs(polygons)))

# 2. Function to extract diffeR metrics per polygon and year ----

extract_differ_metrics <- function(poly, lc_list, years) {
  n_years <- length(lc_list)
  results <- list()

  for (i in 1:(n_years - 1)) {
    r1 <- lc_list[[i]]
    r2 <- lc_list[[i + 1]]
    year_pair <- paste0(years[i + 1], "_", years[i])

    # Crop/mask to polygon on-the-fly
    r1_crop <- crop(r1, poly) |> mask(poly)
    r2_crop <- crop(r2, poly) |> mask(poly)

    # Skip if empty
    if (is.na(global(r1_crop, fun = "sum", na.rm = TRUE))) next

    # Crosstab and difference metrics
    tab <- crosstabm(comp = r1_crop, ref = r2_crop, percent = TRUE)
    diff_stats <- difftablej(tab, analysis = "change")

    # Format result
    diff_stats$fid <- poly$HYBAS_ID
    diff_stats$year_step <- year_pair
    results[[length(results) + 1]] <- diff_stats
  }

  bind_rows(results)
}

# 3. Iterate over all polygons with progress bar ----

pb <- progress_bar$new(
  total = nrow(polygons),
  format = "  Processing [:bar] :percent in :elapsed",
  clear = FALSE, width = 60
)

all_results <- map_dfr(1:nrow(polygons), function(i) {
  pb$tick()
  extract_differ_metrics(polygons[i, ], lc_rasters, years)
})

# 4. Clean and export ----

# Pivot to long format
diff_long <- all_results %>%
  pivot_longer(cols = c("Gain", "Persistence", "Loss", "Quantity", "Exchange", "Shift"),
               names_to = "metric", values_to = "value")

# Export table for plotting or analysis
write.csv(diff_long, here("output_data", "lc_change_metrics.csv"), row.names = FALSE)


```

