---
title: "Current Status"
output: html_notebook
---



```{r setup, message=FALSE, warning=FALSE}
library(terra)
library(sf)
library(dplyr)
library(ggplot2)
library(glue)
library(tidyr)
library(purrr)
library(diffeR)
library(here)
library(stringr)
library(tidytext)
library(rlang)
library(tidyr)
# source the helper functions
source(here("R", "utils_lcc_metrics.R"))
source(here("R","utils_pct_change.R"))
```


#  1 Organize output Dataframes (make sure that they are complete and correclty named)

Right now, 24/04/2025 the problem is on the way the dataframe with the class colors and names is built. The process is still too manual (writing the table) and is easy to get it wrong when it should not be big deal. Ideally, there shoudl be a dialogue box to fill this instead. This should take care of the class names, values and colors neatl
Also, the order of the factors needs to be set, and new classes have to be added. Again, i don't how to do it easily. I mean, it is easy, but annoying to do and easy to get wrong.


#######################################################
Here, i need to clean the directory form a lot of old versions. Only the groupings. If pct not there, it is fine, that is easy to calculate now. 

```{r load polygons, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# # Read spatial polygon dataset
# 
# inpath <- '/Users/sputnik/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_NCP/vector'
# inpath <- here("vector")
# sets <- file.path(inpath, list.files(inpath, pattern= 'gpkg$'))
# 
# #cols <- c("id", "continent", "income_grp", "region_wb", "WWF_biome", "HYBAS_ID", "HYBAS_ID")
# #sets <- cbind(sets, cols)
# # Set an index to select the desired dataset
# t <- 10
# # Select the target dataset from the list 
# set <- sets[t]
# # # load polygons
# lyr <- st_layers(set)
# poly <- st_read(set)
# 
# #get the name of the column with the unique ids of the input vector
# col <- colnames(poly[1])[1]
 
```

# Load and pivot 

Here, the spatial objects with the additional attributes are loaded and reformatted for analysis and chart preparation - **pivot & tidy**. 
Add to the documentation , environment or however that's called the vectors to select and order the columns. 
This should actually be done as a database structure. The columns live somewhere and are summoned upon need from a set of options (list_dir). 
  
```{r chunk1, eval=FALSE, include=FALSE}
# 
#  get pct as standalone data to add to the table
# Most recent version of synthesis data:
 
poly <- st_read('/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_ES_TimeSeries/vector/hydrosheds_lv6_synth.gpkg')

# add areas 
poly <- poly %>% mutate(area_sqkm = as.numeric(st_area(geom)) / 10^6)

 plt <- st_drop_geometry(poly)
 plt <- plt %>%
  dplyr::select(
    1:15, # This is hardcoded for the basins, but has to be adjusted for specific cases/desired columns.
    22,
    contains("pct_ch"), 
    contains("area_sqkm")
  )
 # Edit/Adjust column names 
 names(plt) <- stringr::str_replace_all(names(plt), "_2020_1992", "")
 # remove old N calculation, use the current one made by me
 plt$n_export_pct_ch <- NULL
 
 or_attributes <- plt
 
 ############# PIVOT TABLE ##############################
 plt <- plt %>%
  pivot_longer(
    cols = matches("pct_ch"),  # matches anything containing pct_ch
    names_to = "service",
    names_transform = list(service = as.character),
    values_to = "pct_ch"
  ) %>%
  mutate(service = str_replace(service, "_pct_ch.*", "")) %>% 
  as_tibble()
# #This is annoying and should be improved. Made easier. Here i am adjusting/homologuing variable names to a same standard. Because some of the data has been extracted at different moments there is some, ahem, *inconsistency* in the "service" naming structure that I fixed here.
 
 plt <- plt %>% mutate(service = case_when(
   #service == "n_export" ~ "N_export",
   service == "sed_export" ~ "Sed_export",
   service == "Sed_retention_ratio" ~ "Sed_Ret_Ratio",
    service == "Pop" ~ "Pop_sqkm",
   #service == "N_export_sqkm" ~ "N_Export_2",
   TRUE ~ service
   ))
# 
#Drop Usle 
plt <- plt %>% filter(service!="Usle") # I am keeping the data in the attribute table because it *might* be necessary at some point, but removing from the reformated table. 
# cd <- as_tibble(cbind(service,color)) # I don't know why this is here. 
#Remove invalid / Inf records
plt <- plt%>% filter(!is.na(pct_ch)) %>% filter(pct_ch != Inf)  

ct <- st_read('/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_ES_TimeSeries/vector/cartographic_ee_ee_r264_correspondence.gpkg')

ct <- ct %>% mutate(area_sqkm_country = as.numeric(st_area(geom)) / 10^6)
ct <- st_drop_geometry(ct)
ct <- ct[c(1,37)]
plt <- left_join(plt, ct)
#. Have a method to name versions or something (maybe?)
write.csv(plt, file= here('output_data', "metrics_change_hb_lev_6.csv"))
```

# set class names and colors for plotting 

Here, another dialogue to input the variable names, plot colors and set the factor order.
```{r, fig.height=8, fig.width=14  }
#plt <- as_tibble(read.csv(here('output_data', "metrics_change_hb_lev_6.csv")))
# plt[1] <- NULL 
service <- unique(plt$service)
service <- c("Coastal_Protection","Pollination","N_export","N_Ret_Ratio", "Sed_export", "Sed_Ret_Ratio", "Pop_sqkm","GDP")
 color <- c("#9e9ac8", "#dd1c77","#2c944c", "#2c711c","#08306b", "#02606b", "#d4ac0d", "#85929e")
 service_levels <- service
 cd <- as_tibble(cbind(service,color))
 
 
 plt <- left_join(plt, cd)
```


 This could be a dialogue box!!!
```{r load-plt}
# Calculate pseudo_log val for display purposes. 
plt <- plt %>%
  mutate(
    service = factor(service, levels = service_levels),
    pct_ch_trans = pseudo_log(pct_ch)
  )

# apply filter (5% or 2%) # Remove the smallest basins. 
area_threshold <- quantile(plt$SUB_AREA, probs = 0.05, na.rm = TRUE)
# Filter the dataset
df <- plt %>% filter(SUB_AREA > area_threshold)
# get the filtered locations

# optional. Remove countries by area. 
df <- df %>%
  filter(area_sqkm_country >= 10000)
# Step 3: Select top and bottom 5 by dir_ch_2
top_bottom_lc <- df %>%
  slice_max(dir_ch_2, n = 5, with_ties = FALSE) %>%
  bind_rows(
    bas_lc_filtered %>%
      slice_min(dir_ch_2, n = 5, with_ties = FALSE)
  )
```



## 5. Here, add plots with the differences 


```{r, fig.height=8, fig.width=14}
col <- "HYBAS_ID"

col <- "ee_r264_name"

plot_es_changes <- function(data, label_col,
                            filter_type = "top_bottom", filter_val = 10) {
  label_sym <- sym(label_col)

  # Step 2: Apply filtering based on the chosen type
  if (filter_type == "top_bottom") {
    # Top/bottom n observations per service
    top_bottom <- data %>%
      group_by(service) %>%
      slice_max(pct_ch, n = filter_val, with_ties = FALSE) %>%
      bind_rows(
        data %>%
          group_by(service) %>%
          slice_min(pct_ch, n = filter_val, with_ties = FALSE)
      ) %>%
      ungroup()

  } else if (filter_type == "quantile") {
    # Top/bottom quantile per service (e.g., top/bottom 10%)
    top_bottom <- data %>%
      group_by(service) %>%
      filter(pct_ch >= quantile(pct_ch, 1 - filter_val, na.rm = TRUE) |
             pct_ch <= quantile(pct_ch, filter_val, na.rm = TRUE)) %>%
      ungroup()

  } else if (filter_type == "all") {
    top_bottom <- data
  } else {
    stop("Invalid `filter_type`. Use 'top_bottom', 'quantile', or 'all'.")
  }

  # Step 3: Reorder labels per service
  top_bottom <- top_bottom %>%
    mutate(temp_label = reorder_within(!!label_sym, -pct_ch, service))

  # Step 4: Plot
  ggplot(top_bottom, aes(x = temp_label, y = pct_ch, fill = color)) +
    geom_bar(stat = "identity", show.legend = FALSE) +
    scale_fill_identity() +
    scale_x_reordered() +
    facet_wrap(~ service, scales = "free", ncol = 3) +
    labs(
      #title = paste("% Change 1992–2020 By", cols, sep= " "),
      x = NULL,
      y = "% Change"
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(face = "bold", size = 10),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
    )
}

# Top/bottom 10 countries per service
plot_es_changes(df, label_col = col, filter_type = "top_bottom", filter_val = 10)

# Top/bottom 5% per service
#plot_es_changes(df, label_col = col, filter_type = "quantile", filter_val = 0.5)
# 
# # Show all values
# plot_es_changes(plt, filter_type = "all")
```


## Scatterplots



```{r oterh plot, fig.height=7, fig.width=7, warning=FALSE}
library(dplyr)
library(ggplot2)
library(forcats)

# 1. Filter to top/bottom 10% based on pct_ch
# df_top_bottom <- df %>%
#   group_by(service) %>%  # Do it separately for each service (optional, if needed)
#   filter(pct_ch >= quantile(dir_ch_2, 0.90, na.rm = TRUE) | 
#          pct_ch <= quantile(dir_ch_2, 0.10, na.rm = TRUE)) %>%
#   ungroup()
n <- floor(nrow(df) * 0.1)  # 10% of total rows

df_top_bottom <- df %>%
  slice_max(dir_ch_2, n = n, with_ties = FALSE) %>%
  bind_rows(
    df %>% slice_min(dir_ch_2, n = n, with_ties = FALSE)
  )

df_top_bottom <- df %>%
  group_by(service) %>%
  group_modify(~ bind_rows(
    slice_max(.x, dir_ch_2, n = floor(0.1 * nrow(.x)), with_ties = FALSE),
    slice_min(.x, dir_ch_2, n = floor(0.1 * nrow(.x)), with_ties = FALSE)
  )) %>%
  ungroup()



# 2. Optional: relevel services if you want
service <- c("Coastal_Protection","Pollination","N_export","N_Ret_Ratio", "Sed_export", "Sed_Ret_Ratio", "Pop_sqkm","GDP")
 color <- c("#9e9ac8", "#dd1c77","#2c944c", "#2c711c","#08306b", "#02606b", "#d4ac0d", "#85929e")
 service_levels <- service
df_top_bottom <- df_top_bottom %>%
  mutate(
    service = factor(service, levels = service_levels),
    pct_ch_trans = pseudo_log(pct_ch)  # Apply pseudo-log transformation
  )

# 3. List of LC metrics to analyze
lc_metrics <- c("dir_ch_2")
df_top_bottom <- df_top_bottom %>% filter(service != "GDP") %>% filter(service != "Pop_sqkm")
df_top_bottom <- filter(df_top_bottom, dir_ch_2 != 0)
# 4. Create the scatterplots
for (metric in lc_metrics) {
  p1 <- df_top_bottom %>%
    ggplot(aes_string(x = metric, y = "pct_ch_trans")) +
    geom_hex(bins = 60) +  # Hexbin layer
    scale_fill_viridis_c(option = "D", direction = 1) +  # Use viridis scale
    facet_wrap(~ service, scales = "free",ncol=2) +
    labs(
      title = paste("% Change Natural Cover ", "vs. ES % Change"),
      x = metric,
      y = "% Change in Ecosystem Service",
      fill = "Density"
    ) +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5),
      axis.text = element_text(size = 9)
    )

  print(p1)
}

```


```{r , fig.height=9, fig.width=7, warning=FALSE}

df <- plt %>% filter(SUB_AREA > area_threshold)
#df <- plt %>% filter(id %in% top_bottom_10k$id)
# List of LC metrics to analyze
lc_metrics <- c("dir_ch_2")
pseudo_log <- function(x) sign(x) * log1p(abs(x))

df <- df %>%
  mutate(
    service = factor(service, levels = service_levels),
    pct_ch_trans = pseudo_log(pct_ch)
  )
# Loop over each LC metric and generate a faceted scatterplot
for (metric in lc_metrics) {
  p <- df %>%
    ggplot(aes_string(x = metric, y = "pct_ch_trans", color = "color")) +
    geom_point(alpha = 0.5, size = 1.2) +
    facet_wrap(~ service, scales = "free") +
    scale_color_identity() +
    labs(
      title = paste("Relationship Between", metric, "and ES % Change"),
      x = metric,
      y = "% Change in Ecosystem Service"
    ) +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5),
      axis.text = element_text(size = 9)
    )

  print(p)
}
```
```{r sctp, fig.height=9, fig.width=7, warning=FALSE}


# List of LC metrics to analyze
lc_metrics <- c("dir_ch_2")

df_1 <- df %>% filter(service != "GDP") %>% filter(service != "Pop_sqkm")



 for (metric in lc_metrics) {
  p1 <- df_1 %>%
    ggplot(aes_string(x = metric, y = "pct_ch_trans")) +
    geom_hex(bins = 60) +  # Hexbin layer
    scale_fill_viridis_c(option = "D", direction = -1) +  # Use viridis scale
    facet_wrap(~ service, scales = "free",ncol=2) +
    labs(
      title = paste("Density of", metric, "vs. ES % Change"),
      x = metric,
      y = "% Change in Ecosystem Service",
      fill = "Bin Count"
    ) +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5),
      axis.text = element_text(size = 9)
    )

  print(p1)
}
 
df_split <- split(df_1, df_1$service)


```


# Scatterplots change 2

```{r unnamedchunk, fig.height=9, fig.width=7, warning=FALSE}

# LC metrics of interest
lc_metrics <- c("dir_ch_2")

# Step 1: Remove top 2% outliers for each service
filtered_plt <- plt %>%
  group_by(service) %>%
  mutate(threshold = quantile(pct_ch, 0.98, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(pct_ch <= threshold)

# Step 2: Apply pseudo-log transformation
filtered_plt <- filtered_plt %>%
  mutate(pct_ch_log = sign(pct_ch) * log1p(abs(pct_ch)))

# Step 3: Loop through each LC metric and create faceted plots
for (metric in lc_metrics) {
  p <- ggplot(filtered_plt, aes_string(x = metric, y = "pct_ch_log", color = "color")) +
    geom_point(alpha = 0.5, size = 1.2) +
    facet_wrap(~ service, scales = "free") +
    scale_color_identity() +
    labs(
      title = paste("% Change vs", metric),
      x = metric,
      y = "Log(% Change in ES)"
    ) +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5),
      axis.text = element_text(size = 9)
    )

  print(p)
}
```

```{r eval=FALSE, fig.height=9, fig.width=7, warning=FALSE}


# LC metrics of interest
lc_metrics <- c("dir_ch_2")

# Step 1: Remove top 2% outliers for each service
filtered_plt <- plt %>%
  group_by(service) %>%
  mutate(threshold = quantile(pct_ch, 0.98, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(pct_ch <= threshold)

# Step 3: Loop through each LC metric and create faceted plots
for (metric in lc_metrics) {
  p3 <- ggplot(filtered_plt, aes_string(x = metric, y = "pct_ch", color = "color")) +
    geom_point(alpha = 0.5, size = 1.2) +
    facet_wrap(~ service, scales = "free") +
    scale_color_identity() +
    labs(
      title = paste(" % Change vs", metric),
      x = metric,
      y = "% Change in ES"
    ) +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5),
      axis.text = element_text(size = 9)
    )

  print(p3)
}
```


```{r eval=FALSE, fig.height=8, fig.width=14, include=FALSE}

# apply filter (5 or 2%) # Remove the smalles basins. That help. Also, it seems that some biomes are more prone to artifacts than others!!!
area_threshold <- quantile(plt$SUB_AREA, probs = 0.05, na.rm = TRUE)
# Filter the dataset
df <- plt %>% filter(SUB_AREA > area_threshold)


t_5 <- ggplot(df, aes(x = service, y = pct_ch, fill = color)) +
  geom_boxplot() +  # or geom_violin()
  scale_fill_identity() +
  facet_wrap(~ service, scales = "free_y") +
  labs(
    title = "Distribution of % Change by Service (Excl. Smallest 5% of Basins)",
    x = NULL,
    y = "Percentage Change"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )

library(ggplot2)
library(dplyr)

# Custom pseudo-log transformation
pseudo_log <- function(x) sign(x) * log1p(abs(x))

# Apply  the filtered dataframe
df <- plt %>%
  mutate(pct_ch_trans = pseudo_log(pct_ch))

# Plot with transformed values
p <- ggplot(df, aes(x = service, y = pct_ch_trans, fill = color)) +
  geom_violin(trim = FALSE, scale = "width") +
  scale_fill_identity() +
  facet_wrap(~ service, scales = "free_y") +
  labs(
    title = "Pseudo-log Scaled % Change by Service",
    x = NULL,
    y = "Transformed % Change"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )
p
```




#######################

## 5. Here, add plots with the differences 

```{r eval=FALSE, fig.height=8, fig.width=14, include=FALSE}
print(col)
print(names(df))
stopifnot(col %in% names(df))

plot_es_changes <- function(data, label_col,
                            filter_type = "top_bottom", filter_val = 10) {
  label_sym <- sym(label_col)

  # Step 2: Apply filtering based on the chosen type
  if (filter_type == "top_bottom") {
    # Top/bottom n observations per service
    top_bottom <- data %>%
      group_by(service) %>%
      slice_max(pct_ch, n = filter_val, with_ties = FALSE) %>%
      bind_rows(
        data %>%
          group_by(service) %>%
          slice_min(pct_ch, n = filter_val, with_ties = FALSE)
      ) %>%
      ungroup()

  } else if (filter_type == "quantile") {
    # Top/bottom quantile per service (e.g., top/bottom 10%)
    top_bottom <- data %>%
      group_by(service) %>%
      filter(pct_ch >= quantile(pct_ch, 1 - filter_val, na.rm = TRUE) |
             pct_ch <= quantile(pct_ch, filter_val, na.rm = TRUE)) %>%
      ungroup()

  } else if (filter_type == "all") {
    top_bottom <- data
  } else {
    stop("Invalid `filter_type`. Use 'top_bottom', 'quantile', or 'all'.")
  }

  # Step 3: Reorder labels per service
  top_bottom <- top_bottom %>%
    mutate(temp_label = reorder_within(!!label_sym, -pct_ch, service))

  # Step 4: Plot
  ggplot(top_bottom, aes(x = temp_label, y = pct_ch, fill = color)) +
    geom_bar(stat = "identity", show.legend = FALSE) +
    scale_fill_identity() +
    scale_x_reordered() +
    facet_wrap(~ service, scales = "free", ncol = 3) +
    labs(
      #title = paste("% Change 1992–2020 By", cols, sep= " "),
      x = NULL,
      y = "% Change"
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(face = "bold", size = 10),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
    )
}

# Top/bottom 10 countries per service
plot_es_changes(df, label_col = col, filter_type = "top_bottom", filter_val = 10)

# Top/bottom 5% per service
#plot_es_changes(df, label_col = col, filter_type = "quantile", filter_val = 0.1)

# Show all values
#plot_es_changes(plt, filter_type = "all")
```


## Scatterplots

```{r}


# List of LC metrics to analyze
lc_metrics <- c("Gain_2", "Persistence_2", "Loss_2", "dir_ch_2")

# Loop over each LC metric and generate a faceted scatterplot
for (metric in lc_metrics) {
  p <- plt %>%
    ggplot(aes_string(x = metric, y = "pct_ch", color = "color")) +
    geom_point(alpha = 0.5, size = 1.2) +
    facet_wrap(~ service, scales = "free") +
    scale_color_identity() +
    labs(
      title = paste("Relationship Between", metric, "and ES % Change"),
      x = metric,
      y = "% Change in Ecosystem Service"
    ) +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5),
      axis.text = element_text(size = 9)
    )

  print(p)
}
```


# Scatterplots change 2

```{r, fig.height=8, fig.width=14}


# LC metrics of interest
lc_metrics <- c("Gain_2", "Persistence_2", "Loss_2", "dir_ch_2")

# Step 1: Remove top 2% outliers for each service
filtered_plt <- plt %>%
  group_by(service) %>%
  mutate(threshold = quantile(pct_ch, 0.98, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(pct_ch <= threshold)

# # Step 2: Apply pseudo-log transformation
# filtered_plt <- filtered_plt %>%
#   mutate(pct_ch_log = sign(pct_ch) * log1p(abs(pct_ch)))

# Step 3: Loop through each LC metric and create faceted plots
for (metric in lc_metrics) {
  p <- ggplot(filtered_plt, aes_string(x = metric, y = "pct_ch", color = "color")) +
    geom_point(alpha = 0.5, size = 1.2) +
    facet_wrap(~ service, scales = "free") +
    scale_color_identity() +
    labs(
      title = paste("Log-Transformed % Change vs", metric),
      x = metric,
      y = "Log(% Change in ES)"
    ) +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5),
      axis.text = element_text(size = 9)
    )

  print(p)
}
```

```{r, fig.height=8, fig.width=14}
# library(dplyr)
# library(ggplot2)
# 
# plt <- as_tibble(read.csv(here('output_data', "metrics_change_hb_lev_6.csv")))
# plt[1] <- NULL 
# # apply filter (5 or 2%) # Remove the smalles basins. That help. Also, it seems that some biomes are more prone to artifacts than others!!!
# area_threshold <- quantile(plt$SUB_AREA, probs = 0.05, na.rm = TRUE)
# # Filter the dataset
# df <- plt %>% filter(SUB_AREA > area_threshold)
# 
# 
# t_5 <- ggplot(df, aes(x = service, y = pct_ch, fill = color)) +
#   geom_boxplot() +  # or geom_violin()
#   scale_fill_identity() +
#   facet_wrap(~ service, scales = "free_y") +
#   labs(
#     title = "Distribution of % Change by Service (Excl. Smallest 5% of Basins)",
#     x = NULL,
#     y = "Percentage Change"
#   ) +
#   theme_minimal() +
#   theme(
#     axis.text.x = element_blank(),
#     strip.text = element_text(face = "bold"),
#     legend.position = "none"
#   )
# 
# library(ggplot2)
# library(dplyr)
# 
# # Custom pseudo-log transformation
# pseudo_log <- function(x) sign(x) * log1p(abs(x))
# 
# # Apply to the filtered dataframe
# df <- plt %>% 
#   mutate(pct_ch_trans = pseudo_log(pct_ch))
# 
# # Plot with transformed values
# p <- ggplot(df, aes(x = service, y = pct_ch_trans, fill = color)) +
#   geom_violin(trim = FALSE, scale = "width") +
#   scale_fill_identity() +
#   facet_wrap(~ service, scales = "free_y") +
#   labs(
#     title = "Pseudo-log Scaled % Change by Service",
#     x = NULL,
#     y = "Transformed % Change"
#   ) +
#   theme_minimal() +
#   theme(
#     axis.text.x = element_blank(),
#     strip.text = element_text(face = "bold"),
#     legend.position = "none"
#   )
# p
```






```{r fig.height=8, fig.width=10, warning=FALSE}
# 1. Define desired facet order
df <- plt %>% filter(SUB_AREA > area_threshold)
# df <- df %>%
#   filter(id %in% top_bottom_10k$id)
# 2. Apply pseudo-log transformation
pseudo_log <- function(x) sign(x) * log1p(abs(x))

df <- df %>%
  mutate(
    service = factor(service, levels = service_levels),
    pct_ch_trans = pseudo_log(pct_ch)
  )

# 3. Reorder countries based on dir_ch_2 (descending)
df <- df %>%
  mutate(ee_r264_name = reorder(ee_r264_name, -dir_ch_2))
df. <- df %>% filter(service != "GDP") %>% filter(service != "Pop_sqkm")
# 4. Create plot
p <- ggplot(df., aes(x = ee_r264_name, y = pct_ch_trans, fill = ee_r264_name)) +
  geom_violin(trim = FALSE, scale = "width") +
  facet_wrap(~ service, scales = "free_y") +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "Pseudo-log Scaled % Change by Service ",
    x = NULL,
    y = "Transformed % Change"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )
# 4. Create plot
p2 <- ggplot(df, aes(x = ee_r264_name, y = pct_ch, fill = ee_r264_name)) +
  geom_violin(trim = FALSE, scale = "width") +
  facet_wrap(~ service, scales = "free_y") +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "% Change by Service ",
    x = NULL,
    y = "Transformed % Change"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )
p
p2

```

```{r fig.height=8, fig.width=10, warning=FALSE}
library(dplyr)
library(ggplot2)
library(forcats)

# 1. Filter the dataframe for top and bottom 10 by dir_ch_2
df_top_bottom <- df %>%
  arrange(desc(dir_ch_2)) %>%
  slice_head(n = 10) %>%
  bind_rows(
    df %>%
      arrange(dir_ch_2) %>%
      slice_head(n = 10)
  )

# 2. Reorder countries inside the filtered data
df_top_bottom <- df_top_bottom %>%
  mutate(ee_r264_name = fct_reorder(ee_r264_name, -dir_ch_2))
df_top_bottom <- df_top_bottom %>% filter(service != "GDP") %>% filter(service != "Pop_sqkm")

# 3. Create plot (pseudo-log scale)
p <- ggplot(df_top_bottom, aes(x = ee_r264_name, y = pct_ch_trans, fill = service)) +
  geom_violin(trim = FALSE, scale = "width") +
  facet_wrap(~ service, scales = "free_y") +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "Pseudo-log Scaled % Change by Service (Top and Bottom 10 Countries)",
    x = NULL,
    y = "Transformed % Change"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )

# 4. (Optional) Plot regular % change
p2 <- ggplot(df_top_bottom, aes(x = ee_r264_name, y = pct_ch, fill = service)) +
  geom_violin(trim = FALSE, scale = "width") +
  facet_wrap(~ service, scales = "free_y") +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "% Change by Service (Top and Bottom 10 Countries)",
    x = NULL,
    y = "% Change"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )

# 5. Print
p
# p2  # if you also want the regular % plot

```
