---
title: "Summary ES Analysis"
author: "Jeronimo Rodriguez-Escobar"
date: "2025-02-12"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

# Overview
Systematic extraction of zonal statistics from modeled global **Ecosystem Service (ES)** datasets at the watershed level.

## 1. Objectives
1. Extract zonal statistics from ES global raster datasets using the `exactextractr` package.
2. Integrate extracted statistics as attributes to spatial polygon datasets.
3. Generate visualizations (bar plots, maps) for spatial representation of results.


```{r setup, include=FALSE}
library(sf)             # Spatial vector operations
library(dplyr)          # Data manipulation
library(terra)          # Raster processing
library(exactextractr)  # Zonal statistics
library(ggplot2)        # Visualization
library(forcats)        # Factor reordering
library(tidytext)
library(here)           # Managing paths
library(patchwork)     # Combining plots
library(tidyr)
library(parallel)        # Parallel computing
library(purrr)
library(stringr)
library(gganimate)

```

# 2 Load Watgershed Polygon Data

Obtained from [Hydrosheds](https://www.hydrosheds.org/), we are using level 6 and level 7 for the analysis

```{r load watersheds lev 6}

poly <- st_read(here('vector', 'hydrosheds2.gpkg'), layer='hydrosheds2')
col <- "HYBAS_ID"
cols <- col
```


# 3 Extract % change Sediment Retention

For this we are processing the Potential Sediment Retention data for 1992, 2020 and the calcualted differnece. The formula used is:

Potential of Sediment Retention: (USLE-Export)/USLE
*Note* There is 

```{r extract pot sed Retenion}

# set the path and load the rasters with the differences that we calcualted in step 3
#build paths to files

inpath <- here('input_ES', "Pot_Sed_retention")
tiffes <- file.path(inpath, list.files(paste0(inpath),pattern= 'tif$'))
tiffes <- tiffes[c(3,2,1)]
filename<- basename(tiffes)

raster_name <- c("Pot_Sed_ret_1992", "Pot_Sed_ret_2020", "Pot_Sed_ret_diff")

rasters_list <- lapply(tiffes, rast)
rasters_list <- Map(function(r,new_name){
  names(r) <- new_name
  return(r)
  }, rasters_list, raster_name)

year <- rep(c(1992,2020, "diff"))
# create a table with the content to fill the table
color <- c("#08606b", "#08606b", "#68606b")
cd <- as_tibble(cbind(raster_name, color,year))
rm(filename)
cd


target_stats <- c("mean","stdev")

# Iterate function over the rasters
num.cores <- length(rasters_list)
results_list <- mclapply(rasters_list, function(r) {
  # Perform exact extraction
  # - append_cols = "country" retains the country identifier with each result
  res <- exact_extract(r, poly,
                       fun = target_stats,
                       append_cols = col)
  r_name <- names(r)

  # Add a column labeling which raster these results are from
  res$raster_name <- r_name
  return(res)},mc.cores = num.cores)

# Combine results from all rasters into one data frame
zonal_df <- do.call(rbind, results_list)
#Adjust column so i can assemble the whole thing (yearly and difference values) in the same dataframe



zonal_df <- left_join(zonal_df,cd)
zonal_wide <- pivot_wider(zonal_df, id_cols=all_of(cols), names_from=c(raster_name), values_from = c(mean,stdev))


# Calculate % of change 
zonal_wide <- zonal_wide %>%
  mutate(Pot_Sed_ret_pct_ch = ((mean_Pot_Sed_ret_2020 - mean_Pot_Sed_ret_1992) / mean_Pot_Sed_ret_1992) * 100)


poly <- left_join(poly, zonal_wide, by= cols)
st_write(poly, here('vector', "hydrosheds2.gpkg"), append = FALSE) 
write.csv(zonal_df, here('output_data', paste0('pot_sed_ret_df_', col, '.csv')))# acá viene el problema!
```


```{r get modeled N retention potential} 

inpath <- here('input_ES', 'mod_n_ret')
inpath <- '/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_NCP/input_ES/mod_n_ret'
tiffes <- file.path(inpath, list.files(paste0(inpath),pattern= 'loss'))
filename<- basename(tiffes)
 # here, adjust to call the correct ones and buld thje corresponding lists.

rasters_list <- lapply(tiffes, rast)


# Define service names and colors
year <- c(1992,1995,1998,2001,2004)
filename <- as_tibble(cbind(filename, year))


target_stats <- c("mean", "stdev")
num.cores <- length(rasters_list) # set as many cores as rasters to be evaluated.  sequentially, and there is no progress bar available 
# here, set a way to change the input poly and columns externally instead of editing inside of the functions. That is cumbersome and prone to errors. 

results_list <- mclapply(rasters_list, function(r) {
  # Perform exact extraction
  # - append_cols = "country" retains the country identifier with each result
  res <- exact_extract(r, poly,
                       fun = target_stats,
                       append_cols = cols)
  r_name <- names(r)
  # Add a column labeling which raster these results are from
  res$raster_name <- r_name
  return(res)},mc.cores = num.cores)



results_list <- Map(function(df,y)  {
  df$year <- y
  return(df)
}, results_list, year) 

# Combine results from all rasters into one data frame
zonal_df <- do.call(rbind, results_list)
zonal_df <- zonal_df %>% mutate(service = "n_ret")

zonal_wide <- pivot_wider(zonal_df, id_cols=c(col), names_from=c(year, service), values_from = c(mean, stdev))
library(dplyr)

# Calculate % change between consecutive years
zonal_wide <- zonal_wide %>%
  mutate(
    pct_ch_1995_1992 = 100 * (mean_1995_n_ret - mean_1992_n_ret) / mean_1992_n_ret,
    pct_ch_1998_1995 = 100 * (mean_1998_n_ret - mean_1995_n_ret) / mean_1995_n_ret,
    pct_ch_2001_1998 = 100 * (mean_2001_n_ret - mean_1998_n_ret) / mean_1998_n_ret,
    pct_ch_2004_2001 = 100 * (mean_2004_n_ret - mean_2001_n_ret) / mean_2001_n_ret
  )


poly <- left_join(poly,zonal_wide)

st_write(poly, (here('vector', 'hydrosheds_lev6.gpkg')), append=FALSE)

#keep only colimn wotj the % change (waht if we had some variability measure?)
polyt <- poly[c(1, 24:27)]
#reshape data 
polyt_long <- polyt %>%
  pivot_longer(
    cols = starts_with("pct_ch_"),
    names_to = "year_step",
    values_to = "pct_ch"
  ) %>%
  mutate(
    year_step = gsub("pct_ch_", "", year_step),
    year_step = factor(year_step, levels = unique(year_step))  # ensure correct order
  )

#get top/bottom X% or N units 
polyt_long <- polyt_long %>%
  group_by(year_step) %>%
  mutate(
    change_class = case_when(
      ntile(pct_ch, 10) == 10 ~ "Top 10%",     # highest change
      ntile(pct_ch, 10) == 1  ~ "Bottom 10%",  # lowest change
      TRUE ~ NA_character_
    )
  ) %>%
  ungroup()

polyt_long <- polyt_long %>%
  mutate(year = as.integer(sub("_.*", "", year_step)))

# If year_step is an integer like 1995, convert it to date (optional but recommended)
polyt_long <- polyt_long %>%
  mutate(time = as.Date(paste0(year, "-01-01")))

# Save to GeoPackage or Shapefile (GeoPackage is better for multi-features and attributes)
st_write(polyt_long, here('vector', "hydrosheds_temporal.gpkg"), layer = "sediment_change", delete_dsn = TRUE)



library(ggplot2)

ggplot(polyt_long %>% filter(!is.na(change_class)), 
       aes(fill = change_class)) +
  geom_sf() +
  facet_wrap(~year_step, ncol = 2) +
  scale_fill_manual(values = c("Top 10%" = "darkred", "Bottom 10%" = "blue")) +
  theme_minimal() +
  labs(
    title = "Extreme % Change in Service by Basin",
    fill = "Change Class"
  )


```


```{r }
library(ggplot2)
library(sf)
library(dplyr)
library(tidyr)

# Assuming your long data is already prepared in `polyt_long`
# and contains: geometry, year_step, pct_ch, etc.

# Create a highlight column: Top 10%, Bottom 10%, or Neutral
polyt_long <- polyt_long %>%
  group_by(year_step) %>%
  mutate(
    change_class = case_when(
      ntile(pct_ch, 10) == 10 ~ "Top 10%",
      ntile(pct_ch, 10) == 1  ~ "Bottom 10%",
      TRUE ~ "Neutral"
    )
  ) %>%
  ungroup()

# Order the factor to control legend
polyt_long$change_class <- factor(polyt_long$change_class, levels = c("Top 10%", "Bottom 10%", "Neutral"))

# Define colors: highlight extremes, gray everything else
change_colors <- c(
  "Top 10%" = "#d7191c",      # Red
  "Bottom 10%" = "#2c7bb6",   # Blue
  "Neutral" = "#eeeeee"       # Light gray
)

# Plot
ggplot(polyt_long, aes(fill = change_class)) +
  geom_sf(color = "gray", size = 0.00005) +  # Thin polygon borders
  facet_wrap(~ year_step, ncol = 2) +
  scale_fill_manual(values = change_colors) +
  coord_sf(expand = FALSE) +
  labs(
    title = "% Change in Nitrogen Retention by Basin",
    fill = "Change Class"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "bottom"
  )
```



library(gganimate)

ggplot(polyt_long, aes(fill = pct_ch)) +
  geom_sf() +
  scale_fill_viridis_c(option = "turbo") +
  theme_minimal() +
  labs(title = "Pct Change Over Time: {closest_state}") +
  transition_states(year_step, transition_length = 2, state_length = 1) +
  ease_aes('linear')

```



# 3.1 Visualize change distributiion 

Just a basic plot and actually looks quite normal jsut like that. Waht if we weighted by area?
Add area to the zonal_df object so we can use it to weight (not sure how righ now) but this needs to be incorporated

```{r plot diffs, fig.height=8, fig.width=14}
#filtered
z_fin <- zonal_df %>% filter(year=='diff')
library(ggplot2)

z_fin <- left_join(z_fin, zonal_wide[c(1,8)])


ggplot(z_fin, aes(x = Pot_Sed_ret_pct_ch)) +
  geom_density(fill = "skyblue", alpha = 0.6) +
  labs(
    title = "Density Plot of Value Column",
    x = "Value",
    y = "Density"
  ) +
  theme_minimal()

```


## 5 . Identify Hotspots 

Extract top 10 percent gains/loses. we don't even need something too fancy. Just load the data and extract it. Should try this in Pyton!!!!

```{r ggplot 2020, echo=FALSE, fig.height=8, fig.width=14}
# zonal_df <- load('/Users/rodriguez/Global_ES_TS/global_NCP/output_data/zonal_df_global.RData')
#zonal_df <- load(here('output_data', 'zonal_df_global.RData'))
  # Needed for reorder_within()


plot_ecosystem_services <- function(data, year, col) {
  col_sym <- sym(col)  # Convert column name to symbol for dplyr
  
  # Step 1: Prepare Data (Remove NA values and zero mean values, then reorder names)
  data_prepped <- data %>%
    filter(!is.na(mean) & mean > 0 & year == !!year) %>%  # Exclude cases where mean is 0 and filter year
    mutate(temp_col = reorder_within(!!col_sym, -mean, service))  

  # Step 2: Compute min/max values for selected column
  service_range <- data_prepped %>%
    group_by(service) %>%
    summarize(
      min_val = min(mean, na.rm = TRUE),
      max_val = max(mean, na.rm = TRUE),
      .groups = "drop"
    )

  # Step 3: Create "invisible" data for min/max range
  range_data <- service_range %>%
    pivot_longer(cols = c(min_val, max_val), names_to = "range_type", values_to = "mean") %>%
    mutate(year = !!year, temp_col = "dummy")

  # Step 4: Extract top 10 and bottom 10 per selected column
  top_10 <- data_prepped %>%
    group_by(service) %>%
    slice_max(order_by = mean, n = 10, with_ties = TRUE) 

  bottom_10 <- data_prepped %>%
    group_by(service) %>%
    slice_min(order_by = mean, n = 10, with_ties = TRUE)

  # Step 5: Combine only top 10 and bottom 10
  filtered_data <- bind_rows(top_10, bottom_10) %>%
    arrange(service, desc(mean))

  # Step 6: Reorder selected column to appear correctly
  filtered_data <- filtered_data %>%
    mutate(temp_col = reorder_within(!!col_sym, -mean, service))

  # Step 7: Plot the filtered data
  p <- ggplot(filtered_data, aes(x = temp_col, y = mean, fill = color)) +
    geom_bar(stat = "identity", show.legend = FALSE) +
    scale_fill_identity() +
    facet_wrap(~ service, scales = "free") +
    scale_x_reordered() +  
    labs(
      title = paste("Mean Ecosystem Service Values,", year),
      x = col,
      y = "Mean Value"
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(face = "bold", size = 12),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10)
    )
  
  return(p)
}

# Generate plots for different years using a specified column
p_1992 <- plot_ecosystem_services(zonal_df, 1992, "region_wb")
p_2020 <- plot_ecosystem_services(zonal_df, 2020, "region_wb")

# Display the plots
print(p_1992)
print(p_2020)


```




```{r barplot}
plot_ecosystem_services_diff <- function(data, col) {
  data_prepped <- data %>% filter(!is.na(mean)) %>% mutate(temp_col = reorder_within(!!sym(col), -mean, service))
  
  ggplot(data_prepped, aes(x = temp_col, y = mean, fill = color)) +
    geom_bar(stat = "identity", show.legend = FALSE) +
    scale_fill_identity() +
    facet_wrap(~ service, scales = "free") +
    scale_x_reordered() +  
    labs(title = "Difference in Mean Ecosystem Service Values (1992-2020)",
         x = col, y = "Mean Difference Value") +
    theme_bw()
}

p_diff <- plot_ecosystem_services_diff(zonal_df, col)
print(p_diff)
```





### 3.3 get change in terms of % (do at the end)



This is a WIP. I need to be able to add the numbers as columns at the end....nooo
easier. We are going to calculate them dikrectly fro mthe filtered zonal_df dataframe, but not now. We need to think about the differences calculated and the % of difference. anyway i already have that piece of data. 

```{r extract change in %}

# Function to compute percentage change for all relevant variables in an sf object
calculate_percentage_change <- function(sf_obj) {
  # Identify all variables that exist in both 1992 and 2020
  base_vars <- service#c("Coastal Protection", "Nitrogen Export", "Sediment Export", "Nature Access", "Pollination")
  
  # Iterate through each variable to compute percentage change
  for (var in base_vars) {
    col_1992 <- paste0(var, "_1992")  # Column name for 1992
    col_2020 <- paste0(var, "_2020")  # Column name for 2020
    col_pct  <- paste0(var, "_pct_ch")   # New percentage change column
    
    # Check if the required columns exist in the sf object
    if (all(c(col_1992, col_2020) %in% names(sf_obj))) {
      sf_obj <- sf_obj %>%
        mutate(!!col_pct := ((!!sym(col_2020) - !!sym(col_1992)) / !!sym(col_1992)) * 100)
    }
  }
  return(sf_obj)
}

# Apply function to all sf objects in the list
#pols <- lapply(poly, calculate_percentage_change)
poly <- calculate_percentage_change(poly)
# Check the updated column names

# get pct as standalone data to add to the table
plt <- st_drop_geometry(poly)
plt  <- plt %>%
  select(1, contains("pct"))
################## THIS IS VERY IMPPORTAnt. Instead of struggling  with the multiple dataframes, it is easier to load the vector file swith all the column and pivot longer as necessary. Easier to manage, adjust on the fly!
plt <- plt %>%
  pivot_longer(
    cols = ends_with("pct_ch"),  # Select all columns ending with "pct_ch"
    names_to = "service",        # New column to store the service names
    values_to = "pct_ch"         # New column to store the percentage change values
  )


write.csv(plt, here('output_data', paste0('pct_chg_',cols, '.csv')))

#st_write(poly, here('vector', set), append = FALSE) 
```




## 5. Generate  Maps

Use ggplot2 and the spatial geometry from our sf object to color each polygon by its statistic of interest. Faceting again by each raster is useful to compare spatial patterns.
```{r create maps, include=FALSE}


# First, merge the zonal_df results back into the polygon sf object.
# This ensures each polygon gets a corresponding mean/median value.

# We need a unique identifier. 
# Assuming 'country' is unique per polygon, we can do a left_join:
map_data <- poly %>%
  left_join(zonal_df, by = "name_long")

map_1992 <- map_data %>% filter(year == 1992)
services <- unique(map_1992$service)
map_2020 <-  map_data %>% filter(year == 2020)


# 2) Identify unique services
services <- unique(map_data$service)

# 3) For each service, find min/max across both years
service_range <- map_data %>%
  group_by(service) %>%
  summarize(
    min_val = min(mean, na.rm = TRUE),
    max_val = max(mean, na.rm = TRUE),
    unique_color = unique(color)  # we assume 1 color per service
  )

plot_list_1992 <- list()
plot_list_2020 <- list()

for (s in services) {
  # Extract global range (across both years) for service s
  row_s <- filter(service_range, service == s)
  s_min <- row_s$min_val
  s_max <- row_s$max_val
  s_color <- row_s$unique_color  # The single "unique" color for that service
  
  # Subset to year=1992, that service
  df_1992 <- map_data %>%
    filter(service == s, year == 1992)
  
  # Subset to year=2020, that service
  df_2020 <- map_data %>%
    filter(service == s, year == 2020)
  
  # Build the 1992 map
  p_1992 <- ggplot(df_1992) +
    geom_sf(aes(fill = mean), color = NA) +
    scale_fill_gradientn(
      colors   = c("white", s_color),  # White -> "unique_color"
      limits   = c(s_min, s_max),      # Force the same scale min/max
      na.value = "gray40"
    ) +
    labs(
      title = paste0(s, " (1992)"),
      fill  = "Mean"
    ) +
    theme_minimal()
  
  # Build the 2020 map
  p_2020 <- ggplot(df_2020) +
    geom_sf(aes(fill = mean), color = NA) +
    scale_fill_gradientn(
      colors   = c("white", s_color),
      limits   = c(s_min, s_max),
      na.value = "gray40"
    ) +
    labs(
      title = paste0(s, " (2020)"),
      fill  = "Mean"
    ) +
    theme_minimal()
  
  # Add them to the lists
  plot_list_1992[[s]] <- p_1992
  plot_list_2020[[s]] <- p_2020
}
```

```{r echo=FALSE, fig.height=12, fig.width=14}

# Combine the 1992 plots in a row (or multi-row if you prefer).
combined_1992 <- wrap_plots(plot_list_1992, ncol = 3)

# Combine the 2020 plots similarly.
combined_2020 <- wrap_plots(plot_list_2020, ncol = 3)

combined_1992
combined_2020
```




rasters_list <- lapply(tiffes, rast)
service <- c("Coastal Protection", "Nature Access","Nitrogen Export","Pollination","Sediment Export", "Usle")
service <- gsub(" ", "_", service)




