---
title: "Summary ES"
author: "Jeronimo Rodriguez-Escobar"
date: "2025-02-12"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

  
1.	Systematically extract zonal statistics from ES global rastetsfor specific spatial units (country/territory, groups of countries, biomes, wathersheds, regular grids etc. The basic procerure is built around the R   *exactextractr* package. Reference to exactextractr is availablle in this [GitHub repository](https://github.com/isciences/exactextractr).

2. Add and label the extracted information as additional attributes to the input polygon datasets.

3. Create visualizations to represented the extracted statistics.

Obs: certain situations and cases need t be taken care more carefgully, especially these related with dependencies and, mutiple features corrsponding to the same country, that can yield wierd results. 
A second aspect to consider is related to small island nations and other geographic artifacts thart can 
Pending to do: Think how to deal with the dependencies and complex cases (GB, France, others), but i will take care of this at a later point.

```{r setup, include=FALSE}
# Load required libraries
library(sf)             # For reading and working with vector data
library(dplyr)          # For data manipulation
library(terra)          # For reading raster data (SpatRasters)
library(exactextractr)  # For zonal statistics
library(ggplot2)        # For plotting
library(forcats)        # For factor reordering
library(tidytext)
library(here)
library(patchwork)
library(tidyr)
library(parallel)
```

## 1. Load Polygons

This can be done using spatial vector files stored locally or using existing online available datasets like **GADM** 

```{r load polygons, eval=FALSE, include=FALSE}

# This assumes that the polygon files are stored in the same working directory. and have the names assigned here. Feel free to change it if necessary.  
# The differnet subsets except for biomes are obtained from the dataset provided by Justin to which a "dissolve by attribute" operation has been performed. This is better done in Python/QGis/ArcGis, R is not great at that. 
# By Continent
#set <- 'continent.gpkg' 
# By Subregion
#set <- 'subregion.gpkg'
# By individual territory; "Country" 
#set <- 'Country.gpkg'
# By WWF biome
# set <- 'Biome.gpkg' 


poly <- st_read(here('vector', set)) 

# Set the column with the attribute to group 
# Test this, it might be possible that it is not necesary to dissolve and create the individual vector files in advance, but just change the column. However, i need to check how would it work when joining the new results as attributes. 
# There would be two things to consider here:
# 1. how would it actually look when displaying on a GIS (all the original polygons will be maintained). This is manageable. Waht would this imply for the subsequent analyisis.
# 2. Practicality considerations. The final output, with all services, grouping, metrics, years can end up having too many columns at at some point that can become hard to manage. Just naming columns can be a hassle. 

col <- "WWF_biome"
cols <- col
```


## 2. Load a List of GeoTIFF Rasters

Following products and years are assessed:
```{r load raster es, echo=FALSE}
# Vector of raster file paths
inpath <- here('input_ES')

tiffes <- file.path(inpath, list.files(paste0(inpath),pattern= 'tif$'))
filename<- basename(tiffes)
# Use terra::rast() to load each file as a SpatRaster object
rasters_list <- lapply(tiffes, rast)
# add the names of the services. This is slightly annoying but necessary. We need to make sure that the order is the same of the input data, which depends on the names of the input tiff files and that can change. Check that BEFORE running.  # there is room for improvement here, but life would be easier if the input rasters have a more consisntent naming convention. 
service <- rep(c("Coastal Protection", "Nitrogen Export", "Sediment Export", "Nature Access", "Pollination"), each=2)
year <- rep(c(1992,2020), rep=5)
filename<- as_tibble(cbind(service,filename, year))
print(filename)


# create a table with the content to fill the table
raster_name <- filename$filename
# Maybe it is possible to extract automatically the specific service name being analyzed, but that is contingent of the names ofthe input rasters, and these can be quite complex and inconsistent
service <- rep(c("Coastal Protection", "Nitrogen Export", "Sediment Export", "Nature Access", "Pollination"), each = length(unique(year)))
color <- rep(c("#9e9ac8", "#2c944c", "#08306b", "#A57C00", "#dd1c77"), each=length(unique(year)))
cd <- as_tibble(cbind(raster_name, service, color))
```

# 3. Calcualte Differences 

The difference in the values between both yars is calculatd with a simple substraction and the results are exported to and target destination.


```{r get differences, eval=FALSE, include=FALSE}
outpath <- here("input_ES")
m_index <- length(rasters_list) %/% 2
lab <- unique(service)
l1 <- rasters_list[seq(1,length(rasters_list), by=2)]
l2 <- rasters_list[seq(2,length(rasters_list), by=2)]
 # test with future_map2 next time to get this to run faster. 
lab <- unique(filename$service)

diffs <- map2(l1,l2, function(l1,l2){
  return(l1-l2)
})
# export new rasters 
map(1:length(diffs), function(x) writeRaster(diffs[[x]], paste0(outpath, "/", 'change_calc', '/', lab[x], '_diff.tif')))
```

## 4. Compute Zonal Statistics for each target raster/year 

**Note:** Although Exact exactextractr is used  here ot obtain zonal summary statistics, in this case the mean value for each polygon, it can also return a list of dataframes, one per feature with the cell values from pte input raster. This can be very useful for further analysis. 
Besides, it is also possible to include user-defined functions (somerthing that i am going to use for the USLE analysis).

```{r compute stats 1, eval=FALSE, include=FALSE}
#load the data (this assumes 'tiffs' is the list of paths to the files, and that they are in the desired order. Make sure that the dataframe with the labels is correct (filename object)  There is room for improvement here. 

# Compute zonal stats. Mean for now. Ohers can be added. Cehck extractextractr documentation for a detailed list of stats 
target_stats <- c("mean")
# this is the column in the vector file that we are going to use. 

num.cores <- length(rasters_list) # set as many cores as rasters to be evaluated.  sequentially, and there is no progress bar available 
# here, set a way to change the input poly and columns externally instead of editing inside of the functions. That is cumbersome and prone to errors. 

results_list <- lapply(rasters_list, function(r) {
  # Perform exact extraction
  # - append_cols = "country" retains the country identifier with each result
  res <- exact_extract(r, poly,
                       fun = target_stats,
                       append_cols = cols)
  r_name <- names(r)
  # Add a column labeling which raster these results are from
  res$raster_name <- r_name
  return(res)})#,mc.cores = num.cores)

# Combine results from all rasters into one data frame
zonal_df <- do.call(rbind, results_list)

# create a table with the content to fill the table
raster_name <- unique(zonal_df$raster_name)
service <- rep(c("Coastal Protection", "Nitrogen Export", "Sediment Export", "Nature Access", "Pollination"), each = 2)
color <- rep(c("#9e9ac8", "#2c944c", "#08306b", "#A57C00", "#dd1c77"), each=2)
cd <- as_tibble(cbind(raster_name, service, color))


# add color ands service columns
zonal_df <- left_join(zonal_df,cd)
# Add the year. this uses the original filenames to extract the year, but miht be too hardcoded for generalization 

zonal_df$year <- ifelse(grepl("1992", zonal_df$raster_name), 
                          1992, 
                          ifelse(grepl("2020", zonal_df$raster_name), 2020, NA))


zonal_wide <- pivot_wider(zonal_df, id_cols=all_of(col), names_from=c(service, year), values_from = mean)
# adjust column names #

######### HERE: Include a way/convention to adjust the column names autmatically. Saves work and keeps consistency. I had to do this in Stpe 6, but this is not organized. should be done herer bedore exporting

poly <- left_join(poly, zonal_wide, by= col)
st_write(poly, here('vector', set), append = FALSE) 
write.csv(zonal_df, here('output_data', paste0('zonal_df_',col, '.csv')))

```


## 5. Extract zonal stats for the differences.  

In theory, i could get everything done on a single run together but it somehow gets too complicated. The names of the input rasters don't always follow the same sequence, and sometiomes need to be rearranged. 
There are two main cases: dealing with two or more different  sets of layers (eg one for each year). The second a case with on;ly one set (dealing wit the differneces for example). This is somethign that is really not that complicated and could be organized/simplified here, but needs some work that is not priority right now.

I will have to review this part. Ideally, i should get everything done together but the year issue made it tricky. Easier to do each on its owrn, the problem is that there is still too much manual handling. Eventually this should be generalized.

```{r compute stats diff, eval=FALSE, include=FALSE}

# set the path and load the rasters with the differences that we calcualted in step 3
#build paths to files
inpath <- here('input_ES')
inpath <- paste0(inpath, "/", 'change_calc')

# load the rasters with the calculated differences 
tiffes <- file.path(inpath, list.files(paste0(inpath),pattern= 'tif$'))
filename<- basename(tiffes)


rast_list <- lapply(tiffes, rast)
service <- c("Coastal Protection", "Nature Access","Nitrogen Export","Pollination","Sediment Export")
filename<- as_tibble(cbind(service,filename))
# We'll compute mean 
target_stats <- "mean"

# Number of CPU cores to use. I am still struggling with setting the this using a multicore approach. I have been using mclapply to do this, but the behavior is kind of inconsistent, sometimes it seems to hang without finishing correclty, there is no way to show the progress and error messages are not informative at all. 
#Also, keep in mind that mclapply on;y works with Unix alike systems (Linux, MacOS but is not availalbe in Widnows. Someone (most likely me) will have to take care of this at some point) 

num.cores <- length(rast_list)
results_list <- mclapply(rast_list, function(r) {
  # Perform exact extraction
  # - append_cols = "country" retains the country identifier with each result
  res <- exact_extract(r, poly,
                       fun = target_stats,
                       append_cols = cols)
  r_name <- names(r)

  # Add a column labeling which raster these results are from
  res$raster_name <- r_name
  return(res)}),mc.cores = num.cores)

# Combine results from all rasters into one data frame
zonal_df <- do.call(rbind, results_list)

zonal_df <- left_join(zonal_df,cd)
#zonal_df <- zonal_df[-c(4,6)]
zonal_wide <- pivot_wider(zonal_df, id_cols=all_of(cols), names_from=c(service), values_from = mean)
# adjust column names 
colnames(zonal_wide)[colnames(zonal_wide) != cols] <- paste0(colnames(zonal_wide)[colnames(zonal_wide) != cols], "_diff")

poly <- left_join(poly, zonal_wide, by= cols)
st_write(poly, here('vector', set), append = FALSE) 

write.csv(zonal_df, here('output_data', paste0('diffs_',cols, '.csv')))
```

## 6. Fix column names
```{r fix column names vectors}
inpath <- here('vector')

vecf <- file.path(inpath, list.files(paste0(inpath),pattern= 'gpkg$'))
vecf <- vecf[-4]
filename<- basename(vecf)
grp <- c("Continent", "Country", "Income Group", "World Bank Region", "Biome")

filename<- as_tibble(cbind(grp,filename))
pols <- lapply(vecf, st_read)

# Rename columns in pols[[2]]
pols[[4]] <- pols[[4]] %>%
  rename_with(~ gsub("^mean_", "", gsub("\\.", " ", .x)), starts_with("mean_"))

# Check the new column names
names(pols[[4]])
```




```{r extract change in %}

# Function to compute percentage change for all relevant variables in an sf object
calculate_percentage_change <- function(sf_obj) {
  # Identify all variables that exist in both 1992 and 2020
  base_vars <- c("Coastal Protection", "Nitrogen Export", "Sediment Export", "Nature Access", "Pollination")
  
  # Iterate through each variable to compute percentage change
  for (var in base_vars) {
    col_1992 <- paste0(var, "_1992")  # Column name for 1992
    col_2020 <- paste0(var, "_2020")  # Column name for 2020
    col_pct  <- paste0(var, "_pct_ch")   # New percentage change column
    
    # Check if the required columns exist in the sf object
    if (all(c(col_1992, col_2020) %in% names(sf_obj))) {
      sf_obj <- sf_obj %>%
        mutate(!!col_pct := ((!!sym(col_2020) - !!sym(col_1992)) / !!sym(col_1992)) * 100)
    }
  }
  return(sf_obj)
}

# Apply function to all sf objects in the list
pols <- lapply(pols, calculate_percentage_change)

# Check the updated column names
lapply(pols, names)


# Standardize filenames by removing spaces from the "grp" column
filename <- filename %>%
  mutate(grp = gsub(" ", "_", grp))  # Replace spaces with underscores

# Define the output directory
output_dir <- here("vector")  # Ensure "vector" directory exists

# Save each sf object in pols as a GPKG file
lapply(seq_along(pols), function(i) {
  filepath <- file.path(output_dir, paste0(filename$grp[i], ".gpkg"))
  st_write(pols[[i]], filepath, delete_layer = TRUE)
})

```


## 7 . Get the Barplots

```{r ggplot 2020, echo=FALSE, fig.height=8, fig.width=14}
#zonal_df <- as_tibble(read.csv(here('output_data', 'zonal_df_global.csv')))
#zonal_df <- zonal_df %>% select(id, mean, service, color, year, name_long)
zonal_df <- zonal_df %>% select(BIOME, WWF_biome, mean, service, color, year)
zonal_df <- zonal_df %>% select(id, mean, service, color, year, name_long)

plot_ecosystem_services <- function(data, year, col) {
  col_sym <- sym(col)  # Convert column name to symbol for dplyr
  
  # Step 1: Prepare Data (Remove NA values and zero mean values, then reorder names)
  data_prepped <- data %>%
    filter(!is.na(mean) & mean > 0 & year == !!year) %>%  # Exclude cases where mean is 0 and filter year
    mutate(temp_col = reorder_within(!!col_sym, -mean, service))  

  # Step 2: Compute min/max values for selected column
  service_range <- data_prepped %>%
    group_by(service) %>%
    summarize(
      min_val = min(mean, na.rm = TRUE),
      max_val = max(mean, na.rm = TRUE),
      .groups = "drop"
    )

  # Step 3: Create "invisible" data for min/max range
  range_data <- service_range %>%
    pivot_longer(cols = c(min_val, max_val), names_to = "range_type", values_to = "mean") %>%
    mutate(year = !!year, temp_col = "dummy")

  # Step 4: Extract top 10 and bottom 10 per selected column
  top_10 <- data_prepped %>%
    group_by(service) %>%
    slice_max(order_by = mean, n = 10, with_ties = TRUE) 

  bottom_10 <- data_prepped %>%
    group_by(service) %>%
    slice_min(order_by = mean, n = 10, with_ties = TRUE)

  # Step 5: Combine only top 10 and bottom 10
  filtered_data <- bind_rows(top_10, bottom_10) %>%
    arrange(service, desc(mean))

  # Step 6: Reorder selected column to appear correctly
  filtered_data <- filtered_data %>%
    mutate(temp_col = reorder_within(!!col_sym, -mean, service))

  # Step 7: Plot the filtered data
  p <- ggplot(filtered_data, aes(x = temp_col, y = mean, fill = color)) +
    geom_bar(stat = "identity", show.legend = FALSE) +
    scale_fill_identity() +
    facet_wrap(~ service, scales = "free") +
    scale_x_reordered() +  
    labs(
      title = paste("Mean Ecosystem Service Values,", year),
      x = col,
      y = "Mean Value"
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(face = "bold", size = 12),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 5)
    )
  
  return(p)
}

# Generate plots for different years using a specified column
p_1992 <- plot_ecosystem_services(zonal_df, 1992, col)
p_2020 <- plot_ecosystem_services(zonal_df, 2020, col)

# Display the plots
print(p_1992)
print(p_2020)


```

## 8.Difference barplots 

```{r plot diffs, fig.height=8, fig.width=14}

zonal_df <- as_tibble(read.csv(here('output_data', 'diffs_wb_.csv')))

col <- 'region_wb'

plot_ecosystem_services1 <- function(data, col) {
  col_sym <- sym(col)  # Convert column name to symbol for dplyr
  
  # Step 1: Prepare Data (Remove NA values and zero mean values, then reorder names)
  data_prepped <- data %>%
    filter(!is.na(mean)) %>%  # Exclude cases where mean is 0
    mutate(temp_col = reorder_within(!!col_sym, -mean, service))  

  # Step 2: Compute min/max values for selected column
  service_range <- data_prepped %>%
    group_by(service) %>%
    summarize(
      min_val = min(mean, na.rm = TRUE),
      max_val = max(mean, na.rm = TRUE),
      .groups = "drop"
    )

  # Step 3: Create "invisible" data for min/max range
  range_data <- service_range %>%
    pivot_longer(cols = c(min_val, max_val), names_to = "range_type", values_to = "mean") %>%
    mutate(temp_col = "dummy")

  # Step 4: Extract top 10 and bottom 10 per selected column
  top_10 <- data_prepped %>%
    group_by(service) %>%
    slice_max(order_by = mean, n = 10, with_ties = TRUE) 

  bottom_10 <- data_prepped %>%
    group_by(service) %>%
    slice_min(order_by = mean, n = 10, with_ties = TRUE)

  # Step 5: Combine only top 10 and bottom 10
  filtered_data <- bind_rows(top_10, bottom_10) %>%
    arrange(service, desc(mean))

  # Step 6: Reorder selected column to appear correctly
  filtered_data <- filtered_data %>%
    mutate(temp_col = reorder_within(!!col_sym, -mean, service))

  # Step 7: Plot the filtered data
  p <- ggplot(filtered_data, aes(x = temp_col, y = mean, fill = color)) +
    geom_bar(stat = "identity", show.legend = FALSE) +
    scale_fill_identity() +
    facet_wrap(~ service, scales = "free") +
    scale_x_reordered() +  
    labs(
      title = paste("Difference Mean Ecosystem Service Values, 1992-2020"),
      x = col,
      y = "Mean Diff Value"
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(face = "bold", size = 12),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 7)
    )
  
  return(p)
}


# Generate plots for different years using a specified column
p_diff <- plot_ecosystem_services1(zonal_df, col)


# Display the plots
print(p_diff)

```

## 9. Generate  Maps

Use ggplot2 and the spatial geometry from our sf object to color each polygon by its statistic of interest. Faceting again by each raster is useful to compare spatial patterns.
```{r create maps, eval=FALSE, include=FALSE}

# Very important. This line adds the new attribnutes to the polygons. I could add all here.

col <- "region_wb"

zonal_df <-zonal_df %>% select(c(region_wb,mean, service)) 
zonal_df <- zonal_df %>% rename(diff_mean = mean)

# this is to pivot when there are two dates (each year)
zonal_wide <- pivot_wider(zonal_df, id_cols=!!sym(col), names_from=c(service,year), values_from = c(mean, median))



# this is for when there i only one value (eg differences)
zonal_wide <- pivot_wider(zonal_df, id_cols=!!sym(col), names_from= service, values_from = diff_mean)


#rename the columns to avoid confusions
zonal_wide <- zonal_wide %>%
  rename_with(~ paste0("diff_", .), -1) 

poly <- left_join(poly, zonal_wide, by = col)

st_write(poly,  here('vector', 'wb_poly_means.gpkg'), append=FALSE) 

```
####################################################

```{r mapping f}
# First, merge the zonal_df results back into the polygon sf object.
# This ensures each polygon gets a corresponding mean/median value.

# We need a unique identifier. 
# Assuming 'country' is unique per polygon, we can do a left_join:
map_data <- poly %>%
  left_join(zonal_df, by = "id")

map_1992 <- map_data %>% filter(year == 1992)
services <- unique(map_1992$service)
map_2020 <-  map_data %>% filter(year == 2020)


# 2) Identify unique services
services <- unique(map_data$service)

# 3) For each service, find min/max across both years
service_range <- map_data %>%
  group_by(service) %>%
  summarize(
    min_val = min(mean, na.rm = TRUE),
    max_val = max(mean, na.rm = TRUE),
    unique_color = unique(color)  # we assume 1 color per service
  )

plot_list_1992 <- list()
plot_list_2020 <- list()

for (s in services) {
  # Extract global range (across both years) for service s
  row_s <- filter(service_range, service == s)
  s_min <- row_s$min_val
  s_max <- row_s$max_val
  s_color <- row_s$unique_color  # The single "unique" color for that service
  
  # Subset to year=1992, that service
  df_1992 <- map_data %>%
    filter(service == s, year == 1992)
  
  # Subset to year=2020, that service
  df_2020 <- map_data %>%
    filter(service == s, year == 2020)
  
  # Build the 1992 map
  p_1992 <- ggplot(df_1992) +
    geom_sf(aes(fill = mean), color = NA) +
    scale_fill_gradientn(
      colors   = c("white", s_color),  # White -> "unique_color"
      limits   = c(s_min, s_max),      # Force the same scale min/max
      na.value = "gray40"
    ) +
    labs(
      title = paste0(s, " (1992)"),
      fill  = "Mean"
    ) +
    theme_minimal()
  
  # Build the 2020 map
  p_2020 <- ggplot(df_2020) +
    geom_sf(aes(fill = mean), color = NA) +
    scale_fill_gradientn(
      colors   = c("white", s_color),
      limits   = c(s_min, s_max),
      na.value = "gray40"
    ) +
    labs(
      title = paste0(s, " (2020)"),
      fill  = "Mean"
    ) +
    theme_minimal()
  
  # Add them to the lists
  plot_list_1992[[s]] <- p_1992
  plot_list_2020[[s]] <- p_2020
}
```

```{r eval=FALSE, fig.height=12, fig.width=14, include=FALSE}

# Combine the 1992 plots in a row (or multi-row if you prefer).
combined_1992 <- wrap_plots(plot_list_1992, ncol = 3)

# Combine the 2020 plots similarly.
combined_2020 <- wrap_plots(plot_list_2020, ncol = 3)

combined_1992
combined_2020


'#9e9ac8'
'#2c944c'
```

Next steps:  

Run for: Biomes (done), Continents, income group (done),  wb_region(done), subregion (done) 

Pending: USLE.

Wahtif we try thios at the subregional (using GADM level 2) or hydroshed level? That will be the next step. 



Land cover change -> assess change/area In other words what do i **exactly** need to do here?  What is my target variable? LC change? 

Will have to do it using DiffeR and iterate over all countries, then assemble the dataframes and get percentage of change distributed among the main components. This is not going to be easy but i can get it. However, keep in mind that with all the original classes this becomes unmanageable. Can we produce successive no change maps and check what happens among the change (alloation, swaps and shifts)

Take care of the beneficiaries. -> i think exactextractr has the solution. We can get all this data/syntesis stats including a weighing raster (yay) and also include user defined fucntiosn to extract the target stats. 








