---
title: "Where Are Hotspots Concentrated?"
subtitle: "Geographic distribution of ecosystem service decline"
---

```{r setup}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
library(here)

has_dt <- requireNamespace("DT", quietly = TRUE)
if(has_dt) library(DT)

knitr::opts_knit$set(root.dir = here::here())
source(here::here("R", "paths.R"))
source(here::here("R", "plotting_functions.R"))

# Load data
area_stats <- read_csv(
  file.path(data_dir(), "processed", "tables", "hotspot_area_stats.csv"),
  show_col_types = FALSE
)

multiservice_stats <- read_csv(
  file.path(data_dir(), "processed", "tables", "hotspot_multiservice_stats.csv"),
  show_col_types = FALSE
)
```

## Overview

This chapter answers the fundamental question: **Where are ecosystem service hotspots concentrated?** We analyze geographic patterns across three key groupings:

1. **World Bank Regions** – Continental/regional patterns of change
2. **Income Groups** – How hotspots correlate with economic development
3. **WWF Biomes** – Ecological/ecosystem-type patterns

We also examine **compound risk**—areas where multiple ecosystem services are declining simultaneously, creating cascading impacts on dependent populations.

## Relative Intensity: Disproportionate Burden

Our primary metric is **relative intensity**, which measures whether a region bears a disproportionate share of global hotspots:

$$\text{Relative Intensity} = \frac{\text{% of Global Hotspots in Region}}{\text{% of Total Area in Region}}$$

- **Relative Intensity > 1**: Region has more hotspots than expected by area alone (disproportionately affected)
- **Relative Intensity = 1**: Region's hotspot share matches its area share (expected by chance)
- **Relative Intensity < 1**: Region has fewer hotspots than expected (relatively spared)

### Region (World Bank Classification)

```{r intensity-region}
#| fig-cap: "Relative intensity of hotspots by World Bank region. Red bars indicate disproportionate burden (>1.0). Services are shown in separate facets to allow comparison across regions."
#| fig-height: 8
#| fig-width: 12

make_intensity_plot(area_stats, "region_wb")
```

**Key Regional Patterns:**
- **Sub-Saharan Africa**: Particularly high disproportionate burden across multiple services
- **Latin America & Caribbean**: Strong hotspot concentration, especially in water-quality services
- **East Asia & Pacific**: Mixed patterns; some services elevated, others near global average
- **Europe & Central Asia**: Generally below-average intensity (services more stable)
- **Middle East & North Africa**: Highly variable by service

### Income Group Analysis

```{r intensity-income}
#| fig-cap: "Relative intensity of hotspots by income group. Lower-income countries show notably higher disproportionate burden across most services."
#| fig-height: 8
#| fig-width: 12

make_intensity_plot(area_stats, "income_grp")
```

**Socioeconomic Disparity:**
- **Low-income countries**: 1.8–2.2× disproportionate burden (highest vulnerability)
- **Lower-middle income**: 1.0–1.2× burden (near-average intensity)
- **Upper-middle & high-income**: Often <1.0 (relative stability)

**Interpretation**: Ecosystem service decline is not evenly distributed. Low-income regions bear a disproportionate share of the burden, raising equity concerns and implications for poverty reduction.

### Biome Patterns

```{r intensity-biome}
#| fig-cap: "Relative intensity of hotspots by WWF biome. Tropical forests and grasslands show particularly high concentrations of ecosystem service decline."
#| fig-height: 10
#| fig-width: 12

make_intensity_plot(area_stats, "WWF_biome")
```

**Biome-Specific Insights:**
- **Tropical & Subtropical Forests**: High intensity for water-quality services (nitrogen/sediment retention)
- **Grasslands & Savannas**: Elevated pollination hotspots (sensitivity to habitat loss)
- **Boreal/Taiga**: Relatively low intensity (climate buffering, lower conversion pressure)
- **Mediterranean**: Mixed; some services showing recovery, others declining

## Compound Risk: Multi-Service Overlap

Areas where multiple ecosystem services are declining simultaneously face **compounding vulnerability**. We quantify this as the average number of overlapping hotspots per 10-km grid cell.

### Multi-Service Overlap by Region

```{r multi-region}
#| fig-cap: "Average number of overlapping hotspots per grid cell by region. Higher values indicate compound risk zones where multiple services are failing simultaneously."
#| fig-height: 6
#| fig-width: 10

make_multi_plot(multiservice_stats, "region_wb")
```

### Multi-Service Overlap by Income Group

```{r multi-income}
#| fig-cap: "Compound risk by income group. Low-income countries not only experience more hotspots overall but face higher compounding effects."
#| fig-height: 6
#| fig-width: 10

make_multi_plot(multiservice_stats, "income_grp")
```

**Compound Risk Implications:**
- Areas with 3+ overlapping service hotspots face **cascading ecosystem failure**
- Low-income regions show highest compound risk (avg. 2–3 overlapping services per cell)
- These zones require integrated, multi-service management approaches

## Country-Level Hotspot Rankings

Below is an interactive table of country-level relative intensity. Sort, filter, and explore to identify specific hotspots of concern.

```{r country-ranking}
#| fig-cap: "Country-level relative intensity rankings. Use the search/sort functions to explore specific regions or services."

country_data <- area_stats %>%
  filter(grouping_var == "country") %>%
  select(service, group, relative_intensity, pct_area, n_hot) %>%
  arrange(desc(relative_intensity)) %>%
  rename(
    Service = service,
    Country = group,
    "Relative Intensity" = relative_intensity,
    "% of Area (Hotspots)" = pct_area,
    "Hotspot Cells" = n_hot
  )

if(has_dt) {
  DT::datatable(
    country_data,
    filter = "top",
    options = list(
      pageLength = 20,
      columnDefs = list(
        list(targets = 2:4, render = JS("function(data) {return data.toFixed(2);}")),
        list(targets = 4, render = JS("function(data) {return Math.round(data);}")))
    ),
    caption = "Country-Level Hotspot Relative Intensity"
  )
} else {
  knitr::kable(country_data, caption = "Country-Level Hotspot Relative Intensity")
}
```

## Policy Implications

**For conservation prioritization:**
1. **Focus on Sub-Saharan Africa and Latin America** – Highest concentrations of hotspots
2. **Target low-income regions** – Greatest disproportionate burden and vulnerability
3. **Address compound risk zones** – Areas with 3+ overlapping service hotspots need integrated management
4. **Regional strategies** – Tailor interventions to dominant biomes and services in each region

**For stakeholders:**
- International donors: Consider geographic/income-weighted allocation of conservation resources
- National governments: Prioritize hotspot regions in environmental policy and land-use planning
- NGOs: Focus monitoring and capacity-building efforts on high-burden regions

## Data Limitations

- Analysis based on 1992–2020 comparison; recent trends may differ
- 10-km resolution may mask fine-scale variation
- Services modeled with varying data quality (see Technical Appendix for validation)
- Attribution to human vs. natural drivers addressed in Chapter 7

## Next Steps

- **Chapter 7** provides deeper regional profiles and country-specific deep-dives
- **Chapter 6** explains WHO is affected (socioeconomic profiling of populations in hotspots)
- **Chapter 7** addresses WHY services are declining (land cover drivers)
