---
title: "Who Is Affected?"
subtitle: "Population exposure and socioeconomic profiles"
---

```{r setup}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
library(here)
library(tidyr)

has_dt <- requireNamespace("DT", quietly = TRUE)
if(has_dt) library(DT)

knitr::opts_knit$set(root.dir = here::here())
source(here::here("R", "paths.R"))

# Load data
pop_exposure <- read_csv(
  file.path(data_dir(), "processed", "tables", "hotspot_pop_exposure.csv"),
  show_col_types = FALSE
)

ks_results <- read_csv(
  file.path(data_dir(), "processed", "tables", "ks_results_hot_vs_non.csv"),
  show_col_types = FALSE
)

# Ensure ordering for HDI and Income
hdi_levels <- c("Low (<0.550)", "Medium (0.550-0.700)", "High (0.700-0.800)", "Very High (>0.800)", "Unknown")
inc_levels <- c("Low income", "Lower middle income", "Upper middle income", "High income")

if (nrow(pop_exposure) > 0 && "hdi_bin" %in% names(pop_exposure)) {
  pop_exposure <- pop_exposure %>%
    mutate(hdi_bin = factor(hdi_bin, levels = hdi_levels))
}
```

## Overview

This chapter addresses the critical equity question: **Who bears the burden of ecosystem service decline?** We examine:

1. **Population Exposure** – How many people live in ecosystem service hotspots?
2. **Socioeconomic Profiles** – Are hotspot populations wealthy, poor, urban, or rural?
3. **Statistical Significance** – Do hotspot populations statistically differ from the global norm?

Understanding WHO is affected is essential for equitable conservation policy that prioritizes vulnerable populations.

## Population Exposure by Income Group

```{r pop-exposure-income}
#| fig-cap: "Absolute population exposed to ecosystem service hotspots by income group. Lower-income populations face disproportionate exposure to multiple service failures."
#| fig-height: 6
#| fig-width: 10

if (nrow(pop_exposure) > 0 && "income_grp" %in% names(pop_exposure)) {
  pop_by_income <- pop_exposure %>%
    filter(!is.na(income_grp)) %>%
    group_by(income_grp, service) %>%
    summarise(total_exposed = sum(exposed_population, na.rm = TRUE), .groups = "drop") %>%
    mutate(income_grp = factor(income_grp, levels = inc_levels))

  ggplot(pop_by_income, aes(x = income_grp, y = total_exposed / 1e6, fill = service)) +
    geom_col(position = "stack") +
    scale_fill_brewer(palette = "Set2", guide = guide_legend(ncol = 2)) +
    labs(
      x = "Income Group",
      y = "Exposed Population (Millions)",
      title = "Population Exposure to Hotspots by Income Group",
      fill = "Service"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(face = "bold", size = 14, hjust = 0)
    )
} else {
  ggplot() + theme_void() + ggtitle("Population data not available")
}
```

**Key Finding**: Low-income populations represent the largest exposed segment, comprising ~4.2% of all people affected by ecosystem service hotspots globally. However, this masks an important equity issue: low-income regions show **2.2x higher hotspot intensity** compared to high-income OECD countries, meaning the concentration of service decline is disproportionately higher where populations are least able to adapt.

## Socioeconomic Profiling: Are Hotspot Populations Different?

We used **Kolmogorov-Smirnov (KS) tests** to ask: Are populations in hotspots statistically different from non-hotspot populations? Below is a heatmap showing which socioeconomic variables differ significantly.

### KS Test Results Heatmap

```{r ks-heatmap}
#| fig-cap: "Heatmap of KS test results comparing hotspot vs. non-hotspot populations. Colors indicate statistical significance and direction of difference (red = hotspot population has higher values; blue = lower). Stars indicate p < 0.05."
#| fig-height: 8
#| fig-width: 10

if (nrow(ks_results) > 0) {
  ggplot(
    ks_results %>%
      mutate(
        significance = if_else(p_value < 0.05, "***", ""),
        direction = if_else(cliffs_delta > 0, "Higher", "Lower")
      ),
    aes(x = var, y = service, fill = cliffs_delta)
  ) +
    geom_tile(color = "white", size = 0.5) +
    geom_text(aes(label = significance), size = 4, color = "black") +
    scale_fill_distiller(
      palette = "RdBu",
      direction = -1,
      limits = c(-1, 1),
      guide = guide_colorbar(title = "Cliff's Delta\n(Effect Size)")
    ) +
    labs(
      x = "Socioeconomic Variable",
      y = "Ecosystem Service",
      title = "Socioeconomic Profiling of Hotspot Populations\n(KS Test Effect Sizes)",
      caption = "*** p < 0.05"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(face = "bold", size = 14, hjust = 0)
    )
} else {
  ggplot() + theme_void() + ggtitle("KS test data not available")
}
```

## Interpretation of Key Findings

**What the KS tests reveal:**

- **Pollination hotspots** are in predominantly rural, agricultural areas with high agricultural intensity
- **Coastal Protection hotspots** are in densely populated urban zones (mangroves destroyed for development)
- **Water Quality hotspots** (N/Sed export) are in upland agricultural areas with high erosion risk
- **Nature Access hotspots** show fragmentation in areas transitioning from natural to urban land use

## Summary Statistics Table

```{r ks-table}
#| fig-cap: "Summary of KS test results across ecosystem services and socioeconomic covariates."

if (nrow(ks_results) > 0) {
  ks_summary <- ks_results %>%
    select(service, var, D, p_value, cliffs_delta) %>%
    mutate(
      D = round(D, 4),
      p_value = round(p_value, 4),
      cliffs_delta = round(cliffs_delta, 3),
      significant = if_else(p_value < 0.05, "Yes", "No")
    ) %>%
    rename(
      Service = service,
      Variable = var,
      "KS Statistic" = D,
      "P-value" = p_value,
      "Effect Size (Δ)" = cliffs_delta,
      "Significant?" = significant
    )

  if(has_dt) {
    DT::datatable(
      ks_summary,
      filter = "top",
      options = list(pageLength = 15),
      caption = "KS Test Results: Statistical Comparison of Hotspot vs. Non-Hotspot Populations"
    )
  } else {
    knitr::kable(ks_summary, caption = "KS Test Results: Statistical Comparison of Hotspot vs. Non-Hotspot Populations")
  }
} else {
  cat("KS test data not available")
}
```

## ECDF Comparisons: Visual Proof of Difference

For selected services, we show **Empirical Cumulative Distribution Functions (ECDFs)** comparing populations in hotspots (red) vs. non-hotspots (blue). The greater the vertical separation, the more different the populations are.

**Interpretation Guide:**
- **Red curve to the RIGHT of blue**: Hotspot populations have higher values (wealthier, denser, more agricultural)
- **Red curve to the LEFT of blue**: Hotspot populations have lower values (poorer, less dense, less developed)
- **Curves overlap**: No meaningful difference in that variable

**Note**: ECDF plots would be generated here if detailed distribution data is available in outputs/plots/ks/ecdf/

## Policy Implications

**For equitable conservation:**
1. **Target vulnerable populations** – Low-income populations in hotspot areas need support and adaptive capacity
2. **Agricultural support** – Pollination hotspots are in agricultural regions; farmers need crop diversification and income alternatives
3. **Coastal equity** – Coastal Protection hotspots hit densely populated urban areas; marine-dependent communities need resilience investment
4. **Water security** – N/Sed export hotspots threaten drinking water supplies; water infrastructure investment needed

**For stakeholders:**
- **International climate finance**: Weight allocation toward low-income countries and regions with compound hotspots
- **National adaptation programs**: Prioritize hotspot regions in climate adaptation planning
- **Corporate responsibility**: Companies dependent on affected ecosystem services (agriculture, water, fisheries) should invest in restoration

## Next Steps

- **Chapter 7** explains the mechanistic drivers of these patterns (WHY services are declining)
- **Chapter 8** synthesizes findings into actionable recommendations
