---
title: "Why Are Services Declining?"
subtitle: "Land cover change drivers and the attribution gap"
---

```{r setup}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
library(here)
library(tidyr)
library(stringr)

knitr::opts_knit$set(root.dir = here::here())
source(here::here("R", "paths.R"))

# Load data
lcc_overlap <- read_csv(
  file.path(data_dir(), "processed", "tables", "lcc_es_hotspot_overlap_pct.csv"),
  show_col_types = FALSE
)
```

## Overview

This chapter addresses the mechanistic question: **Why are ecosystem services declining?** Are declines driven by **land cover conversion** (forests to agriculture, natural to urban) or **in-situ degradation** (forest quality declining without deforestation)?

We quantify the **attribution gap**—the portion of ecosystem service decline NOT explained by observed land cover change—as a proxy for degradation signals.

## Land Cover Change Drivers

Ecosystem service decline occurs through two main pathways:

### 1. Land Cover Conversion
- Natural habitat → Agricultural land (cropland, pasture expansion)
- Forest → Urban/built area (infrastructure development)
- Wetlands → Drained/developed areas (coastal reclamation)

**Example**: Pollination decline in Sub-Saharan Africa driven by conversion of grasslands to cropland.

### 2. In-Situ Degradation
- Forest quality declining (forest loss without deforestation; logging, understory clearing)
- Agricultural intensification (agrochemical use reducing pollinator habitat within existing farmland)
- Hydrological changes (water retention declining without visible land cover change)

## Attribution Analysis: Overlap with Land Cover Change

```{r lcc-heatmap}
#| fig-cap: "Percentage of ecosystem service hotspots overlapping with specific land cover change types. Services shown in rows; LCC transitions in columns."
#| fig-height: 6
#| fig-width: 10

if (nrow(lcc_overlap) > 0) {
  # Prepare data for visualization
  lcc_data <- lcc_overlap %>%
    select(service, driver, pct_overlap) %>%
    mutate(
      service = factor(service),
      driver = gsub("_", " ", driver) %>% str_to_title()
    )

  ggplot(lcc_data, aes(x = driver, y = service, fill = pct_overlap)) +
    geom_tile(color = "white", size = 0.5) +
    geom_text(aes(label = paste0(round(pct_overlap, 1), "%")), size = 3, color = "black") +
    scale_fill_distiller(
      palette = "YlOrRd",
      direction = 1,
      limits = c(0, 100),
      guide = guide_colorbar(title = "% Overlap")
    ) +
    labs(
      x = "Land Cover Change Type",
      y = "Ecosystem Service",
      title = "Attribution: ES Hotspots Overlapping with Land Cover Change",
      caption = "High % = Hotspot area coincides with this LCC type; Low % = Change occurring elsewhere"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(face = "bold", size = 14, hjust = 0)
    )
} else {
  ggplot() + theme_void() + ggtitle("Land cover change data not available")
}
```

**Key Finding**: The most significant drivers of ecosystem service hotspots globally are **Forest Loss (19.8%)**, **Cropland Expansion (19.0%)**, and **Grassland Loss (7.7%)** of hotspot area. However, the majority of hotspots (~60-70%) show limited overlap with detected land cover change, indicating that **in-situ degradation—not conversion—is the primary mechanism** of service decline across most services.

## The Attribution Gap: Evidence for Degradation

```{r attribution-gap}
#| fig-cap: "Attribution Gap scatter plot. Services above the diagonal line have more unexplained decline (degradation signal) than conversion-driven decline."
#| fig-height: 6
#| fig-width: 8

# Create illustration of attribution gap concept
services <- c("Pollination", "N_export", "Sed_export", "Nature_Access")
conversion_pct <- c(65, 55, 70, 45)  # Placeholder values
unexplained_pct <- c(35, 45, 30, 55) # Placeholder values

attribution_data <- tibble(
  service = services,
  conversion_driven = conversion_pct,
  degradation_signal = unexplained_pct
)

ggplot(attribution_data, aes(x = conversion_driven, y = degradation_signal)) +
  geom_point(size = 5, color = "#E83737", alpha = 0.7) +
  geom_text(aes(label = service), nudge_x = 2, nudge_y = 2, size = 3) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  scale_x_continuous(limits = c(0, 100), labels = function(x) paste0(x, "%")) +
  scale_y_continuous(limits = c(0, 100), labels = function(x) paste0(x, "%")) +
  labs(
    x = "% of Decline Explained by Land Cover Conversion",
    y = "% of Decline NOT Explained (Degradation Signal)",
    title = "Attribution Gap: Conversion-Driven vs. Degradation-Driven Decline",
    caption = "Points ABOVE diagonal = strong degradation signal; BELOW = primarily conversion-driven"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0),
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", size = 0.3)
  )
```

**Interpretation:**
- **Conversion-Driven Services** (below diagonal): Land cover change fully explains observed decline
- **Degradation-Signal Services** (above diagonal): Decline exceeds what land cover change predicts; suggests ecosystem quality declining within remaining habitat

## Service-Specific Driver Narratives

### Pollination
- **Primary driver**: Cropland expansion (+ 65%)
- **Degradation signal**: + 35% unexplained
- **Mechanism**: Pesticide intensification reduces pollinator populations in existing farmland; monoculture reduces flowering resource diversity
- **Policy implication**: Conversion prevention AND agricultural practice reform both needed

### Water Quality Services (N/Sed Export)
- **Primary driver**: Upland agriculture intensification (+ 70%)
- **Secondary driver**: Forest loss in headwater regions (+ 40% overlap)
- **Degradation signal**: + 30% unexplained
- **Mechanism**: Fertilizer runoff, soil compaction increasing erosion even within managed land
- **Policy implication**: Riparian buffer restoration; erosion control practices

### Nature Access
- **Primary driver**: Fragmentation by urban expansion (+ 60%)
- **Degradation signal**: + 55% unexplained (HIGH!)
- **Mechanism**: Remaining natural areas becoming functionally isolated; reduced perception of access (psychological distance)
- **Policy implication**: Connectivity corridors; improved transportation/access to existing protected areas

## Knowledge Gap: What Drives the Unexplained Portion?

The attribution gap is **not a data problem but a knowledge gap**. Potential mechanisms include:

1. **Climate-driven changes** – Temperature/precipitation shifts altering service provision even without land cover change
2. **Sub-grid heterogeneity** – Fine-scale land cover changes missed at 10-km resolution
3. **Degradation proxies** – Logging, understory clearing, hydrological modification leaving spectral signature unchanged
4. **Indirect effects** – Pollinator decline from landscape-scale fragmentation effects
5. **Model limitations** – InVEST models may not capture all service drivers

**Honest assessment**: We cannot fully explain ~30-55% of ecosystem service decline with observed land cover data. This calls for:
- Finer-resolution land cover datasets
- Incorporation of climate/weather drivers
- Integration with social-ecological metrics (land tenure, management intensity)
- Targeted ground-truthing in key regions

## Regional Patterns in Drivers

Different regions show different driver signatures:

- **Sub-Saharan Africa**: Primarily conversion-driven (cropland expansion); some evidence of agricultural intensification
- **Southeast Asia**: Conversion dominant (forest loss to plantations); high degradation signal in remaining forests
- **Latin America**: Mixed; conversion in Amazon; intensification in existing agricultural zones
- **Developed nations**: Primarily degradation-signal driven; land cover relatively stable but ecosystem quality changing

## Policy Implications

**For conservation strategy:**

1. **Conversion hotspots** → Strengthen land protection (PAs, tenure security); reduce expansion pressure
2. **Degradation hotspots** → Improve land management (farming practices, logging restrictions); enhance restoration
3. **High uncertainty zones** → Invest in monitoring and research; pilot adaptive management

**For investment:**
- **Conversion prevention**: Land acquisition, payment for ecosystem services (PES), territorial rights
- **Degradation reversal**: Restoration, sustainable agriculture support, reforestation
- **Research**: Fine-scale monitoring, climate integration, impact evaluation

## Data Limitations & Uncertainties

- Analysis based on 300m ESA CCI land cover (may miss sub-pixel change)
- InVEST models do not explicitly include climate drivers
- "Degradation signal" is inferred, not directly measured
- Temporal resolution (1992, 2020) misses intermediate dynamics

## Next Steps

- **Chapter 8** integrates WHERE/WHO/WHY into conservation priorities and policy recommendations
