---
title: "Comparison of V1 vs V2 Hotspots"
author: "Jeronimo Rodriguez Escobar"
date: last-modified
format:
  html:
    toc: true
    df-print: kable
editor: source
---

## Overview

This script compares the hotspot results from **V1** (Original Absolute/Relative Change) and **V2** (Symmetric Percentage Change). It checks for changes in the total number of hotspots and calculates the spatial overlap (Jaccard Index) to quantify how much the hotspot locations have shifted.

```{r setup}
#| message: false
#| warning: false

library(dplyr)
library(readr)
library(sf)
library(here)
library(ggplot2)
library(tidyr)
library(devtools)

# Initialize paths
knitr::opts_knit$set(root.dir = here::here())
source(here::here("R", "paths.R"))
devtools::load_all(quiet = TRUE)

# Define paths
v1_dir <- file.path(data_dir(), "processed", "hotspots")
v2_dir <- file.path(data_dir(), "processed", "hotspots_v2")

v1_index_path <- file.path(v1_dir, "_hotspots_index.csv")
v2_index_path <- file.path(v2_dir, "_hotspots_index_v2.csv")
```

## 1. Compare Global Counts

```{r compare-counts}
if (!file.exists(v1_index_path) || !file.exists(v2_index_path)) {
  stop("One or both index files are missing. Ensure both V1 and V2 extractions have run.")
}

idx1 <- read_csv(v1_index_path, show_col_types = FALSE) %>% mutate(version = "v1")
idx2 <- read_csv(v2_index_path, show_col_types = FALSE) %>% mutate(version = "v2")

# Combine and pivot
combined <- bind_rows(idx1, idx2)

# Filter for Global scope summary
global_comp <- combined %>%
  filter(scope == "global") %>%
  select(metric, version, n_hot) %>%
  pivot_wider(names_from = version, values_from = n_hot) %>%
  mutate(
    diff = v2 - v1,
    pct_diff = round((diff / v1) * 100, 1)
  )

print(global_comp)
```

## 2. Spatial Overlap (Jaccard Index)

We calculate the Jaccard Index ($J = \frac{|A \cap B|}{|A \cup B|}$) for each service to see how much the hotspot locations overlap between versions.

```{r spatial-overlap}
#| message: false
#| warning: false

calc_overlap <- function(metric_type) {
  # Paths to global GPKGs
  gpkg1 <- file.path(v1_dir, metric_type, "global", paste0("hotspots_global_", metric_type, ".gpkg"))
  gpkg2 <- file.path(v2_dir, metric_type, "global", paste0("hotspots_global_", metric_type, ".gpkg"))

  if (!file.exists(gpkg1) || !file.exists(gpkg2)) {
    message("Missing GPKGs for metric: ", metric_type)
    return(NULL)
  }

  sf1 <- st_read(gpkg1, quiet = TRUE)
  sf2 <- st_read(gpkg2, quiet = TRUE)

  # Normalize ID columns (V2 uses grid_fid, V1 might use fid)
  if ("grid_fid" %in% names(sf1)) sf1 <- rename(sf1, fid = grid_fid)
  if ("grid_fid" %in% names(sf2)) sf2 <- rename(sf2, fid = grid_fid)

  # Identify common services (columns present in both that are not metadata)
  meta_cols <- c("fid", "geom", "geometry", "n_hot", "n_hot_deg_combo", "n_hot_rec_combo")
  services <- intersect(setdiff(names(sf1), meta_cols), setdiff(names(sf2), meta_cols))

  results <- list()

  for (svc in services) {
    fids1 <- sf1$fid[sf1[[svc]] == 1]
    fids2 <- sf2$fid[sf2[[svc]] == 1]

    intersection <- length(intersect(fids1, fids2))
    union <- length(union(fids1, fids2))
    jaccard <- if (union > 0) intersection / union else 0

    results[[svc]] <- tibble(
      service = svc,
      metric = metric_type,
      v1_count = length(fids1),
      v2_count = length(fids2),
      overlap_count = intersection,
      jaccard_index = round(jaccard, 3)
    )
  }
  bind_rows(results)
}

overlap_abs <- calc_overlap("abs")
overlap_pct <- calc_overlap("pct")
overlap_all <- bind_rows(overlap_abs, overlap_pct)

print(overlap_all)
```

### Visualization

```{r plot-overlap}
#| fig.width: 10
#| fig.height: 6

if (!is.null(overlap_all) && nrow(overlap_all) > 0) {
  ggplot(overlap_all, aes(x = reorder(service, jaccard_index), y = jaccard_index, fill = metric)) +
    geom_col(position = "dodge") +
    coord_flip() +
    scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
    labs(
      title = "Stability of Hotspots: V1 vs V2 (Jaccard Index)",
      subtitle = "Higher values indicate high spatial agreement between versions",
      y = "Spatial Overlap (Jaccard Index)",
      x = "Ecosystem Service"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
} else {
  message("No overlap data to plot.")
}
```