---
title: "Global NCP Hotspots: Interactive EDA"
author: "Jerónimo Rodríguez Escobar"
format: 
  dashboard:
    theme: flatly
    orientation: rows
    scrollable: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(tidyr)
library(here)
library(readr)
library(scales)
library(tidytext)

has_dt <- requireNamespace("DT", quietly = TRUE)
if(has_dt) library(DT)

knitr::opts_knit$set(root.dir = here::here())
source(here::here("R", "paths.R"))

# Safely load the CSVs
load_csv <- function(filename) {
  path <- file.path(data_dir(), "processed", "tables", filename)
  if (file.exists(path)) {
    read_csv(path, show_col_types = FALSE)
  } else {
    tibble::tibble(service = character(), message = "Data not found")
  }
}

area_stats         <- load_csv("hotspot_area_stats.csv")
multiservice_stats <- load_csv("hotspot_multiservice_stats.csv")
pop_exposure       <- load_csv("hotspot_pop_exposure.csv")
lcc_stats          <- load_csv("lcc_es_hotspot_overlap_pct.csv")
ks_res             <- load_csv("ks_results_hot_vs_non.csv")

# Ensure ordering for HDI and Income
hdi_levels <- c("Low (<0.550)", "Medium (0.550-0.700)", "High (0.700-0.800)", "Very High (>0.800)", "Unknown")
inc_levels <- c("Low income", "Lower middle income", "Upper middle income", "High income")

if(nrow(pop_exposure) > 0 && "hdi_bin" %in% names(pop_exposure)) {
  pop_exposure <- pop_exposure %>%
    mutate(hdi_bin = factor(hdi_bin, levels = hdi_levels))
}

# Biome Short Names
biome_labels <- c(
  'Tropical & Subtropical Moist Broadleaf Forests' = 'Trop/Subtrop Moist Broadleaf',
  'Tropical & Subtropical Dry Broadleaf Forests' = 'Trop/Subtrop Dry Broadleaf',
  'Tropical & Subtropical Coniferous Forests' = 'Trop/Subtrop Coniferous',
  'Temperate Broadleaf & Mixed Forests' = 'Temp Broadleaf/Mixed',
  'Temperate Coniferous Forests' = 'Temp Coniferous',
  'Boreal Forests/Taiga' = 'Boreal/Taiga',
  'Tropical & Subtropical Grasslands, Savannas & Shrublands' = 'Trop/Subtrop Grass/Sav/Shrub',
  'Temperate Grasslands, Savannas & Shrublands' = 'Temp Grass/Sav/Shrub',
  'Flooded Grasslands & Savannas' = 'Flooded Grass/Savannas',
  'Montane Grasslands & Shrublands' = 'Montane Grass/Shrub',
  'Tundra' = 'Tundra',
  'Mediterranean Forests, Woodlands & Scrub' = 'Mediterranean',
  'Deserts & Xeric Shrublands' = 'Deserts & Xeric Shrub',
  'Mangroves' = 'Mangroves'
)

# Plot Helper Functions
make_intensity_plot <- function(df, grp_var) {
  if(nrow(df) == 0) return(ggplot() + theme_void() + ggtitle("No data"))
  d <- df %>% filter(grouping_var == grp_var) %>% filter(!is.na(group))
  if(nrow(d) == 0) return(ggplot() + theme_void() + ggtitle("No data for this grouping"))
  
  if (grp_var == "WWF_biome") {
    d <- d %>% mutate(group = ifelse(group %in% names(biome_labels), biome_labels[group], group))
  }
  
  # --- TEMPORARY DEBUGGING STEP ---
  if (grp_var == "WWF_biome") {
    message("Unique biomes found in data for intensity plot:")
    print(sort(unique(d$group)))
  }
  # ------------------------------

  # Set factor levels for consistent alphabetical ordering across all facets.
  # `rev(sort(...))` ensures 'A' is at the top of the plot.
  all_groups <- sort(unique(d$group))
  d <- d %>% mutate(group = factor(group, levels = rev(all_groups)))
  
  ggplot(d, aes(x = relative_intensity, y = group, fill = relative_intensity > 1)) +
    geom_col() +
    geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
    scale_fill_manual(values = c("TRUE" = "#E83737", "FALSE" = "gray70")) +
    facet_wrap(~service, scales = "free_x", ncol = 2) +
    labs(x = "Relative Intensity Score (Share of Hotspots / Share of Area)", y = NULL) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none", panel.grid.minor.y = element_blank())
}

make_multi_plot <- function(df, grp_var) {
  if(nrow(df) == 0) return(ggplot() + theme_void() + ggtitle("No data"))
  d <- df %>% filter(grouping == grp_var) %>% filter(!is.na(group))
  if(nrow(d) == 0) return(ggplot() + theme_void() + ggtitle("No data for this grouping"))
  
  if (grp_var == "WWF_biome") {
    d <- d %>% mutate(group = ifelse(group %in% names(biome_labels), biome_labels[group], group))
  }
  
  # Strictly alphabetical
  d <- d %>% mutate(group = factor(group, levels = rev(sort(unique(as.character(group))))))
  
  ggplot(d, aes(x = mean_hotspots, y = group, fill = mean_hotspots)) +
    geom_col() +
    scale_fill_distiller(palette = "Reds", direction = 1) +
    labs(x = "Average Number of Overlapping Hotspots per 10km Cell", y = NULL) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none", panel.grid.minor.y = element_blank())
}

make_country_intensity_plot <- function(df) {
  d <- df %>% filter(grouping_var == "country") %>% filter(!is.na(group))
  if(nrow(d) == 0) return(ggplot() + theme_void() + ggtitle("No data"))
  
  # Order descending by intensity within each facet
  d <- d %>% mutate(group_ordered = tidytext::reorder_within(group, relative_intensity, service))
  
  ggplot(d, aes(x = relative_intensity, y = group_ordered, fill = relative_intensity > 1)) +
    geom_col() +
    geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
    scale_fill_manual(values = c("TRUE" = "#E83737", "FALSE" = "gray70")) +
    facet_wrap(~service, scales = "free_y", ncol = 2) +
    tidytext::scale_y_reordered() +
    labs(x = "Relative Intensity Score", y = NULL) +
    theme_minimal() +
    theme(legend.position = "none", axis.text.y = element_text(size = 7), panel.grid.minor.y = element_blank())
}

make_country_multi_plot <- function(df) {
  d <- df %>% filter(grouping == "country") %>% filter(!is.na(group))
  if(nrow(d) == 0) return(ggplot() + theme_void() + ggtitle("No data"))
  
  # Order strictly descending by value
  d <- d %>% arrange(mean_hotspots) %>% mutate(group = factor(group, levels = unique(group)))
  
  ggplot(d, aes(x = mean_hotspots, y = group, fill = mean_hotspots)) +
    geom_col() +
    scale_fill_distiller(palette = "Reds", direction = 1) +
    labs(x = "Average Number of Overlapping Hotspots", y = NULL) +
    theme_minimal() +
    theme(legend.position = "none", axis.text.y = element_text(size = 7), panel.grid.minor.y = element_blank())
}
```

# 🌍 Hotspot Intensity (WHERE)

::: columns
::: {.column width="60%"}
### Disproportionate Burden (Relative Intensity)

::: {.panel-tabset}

#### Region (WB)
```{r}
#| fig-height: 8
#| fig-width: 12
make_intensity_plot(area_stats, "region_wb")
```

#### Income Group
```{r}
#| fig-height: 8
#| fig-width: 12
make_intensity_plot(area_stats, "income_grp")
```

#### Biome
```{r}
#| fig-height: 16
#| fig-width: 12
make_intensity_plot(area_stats, "WWF_biome")
```

:::
:::
::: {.column width="40%"}
### Compound Risk: Multi-service Overlap

::: {.panel-tabset}

#### Region (WB)
```{r}
#| fig-height: 6
#| fig-width: 10
make_multi_plot(multiservice_stats, "region_wb")
```

#### Income Group
```{r}
#| fig-height: 6
#| fig-width: 10
make_multi_plot(multiservice_stats, "income_grp")
```

#### Biome
```{r}
#| fig-height: 8
#| fig-width: 10
make_multi_plot(multiservice_stats, "WWF_biome")
```

:::

# 🗺️ Country Profiles (All)

::: columns
::: {.column width="60%"}
#### Disproportionate Burden (All Countries)
```{r}
#| fig-height: 90
#| fig-width: 12
make_country_intensity_plot(area_stats)
```
:::
::: {.column width="40%"}
#### Compound Risk (All Countries)
```{r}
#| fig-height: 90
#| fig-width: 8
make_country_multi_plot(multiservice_stats)
```
:::
:::

# 👥 Population Exposure (WHO)

### Absolute Population Exposure by Income Group
```{r}
#| fig-height: 10
#| fig-width: 10
if(nrow(pop_exposure) > 0 && "income_grp" %in% names(pop_exposure)) {
  pop_inc <- pop_exposure %>%
    filter(!is.na(income_grp)) %>%
    group_by(service, income_grp) %>%
    summarise(exposed_population = sum(exposed_population, na.rm = TRUE), .groups = "drop") %>%
    mutate(
      income_grp_label = stringr::str_remove(income_grp, "^\\d+\\.?\\s*"),
      income_grp_label = factor(income_grp_label, levels = inc_levels)
    )
  
  ggplot(pop_inc, aes(x = income_grp_label, y = exposed_population, fill = income_grp_label)) +
    geom_col() +
    facet_wrap(~ service, scales = "free_y", ncol = 2) +
    scale_y_continuous(labels = label_number(scale = 1e-6, suffix = "M")) +
    scale_fill_brewer(palette = "RdYlBu", direction = -1) +
    labs(x = NULL, y = "Total Exposed Population") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
} else {
  "Data not available."
}
```

### Absolute Population Exposure by HDI Bin
```{r}
#| fig-height: 10
#| fig-width: 10
if(nrow(pop_exposure) > 0 && "hdi_bin" %in% names(pop_exposure)) {
  pop_hdi <- pop_exposure %>%
    filter(!is.na(hdi_bin)) %>%
    group_by(service, hdi_bin) %>%
    summarise(exposed_population = sum(exposed_population, na.rm = TRUE), .groups = "drop")
  
  ggplot(pop_hdi, aes(x = hdi_bin, y = exposed_population, fill = hdi_bin)) +
    geom_col() +
    facet_wrap(~ service, scales = "free_y", ncol = 2) +
    scale_y_continuous(labels = label_number(scale = 1e-6, suffix = "M")) +
    scale_fill_brewer(palette = "YlOrRd", direction = 1) +
    labs(x = NULL, y = "Total Exposed Population") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
} else {
  "Data not available."
}
```

# 📊 Socioeconomic Drivers (KS Tests)

### Directionality (Cliff's Delta)
```{r}
#| fig-height: 12
#| fig-width: 8
if(nrow(ks_res) > 0 && "cliffs_delta" %in% names(ks_res)) {
  ks_res %>%
    mutate(var_clean = gsub("_", " ", var)) %>%
    ggplot(aes(x = service, y = cliffs_delta, fill = cliffs_delta)) +
    geom_col() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    facet_wrap(~var_clean, scales = "free_y", ncol = 1) +
    coord_flip() +
    labs(x = NULL, y = "Cliff's Delta (Red = Hotspots have higher values)") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none")
} else {
  "Data not available."
}
```

### Statistical Results Table
```{r}
if(nrow(ks_res) > 0) {
  ks_display <- ks_res %>%
    select(Service = service, Variable = var, `KS Stat (D)` = D, `Cliff's Delta` = cliffs_delta, `P-Value` = p_adj) %>%
    mutate(across(is.numeric, ~round(., 4))) %>%
    arrange(desc(abs(`Cliff's Delta`)))
    
  if(has_dt) {
    datatable(ks_display, options = list(pageLength = 20, searchHighlight = TRUE, scrollX = TRUE), filter = "top")
  } else {
    knitr::kable(ks_display)
  }
} else {
  "Data not available."
}
```

# 🌲 LCC Attribution (WHY)

### The Attribution Gap: Drivers of Ecosystem Service Decline
```{r}
#| fig-height: 12
#| fig-width: 14
if(nrow(lcc_stats) > 0) {
  lcc_plot_data <- lcc_stats %>%
    rename_with(~"overlap_pct", any_of(c("pct_overlap", "percent", "value", "pct_hotspot_in_lcc"))) %>%
    rename_with(~"Attribution", any_of(c("attribution", "Driver", "driver", "Category"))) %>%
    filter(Attribution != "Change Occuring Off-Pixel")
    
  ggplot(lcc_plot_data, aes(x = overlap_pct, y = reorder(Attribution, overlap_pct), fill = Attribution)) +
    geom_col() +
    facet_wrap(~ service, ncol = 3) +
    scale_x_continuous(labels = percent_format()) +
    labs(x = "% of Hotspot Area Overlapping with Driver", y = NULL) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none")
} else {
  "Data not available."
}
```