---
title: "Hotspot Intensity Analysis"
subtitle: "Quantifying the spatial extent of global hotspots"
author: "Jerónimo Rodríguez Escobar"
date: last-modified
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
editor: source
---

## Overview

This notebook calculates the **Hotspot Intensity (Coverage)**: the percentage of land area within each spatial unit (e.g., Region, Biome) that is classified as a "Global Hotspot".

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
library(devtools)

# Initialize paths and load project functions
knitr::opts_knit$set(root.dir = here::here())
source(here::here("R","paths.R"))
devtools::load_all(quiet = TRUE)

options(dplyr.summarise.inform = FALSE)
```

## Load Data

We load the consolidated long-format data (`plt_long.rds`) generated by `hotspot_extraction.qmd`.

```{r}
#| label: load-data

rds_path <- file.path(data_dir(), "processed", "plt_long.rds")

if (file.exists(rds_path)) {
  message("Loading data from: ", rds_path)
  plt_long <- readRDS(rds_path)
} else {
  stop("plt_long.rds not found. Please run analysis/hotspot_extraction.qmd first to generate it.")
}
```

## Configuration

Using the standard hotspot definition rules.

```{r}
#| label: config

HOTS_CFG <- list(
  pct_cutoff      = 0.05,
  threshold_mode  = "percent",
  rule_mode       = "vectors",
  loss = c("Nature_Access","Pollination","N_Ret_Ratio","Sed_Ret_Ratio","C_Risk_Red_Ratio"),
  gain = c("Sed_export","N_export","C_Risk"),
  groupings = c("income_grp","region_wb","continent","region_un","WWF_biome","nev_name"),
  svc_order = c(
    "C_Risk","N_export","Sed_export",
    "C_Risk_Red_Ratio","N_Ret_Ratio","Sed_Ret_Ratio",
    "Pollination","Nature_Access"
  )
)
```

## Analysis: Calculate Hotspot Intensity

```{r}
#| label: calc-intensity

# 1. Identify Global Hotspots
# We use the global scope to define what a "hotspot" is based on the full dataset distribution.
hs_global_obj <- extract_hotspots(
  df = plt_long,
  value_col = "pct_chg", # Using Percent Change as the primary metric
  pct_cutoff = HOTS_CFG$pct_cutoff,
  threshold_mode = HOTS_CFG$threshold_mode,
  rule_mode = HOTS_CFG$rule_mode,
  loss_services = HOTS_CFG$loss,
  gain_services = HOTS_CFG$gain,
  combos = NULL, # Focus on individual services
  id_cols = c("c_fid")
)

# 2. Flag Hotspot Pixels
hs_flags <- hs_global_obj$hotspots_df %>%
  dplyr::select(c_fid, service) %>%
  dplyr::mutate(is_hotspot = 1)

# Calculate global total hotspots per service (for "Share of Global" metric)
global_totals <- hs_flags %>%
  dplyr::count(service, name = "global_n_hot")

# 3. Calculate Coverage % per Grouping
# Ensure we only select groupings present in the data
valid_groupings <- intersect(HOTS_CFG$groupings, names(plt_long))
df_area_calc <- plt_long %>%
  dplyr::filter(!is.na(pct_chg)) %>%
  dplyr::select(c_fid, service, dplyr::all_of(valid_groupings)) %>%
  dplyr::left_join(hs_flags, by = c("c_fid", "service")) %>%
  dplyr::mutate(is_hotspot = tidyr::replace_na(is_hotspot, 0))

area_stats_list <- list()

# Services and regions to exclude from plots/stats
exclude_services <- c("C_Risk_NoHab", "USLE")
exclude_regions  <- c("Seven seas (Open Ocean)", "Seven seas (open ocean)", "Antarctica")

for (g in valid_groupings) {
  if (!g %in% names(df_area_calc)) next

  stats <- df_area_calc %>%
    dplyr::filter(!is.na(.data[[g]])) %>%
    dplyr::filter(!service %in% exclude_services) %>%
    dplyr::filter(!.data[[g]] %in% exclude_regions) %>%
    dplyr::group_by(service, group = .data[[g]]) %>%
    dplyr::summarise(
      n_total = dplyr::n(),
      n_hot   = sum(is_hotspot),
      pct_area = (n_hot / n_total) * 100, # Coverage: % of Region that is Hotspot
      .groups = "drop"
    ) %>%
    dplyr::left_join(global_totals, by = "service") %>%
    dplyr::mutate(
      pct_share = (n_hot / global_n_hot) * 100 # Distribution: % of Global Hotspots in Region
    ) %>%
    dplyr::mutate(grouping_var = g) %>%
    dplyr::arrange(desc(pct_area))

  area_stats_list[[g]] <- stats
}

hotspot_area_stats <- dplyr::bind_rows(area_stats_list)

# Save summary table
out_csv_path <- file.path(data_dir(), "processed", "hotspot_area_stats.csv")
readr::write_csv(hotspot_area_stats, out_csv_path)
message("Hotspot area statistics saved to: ", out_csv_path)

# Preview
head(hotspot_area_stats, 10)
```

## Visualization

```{r}
#| label: plot-intensity
#| message: false

plot_area_stats <- function(df, g_var, metric = "pct_area") {
  metric_label <- if (metric == "pct_area") "% Area is Hotspot (Intensity)" else "% Share of Global Hotspots (Distribution)"
  file_suffix  <- if (metric == "pct_area") "coverage" else "share"

  d_plot <- df %>% dplyr::filter(grouping_var == g_var)

  # Ensure service order
  d_plot <- d_plot %>%
    dplyr::mutate(service = factor(service, levels = HOTS_CFG$svc_order)) %>%
    dplyr::filter(!is.na(service))

  # If there are many groups (e.g. countries), filter to top 15 per service
  n_groups <- dplyr::n_distinct(d_plot$group)
  is_large_group <- n_groups > 20

  if (is_large_group) {
    d_plot <- d_plot %>%
      dplyr::group_by(service) %>%
      dplyr::slice_max(order_by = .data[[metric]], n = 15) %>%
      dplyr::ungroup()
    subtitle_add <- " (Top 15 per service)"

    # Large groups: Free Y axis, sorted by magnitude
    d_plot <- d_plot %>%
      dplyr::mutate(group_ordered = stats::reorder(paste(group, service, sep = "___"), .data[[metric]]))
    scales_val <- "free_y"
  } else {
    subtitle_add <- ""

    # Small groups: Shared Y axis, sorted alphabetically (A top)
    d_plot <- d_plot %>%
      dplyr::arrange(dplyr::desc(group)) %>%
      dplyr::mutate(group_ordered = factor(group, levels = unique(group)))
    scales_val <- "free_x"
  }

  p <- ggplot2::ggplot(d_plot, ggplot2::aes(x = .data[[metric]], y = group_ordered, fill = service)) +
    ggplot2::geom_col(show.legend = FALSE) +
    ggplot2::facet_wrap(~ service, scales = scales_val, ncol = 3) +
    ggplot2::labs(
      title = paste0("Hotspot ", if (metric == "pct_area") "Intensity" else "Distribution", " by ", g_var),
      subtitle = paste0(metric_label, subtitle_add),
      x = metric_label, y = NULL
    ) +
    ggplot2::theme_minimal()

  if (is_large_group) {
    p <- p + ggplot2::scale_y_discrete(labels = function(x) sub("___.*", "", x))
  }

  # Save plot
  out_file <- file.path(out_plots(), paste0("hotspot_", file_suffix, "_", tolower(g_var), ".png"))
  ggsave(out_file, p, width = 12, height = 8, bg = "white")
  message("Saved plot: ", out_file)
  return(p)
}

# Generate plots for all groupings
plots_intensity <- lapply(names(area_stats_list), function(g) plot_area_stats(hotspot_area_stats, g, "pct_area"))
plots_share     <- lapply(names(area_stats_list), function(g) plot_area_stats(hotspot_area_stats, g, "pct_share"))

# Display the first plot (e.g., Income Group)
if (length(plots_intensity) > 0) plots_intensity[[1]]
```