---
title: "Hotspot Spatial Synthesis"
subtitle: "Intensity, Relative Intensity, and Multi-service Overlap"
author: "Jerónimo Rodríguez Escobar"
date: last-modified
format:
#  html:
#    toc: true
#    number-sections: true
#    theme: cosmo
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
    df-print: kable
editor: source
params:
  input_gpkg: "10k_change_calc.gpkg"
  input_plt_long: "plt_long.rds"
  output_suffix: "" # e.g., "_v2"
---

## Overview

This notebook consolidates all spatial summary statistics for global hotspots into a single, reproducible pipeline. It calculates and visualizes three key metrics of geographic concentration:

1.  **Hotspot Intensity (Coverage)**
    -   **Question**: What percentage of a region's own land area is experiencing a crisis for a given service?
    -   **Formula**: `(Hotspot Area in Region / Total Area of Region) * 100`

2.  **Global Share (Distribution)**
    -   **Question**: What percentage of all global hotspots for a service are located within this specific region?
    -   **Formula**: `(Hotspot Area in Region / Total Global Hotspot Area) * 100`

3.  **Relative Intensity (Enrichment Score)**
    -   **Question**: Is a region bearing a disproportionate burden of hotspots compared to its size?
    -   **Formula**: `Global Share / Expected Share`, where `Expected Share` is the region's percentage of the total global land area.
    -   **Interpretation**: A score > 1 indicates the region is a "hotspot of hotspots"—it contains more of the crisis than would be expected based on its size alone.

4.  **Multi-service Hotspots ("Hotness")**
    -   **Question**: How many overlapping crises are occurring in the same place?
    -   **Metrics**: We calculate the average number of overlapping hotspots per gridcell within a region, and the distribution of cells with 1, 2, or 3+ hotspots.

5.  **Absolute Population Exposure (Impact Tier)**
    -   **Question**: How many people are directly living within these top 5% areas of ES decline?
    -   **Metrics**: Total population count within hotspots, segmented by Human Development Index (HDI) vulnerability bins and World Bank Income Group.

## Setup

## Data Dictionary for Output Tables

This notebook produces three key summary tables in the `processed/tables/` directory. Their columns are defined below.

### `hotspot_area_stats.csv`

| Column | Meaning |
|:---|:---|
| `service` | The ecosystem service being analyzed. |
| `group` | The specific value of the grouping variable (e.g., 'East Asia & Pacific', 'Low income'). |
| `n_total` | Total number of valid grid cells within that `group`. |
| `n_hot` | Number of hotspot cells within that `group`. |
| `pct_area` | **Hotspot Intensity (Coverage)**. The percentage of the group's area that is a hotspot. Formula: `(n_hot / n_total) * 100`. |
| `global_n_hot` | The total number of hotspots for that service across the entire globe. |
| `pct_share` | **Global Share**. The percentage of all global hotspots for a service that are located within this `group`. Formula: `(n_hot / global_n_hot) * 100`. |
| `grouping_var` | The grouping variable used for the aggregation (e.g., 'region_wb', 'income_grp'). |
| `expected_share` | The group's percentage of the total global land area. This is the share of hotspots the group *would* have if hotspots were distributed evenly. |
| `relative_intensity` | **Enrichment Score**. A measure of disproportionate burden. It is the ratio of the group's actual share of hotspots to its expected share. A score > 1 indicates the region is a "hotspot of hotspots." Formula: `pct_share / expected_share`. |

### `hotspot_multiservice_stats.csv`

| Column | Meaning |
|:---|:---|
| `group` | The specific value of the grouping variable. |
| `mean_hotspots` | **"Hotness"**. The average number of overlapping service hotspots per grid cell within that group. |
| `pct_multi` | The percentage of grid cells within the group that have more than one hotspot. |
| `n_pixels` | The total number of grid cells in the group. |
| `grouping` | The grouping variable used for the aggregation. |

### `hotspot_pop_exposure.csv`

| Column | Meaning |
|:---|:---|
| `service` | The ecosystem service being analyzed. |
| `income_grp` | The World Bank Income Group of the grid cell. |
| `hdi_bin` | The Human Development Index (HDI) bin of the grid cell. |
| `exposed_population` | The total population (from GHS-POP) living within the hotspot cells for that specific service and socioeconomic group. |
| `hotspot_cells` | The number of hotspot cells corresponding to that service and group. |

---

```{r}
#| label: setup
#| message: false
#| warning: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
library(devtools)

# Initialize paths and load project functions
knitr::opts_knit$set(root.dir = here::here())
source(here::here("R","paths.R"))
devtools::load_all(quiet = TRUE)

options(dplyr.summarise.inform = FALSE)
```

## Cleanup Legacy Outputs
```{r}
#| label: cleanup-legacy

# 1. Clear summary CSVs
csvs_to_remove <- c(
file.path(data_dir(), "processed", "hotspot_area_stats.csv"),
file.path(data_dir(), "processed", "hotspot_multiservice_stats.csv")
)
for (f in csvs_to_remove) if (file.exists(f)) file.remove(f)

# 2. Clear intensity plots directory
intensity_dir <- file.path(out_plots(), paste0("intensity", params$output_suffix))
if (dir.exists(intensity_dir)) {
  unlink(list.files(intensity_dir, full.names = TRUE))
  message("Cleared legacy intensity plots in: ", intensity_dir)
}

# 3. Clear hotness plots from base plots directory
old_hotness <- list.files(out_plots(), pattern = "^hotness_.*\\.png$", full.names = TRUE)
if (length(old_hotness) > 0) {
  file.remove(old_hotness)
  message("Removed ", length(old_hotness), " old hotness plots.")
}
```

## Load Data

```{r}
#| label: load-data

# 1. Load Hotspot Data (plt_long)
rds_path <- here::here("home", "jeronimo", "data", "global_ncp", "processed", params$input_plt_long)
if (file.exists(rds_path)) {
  message("Loading data from: ", rds_path)
  plt_long <- readRDS(rds_path)
} else {
  stop(params$input_plt_long, " not found. Please run analysis/hotspot_extraction.qmd first.")
}

# 2. Load Canonical Grid (for accurate denominators, including areas with no service)
gpkg_grid <- file.path(data_dir(), "processed", params$input_gpkg)
if (file.exists(gpkg_grid)) {
  message("Loading canonical grid from: ", gpkg_grid)
  grid_df <- sf::st_read(gpkg_grid, quiet = TRUE)
  grid_df <- sf::st_drop_geometry(grid_df)
  if ("orig_fid" %in% names(grid_df)) grid_df <- dplyr::rename(grid_df, fid = orig_fid)
  if ("grid_fid" %in% names(grid_df) && !"fid" %in% names(grid_df)) grid_df <- dplyr::rename(grid_df, fid = grid_fid)
  if ("id" %in% names(grid_df) && !"fid" %in% names(grid_df)) grid_df <- dplyr::rename(grid_df, fid = id)

  if (!"fid" %in% names(grid_df)) stop("Fatal Error: grid_df is missing a stable identifier (fid/orig_fid/grid_fid).")

  # FIX: Ensure missing grouping attributes are rejoined from the master grid
  missing_groupings <- setdiff(c("income_grp", "region_wb", "WWF_biome"), names(grid_df))
  if (length(missing_groupings) > 0) {
    message("Joining missing attributes from master grid: ", paste(missing_groupings, collapse = ", "))
    master_grid <- sf::st_read(file.path(data_dir(), "vector_basedata", "AOOGrid_10x10km_land_4326_clean.gpkg"), quiet = TRUE)
    master_grid <- sf::st_drop_geometry(master_grid)
    if ("id" %in% names(master_grid) && !"fid" %in% names(master_grid)) {
      master_grid <- dplyr::rename(master_grid, fid = id)
    }
    # This is the critical missing piece: ensure master_grid has a 'fid'
    if (!"fid" %in% names(master_grid)) {
      master_grid$fid <- seq_len(nrow(master_grid))
    }
    grid_attrs <- master_grid %>% dplyr::select(dplyr::any_of(c("fid", missing_groupings)))
    grid_df <- grid_df %>% dplyr::left_join(grid_attrs, by = "fid")
  }
} else {
  stop("Canonical grid file not found: ", gpkg_grid)
}
```

```{r}
#| label: global-filters
#| echo: false

# --- GLOBAL FILTER: Exclude "High income: nonOECD" from all analyses in this script ---
if ("income_grp" %in% names(grid_df)) {
  message(glue::glue("Globally excluding '2. High income: nonOECD' from all synthesis calculations.
                     Original grid_df rows: {nrow(grid_df)}
                     Original plt_long rows: {nrow(plt_long)}"))
  grid_df <- grid_df %>%
    dplyr::filter(is.na(income_grp) | income_grp != "2. High income: nonOECD")

  if ("income_grp" %in% names(plt_long)) {
    plt_long <- plt_long %>%
      dplyr::filter(is.na(income_grp) | income_grp != "2. High income: nonOECD")
  } else {
    plt_long <- plt_long %>%
      dplyr::filter(fid %in% grid_df$fid)
  }
  message(glue::glue("Filtered grid_df rows: {nrow(grid_df)}, Filtered plt_long rows: {nrow(plt_long)}"))
}
```

## Configuration & Exclusions

```{r}
#| label: config

HOTS_CFG <- list(
  pct_cutoff      = 0.05,
  threshold_mode  = "percent",
  rule_mode       = "vectors",
  loss = c("Nature_Access","Pollination","N_Ret_Ratio","Sed_Ret_Ratio","C_Risk_Red_Ratio"),
  gain = c("Sed_export","N_export","C_Risk"),
  groupings = c("income_grp","region_wb","WWF_biome"),
  combos = list(
    deg_combo = c("Nature_Access","Pollination","N_export","Sed_export","C_Risk"),
    rec_combo = c("Nature_Access","Pollination","N_Ret_Ratio","Sed_Ret_Ratio","C_Risk_Red_Ratio")
  ),
  svc_order = c(
    "C_Risk","N_export","Sed_export",
    "C_Risk_Red_Ratio","N_Ret_Ratio","Sed_Ret_Ratio",
    "Pollination","Nature_Access"
  )
)

# Master Exclusion Lists (Applies to all downstream Intensity & Hotness calculations)
exclude_services <- c("C_Risk_NoHab", "USLE")
exclude_regions  <- c(
  "Seven seas (Open Ocean)", "Seven seas (open ocean)", "Antarctica",
  "Lakes", "Rock & Ice"
)

valid_groupings <- intersect(HOTS_CFG$groupings, names(grid_df))
```

## Extract Global Hotspots

```{r}
#| label: extract-hotspots

hs_global_obj <- extract_hotspots(
  df = plt_long,
  value_col = "pct_chg",
  pct_cutoff = HOTS_CFG$pct_cutoff,
  threshold_mode = HOTS_CFG$threshold_mode,
  rule_mode = HOTS_CFG$rule_mode,
  loss_services = HOTS_CFG$loss,
  gain_services = HOTS_CFG$gain,
  combos = HOTS_CFG$combos,
  id_cols = c("fid")
)
```

## Part 1: Intensity, Share, and Relative Intensity

```{r}
#| label: calc-intensity

hs_flags <- hs_global_obj$hotspots_df %>%
  dplyr::select(fid, service) %>%
  dplyr::mutate(is_hotspot = 1)

global_totals <- hs_flags %>%
  dplyr::count(service, name = "global_n_hot")

area_stats_list <- list()

for (g in valid_groupings) {
  stats_all <- lapply(HOTS_CFG$svc_order, function(svc) {
    grid_svc <- grid_df %>%
      dplyr::select(fid, dplyr::all_of(g)) %>%
      dplyr::mutate(service = as.character(svc)) %>%
      dplyr::left_join(hs_flags %>% dplyr::filter(service == svc) %>% dplyr::mutate(service = as.character(service)), by = c("fid", "service")) %>%
      dplyr::mutate(is_hotspot = tidyr::replace_na(is_hotspot, 0)) %>%
      # Apply master exclusions
      dplyr::filter(!service %in% exclude_services) %>%
      dplyr::filter(!.data[[g]] %in% exclude_regions) %>%
      dplyr::filter(!is.na(.data[[g]]))

    grid_svc %>%
      dplyr::group_by(service, group = .data[[g]]) %>%
      dplyr::summarise(
        n_total = dplyr::n(),
        n_hot   = sum(is_hotspot),
        pct_area = (n_hot / n_total) * 100,
        .groups = "drop"
      )
  })

  stats <- dplyr::bind_rows(stats_all) %>%
    dplyr::left_join(global_totals, by = "service") %>%
    dplyr::group_by(service) %>%
    dplyr::mutate(
      pct_share = (n_hot / global_n_hot) * 100,
      grouping_var = g,
      expected_share = (n_total / sum(n_total)) * 100,
      relative_intensity = ifelse(expected_share > 0, pct_share / expected_share, NA)
    ) %>%
    dplyr::ungroup() %>%
    dplyr::arrange(desc(pct_area))

  area_stats_list[[g]] <- stats
}

hotspot_area_stats <- dplyr::bind_rows(area_stats_list)

out_csv_path <- file.path(data_dir(), "processed", "tables", "hotspot_area_stats.csv")
dir.create(dirname(out_csv_path), recursive = TRUE, showWarnings = FALSE)
readr::write_csv(hotspot_area_stats, out_csv_path)
message("Hotspot area statistics saved to: ", out_csv_path)
```

```{r}
#| label: plot-intensity
#| message: false

biome_labels <- c(
  'Tropical & Subtropical Moist Broadleaf Forests' = 'Trop/Subtrop Moist Broadleaf',
  'Tropical & Subtropical Dry Broadleaf Forests' = 'Trop/Subtrop Dry Broadleaf',
  'Tropical & Subtropical Coniferous Forests' = 'Trop/Subtrop Coniferous',
  'Temperate Broadleaf & Mixed Forests' = 'Temp Broadleaf/Mixed',
  'Temperate Coniferous Forests' = 'Temp Coniferous',
  'Boreal Forests/Taiga' = 'Boreal/Taiga',
  'Tropical & Subtropical Grasslands, Savannas & Shrublands' = 'Trop/Subtrop Grass/Sav/Shrub',
  'Temperate Grasslands, Savannas & Shrublands' = 'Temp Grass/Sav/Shrub',
  'Flooded Grasslands & Savannas' = 'Flooded Grass/Savannas',
  'Montane Grasslands & Shrublands' = 'Montane Grass/Shrub',
  'Tundra' = 'Tundra',
  'Mediterranean Forests, Woodlands & Scrub' = 'Mediterranean',
  'Deserts & Xeric Shrublands' = 'Deserts & Xeric Shrub',
  'Mangroves' = 'Mangroves'
)

get_short_name <- function(g, grp_name) {
  g <- as.character(g)
  if (any(grp_name %in% c("WWF_biome", "biome"))) {
    res <- biome_labels[g]
    unname(ifelse(is.na(res), g, res))
  } else {
    g
  }
}

library(tidytext)

out_intensity_dir <- file.path(out_plots(), paste0("intensity", params$output_suffix))
dir.create(out_intensity_dir, recursive = TRUE, showWarnings = FALSE)

for (g_var in valid_groupings) {
  plot_data <- hotspot_area_stats %>%
    dplyr::filter(grouping_var == g_var) %>%
    dplyr::mutate(service = factor(service, levels = HOTS_CFG$svc_order)) %>%
    dplyr::filter(!is.na(service)) %>%
    dplyr::mutate(group = factor(group, levels = rev(sort(unique(as.character(group))))))

  # Prevent Quarto crash if left_join generated NaNs
  plot_data <- plot_data %>%
    dplyr::mutate(
      pct_share = tidyr::replace_na(pct_share, 0),
      relative_intensity = tidyr::replace_na(relative_intensity, 0)
    )
  if(nrow(plot_data) == 0) next

  unique_groups <- sort(unique(as.character(plot_data$group)))
  group_id_map <- setNames(seq_along(unique_groups), unique_groups)
  unique_groups_desc <- sort(unique_groups, decreasing = TRUE)
  legend_levels <- paste0("[", group_id_map[unique_groups_desc], "] ", get_short_name(unique_groups_desc, g_var))

  plot_data <- plot_data %>%
    dplyr::mutate(
      group = factor(group, levels = unique_groups_desc),
      grp_num_id = as.character(group_id_map[as.character(group)]),
      grp_legend_label = factor(paste0("[", grp_num_id, "] ", get_short_name(as.character(group), g_var)), levels = legend_levels)
    )

  legend_cols <- 3 # As requested

  # 1. Coverage Plot
  p_cov <- ggplot(plot_data, aes(x = pct_area, y = group)) +
    geom_col(fill = "#E83737") +
    # HACK: Add an invisible geom_point to generate a color aesthetic, which we can then use to build a manual legend for the y-axis labels.
    geom_point(aes(color = grp_legend_label), x = -Inf, shape = 15, size = 0, alpha = 0) +
    coord_cartesian(clip = "off") +
    facet_wrap(~ service, scales = "free_x", ncol = 3) +
    labs(title = paste("Hotspot Coverage by", g_var), x = "% of Area that is a Hotspot", y = NULL) +
    scale_y_discrete(labels = function(x) group_id_map[x]) +
    theme_minimal() +
    theme(
          axis.ticks.y = element_blank(), 
          legend.position = "bottom", 
          legend.title = element_blank(), 
          legend.text = element_text(size = 10),
          axis.text = element_text(size = 10)) +
    guides(color = guide_legend(ncol = legend_cols, override.aes = list(size = 4, alpha = 1), reverse = TRUE))

  ggsave(file.path(out_intensity_dir, paste0("hotspot_coverage_", tolower(g_var), ".png")), p_cov, width=14, height=10, bg="white")

  # 2. Share Plot
  p_share <- ggplot(plot_data, aes(x = pct_share, y = group)) +
    geom_col(fill = "#E83737") +
    # HACK: Add an invisible geom_point to generate a color aesthetic for the legend.
    geom_point(aes(color = grp_legend_label), x = -Inf, shape = 15, size = 0, alpha = 0) +
    coord_cartesian(clip = "off") +
    facet_wrap(~ service, scales = "free_x", ncol = 3) +
    labs(title = paste("Global Hotspot Share by", g_var), x = "% of Global Hotspots", y = NULL) +
    scale_y_discrete(labels = function(x) group_id_map[x]) +
    theme_minimal() +
    theme(
          axis.ticks.y = element_blank(), 
          legend.position = "bottom", 
          legend.title = element_blank(), 
          legend.text = element_text(size = 10),
          axis.text = element_text(size = 10)) +
    guides(color = guide_legend(ncol = legend_cols, override.aes = list(size = 4, alpha = 1), reverse = TRUE))

  ggsave(file.path(out_intensity_dir, paste0("hotspot_share_", tolower(g_var), ".png")), p_share, width=14, height=10, bg="white")

  # 3. Relative Intensity Plot
  p_enrich <- ggplot(plot_data, aes(x = relative_intensity, y = group)) +
    geom_col(aes(alpha = relative_intensity > 1), fill = "#E83737") + geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
    # HACK: Add an invisible geom_point to generate a color aesthetic for the legend.
    geom_point(aes(color = grp_legend_label), x = -Inf, shape = 15, size = 0, alpha = 0) +
    coord_cartesian(clip = "off") +
    facet_wrap(~ service, scales = "free_x", ncol = 3) +
    scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.4), guide = "none") +
    labs(title = paste("Relative Hotspot Intensity by", g_var), x = "Intensity Score (Regional Hotspot % ÷ Global Average [~5%])", y = NULL) +
    scale_y_discrete(labels = function(x) group_id_map[x]) +
    theme_minimal() +
    theme(
          axis.ticks.y = element_blank(), 
          legend.position = "bottom", 
          legend.title = element_blank(), 
          legend.text = element_text(size = 10),
          axis.text = element_text(size = 10)) +
    guides(color = guide_legend(ncol = legend_cols, override.aes = list(size = 4, alpha = 1), reverse = TRUE))

  ggsave(file.path(out_intensity_dir, paste0("hotspot_relative_intensity_", tolower(g_var), ".png")), p_enrich, width=14, height=10, bg="white")
}
```

## Part 2: Multi-service Overlap (Hotness)

```{r}
#| label: calc-overlap

hotspot_counts <- hs_global_obj$hotspots_df %>%
  dplyr::group_by(fid) %>%
  dplyr::summarise(hotspot_count = dplyr::n(), .groups = "drop")

df_overlap <- grid_df %>%
  dplyr::select(fid, dplyr::all_of(valid_groupings)) %>%
  dplyr::distinct(fid, .keep_all = TRUE) %>%
  dplyr::left_join(hotspot_counts, by = "fid") %>%
  dplyr::mutate(hotspot_count = tidyr::replace_na(hotspot_count, 0))

calc_hotness <- function(df, g_var) {
  df %>%
    dplyr::filter(!is.na(.data[[g_var]])) %>%
    # Apply master exclusions strictly here too
    dplyr::filter(!.data[[g_var]] %in% exclude_regions) %>%
    dplyr::group_by(group = .data[[g_var]]) %>%
    dplyr::summarise(
      mean_hotspots = mean(hotspot_count),
      pct_multi = mean(hotspot_count > 1) * 100,
      n_pixels = dplyr::n()
    ) %>%
    dplyr::mutate(grouping = g_var) %>%
    dplyr::arrange(desc(mean_hotspots))
}

hotness_stats <- dplyr::bind_rows(lapply(valid_groupings, function(g) calc_hotness(df_overlap, g)))
out_csv <- file.path(data_dir(), "processed", "tables", "hotspot_multiservice_stats.csv")
dir.create(dirname(out_csv), recursive = TRUE, showWarnings = FALSE)
readr::write_csv(hotness_stats, out_csv)
```

```{r}
#| label: plot-overlap

plot_hotness <- function(df, g_var) {
  d_plot <- df %>% dplyr::filter(grouping == g_var) %>% dplyr::slice_head(n = 15) %>%
    dplyr::mutate(group = factor(group, levels = rev(sort(unique(as.character(group))))))

  p <- ggplot2::ggplot(d_plot, ggplot2::aes(x = group, y = mean_hotspots, fill = mean_hotspots)) +
    ggplot2::geom_col() +
    ggplot2::scale_fill_distiller(palette = "Reds", direction = 1, name = "Avg.\nHotspots") +
    ggplot2::coord_flip() +
    ggplot2::labs(title = paste0("Average Hotspot Overlap ('Hotness') by ", g_var), subtitle = "Average number of overlapping service hotspots per grid cell", x = NULL, y = "Avg. Hotspot Count") +
    ggplot2::theme_minimal() +
    ggplot2::theme(axis.text.y = element_text(size = 8))

  ggsave(file.path(out_plots(), paste0("hotness_", tolower(g_var), ".png")), p, width = 10, height = 6, bg = "white")
  return(p)
}

# Plot B: Stacked Distribution (1, 2, 3+ services)
plot_stacked_overlap <- function(df, g_var) {
  d_summ <- df %>%
    dplyr::filter(!is.na(.data[[g_var]])) %>%
    dplyr::filter(!.data[[g_var]] %in% exclude_regions) %>%
    dplyr::filter(hotspot_count > 0) %>% # Conditional on being a hotspot
    dplyr::mutate(
      hotspots_capped = pmin(as.numeric(hotspot_count), 4),
      cat = factor(hotspots_capped, levels = 1:4, labels = c("1", "2", "3", "4+"))
    ) %>%
    dplyr::count(group = .data[[g_var]], cat) %>%
    dplyr::group_by(group) %>%
    dplyr::mutate(pct = n / sum(n)) %>%
    dplyr::ungroup()

  # Filter top 15 groups by severe multi-service share (3 or 4+ combined)
  top_groups <- d_summ %>% dplyr::filter(cat %in% c("3", "4+")) %>% dplyr::group_by(group) %>% dplyr::summarise(pct = sum(pct)) %>% dplyr::arrange(desc(pct)) %>% dplyr::slice_head(n = 15) %>% dplyr::pull(group)
  if (length(unique(d_summ$group)) > 20) {
    if (length(top_groups) < 15) {
      extra_groups <- setdiff(unique(d_summ$group), top_groups)
      top_groups <- c(top_groups, head(extra_groups, 15 - length(top_groups)))
    }
    d_summ <- d_summ %>% dplyr::filter(group %in% top_groups)
  }

  # Order alphabetically (reversed for coord_flip to show A at top)
  d_summ <- d_summ %>% dplyr::mutate(group = factor(group, levels = rev(sort(unique(as.character(group))))))

  p <- ggplot2::ggplot(d_summ, ggplot2::aes(x = group, y = pct, fill = cat)) +
    ggplot2::geom_col(position = "fill") +
    ggplot2::scale_fill_manual(
      name = "Overlapping\nHotspots",
      values = c(
        "1" = "#FFD54F",
        "2" = "#FB8C00",
        "3" = "#E53935",
        "4+" = "#800026"
      )
    ) +
    ggplot2::coord_flip() +
    ggplot2::labs(
      title = paste0("Multi-service Hotspot Distribution by ", g_var),
      subtitle = "Share of hotspot gridcells by number of overlapping services",
      x = NULL, y = "Proportion of hotspot gridcells"
    ) +
    ggplot2::theme_minimal()
  ggsave(file.path(out_plots(), paste0("hotness_dist_", tolower(g_var), ".png")), p, width = 10, height = 6, bg = "white")
  return(p)
}

plots_hotness <- lapply(valid_groupings, function(g) plot_hotness(hotness_stats, g))
plots_dist    <- lapply(valid_groupings, function(g) plot_stacked_overlap(df_overlap, g))
if (length(plots_hotness) > 0) plots_hotness[]
```

## Part 3: Absolute Population Exposure

```{r}
#| label: calc-pop-exposure
#| message: false

::: callout-note
## Note on Aggregation (Sum vs. Mean)

This analysis uses socioeconomic variables aggregated as both sums (e.g., `GHS_POP_E2020_GLOBE_sum` for total population) and means (e.g., `hdi_raster_predictions_2020_mean` for average HDI). This is a deliberate and methodologically sound choice.

For reporting absolute exposure, using the `_sum` of population is critical to capture the total number of people affected. For statistical comparisons like the KS test, the choice between sum and mean has no impact on the results because our equal-area grid ensures they are perfectly proportional. For more details, see the "Aggregation Logic: Sum vs. Mean" section in `docs/methodology.md`.
:::

# Define the canonical socioeconomic variables to use for exposure
pop_var <- "GHS_POP_E2020_GLOBE_sum"
hdi_var <- "hdi_raster_predictions_2020_mean"
gdp_var <- "rast_gdpTot_1990_2020_30arcsec_2020_sum"
gini_var <- "rast_adm1_gini_disp_2020_mean"

req_vars <- c(pop_var, hdi_var, gdp_var, gini_var, "income_grp")

missing_vars <- setdiff(req_vars, names(grid_df))

# Check if all required variables are present in grid_df
if (length(missing_vars) == 0) {
  pop_exposure <- hs_flags %>%
    dplyr::left_join(grid_df %>% dplyr::select(fid, dplyr::all_of(req_vars)), by = "fid") %>%
    dplyr::filter(!is.na(.data[[pop_var]]), !is.na(.data[[hdi_var]])) %>%
    dplyr::mutate(
      hdi_bin = dplyr::case_when(
        .data[[hdi_var]] < 0.550 ~ "Low (<0.550)",
        .data[[hdi_var]] >= 0.550 & .data[[hdi_var]] < 0.700 ~ "Medium (0.550-0.700)",
        .data[[hdi_var]] >= 0.700 & .data[[hdi_var]] <= 0.800 ~ "High (0.700-0.800)",
        .data[[hdi_var]] > 0.800 ~ "Very High (>0.800)",
        TRUE ~ "Unknown"
      ),
      hdi_bin = factor(hdi_bin, levels = c("Low (<0.550)", "Medium (0.550-0.700)", "High (0.700-0.800)", "Very High (>0.800)", "Unknown"))
    ) %>%
    dplyr::mutate(
      gdp_bin = ntile(.data[[gdp_var]], 4), # Quartiles for GDP
      gdp_bin = factor(gdp_bin, labels = c("Q1 (Lowest GDP)", "Q2", "Q3", "Q4 (Highest GDP)"))
    ) %>%
    dplyr::mutate(
      gini_bin = ntile(.data[[gini_var]], 4), # Quartiles for Gini
      gini_bin = factor(gini_bin, labels = c("Q1 (Lowest Gini)", "Q2", "Q3", "Q4 (Highest Gini)"))
    ) %>%
    dplyr::group_by(service, income_grp, hdi_bin) %>%
    dplyr::summarise(
      exposed_population = sum(.data[[pop_var]], na.rm = TRUE),
      hotspot_cells = dplyr::n(),
      .groups = "drop"
    ) %>%
    dplyr::arrange(service, income_grp, hdi_bin)

  # Also calculate for GDP bins
  pop_exposure_gdp <- hs_flags %>%
    dplyr::left_join(grid_df %>% dplyr::select(fid, dplyr::all_of(req_vars)), by = "fid") %>%
    dplyr::filter(!is.na(.data[[pop_var]]), !is.na(.data[[gdp_var]])) %>%
    dplyr::mutate(gdp_bin = ntile(.data[[gdp_var]], 4), # Quartiles for GDP
                  gdp_bin = factor(gdp_bin, labels = c("Q1 (Lowest GDP)", "Q2", "Q3", "Q4 (Highest GDP)"))) %>%
    dplyr::group_by(service, gdp_bin) %>%
    dplyr::summarise(exposed_population = sum(.data[[pop_var]], na.rm = TRUE),
                     hotspot_cells = dplyr::n(), .groups = "drop") %>%
    dplyr::arrange(service, gdp_bin)

  # Also calculate for Gini bins
  pop_exposure_gini <- hs_flags %>%
    dplyr::left_join(grid_df %>% dplyr::select(fid, dplyr::all_of(req_vars)), by = "fid") %>%
    dplyr::filter(!is.na(.data[[pop_var]]), !is.na(.data[[gini_var]])) %>%
    dplyr::mutate(gini_bin = ntile(.data[[gini_var]], 4), # Quartiles for Gini
                  gini_bin = factor(gini_bin, labels = c("Q1 (Lowest Gini)", "Q2", "Q3", "Q4 (Highest Gini)"))) %>%
    dplyr::group_by(service, gini_bin) %>%
    dplyr::summarise(exposed_population = sum(.data[[pop_var]], na.rm = TRUE),
                     hotspot_cells = dplyr::n(), .groups = "drop") %>%
    dplyr::arrange(service, gini_bin)

  out_pop_csv <- file.path(data_dir(), "processed", "tables", "hotspot_pop_exposure.csv")
  dir.create(dirname(out_pop_csv), recursive = TRUE, showWarnings = FALSE)
  readr::write_csv(pop_exposure, out_pop_csv)
  message("Absolute Population Exposure statistics saved to: ", out_pop_csv)

} else {
  warning("Missing required variables for population exposure calculation: ", paste(missing_vars, collapse = ", "))
}
```


```{r}
#| label: plot-pop-exposure
#| message: false
#| echo: false

pop_exposure_csv <- file.path(data_dir(), "processed", "tables", "hotspot_pop_exposure.csv")
if (file.exists(pop_exposure_csv)) {
  pop_df <- readr::read_csv(pop_exposure_csv, show_col_types = FALSE)

  facet_names <- c(
    "C_Risk" = "Coastal Risk",
    "N_export" = "Nitrogen Export",
    "Sed_export" = "Sediment Export",
    "C_Risk_Red_Ratio" = "Coastal Risk Red. Ratio",
    "N_Ret_Ratio" = "N Retention Ratio",
    "Sed_Ret_Ratio" = "Sed Retention Ratio",
    "Pollination" = "Pollination",
    "Nature_Access" = "Nature Access"
  )

  # 1. Plot by Income Group - using a red intensity scale
  p_pop_inc <- ggplot(pop_df, aes(x = income_grp, y = exposed_population, fill = exposed_population)) +
    geom_col() +
    facet_wrap(~ service, scales = "free_y", ncol = 3, labeller = as_labeller(facet_names)) + # Changed to free_y for better comparison
    scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M")) +
    scale_x_discrete(labels = function(x) stringr::str_remove(x, "^\\d\\.\\s*")) +
    scale_fill_distiller(palette = "Reds", direction = 1, name = "Population", labels = scales::label_number(scale = 1e-6, suffix = "M"), breaks = scales::pretty_breaks(n = 5)) + # Added breaks
    labs(title = "Absolute Population Exposure in Hotspots",
         subtitle = "By World Bank Income Group",
         y = "Exposed Population", x = NULL) +
    theme_minimal() +
    theme(legend.position = "right", 
          axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
          axis.text.y = element_text(size = 12),
          axis.title = element_text(size = 14),
          legend.text = element_text(size = 10),
          legend.title = element_text(size = 12))

  ggsave(file.path(out_plots(), paste0("pop_exposure_income", params$output_suffix, ".png")), p_pop_inc, width = 14, height = 10, bg = "white")

  # 2. Plot by HDI Bin
  hdi_levels <- c("Low (<0.550)", "Medium (0.550-0.700)", "High (0.700-0.800)", "Very High (>0.800)", "Unknown")

  pop_df_hdi <- pop_df %>%
    dplyr::group_by(service, hdi_bin) %>%
    dplyr::summarise(exposed_population = sum(exposed_population, na.rm = TRUE), .groups = "drop") %>%
    dplyr::mutate(hdi_bin = factor(hdi_bin, levels = hdi_levels))
  
  p_pop_hdi <- ggplot(pop_df_hdi, aes(x = hdi_bin, y = exposed_population, fill = exposed_population)) +
    geom_col() +
    facet_wrap(~ service, scales = "free_y", ncol = 3, labeller = as_labeller(facet_names)) + # Changed to free_y
    scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M")) +
    scale_fill_distiller(palette = "Reds", direction = 1, name = "Population", labels = scales::label_number(scale = 1e-6, suffix = "M")) +
    labs(title = "Absolute Population Exposure in Hotspots",
         subtitle = "By UN Human Development Index (HDI)",
         y = "Exposed Population", x = NULL) +
    theme_minimal() +
    theme(legend.position = "right", 
          axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
          axis.text.y = element_text(size = 12),
          axis.title = element_text(size = 14),
          legend.text = element_text(size = 10),
          legend.title = element_text(size = 12))
  ggsave(file.path(out_plots(), paste0("pop_exposure_hdi", params$output_suffix, ".png")), p_pop_hdi, width = 14, height = 10, bg = "white")

  # 3. Plot by GDP Bin
  p_pop_gdp <- ggplot(pop_exposure_gdp, aes(x = gdp_bin, y = exposed_population, fill = exposed_population)) +
    geom_col() +
    facet_wrap(~ service, scales = "free_y", ncol = 3, labeller = as_labeller(facet_names)) +
    scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M")) +
    scale_fill_distiller(palette = "Reds", direction = 1, name = "Population", labels = scales::label_number(scale = 1e-6, suffix = "M")) +
    labs(title = "Absolute Population Exposure in Hotspots",
         subtitle = "By GDP Quartile",
         y = "Exposed Population", x = NULL) +
    theme_minimal() +
    theme(legend.position = "right", 
          axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
          axis.text.y = element_text(size = 12),
          axis.title = element_text(size = 14),
          legend.text = element_text(size = 10),
          legend.title = element_text(size = 12))
  ggsave(file.path(out_plots(), paste0("pop_exposure_gdp", params$output_suffix, ".png")), p_pop_gdp, width = 14, height = 10, bg = "white")

  # 4. Plot by Gini Bin
  p_pop_gini <- ggplot(pop_exposure_gini, aes(x = gini_bin, y = exposed_population, fill = exposed_population)) +
    geom_col() +
    facet_wrap(~ service, scales = "free_y", ncol = 3, labeller = as_labeller(facet_names)) +
    scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M")) +
    scale_fill_distiller(palette = "Reds", direction = 1, name = "Population", labels = scales::label_number(scale = 1e-6, suffix = "M")) +
    labs(title = "Absolute Population Exposure in Hotspots",
         subtitle = "By Gini Quartile",
         y = "Exposed Population", x = NULL) +
    theme_minimal() +
    theme(legend.position = "right", 
          axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
          axis.text.y = element_text(size = 12),
          axis.title = element_text(size = 14),
          legend.text = element_text(size = 10),
          legend.title = element_text(size = 12))
  ggsave(file.path(out_plots(), paste0("pop_exposure_gini", params$output_suffix, ".png")), p_pop_gini, width = 14, height = 10, bg = "white")

  plots_pop <- list(p_pop_inc, p_pop_hdi, p_pop_gdp, p_pop_gini)
  if (length(plots_pop) > 0) plots_pop[]
}
```