---
title: "Ecosystem Service Change & Hotspot Extraction"
subtitle: "v{{< meta analysis_version >}}"
author: "Jerónimo Rodríguez Escobar"
email: "jeronimo.rodriguez@wwfus.org"
date: last-modified
date-format: "YYYY-MM-DD HH:mm zzz"
analysis_version: "v1.3.1"
format:
#  html:
#    toc: true
#   number-sections: true
#   code-fold: false
#  toc-title: "Contents"
#   theme: cosmo
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
    df-print: kable
editor: source

params:
  analysis_version: "v1.3.1"
  input_gpkg: "10k_change_calc.gpkg"
  output_suffix: "" # e.g., "_v2"
  force_export: false
---

## Overview

Derives hotspot flags and produces bar/violin plots summarizing change by subregion.
Assumes process_data.qmd has been run first to produce the canonical grid + change dataset (10k_change_calc.gpkg).
Key steps: load the canonical grid, pivot change fields to long form, define hotspots using the configured thresholds, export hotspot layers, and save plots to outputs/plots/.

::: {.callout-tip icon="true"}
## Future Tasks & Ideas

Here are some ideas and future tasks for this analysis:

1.  **Adapt analysis for multi-temporal data:** Adapt analysis to handle updated modeled ES layers and multiple points in time (beyond bi-temporal T0, T1). Strategize for incorporating multi-temporal data.
2.  **Quantify hotspot vs. non-hotspot change:** Develop a method to quantify and visualize the share of total change (from bar plots) that occurs within hotspots versus outside of them, possibly using stacked bar plots.
3.  **Simplify Spatial Joins (V2 Technical Debt):** Remove the on-the-fly spatial coordinate joins (X/Y) used to bypass upstream row-shifting. Once `summary_pipeline_landgrid.py` is updated to preserve true FIDs, revert to standard tabular `left_join(by = "fid")`.
:::

## Setup and Libraries

Load all core packages (tidyverse, spatial, plotting helpers) and initialize reproducibility settings. This chunk also sources `R/paths.R` and runs `devtools::load_all()` so downstream chunks can reuse helper functions.

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false

# Core tidy + spatial
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(forcats)
library(ggplot2)
library(scales)
library(sf)
library(terra)
library(exactextractr)
library(here)

# Plot helpers
library(viridisLite)
library(ggnewscale)
library(ragg)

# Dev/project helpers
library(devtools)   # for load_all()
# Paths helper (set GLOBAL_NCP_DATA in ~/.Renviron first)
knitr::opts_knit$set(root.dir = here::here())
source(here::here("R","paths.R"))
# Optional viewers (enable only if you use them)
# library(httpgd)
# library(leaflet)
# library(htmltools)

# ---- Global options / performance ---------------------------------------
options(dplyr.summarise.inform = FALSE)
sf::sf_use_s2(TRUE)
Sys.setenv(GDAL_NUM_THREADS = "ALL_CPUS", PROJ_NETWORK = "ON")
# terraOptions(tempdir = file.path(tempdir(), "terra_tmp"))  # comment out, or:
dir.create(file.path(tempdir(), "terra_tmp"), recursive = TRUE, showWarnings = FALSE)



set.seed(1)

# ---- Project wiring ------------------------------------------------------
# Load package-style functions from R/ (if this is a package-ish repo)
devtools::load_all(quiet = TRUE)

# ---- Python (enable when needed) ----------------------------------------
library(reticulate)
use_virtualenv("/home/jeronimo/venvs/coastal_snap_env", required = TRUE)

```

## Metadata Banner

Published: `r format(Sys.time(), "%Y-%m-%d %H:%M %Z")`

```{r}
#| label: run-metadata
#| echo: false
#| message: false
#| warning: false
#| results: 'asis'

# Prints version/time/git/data-root info so renders are traceable (no heavy compute).

analysis_version <- tryCatch(params$analysis_version, error = function(e) NULL)
if (is.null(analysis_version) || is.na(analysis_version)) {
  analysis_version <- Sys.getenv("ANALYSIS_VERSION", unset = "dev")
}

# Git info (robust to non-git folders)
git_branch <- tryCatch(
  system2("git", c("rev-parse", "--abbrev-ref", "HEAD"), stdout = TRUE),
  error = function(e) NA_character_
)
git_commit <- tryCatch(
  system2("git", c("rev-parse", "--short", "HEAD"), stdout = TRUE),
  error = function(e) NA_character_
)

# Data root from your paths helper, with fallbacks
data_root <- tryCatch(data_dir(), error = function(e) NULL)
if (is.null(data_root)) {
  data_root <- Sys.getenv("GLOBAL_NCP_DATA", unset = "")
}
message("data_dir(): ", data_dir())
message("Outputs root: ", out_plots())
cat(paste0(
  "::: callout-note\n",
  "**Run metadata**\n\n",
  "- Analysis version: ", analysis_version, "\n",
  "- Rendered: ", format(Sys.time(), "%Y-%m-%d %H:%M %Z"), "\n",
  "- Git: ", if (!is.na(git_branch)) git_branch else "NA", " @ ",
                 if (!is.na(git_commit)) git_commit else "NA", "\n",
  "- Data root: ", if (nzchar(data_root)) data_root else "unset", "\n",
  ":::"
))
```

```{r}
#| label: svc-order-global
#| include: false
#| echo: false
if (!exists("svc_order", inherits = TRUE)) {
  svc_order <- c(
    "C_Risk","N_export","Sed_export",
    "C_Risk_Red_Ratio","N_Ret_Ratio","Sed_Ret_Ratio",
    "Pollination","Nature_Access"
  )
}
if (!exists("canonical_lookup", inherits = TRUE)) {
  canonical_lookup <- c(
    sed_export       = "Sed_export",
    n_export         = "N_export",
    n_retention      = "N_retention",
    nature_access    = "Nature_Access",
    pollination      = "Pollination",
    usle             = "USLE",
    n_ret_ratio      = "N_Ret_Ratio",
    sed_ret_ratio    = "Sed_Ret_Ratio",
    coastal_protection_rt            = "C_Risk",
    coastal_protection_rt_nohab_all  = "C_Risk_NoHab",
    coastal_protection_rt_ratio      = "C_Risk_Red_Ratio",
    rt_ratio         = "C_Risk_Red_Ratio",
    rt               = "C_Risk",
    c_risk           = "C_Risk",
    c_risk_red_ratio = "C_Risk_Red_Ratio",
    rt_service       = "C_Prot_service",
    rt_nohab         = "Rt_nohab"
  )
}
```



## Produce signed change bars

Bars are computed directly from `processed/10k_change_calc.gpkg`. We focus specifically on "active areas" by dropping true zeros, ensuring that our regional summaries reflect areas where the ecosystem service is actually provided or changing. Both absolute and relative changes are plotted with trimmed tails and a dashed global mean. We no longer compute "zeros kept" variants, as discrepancies in sign direction (MAUP/Simpson's Paradox) are a known property of spatial aggregation rather than an artifact of zero-handling.

What this chunk does:
- Read the processed GPKG and normalize IDs (`fid`, `c_fid`), dropping any `c_fid.x/y` artifacts.
- Build a service mapping (`svc_map`) so raw column bases (with/without `_mean`) map to the canonical services; construct `col_pct`/`col_abs` names.
- `agg_by_group()`: intersect requested `groupings` with available columns; for each grouping × service, pick the pct/abs change column, drop zeros (`drop_zero = TRUE`), trim extremes at `cut_q`, then compute mean change by group. Returns tidy rows.
- Run `agg_by_group()` to produce `signed_summary` and a later chunk computes global references `glob_refs`.

```{r}
#| label: make-signed-bars-data
#| eval: false
#| echo: false
#| message: true
#| warning: false
#| cache: false

services <- svc_order
groupings <- c("income_grp","region_wb","WWF_biome")#, "nev_name") (for country if necessary)
cut_q <- 0.999
handle_inf <- "na"  # options: "na", "cap"
verbose <- TRUE
# helper for lightweight progress messages

vmsg <- function(...) if (isTRUE(verbose)) message(...)

# Load canonical grid + change data from Consolidation.qmd output.
# This file is the single source of truth for grid attributes and grouping joins.
gpkg <- file.path(data_dir(), "processed", params$input_gpkg)
if (!file.exists(gpkg)) {
  stop("Consolidation.qmd must be run first to produce 10k_change_calc.gpkg\nExpected at: ", gpkg)
}

vmsg("Reading canonical grid + change: ", gpkg)
sf_f <- sf::st_read(gpkg, quiet = TRUE)
sf_f <- sf::st_drop_geometry(sf_f)
sf_f <- dplyr::ungroup(sf_f)

# Normalize names to lowercase match canonical (robustness against casing)
for (target in c("income_grp", "region_wb", "continent", "region_un", "WWF_biome", "nev_name")) {
  match_idx <- which(tolower(names(sf_f)) == tolower(target))
  if (length(match_idx) == 1) names(sf_f)[match_idx] <- target
}

# Safety: Clamp any Infinite values (from 0 baseline) to +/- 200%
# This ensures emerging/disappearing services are captured as max change, not dropped.
sf_f <- sf_f %>%
  mutate(across(ends_with("_pct_chg"), ~ case_when(
    is.infinite(.) & . > 0 ~ 200,
    is.infinite(.) & . < 0 ~ -200,
    TRUE ~ .
  )))

# Handle grid_fid if present (renaming to fid for consistency)
if ("grid_fid" %in% names(sf_f) && !"fid" %in% names(sf_f)) {
  sf_f <- dplyr::rename(sf_f, fid = grid_fid)
}

# Ensure required keys present; rebuild if missing (e.g. if dropped during IO)
# TODO: Robustify grid alignment. Relying on row-order (seq_len) for fid generation is fragile
# if the input GPKG has been filtered or sorted differently from the base attributes.
if (!"fid" %in% names(sf_f)) {
  sf_f$fid <- seq_len(nrow(sf_f))
}

# Diagnostic: Check loaded columns
vmsg("Loaded columns: ", paste(head(names(sf_f), 10), collapse=", "), " ...")
vmsg("Total columns: ", ncol(sf_f))

# Diagnostic: Check NAs in grouping columns
for (g in groupings) {
  if (g %in% names(sf_f)) {
    n_na <- sum(is.na(sf_f[[g]]))
    # Force print to standard output so it's visible in error logs
    cat("Grouping '", g, "': ", n_na, " NAs out of ", nrow(sf_f), " rows (",
        round(n_na/nrow(sf_f)*100, 1), "%)\n", sep="")
    # Print sample values to verify content
    vals <- head(unique(na.omit(sf_f[[g]])), 5)
    cat("  Sample values: ", paste(vals, collapse=", "), "\n")
  } else {
    cat("Grouping '", g, "' NOT FOUND in sf_f.\n")
  }
}

available_groupings <- intersect(groupings, names(sf_f))
if (length(available_groupings) == 0) warning("No requested groupings found in data!")
vmsg("Groupings requested: ", paste(groupings, collapse=", "))
vmsg("Groupings found in data: ", paste(available_groupings, collapse=", "))

# Ensure required keys present (Consolidation.qmd guarantees this)
stopifnot("fid" %in% names(sf_f))

# build mapping; include both raw base and no-_mean base, with ready-to-use columns
chg_cols <- grep("_(abs|pct)_chg$", names(sf_f), value = TRUE)
vmsg("Found ", length(chg_cols), " change columns; mapping services...")
services_raw   <- unique(sub("_(abs|pct)_chg$", "", chg_cols))     # as in file (may include _mean)
services_raw   <- services_raw[!grepl("^(stdev|valid_count)_", services_raw)]
services_clean <- stringr::str_remove(services_raw, "_(mean|sum)$")
services_clean <- stringr::str_remove(services_clean, "^(mean|sum)_")
services_lower <- tolower(services_clean)
svc_map <- tibble::tibble(
  col_base  = c(services_raw, services_clean),
  canonical = dplyr::recode(c(services_lower, services_lower),
                            !!!canonical_lookup, .default = c(services_clean, services_clean))
) |>
  dplyr::distinct() |>
  dplyr::mutate(col_pct = paste0(col_base, "_pct_chg"),
                col_abs = paste0(col_base, "_abs_chg"))

agg_by_group <- function(df, services, groupings, cut_q = 0.999,
                         handle_inf = c("na","cap"), drop_zero = FALSE) {
  handle_inf <- match.arg(handle_inf)
  groupings <- intersect(groupings, names(df))
  vmsg("Groupings available: ", paste(groupings, collapse = ", "))
  out <- list()
  for (g in groupings) {
    vmsg("  Grouping: ", g)
    for (svc in services) {
      map_rows <- dplyr::filter(svc_map, canonical == svc)
      if (!nrow(map_rows)) next
      cols_pct <- map_rows$col_pct[map_rows$col_pct %in% names(df)]
      cols_abs <- map_rows$col_abs[map_rows$col_abs %in% names(df)]
      if (!length(cols_pct) && !length(cols_abs)) next

      add_one <- function(col, metric) {
        v <- df[[col]]
        if (metric == "pct" && any(is.infinite(v))) {
          if (handle_inf == "na") v[is.infinite(v)] <- NA_real_
        }
        if (isTRUE(drop_zero)) v[v == 0] <- NA_real_
        cap <- stats::quantile(abs(v), cut_q, na.rm = TRUE)
        v_trim <- pmax(pmin(v, cap), -cap)
        tibble::tibble(
          service = svc,
          group   = df[[g]],
          metric  = metric,
          val     = v_trim
        )
      }

      rows <- list()
      if (length(cols_pct)) rows[[length(rows)+1]] <- add_one(cols_pct[1], "pct")
      if (length(cols_abs)) rows[[length(rows)+1]] <- add_one(cols_abs[1], "abs")
      rows <- dplyr::bind_rows(rows)
      if (!nrow(rows)) next
      rows <- rows |>
        dplyr::filter(!is.na(group)) |>
        dplyr::group_by(service, metric, group) |>
        dplyr::summarise(mean_chg = mean(val, na.rm = TRUE), .groups = "drop") |>
        dplyr::mutate(grouping = g)
      out[[length(out)+1]] <- rows
    }
  }
  dplyr::bind_rows(out)
}

cat("Dimensions of sf_f before filtering: ", paste(dim(sf_f), collapse = " x "), "\n")
original_rows <- nrow(sf_f)

# Defer filtering to here (just before aggregation) to protect FID alignment upstream
# Create a working subset so sf_f remains full (unfiltered) for downstream chunks (pivot-plt-long)
sf_subset <- sf_f

# Use explicit NA check to avoid dropping rows where attributes are missing
if ("continent" %in% names(sf_subset)) {
  sf_subset <- dplyr::filter(sf_subset, is.na(continent) | !continent %in% c("Antarctica", "Seven seas (Open Ocean)"))
}
if ("region_wb" %in% names(sf_subset)) {
  sf_subset <- dplyr::filter(sf_subset, is.na(region_wb) | !region_wb %in% c("Antarctica"))
}
if ("WWF_biome" %in% names(sf_subset)) {
  sf_subset <- dplyr::filter(sf_subset, is.na(WWF_biome) | !WWF_biome %in% c("Lakes", "Rock & Ice"))
}

cat("Dimensions of sf_subset after filtering: ", paste(dim(sf_subset), collapse = " x "), "\n")
cat("Filtered out ", original_rows - nrow(sf_subset), " rows.\n")

if (nrow(sf_subset) == 0) {
  stop("CRITICAL: All rows were filtered out! Check if grouping columns are fully NA or filters are too aggressive.")
}

signed_summary <- agg_by_group(sf_subset, services, groupings, cut_q = cut_q,
                               handle_inf = handle_inf, drop_zero = TRUE)

if (nrow(signed_summary) == 0) {
  vmsg("Svc map entries: ", nrow(svc_map))
  stop("CRITICAL: signed_summary is empty. No plots will be generated. Check if grouping columns are present in input or AOOGrid.")
}

if (nrow(signed_summary) > 0) {
  print(dplyr::count(signed_summary, grouping, metric, service))
}
```

```{r}
#| label: compute-global-refs
#| eval: false
#| message: false
#| warning: false
#| echo: false

compute_global_refs <- function(df, services, cut_q = 0.999,
                                handle_inf = c("na","cap"), drop_zero = FALSE) {
  handle_inf <- match.arg(handle_inf)
  vmsg("Computing global refs: drop_zero = ", drop_zero)
  out <- list()
  for (svc in services) {
    map_rows <- dplyr::filter(svc_map, canonical == svc)
    if (!nrow(map_rows)) next
    for (metric in c("pct","abs")) {
      cols <- if (metric == "pct") map_rows$col_pct else map_rows$col_abs
      cols <- cols[cols %in% names(df)]
      if (!length(cols)) next
      v <- df[[cols[1]]]
      if (metric == "pct" && any(is.infinite(v))) {
        if (handle_inf == "na") v[is.infinite(v)] <- NA_real_
      }
      if (isTRUE(drop_zero)) v[v == 0] <- NA_real_
      cap <- stats::quantile(abs(v), cut_q, na.rm = TRUE)
      v_trim <- pmax(pmin(v, cap), -cap)
      out[[length(out)+1]] <- tibble::tibble(
        service = svc,
        metric  = metric,
        ref     = mean(v_trim, na.rm = TRUE)
      )
    }
  }
  dplyr::bind_rows(out)
}

glob_refs <- compute_global_refs(sf_f, services, cut_q = cut_q,
                                 handle_inf = handle_inf, drop_zero = TRUE)
```

```{r}
#| label: make-signed-bars-plot
#| eval: false
#| message: false
#| warning: false
#| echo: false

plot_signed_bars <- function(df, grouping, metric, refs, out_dir_base = "outputs/plots") {
  d <- dplyr::filter(df, grouping == !!grouping, metric == !!metric)
  if (!nrow(d)) return(invisible(NULL))
  d$service <- factor(d$service, levels = services)
  d$group   <- factor(d$group, levels = rev(sort(unique(d$group))))
  refs_use <- dplyr::filter(refs, metric == !!metric)
  refs_use$service <- factor(refs_use$service, levels = services)

  # Handle output directory with suffix
  out_dir <- out_dir_base

  p <- ggplot(d, aes(x = mean_chg, y = group, fill = group)) +
    geom_col(show.legend = FALSE) +
    geom_vline(xintercept = 0, color = "#7f7f7f", linewidth = 0.6) +
    geom_vline(data = refs_use, aes(xintercept = ref),
               linetype = "dashed", color = "#4a4a4a", linewidth = 0.5) +
    facet_wrap(~ service, scales = "free_x", ncol = 3) +
    labs(title = paste0("Regional Mean Change — ", grouping, " [", metric, "]"),
         subtitle = "Averaged over active areas only; dashed line = global mean",
         y = NULL,
         x = if (metric == "pct") "Mean % change (trimmed, signed)" else "Mean absolute change (trimmed, signed)") +
    theme_minimal(base_size = 12) +
    theme(strip.background = element_rect(fill = "#f3f4f6", color = NA),
          strip.text = element_text(face = "bold"))
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

  fp <- file.path(out_dir, paste0("bars_signed_", tolower(grouping), "_", metric, ".png"))
  ggsave(fp, p, width = 12, height = 8, dpi = 300, bg = "white")

  message("Saved regional bars: ", fp)
}

# Write plots
for (g in groupings) {
  for (m in c("abs", "pct")) {
    plot_signed_bars(signed_summary, grouping = g, metric = m,
                     refs = glob_refs,
                     out_dir_base = file.path("outputs", "plots", paste0("signed_bars", params$output_suffix)))
  }
}
```


## Load cached long table (plt_long) for hotspots/KS

::: callout-note
**Data sharing and external validation**

The pivoted long table (`plt_long`) is exported as an RDS file to the processed data directory (e.g., `/home/jeronimo/data/global_ncp/processed/plt_long.rds`). This file contains all per-cell, per-service change metrics and is the canonical input for hotspot and KS analyses.

- **For R users:** Load with:
  ```r
  source("R/config.R") # to load cfg object
  plt_long <- readRDS(file.path(cfg$paths$processed_data_dir, "plt_long.rds"))
  ```
- **For Python users:** Use `pyreadr` (`pip install pyreadr`), e.g.:
  ```python
  import pyreadr
  # Note: You may need to define the path manually if not running in an R-integrated environment
  processed_data_dir = "/home/jeronimo/data/global_ncp/processed"
  result = pyreadr.read_r(f'{processed_data_dir}/plt_long.rds')
  df = result[None]
  ```
- **For other tools:** RDS is a binary format; consider converting to CSV or Feather for broader compatibility if needed.

This object is suitable for external review, reproducibility checks, and further interpretation in any suitable software.
:::

Hotspot export and KS still expect `plt_long`. To avoid the heavy pivot here, we load the cached table from the processed data directory. If it’s missing, this script can regenerate it.

```{r}
#| label: pivot-plt-long
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| cache: false

# Generate or load plt_long
rds_path <- file.path(cfg$paths$processed_data_dir, paste0("plt_long", params$output_suffix, ".rds"))

need_build <- TRUE
if (file.exists(rds_path) && !isTRUE(params$force_export)) {
  message("Loading cached plt_long from: ", rds_path)
  plt_long <- readRDS(rds_path)
  # Check if we have the ratio services. If not, force rebuild.
  if ("N_Ret_Ratio" %in% unique(plt_long$service)) {
    need_build <- FALSE
  } else {
    message("Cached plt_long is missing ratio services. Rebuilding...")
  }
}

if (need_build) {
  # Generate plt_long directly from sf_f to ensure freshness and consistency
  # This replaces the need to run scratch/pivot_long.R separately.

# 1. Ensure source data (sf_f) is available
if (!exists("sf_f")) {
  gpkg <- file.path(data_dir(), "processed", params$input_gpkg)
  if (!file.exists(gpkg)) {
    stop("Canonical change file not found: ", gpkg)
  }
  message("Loading canonical change data for pivoting...")
  sf_f <- sf::st_read(gpkg, quiet = TRUE)

  sf_f <- sf::st_drop_geometry(sf_f)
  if ("orig_fid" %in% names(sf_f)) sf_f <- dplyr::rename(sf_f, fid = orig_fid)
  if ("grid_fid" %in% names(sf_f) && !"fid" %in% names(sf_f)) sf_f <- dplyr::rename(sf_f, fid = grid_fid)
  if (!"fid" %in% names(sf_f)) sf_f$fid <- seq_len(nrow(sf_f))

  # Robustness: Ensure grouping columns exist before pivoting
  # (Removed redundant AOOGrid join; input GPKG is authoritative)
}

# 2. Define grouping columns to keep
#    (Must match what is used in make-signed-bars-data and hotspots_config)
pivot_groupings <- c("c_fid", "iso3", "continent", "region_un", "subregion", "nev_name", "region_wb", "income_grp", "WWF_biome")
pivot_groupings <- intersect(pivot_groupings, names(sf_f))

# 3. Identify change columns
chg_cols <- grep("_(abs|pct)_chg$", names(sf_f), value = TRUE)
if (length(chg_cols) == 0) {
  stop("CRITICAL: No change columns found matching '_(abs|pct)_chg$'. Check input GPKG columns.")
}
service_suffix_pattern <- "_(mean|sum)$"
service_prefix_pattern <- "^(mean|sum|stdev|valid_count)_"

# Optimization: Deduplicate columns mapping to the same service before pivoting
matches <- stringr::str_match(chg_cols, "^(.*)_(abs|pct)_chg$")
col_map <- tibble::tibble(
  col  = chg_cols,
  base = matches[,2],
  type = matches[,3]
) |>
  dplyr::filter(!stringr::str_detect(base, "^(stdev|valid_count)_")) |>
  dplyr::mutate(service_norm = stringr::str_remove(base, service_suffix_pattern)) |>
  dplyr::mutate(service_norm = stringr::str_remove(service_norm, service_prefix_pattern)) |>
  dplyr::group_by(service_norm, type) |>
  dplyr::slice(1) |>
  dplyr::ungroup()
chg_cols <- col_map$col

message("Pivoting ", length(chg_cols), " change columns to long format...")

# 4. Pivot to long format
#    Structure: fid, c_fid, <groupings>, service, abs_chg, pct_chg
plt_long <- sf_f |>
  dplyr::select(dplyr::any_of(c("fid", pivot_groupings)), dplyr::all_of(chg_cols)) |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(chg_cols),
    names_to = c("service", ".value"),
    names_pattern = "^(.*)_(abs|pct)_chg$"
  ) |>
  # Normalize service names to canonical forms (matching HOTS_CFG)
  dplyr::mutate(service = stringr::str_remove(service, service_suffix_pattern)) |>
  dplyr::mutate(service = stringr::str_remove(service, service_prefix_pattern)) |>
  dplyr::mutate(service = tolower(service)) |>
  dplyr::mutate(service = dplyr::recode(service, !!!canonical_lookup)) |>
  dplyr::rename(abs_chg = abs, pct_chg = pct) |>
  dplyr::mutate(service = as.factor(service)) |>
  # Safety: Clamp any Infinite values (from 0 baseline) to +/- 200%
  # This ensures emerging/disappearing services are captured as max change, not dropped.
  dplyr::mutate(pct_chg = dplyr::case_when(
    is.infinite(pct_chg) & pct_chg > 0 ~ 200,
    is.infinite(pct_chg) & pct_chg < 0 ~ -200,
    TRUE ~ pct_chg
  )) |>
  # Scrub floating point dust from Python raster overlap that creates fake 200% hotspots
  dplyr::mutate(
    pct_chg = ifelse(!is.na(abs_chg) & abs(abs_chg) < 1e-6, 0, pct_chg)
  )

# Diagnostic: Check for NAs in pivoted data
if (nrow(plt_long) == 0) {
  stop("CRITICAL: plt_long is empty after pivot. The spatial coordinate join likely failed (0 matches) due to a CRS mismatch.")
}

na_pct <- sum(is.na(plt_long$pct_chg))
if (na_pct == nrow(plt_long)) {
  stop("CRITICAL: All pct_chg values in plt_long are NA. Check input data (10k_change_calc.gpkg).")
}

# 5. Cache the result (optional but helpful for debugging)
rds_path <- file.path(cfg$paths$processed_data_dir, paste0("plt_long", params$output_suffix, ".rds"))
dir.create(dirname(rds_path), recursive = TRUE, showWarnings = FALSE)
saveRDS(plt_long, rds_path)
message("plt_long generated and cached to: ", rds_path)
}
```

## Hotspot extraction workflow (global + subregional)

We identify per-service hotspots using a 5% percentile rule and **direction vectors**: - `loss_services` (e.g., Nature_Access, Pollination, N/Sed_Ret_Ratio, C_Risk_Red_Ratio) → we flag the **lowest** values; - `gain_services` (Sed_export, N_export, C_Risk) → we flag the **highest** values.
**Definition of a Hotspot:**
A hotspot is defined as a grid cell representing the **most detrimental 5% of change** globally for a given service.

*   **For Beneficial Services** (e.g., Nature Access, Pollination, Retention Ratios): Hotspots are cells with the **steepest declines** (bottom 5% of the distribution).
*   **For Detrimental Flows/Risks** (e.g., N/Sed Export, Coastal Risk): Hotspots are cells with the **steepest increases** (top 5% of the distribution).

We run this **once globally** and then **once per subregion** (World Bank region, income group, WWF biome). For each run and for each metric (absolute and percent change) we write a compact **GPKG** containing only the hotspot cells, plus a CSV index:

-   Output root: `processed/hotspots/`
    -   `abs/global/hotspots_global_abs.gpkg`
    -   `pct/global/hotspots_global_pct.gpkg`
    -   `abs/<group_col>/hotspots_<group_col>_<group_val>_abs.gpkg`
    -   `pct/<group_col>/hotspots_<group_col>_<group_val>_pct.gpkg`
-   Index: `processed/hotspots/_hotspots_index.csv` (columns: scope, group_col, group_val, metric, n_hot, gpkg).

These files are meant for QGIS/QA and downstream stats (e.g., KS) without recomputing hotspots.


### Hotspot rules & export configuration

The analysis uses a single, central configuration so the hotspot rules are consistent everywhere:

Thresholding: we flag hotspots using the top/bottom tails of the distribution per service. Here we use a percentile cutoff (e.g., 5%) rather than a fixed count.

Direction of concern: services in loss are “worse when they go down” (we keep the lowest tail); services in gain are “worse when they go up” (we keep the highest tail).

Combos (optional): grouped service sets that we count per cell for quick composite summaries.

Export switches: choose whether to write GPKGs and/or the CSV index.

```{r}
#| label: hotspots_config
#| include: true
#| eval: true
#| echo: false
HOTS_CFG <- list(
  analysis_name   = "global_NCP_hotspots",
  pct_cutoff      = 0.05,
  threshold_mode  = "percent",
  rule_mode       = "vectors",
  loss = c("Nature_Access","Pollination","N_Ret_Ratio","Sed_Ret_Ratio","C_Risk_Red_Ratio"),
  gain = c("Sed_export","N_export","C_Risk"),
  combos = list(
    deg_combo = c("Nature_Access","Pollination","N_export","Sed_export","C_Risk"),
    rec_combo = c("Nature_Access","Pollination","N_Ret_Ratio","Sed_Ret_Ratio","C_Risk_Red_Ratio")
  ),
# centralize the grouping columns here
  groupings = c("income_grp","region_wb","WWF_biome","nev_name"),
  # IO
  write_layers = TRUE,
  write_index  = TRUE,
  out_dir      = file.path(data_dir(), "processed", paste0("hotspots", params$output_suffix))
)
```

```{r}
#| label: show_hotspots_config
#| echo: false
loss_txt <- paste(HOTS_CFG$loss, collapse = ", ")
gain_txt <- paste(HOTS_CFG$gain, collapse = ", ")
grp_txt <- paste(HOTS_CFG$groupings, collapse = ", ")
combo_txt <- if (length(HOTS_CFG$combos))
  paste(paste0("**", names(HOTS_CFG$combos), "**: ",
               vapply(HOTS_CFG$combos, \(v) paste(v, collapse=", "), "")),
        collapse = "<br>") else "None"

cat(paste0(
"::: callout-note\n",
"**Hotspot configuration**\n\n",
"- Cutoff: ", HOTS_CFG$pct_cutoff * 100, "% (", HOTS_CFG$threshold_mode, ")\n",
"- Rule mode: ", HOTS_CFG$rule_mode, "\n",
"- Loss services: ", loss_txt, "\n",
"- Gain services: ", gain_txt, "\n",
"- Combos: ", combo_txt, "\n",
"…\n- Groupings: ", grp_txt, "\n…",
"- Write layers: ", HOTS_CFG$write_layers, " | Write index: ", HOTS_CFG$write_index, "\n",
"- Output dir: `", HOTS_CFG$out_dir, "`\n",
":::"
))
```

### Validate hotspot configuration

```{r}
#| label: validate_hotspots_config
#| message: false
#| warning: false
#| eval: true
#| echo: false
# stopifnot(exists("plt_long"))
# svc_all <- unique(plt_long$service)
# if (length(intersect(HOTS_CFG$loss, HOTS_CFG$gain)) > 0) {
#   stop("A service appears in BOTH `loss` and `gain`. Fix HOTS_CFG.")
# }
# miss_loss <- setdiff(HOTS_CFG$loss, svc_all)
# miss_gain <- setdiff(HOTS_CFG$gain, svc_all)
# if (length(miss_loss) > 0 || length(miss_gain) > 0) {
#   warning("Services in HOTS_CFG not found in `plt_long$service`:\n",
#           if (length(miss_loss)) paste0("  - missing loss: ", paste(miss_loss, collapse=", "), "\n"),
#           if (length(miss_gain)) paste0("  - missing gain: ", paste(miss_gain, collapse=", "), "\n"))
# }
```

```{r}
#| label: ensure-plt-long-loaded
#| echo: false
if (!exists("plt_long")) {
  rds_path <- file.path(data_dir(), "processed", paste0("plt_long", params$output_suffix, ".rds"))
  if (file.exists(rds_path)) {
    message("Loading cached plt_long from: ", rds_path)
    plt_long <- readRDS(rds_path)
  } else {
    stop("Cached 'plt_long.rds' not found. Please enable the pivot chunk to generate it.")
  }
}
```

### Export hotspot layers

::: callout-note
**Hotspot export module**\
- Computes hotspot cells once (global + by subregion) for ABS and PCT change.\
- Writes compact GPKGs for mapping/QA and maintains `_hotspots_index.csv`.\
- Prereqs: `plt_long` in memory, `HOTS_CFG` defined (loss/gain/combos/etc.).
:::

```{r}
#| label: hotspots_setup
#| eval: true
#| message: true
#| warning: false
#| echo: false
#| include: false
#| cache: false

stopifnot(exists("plt_long"))

cat("Hotspot export target directory: ", HOTS_CFG$out_dir, "\n", file = stderr())
cat("plt_long rows: ", nrow(plt_long), "\n", file = stderr())

# Diagnostic: Check service name alignment
svc_in_data <- unique(as.character(plt_long$service))
svc_in_cfg <- unique(c(HOTS_CFG$loss, HOTS_CFG$gain))
cat("Services in data: ", paste(head(svc_in_data, 10), collapse=", "), "\n", file = stderr())
cat("Services in CFG:  ", paste(head(svc_in_cfg, 10), collapse=", "), "\n", file = stderr())
cat("Services missing from CFG: ", paste(setdiff(svc_in_data, svc_in_cfg), collapse=", "), "\n", file = stderr())

# CRITICAL CHECK: If no services match, stop immediately to avoid silent failure
if (length(intersect(svc_in_data, svc_in_cfg)) == 0) {
  stop("CRITICAL ERROR: No overlap between services in data and services in HOTS_CFG. Check column names and canonical_lookup.")
}

# ---- Geometry: prefer an in-memory slim grid, else build it --------------
if (exists("grid_sf") && inherits(grid_sf, "sf")) {
  geom_sf <- dplyr::select(grid_sf, dplyr::any_of(c("fid")))
} else {
  master_path <- file.path(data_dir(), "vector_basedata", "AOOGrid_10x10km_land.gpkg")
  stopifnot(file.exists(master_path))
  geom_sf <- sf::st_read(master_path, quiet = TRUE)
  if ("id" %in% names(geom_sf) && !"fid" %in% names(geom_sf)) {
    geom_sf <- dplyr::rename(geom_sf, fid = id)
  }

  # ASSIGN FID HERE BEFORE FILTERING
  if (!"fid" %in% names(geom_sf)) {
    geom_sf$fid <- seq_len(nrow(geom_sf))
  }

  geom_sf <- geom_sf |> dplyr::select(dplyr::any_of(c("fid", "iso3", "continent", "region_un", "subregion", "nev_name", "region_wb", "income_grp", "WWF_biome")))
}

# Ensure we filter Antarctica to match the logic used for plt_long (sf_f)
if ("continent" %in% names(geom_sf)) {
  geom_sf <- dplyr::filter(geom_sf, !continent %in% c("Antarctica", "Seven seas (Open Ocean)"))
}
if ("WWF_biome" %in% names(geom_sf)) {
  geom_sf <- dplyr::filter(geom_sf, !WWF_biome %in% c("Lakes", "Rock & Ice"))
}

if (!"fid" %in% names(geom_sf)) {
  geom_sf$fid <- seq_len(nrow(geom_sf))
}
stopifnot("fid" %in% names(geom_sf), !any(duplicated(geom_sf$fid)))

# Join Land Cover Transitions (diffeR output from LC_change.qmd)
lc_gpkg <- file.path(data_dir(), "processed", "10k_lcc_granular_metrics.gpkg")
if (file.exists(lc_gpkg)) {
  # Load granular land cover metrics
  lc_sf <- sf::st_read(lc_gpkg, quiet = TRUE)
  lc_df <- sf::st_drop_geometry(lc_sf)

  # Handle column rename from LC_change.qmd (fid -> grid_fid)
  if ("grid_fid" %in% names(lc_df) && !"fid" %in% names(lc_df)) {
    lc_df <- dplyr::rename(lc_df, fid = grid_fid)
  }

  # Identify specific driver columns (Granular Analysis)
  drivers <- list(
    "Forest_Loss" = "ForestLoss_Loss_Forest_1992_2020",
    "Urban_Exp"   = "Expansion_Gain_Urban_1992_2020",
    "Crop_Exp"    = "Expansion_Gain_Cropland_1992_2020"
  )

  # Fallback to pattern matching
  if (!any(unlist(drivers) %in% names(lc_df))) {
    drivers$Forest_Loss <- grep("ForestLoss_Loss_Forest", names(lc_df), value = TRUE)[1]
    drivers$Urban_Exp <- grep("Expansion_Gain_Urban", names(lc_df), value = TRUE)[1]
    drivers$Crop_Exp <- grep("Expansion_Gain_Cropland", names(lc_df), value = TRUE)[1]
    drivers <- drivers[!is.na(drivers)]
  }

  cols_to_keep <- unlist(drivers)
  cols_to_keep <- cols_to_keep[cols_to_keep %in% names(lc_df)]

  if (length(cols_to_keep) > 0) {
    lc_df_subset <- lc_df %>% dplyr::select(fid, dplyr::all_of(cols_to_keep))
    geom_sf <- dplyr::left_join(geom_sf, lc_df_subset, by = "fid")
    cat("Joined Granular Land Cover metrics to geometry.\n", file = stderr())
  }
}

# Diagnostic: check ID overlap
n_overlap <- length(intersect(unique(plt_long$fid), unique(geom_sf$fid)))
cat("Diagnostic: plt_long unique fids=", length(unique(plt_long$fid)), " geom_sf unique fids=", length(unique(geom_sf$fid)), " Overlap=", n_overlap, "\n", file = stderr())

# Alignment: Ensure plt_long strictly matches the geometry domain
# (e.g. if plt_long includes Antarctica but geom_sf filtered it out)
valid_fids <- unique(geom_sf$fid)
if (!all(plt_long$fid %in% valid_fids)) {
  cat("Aligning plt_long to geometry domain. Filtering out mismatched FIDs...\n", file = stderr())
  plt_long <- dplyr::filter(plt_long, fid %in% valid_fids)
}

# ---- Helper: safe slug for filenames -------------------------------------
slug <- function(x) {
  x <- gsub("[^A-Za-z0-9]+", "_", x)
  x <- gsub("_+", "_", x)
  sub("^_|_$", "", x)
}

# ---- Core runner (uses central HOTS_CFG) ----------------------------------
run_one_hotset <- function(df, value_col, scope,
                           group_col = NA_character_, group_val = NA_character_,
                           sf_obj = geom_sf,
                           cfg = HOTS_CFG) {
  # Filter to valid values (domain) so n_total reflects the service area, not global land
  df <- df[!is.na(df[[value_col]]), , drop = FALSE]

  # Optional subsetting by a specific group value
  if (!is.na(group_col) && !is.na(group_val)) {
    df <- df[df[[group_col]] %in% group_val, , drop = FALSE]
  }
  n_total <- nrow(df)
  if (nrow(df) == 0L) {
    return(tibble::tibble(scope, group_col, group_val = as.character(group_val),
                          metric = if (identical(value_col, "abs_chg")) "abs" else "pct",
                          n_hot = 0L, n_total = 0L, pct_hot = 0, gpkg = NA_character_))
  }

  # Safety: geometry must have all fids present in df
  stopifnot("fid" %in% names(df), "fid" %in% names(sf_obj))
  if (!all(df$fid %in% sf_obj$fid)) {
    missing <- setdiff(unique(df$fid), sf_obj$fid)
    stop(sprintf("Geometry is missing %d fid(s), e.g. %s",
                 length(missing), paste(head(missing, 5), collapse = ", ")))
  }

  # Single source of truth for rules/directions/combos
  hs <- extract_hotspots(
    df             = df,
    value_col      = value_col,
    pct_cutoff     = cfg$pct_cutoff,
    threshold_mode = cfg$threshold_mode,
    rule_mode      = cfg$rule_mode,
    loss_services  = cfg$loss,
    gain_services  = cfg$gain,
    combos         = cfg$combos,
    id_cols        = c("fid", intersect(names(df), c("c_fid", "iso3", "continent", "region_un", "subregion", "nev_name", "region_wb", "income_grp", "WWF_biome"))),
    sf_obj         = sf_obj,
    write_sf_path  = NULL,
    clean_names    = TRUE
  )

  # Output layout
  out_root <- cfg$out_dir
  metric_stub <- if (identical(value_col, "abs_chg")) "abs" else "pct"
  folder <- if (is.na(group_col)) file.path(out_root, metric_stub, "global")
            else                   file.path(out_root, metric_stub, tolower(group_col))
  dir.create(folder, recursive = TRUE, showWarnings = FALSE)

  file_stub <- if (is.na(group_col)) {
    sprintf("hotspots_global_%s", metric_stub)
  } else {
    sprintf("hotspots_%s_%s_%s", tolower(group_col), slug(group_val), metric_stub)
  }
  out_gpkg <- file.path(folder, paste0(file_stub, ".gpkg"))

  # Write only hotspot features
  if (!is.null(hs$hotspots_sf) && nrow(hs$hotspots_sf) > 0) {
     n_hot <- nrow(hs$hotspots_sf)

     if (file.exists(out_gpkg) && !isTRUE(params$force_export)) {
       cat("[DIAGNOSTIC] Skipping write (file exists): ", out_gpkg, "\n", file = stderr())
     } else {
       if (isTRUE(params$force_export) && file.exists(out_gpkg)) {
         cat("[DIAGNOSTIC] Overwriting (force_export=TRUE): ", out_gpkg, "\n", file = stderr())
       } else {
         cat("[DIAGNOSTIC] Writing ", n_hot, " hotspots to: ", out_gpkg, "\n", file = stderr())
       }

       # Rename 'fid' to 'grid_fid' to avoid GPKG driver conflict (reserved primary key)
       sf_to_write <- hs$hotspots_sf
       if ("fid" %in% names(sf_to_write)) {
         sf_to_write <- dplyr::rename(sf_to_write, grid_fid = fid)
       }
       sf::st_write(sf_to_write, out_gpkg, quiet = TRUE, delete_dsn = TRUE)
       cat("[DIAGNOSTIC] Write complete.\n", file = stderr())
     }
  } else {
    cat("No hotspots found for ", metric_stub, " scope=", scope, " group=", group_val, "\n", file = stderr())
    out_gpkg <- NA_character_; n_hot <- 0L
  }

  tibble::tibble(
    scope      = scope,
    group_col  = ifelse(is.na(group_col), NA_character_, group_col),
    group_val  = ifelse(is.na(group_val), NA_character_, as.character(group_val)),
    metric     = metric_stub,
    n_hot      = n_hot,
    n_total    = n_total,
    pct_hot    = if (n_total > 0) (n_hot / n_total) * 100 else 0,
    gpkg       = out_gpkg
  )
}
```

```{r}
#| label: hotspots_execution
#| eval: true
#| include: false

# ---- Execute: global + subregional runs -----------------------------------

# Pull groupings from config, with a safe fallback
groupings <- if (!is.null(HOTS_CFG$groupings)) {
  HOTS_CFG$groupings
} else {
  c("income_grp","region_wb","continent","region_un","WWF_biome","nev_name")
}

# Run both metrics (hotspot export)
metrics_hs <- c("abs_chg","pct_chg")
index_rows <- lapply(metrics_hs, function(m) run_one_hotset(plt_long, value_col = m, scope = "global"))

# Subregional runs
for (gc in groupings) {
  if (!gc %in% names(plt_long)) next
  vals_chr <- as.character(stats::na.omit(unique(plt_long[[gc]])))
  # Explicitly filter out Antarctica from region_wb as it contains no modeled data
  if (gc == "region_wb") {
    vals_chr <- vals_chr[vals_chr != "Antarctica"]
  }
  cat("Processing grouping: ", gc, " with ", length(vals_chr), " values.\n", file = stderr())
  if (!length(vals_chr)) next

  for (m in metrics_hs) {
    index_rows <- append(index_rows, list(
      purrr::map_dfr(vals_chr, \(v) run_one_hotset(
        df        = plt_long,
        value_col = m,
        scope     = "by_group",
        group_col = gc,
        group_val = v
      ))
    ))
  }
}

hot_index <- dplyr::bind_rows(index_rows)

# Ensure the output dir exists before writing the CSV index
csv_dir <- HOTS_CFG$out_dir
dir.create(csv_dir, recursive = TRUE, showWarnings = FALSE)
readr::write_csv(hot_index, file.path(csv_dir, paste0("_hotspots_index", params$output_suffix, ".csv")))

# Small console summary
dplyr::glimpse(hot_index, width = 120)

```




```{r}
#| label: display-barplots
#| eval: false
#| echo: false
#| results: 'asis'
#| cache: false
# Show barplots directly
bar_dir <- file.path("outputs", "plots", paste0("signed_bars", params$output_suffix))
pngs <- list.files(bar_dir, pattern = "^bars_signed_.*\\.png$", full.names = FALSE)

if (length(pngs)) {
  rel_paths <- file.path("..", bar_dir, pngs)
  md <- c(
    "### Regional Mean Change (dashed = global mean)",
    paste(sprintf("![](%s)", rel_paths), collapse = "\n\n")
  )
  knitr::asis_output(paste(md, collapse = "\n\n"))
} else {
  knitr::asis_output("> No signed bar plots were generated; check upstream chunks.")
}
```

```{r}
#| label: load-cached-plt-long
#| echo: false
#| message: true

if (!exists("plt_long")) {
  rds_path <- file.path(cfg$paths$processed_data_dir, paste0("plt_long", params$output_suffix, ".rds"))
  if (file.exists(rds_path)) {
    message("Loading cached plt_long from: ", rds_path)
    plt_long <- readRDS(rds_path)
  } else {
    stop("Cached 'plt_long.rds' not found and the 'pivot' chunk was not run. Please run the pivot chunk once to generate the file.")
  }
}
```

## Hotspot boxplots

Summarize hotspot distributions by group using saved PNGs; skip computation if group columns are absent. Helper `run_hotspot_boxplots_by()` now lives in `R/hotspot_violins.R`, so we just call it from this report.

Outputs: PNGs are written to `outputs/plots/{abs|pct}/<group_col>/boxplots_*.png` and mirrored into `outputs/plots/latest/boxplots/` for embedding.

```{r}
#| label: make-hotspot-boxplots
#| message: false
#| warning: false
#| echo: false
#| cache: false
#| results: "hold"
#| eval: true

stopifnot(exists("plt_long"))
groupings <- HOTS_CFG$groupings

# Filter out Antarctica and Seven Seas for plotting
if ("continent" %in% names(plt_long)) {
  plt_long <- dplyr::filter(plt_long, !continent %in% c("Antarctica", "Seven seas (Open Ocean)"))
}
if ("WWF_biome" %in% names(plt_long)) {
  plt_long <- dplyr::filter(plt_long, !WWF_biome %in% c("Lakes", "Rock & Ice"))
}
if ("region_wb" %in% names(plt_long)) {
  plt_long <- dplyr::filter(plt_long, !region_wb %in% c("Antarctica"))
}

# Define service lists for directing plotting logic
volumetric_services <- c("Nature_Access", "Pollination", "Sed_export", "N_export")
ratio_services <- c("N_Ret_Ratio", "Sed_Ret_Ratio")
coastal_services <- c("C_Risk", "C_Risk_Red_Ratio")

missing_groupings <- setdiff(groupings, names(plt_long))
if (length(missing_groupings)) {
  warning("Grouping column(s) not found in plt_long and will be skipped: ",
          paste(missing_groupings, collapse = ", "))
}
groupings <- intersect(groupings, names(plt_long))
message("make-hotspot-boxplots groupings: ", if (length(groupings)) paste(groupings, collapse = ", ") else "none")
boxplot_order <- levels(plt_long$service)
if (is.null(boxplot_order)) {
  boxplot_order <- unique(plt_long$service)
}
boxplot_order <- as.character(boxplot_order)
if (length(groupings) == 0) {
  warning("No grouping columns found in plt_long; skipping hotspot boxplots.")
} else {

  # Cleanup legacy absolute coastal plots to avoid confusion in outputs
  legacy_abs_coastal <- list.files(out_plots(), pattern = ".*_abs_coastal\\.png$", recursive = TRUE, full.names = TRUE)
  if (length(legacy_abs_coastal) > 0) invisible(file.remove(legacy_abs_coastal))

  for (gc in groupings) {
    # Check for valid data before attempting to plot
    n_valid <- sum(!is.na(plt_long[[gc]]))
    if (n_valid == 0) {
      message("Skipping boxplots for ", gc, ": all values are NA.")
      next
    }

    message("Rendering hotspot boxplots for: ", gc)
    tryCatch({
      run_hotspot_boxplots_by(
        # Exclude coastal services here because they are handled in their own dedicated plot below
        df_long             = plt_long %>% dplyr::filter(!service %in% coastal_services),
        group_col           = gc,
        loss                = HOTS_CFG$loss,
        gain                = HOTS_CFG$gain,
        pct_cutoff          = HOTS_CFG$pct_cutoff,
        threshold_mode      = HOTS_CFG$threshold_mode,
        svc_order           = boxplot_order,
        keep_only_ordered   = TRUE,
        volumetric_services = volumetric_services,
        ratio_services      = ratio_services,
        top_bottom_n        = if (gc == "nev_name") 10 else NULL
      )
    }, error = function(e) {
      message("Error generating boxplots for ", gc, ": ", e$message)
    })

    # ------------------------------------------------------------------
    # Custom plotting for Coastal Protection (Free Scales)
    # ------------------------------------------------------------------
    # This addresses the issue where C_Risk (index) and C_Risk_Red_Ratio (ratio)
    # appear as flat lines when faceted with other ratio services due to scale differences.

    # Ensure we have coastal services in the data
    curr_coastal <- intersect(coastal_services, unique(plt_long$service))

    if (length(curr_coastal) > 0) {
      message("Rendering dedicated coastal boxplots for: ", gc)

      # We plot only Percent change because absolute change doesn't make sense for these indices
      for (metric in c("pct_chg")) {
        metric_label <- if (metric == "pct_chg") "Percent Change (%)" else "Absolute Change"
        file_tag     <- if (metric == "pct_chg") "pct" else "abs"

        plot_data_list <- list()

        for (svc in curr_coastal) {
          # Filter data for this service and group
          d_svc <- plt_long %>%
            dplyr::filter(service == svc, !is.na(.data[[metric]]), !is.na(.data[[gc]]))

          if (nrow(d_svc) == 0) next

          # Identify Hotspots (using PCT cutoff always, per HOTS_CFG)
          # Note: We use the global percentile logic here for simplicity of visualization
          is_loss <- svc %in% HOTS_CFG$loss

          # Calculate threshold on PCT change (standard definition)
          cutoff <- if (is_loss) quantile(d_svc$pct_chg, HOTS_CFG$pct_cutoff, na.rm = TRUE)
                    else         quantile(d_svc$pct_chg, 1 - HOTS_CFG$pct_cutoff, na.rm = TRUE)

          d_hot <- if (is_loss) d_svc %>% dplyr::filter(pct_chg <= cutoff)
                   else         d_svc %>% dplyr::filter(pct_chg >= cutoff)

          if (nrow(d_hot) > 0) {
            d_hot$Service <- svc
            plot_data_list[[svc]] <- d_hot
          }
        }

        if (length(plot_data_list) > 0) {
          coast_df <- dplyr::bind_rows(plot_data_list)

          # Handle Top 10 logic for Coastal if grouping is country
          x_var <- gc
          facet_scales <- "free_y"
          if (gc == "nev_name") {
            coast_df <- coast_df %>%
              dplyr::group_by(Service, .data[[gc]]) %>%
              dplyr::mutate(med_val = median(pct_chg, na.rm = TRUE), v_count = n()) %>%
              dplyr::ungroup() %>%
              dplyr::filter(v_count >= 10) %>%
              dplyr::arrange(Service, med_val) %>%
              dplyr::group_by(Service) %>%
              dplyr::filter(dplyr::dense_rank(med_val) <= 10 | dplyr::dense_rank(med_val) >= max(dplyr::dense_rank(med_val)) - 9) %>%
              dplyr::ungroup() %>%
              dplyr::mutate(plot_label = stats::reorder(paste(.data[[gc]], Service, sep="__"), med_val))

            x_var <- "plot_label"
            facet_scales <- "free"
          }

          p <- ggplot(coast_df, aes(x = .data[[x_var]], y = .data[[metric]])) +
            geom_boxplot(fill = "gray95", outlier.shape = 16, outlier.alpha = 0.3) +
            facet_wrap(~Service, scales = facet_scales, ncol = 1) +
            labs(title = paste("Hotspot Change Distribution:", gc),
                 subtitle = paste("Coastal Services (Free Scale) -", metric_label),
                 y = metric_label, x = NULL) +
            theme_minimal() +
            theme(axis.text.x = element_text(angle = 45, hjust = 1))

          if (gc == "nev_name") {
             p <- p + ggplot2::scale_x_discrete(labels = function(x) gsub("__.*$", "", x)) + ggplot2::coord_flip()
          }

          # Save to standard output structure so downstream copy picks it up
          out_sub <- file.path(out_plots(), "coastal", gc)
          dir.create(out_sub, recursive = TRUE, showWarnings = FALSE)
          fname <- file.path(out_sub, paste0("boxplots_", gc, "_", file_tag, "_coastal.png"))
          ggsave(fname, p, width = 8, height = 8, bg = "white")
        }
      }
    }
  }
}

```

```{r}
#| label: display-boxplots
#| echo: false
#| results: 'asis'
#| cache: false
box_dir <- "outputs/plots"
box_files <- list.files(box_dir, pattern = "boxplots_.*\\.png$", recursive = TRUE, full.names = FALSE)
box_files <- box_files[!grepl("^latest/", box_files)]

if (length(box_files)) {
  rel_paths <- file.path("..", box_dir, sort(box_files))
  md <- paste(sprintf("![](%s)", rel_paths), collapse = "\n\n")
  knitr::asis_output(md)
} else {
  knitr::asis_output("> No boxplots were generated.")
}
```

## Attribution to Drivers of Change

Here we integrate Land Cover change, specifically focusing on Land Conversion (Loss of Natural Land). We define Conversion Hotspots as the top 5% of grid cells with the highest gross loss of natural land (Loss_2). We then calculate the overlap: What percentage of ES hotspots are also Conversion Hotspots?
```{r}
#| label: lc-hotspot-overlap-setup
#| eval: true
#| echo: false
#| message: false
#| warning: false

# 1. Load LC Data
lc_gpkg <- file.path(data_dir(), "processed", "10k_lcc_granular_metrics.gpkg")
lc_data_ready <- FALSE
lc_df <- NULL
loss_col <- NULL
  net_col <- NULL
if (file.exists(lc_gpkg) && exists("plt_long")) {
  # Load data with geometry to allow export of driver hotspots later
  message("SUCCESS: Found lc_gpkg and plt_long. Proceeding with driver analysis setup.")
  lc_sf_full <- sf::st_read(lc_gpkg, quiet = TRUE)
  lc_df <- sf::st_drop_geometry(lc_sf_full)

  # Handle column rename from LC_change.qmd (fid -> grid_fid)
  if ("grid_fid" %in% names(lc_df) && !"fid" %in% names(lc_df)) {
    lc_df <- dplyr::rename(lc_df, fid = grid_fid)
    lc_sf_full <- dplyr::rename(lc_sf_full, fid = grid_fid)
  }

  # Identify specific driver columns (Granular Analysis)
  # We look for Forest Loss and Urban/Cropland Expansion
  # Explicitly targeting 1992-2020 transitions based on LCC_column_dictionary.md
  drivers <- list(
    "Forest_Loss" = "ForestLoss_Loss_Forest_1992_2020",
    "Crop_Exp"    = "Expansion_Gain_Cropland_1992_2020"
  )

  # Filter out columns that don't exist in the data
  drivers <- drivers[sapply(drivers, function(x) x %in% names(lc_df))]

  # Fallback: If exact match failed, try pattern matching (ignoring year suffixes)
  if (length(drivers) == 0) {
    message("Exact match failed. Trying pattern matching...")
    fl_col <- grep("ForestLoss_Loss_Forest", names(lc_df), value = TRUE)
    if (length(fl_col) > 0) drivers[["Forest_Loss"]] <- fl_col[1]

    ce_col <- grep("Expansion_Gain_Cropland", names(lc_df), value = TRUE)
    if (length(ce_col) > 0) drivers[["Crop_Exp"]] <- ce_col[1]
  }

  # Identify Net Forest Change for correlation analysis
  net_col <- "ForestLoss_dir_ch_Forest_1992_2020"
  if (!net_col %in% names(lc_df)) {
    net_col_candidates <- grep("ForestLoss_dir_ch_Forest", names(lc_df), value = TRUE)
    if (length(net_col_candidates) > 0) net_col <- net_col_candidates[1] else net_col <- NULL
  }

  if (length(drivers) > 0) {
    message("Found granular LC drivers: ", paste(names(drivers), collapse = ", "))
    lc_data_ready <- TRUE
  } else {
    warning("No granular driver columns found in LC data.")
    message("Available columns in LC data: ", paste(head(names(lc_df), 20), collapse = ", "))
  }
} else {
  msg <- paste0(
    "Skipping Driver Analysis. Reason: ",
    if (!file.exists(lc_gpkg)) paste0("LC data file not found at '", lc_gpkg, "'") else "",
    if (!file.exists(lc_gpkg) && !exists("plt_long")) " and " else "",
    if (!exists("plt_long")) "plt_long object not found in memory." else ""
  )
  warning(msg)
}
```

```{r}
#| label: lc-hotspot-overlap-calc
#| eval: true
#| echo: false
#| message: false
#| warning: false

overlap_stats <- NULL
if (lc_data_ready) {
  overlap_list <- list()

  for (d_name in names(drivers)) {
    d_col <- drivers[[d_name]]

    # Define Driver Hotspots (Top 5% of change)
    # For Loss columns (ForestLoss), high values = high loss.
    # For Gain columns (Urban/Crop), high values = high expansion.
    cutoff <- quantile(lc_df[[d_col]], 0.95, na.rm = TRUE)
    driver_hotspots <- lc_df %>%
      filter(.data[[d_col]] > cutoff) %>%
      pull(fid)

    # Export Driver Hotspots (Top 5%) to GPKG
    driver_out_dir <- file.path(data_dir(), "processed", paste0("hotspots", params$output_suffix), "drivers")
    if (!dir.exists(driver_out_dir)) dir.create(driver_out_dir, recursive = TRUE)

    driver_layer_name <- paste0("hotspots_", d_name, ".gpkg")
    out_path <- file.path(driver_out_dir, driver_layer_name)

    if (file.exists(out_path) && !isTRUE(params$force_export)) {
      message("Skipping driver hotspot export (file exists): ", out_path)
    } else {
      if (isTRUE(params$force_export) && file.exists(out_path)) {
        message("Overwriting driver hotspot (force_export=TRUE): ", out_path)
      } else {
        message("Exported driver hotspot layer: ", out_path)
      }
      # Rename fid -> grid_fid to avoid GPKG primary key conflict
      lc_sf_full %>%
        filter(fid %in% driver_hotspots) %>%
        select(grid_fid = fid, value = all_of(d_col)) %>%
        st_write(out_path, quiet = TRUE, delete_dsn = TRUE)
    }

    # Calculate Overlap
    stats <- plt_long %>%
      filter(service %in% c(HOTS_CFG$loss, HOTS_CFG$gain)) %>%
      group_by(service) %>%
      summarize(
        n_es_hotspots = sum(
          (service %in% HOTS_CFG$loss & pct_chg <= quantile(pct_chg, HOTS_CFG$pct_cutoff, na.rm=TRUE)) |
          (service %in% HOTS_CFG$gain & pct_chg >= quantile(pct_chg, 1-HOTS_CFG$pct_cutoff, na.rm=TRUE)),
          na.rm = TRUE
        ),
        n_overlap = sum(
          fid %in% driver_hotspots &
          ((service %in% HOTS_CFG$loss & pct_chg <= quantile(pct_chg, HOTS_CFG$pct_cutoff, na.rm=TRUE)) |
           (service %in% HOTS_CFG$gain & pct_chg >= quantile(pct_chg, 1-HOTS_CFG$pct_cutoff, na.rm=TRUE))),
          na.rm = TRUE
        )
      ) %>%
      mutate(
        pct_overlap = (n_overlap / n_es_hotspots) * 100,
        driver = d_name
      )

    overlap_list[[d_name]] <- stats
  }

  overlap_stats <- bind_rows(overlap_list) %>%
    arrange(desc(pct_overlap))
}
```

```{r}
#| label: lc-hotspot-overlap-plot
#| eval: true
#| echo: false
#| message: true
#| warning: false

if (!is.null(overlap_stats)) {
  # Make driver labels presentation-ready and enforce canonical ordering
  canon_order <- rev(c("C_Risk", "N_export", "Sed_export",
                       "C_Risk_Red_Ratio", "N_Ret_Ratio", "Sed_Ret_Ratio",
                       "Pollination", "Nature_Access"))

  overlap_plot_data <- overlap_stats %>%
    mutate(
      driver_label = case_when(
        driver == "Forest_Loss" ~ "Forest Loss",
        driver == "Crop_Exp" ~ "Cropland Expansion",
        driver == "Urban_Exp" ~ "Urban Expansion",
        TRUE ~ driver
      ),
      service = factor(service, levels = canon_order)
    ) %>%
    filter(!is.na(service))

  # Plot
  p_bar <- ggplot(overlap_plot_data, aes(x = service, y = pct_overlap, fill = driver_label)) +
   geom_col(position = "dodge") +
   scale_fill_brewer(palette = "Set2") +
   coord_flip() +
   labs(title = "Attribution of ES Hotspots to Land Cover Drivers",
        subtitle = "% of ES Hotspots overlapping with Top 5% of Driver Change",
        x = "Service", y = "% Overlap", fill = "Driver") +
   theme_minimal()
  print(p_bar)

  out_dir <- file.path("outputs", "plots", paste0("drivers", params$output_suffix))
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  out_path_bar <- file.path(out_dir, "bar_driver_overlap.png")
  ggsave(out_path_bar, p_bar, width = 10, height = 6, bg = "white", dpi = 300)

  # Add a Heatmap for instant readability
  p_heat <- ggplot(overlap_plot_data, aes(x = driver_label, y = service, fill = pct_overlap)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = sprintf("%.1f%%", pct_overlap)), color = "black", size = 4) +
    scale_fill_distiller(palette = "Reds", direction = 1, name = "% Overlap") +
    labs(title = "Attribution Gap: ES Hotspots vs. LCC Drivers",
         subtitle = "Percentage of ES Hotspots that overlap with Top 5% of Land Conversion",
         x = "Land Cover Driver", y = "Ecosystem Service") +
    theme_minimal(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid = element_blank())

  out_path_heat <- file.path(out_dir, "heatmap_driver_overlap.png")
  ggsave(out_path_heat, p_heat, width = 8, height = 6, bg = "white", dpi = 300)

  message("Saved overlap charts to: ", out_dir)
} else {
  message("Skipping overlap analysis (data not ready).")
}
```

```{r}
#| label: lc-es-correlation
#| eval: true
#| echo: false
#| message: true
#| warning: false

if (lc_data_ready && !is.null(net_col)) {
  # Join ES change (plt_long) with Net LC Change
  # We correlate ES Change (pct_chg) with Net Natural Change (dir_ch_2)
  # Hypothesis: Loss of Natural (negative dir_ch) -> Loss of ES (negative pct_chg for loss services)

  cor_data <- plt_long %>%
    dplyr::inner_join(lc_df %>% dplyr::select(fid, dplyr::all_of(net_col)), by = "fid") %>%
    dplyr::filter(!is.na(pct_chg), !is.na(.data[[net_col]])) %>%
    dplyr::filter(!service %in% c("USLE", "N_retention"))

  # Ensure canonical ordering for facets
  canon_order <- c("C_Risk", "N_export", "Sed_export", "C_Risk_Red_Ratio", "N_Ret_Ratio", "Sed_Ret_Ratio", "Pollination", "Nature_Access")
  cor_data$service <- factor(cor_data$service, levels = canon_order)

  # Calculate correlation per service
  cor_stats <- cor_data %>%
    dplyr::group_by(service) %>%
    dplyr::summarize(
      pearson_r = cor(pct_chg, .data[[net_col]], method = "pearson"),
      spearman_rho = cor(pct_chg, .data[[net_col]], method = "spearman"),
      n = n()
    ) %>%
    dplyr::arrange(desc(abs(spearman_rho)))

  print(knitr::kable(cor_stats, digits = 3, caption = "Correlation between ES % Change and Net Natural Land Change"))

  # Faceted plot for all services showing point density
  p_cor <- ggplot(cor_data, aes(x = .data[[net_col]], y = pct_chg)) +
    geom_bin2d(bins = 60) +
    scale_fill_viridis_c(trans = "log10", option = "viridis", labels = scales::comma) +
    geom_smooth(method = "lm", color = "tomato", linetype = "22", linewidth = 0.5) +
    facet_wrap(~ service, scales = "free_y", ncol = 3) +
    labs(title = "ES Change vs. Net Natural Change",
         x = "Net Change in Natural Land (%)",
         y = "ES Change (%)",
         fill = "Count\n(log10)") +
    theme_minimal() +
    theme(axis.title = element_text(size = 10),
          axis.text = element_text(size = 8))
    theme(axis.title = element_text(size = 8),
          axis.text = element_text(size = 8),
          strip.text = element_text(size = 10),
          legend.title = element_text(size = 9),
          legend.text = element_text(size = 7))

  print(p_cor)

  out_dir <- file.path("outputs", "plots", paste0("drivers", params$output_suffix))
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  ggsave(file.path(out_dir, "scatter_Net_Natural_Change.png"), p_cor, width = 12, height = 8, bg = "white", dpi = 300)
}
```

```{r}
#| label: lc-es-scatterplots
#| eval: true
#| echo: false
#| message: true
#| warning: false

if (lc_data_ready) {
  for (d_name in names(drivers)) {
    d_col <- drivers[[d_name]]

    # Filter data
    plot_data <- plt_long %>%
      dplyr::filter(!service %in% c("USLE", "N_retention")) %>%
      dplyr::inner_join(lc_df %>% dplyr::select(fid, driver_val = dplyr::all_of(d_col)), by = "fid") %>%
      dplyr::filter(driver_val > 0, !is.na(pct_chg)) # Focus on areas where change occurred

    # Ensure canonical ordering for facets
    canon_order <- c("C_Risk", "N_export", "Sed_export", "C_Risk_Red_Ratio", "N_Ret_Ratio", "Sed_Ret_Ratio", "Pollination", "Nature_Access")
    plot_data$service <- factor(plot_data$service, levels = canon_order)

    if (nrow(plot_data) > 100) {
      p_scatter <- ggplot(plot_data, aes(x = driver_val, y = pct_chg)) +
        geom_bin2d(bins = 60) +
        scale_fill_viridis_c(trans = "log10", option = "viridis", labels = scales::comma) +
        geom_smooth(method = "lm", color = "tomato", linetype = "22", linewidth = 0.5) +
        facet_wrap(~ service, scales = "free_y", ncol = 3) +
        labs(title = paste("Driver vs ES Change:", d_name),
             x = paste(d_name, "(% Area)"),
             y = "ES Change (%)",
             fill = "Count\n(log10)") +
        theme_minimal() +
        theme(axis.title = element_text(size = 10),
              axis.text = element_text(size = 8))
        theme(axis.title = element_text(size = 8),
              axis.text = element_text(size = 8),
          strip.text = element_text(size = 10),
          legend.title = element_text(size = 9),
          legend.text = element_text(size = 7))

      print(p_scatter)

      out_dir <- file.path("outputs", "plots", paste0("drivers", params$output_suffix))
      dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
      ggsave(file.path(out_dir, paste0("scatter_", d_name, ".png")), p_scatter, width = 12, height = 8, bg = "white", dpi = 300)
    }
  }
}
```

```{r}
#| label: lc-hotspots-export-by-group
#| eval: true
#| echo: false
#| message: true
#| warning: false

if (lc_data_ready) {
  message("Generating Land Cover Hotspots by Grouping...")

  # 1. Prepare LC Long Data (Drivers as 'services')
  # We need to join LC metrics with the grouping columns from plt_long (or sf_f)
  group_cols <- HOTS_CFG$groupings

  # Get attributes from plt_long (assuming it's loaded)
  if (exists("plt_long")) {
    # Get attributes from plt_long, but if grouping columns are missing, join them from the base grid.
    if (!all(group_cols %in% names(plt_long))) {
      message("Grouping columns missing from plt_long. Joining from base grid attributes.")
      grid_attrs_path <- file.path(data_dir(), "interim", "..", "vector_basedata", "AOOGrid_10x10km_land_4326_clean.gpkg")
      if (!file.exists(grid_attrs_path)) stop("Base grid attribute file not found.")

      grid_attrs_df <- sf::st_read(grid_attrs_path, quiet = TRUE) %>%
        sf::st_drop_geometry() %>%
        dplyr::select(any_of(c("fid", group_cols)))

      if (!"fid" %in% names(grid_attrs_df)) {
        grid_attrs_df$fid <- seq_len(nrow(grid_attrs_df))
      }

      attrs <- plt_long %>%
        dplyr::select(fid) %>%
        dplyr::distinct(fid, .keep_all = TRUE) %>%
        dplyr::left_join(grid_attrs_df, by = "fid")

    } else {
      attrs <- plt_long %>%
        dplyr::select(fid, dplyr::all_of(group_cols)) %>%
        dplyr::distinct(fid, .keep_all = TRUE)
    }

    # Select driver columns from lc_df
    driver_cols <- unlist(drivers)

    lc_subset <- lc_df %>%
      select(fid, all_of(unname(driver_cols))) %>%
      inner_join(attrs, by = "fid")

    # Pivot to long format for the runner
    lc_long <- lc_subset %>%
      pivot_longer(cols = all_of(unname(driver_cols)), names_to = "driver_key", values_to = "pct_chg") %>%
      mutate(
        # Map full column name back to short driver name
        service = names(drivers)[match(driver_key, drivers)],
        abs_chg = NA_real_ # LC metrics are usually % or area, treating as pct here
      ) %>%
      filter(!is.na(pct_chg))

    # 2. Define a temporary config for LC
    # We treat all drivers as "Gain" (high value = hotspot)
    LC_CFG <- HOTS_CFG
    LC_CFG$loss <- character(0)
    LC_CFG$gain <- names(drivers)
    LC_CFG$out_dir <- file.path(data_dir(), "processed", paste0("hotspots", params$output_suffix), "drivers_by_group")

    # 3. Run Extraction Loop (Global + Groups)
    # Global
    run_one_hotset(lc_long, value_col = "pct_chg", scope = "global", sf_obj = geom_sf, cfg = LC_CFG)

    # By Group
    for (gc in group_cols) {
      vals <- unique(na.omit(lc_long[[gc]]))
      # Filter out Antarctica/Seven Seas if present in vals (usually handled in run_one_hotset via df filter but good to be safe)
      for (v in vals) {
        run_one_hotset(lc_long, value_col = "pct_chg", scope = "by_group", group_col = gc, group_val = v, sf_obj = geom_sf, cfg = LC_CFG)
      }
    }
    message("LC Hotspots by group exported to: ", LC_CFG$out_dir)
  }
}
```


```{r}
#| label: lc-socio-correlation
#| eval: true
#| echo: false
#| message: true
#| warning: true

if (lc_data_ready) {
  # Load Socioeconomic data from the input GPKG
  gpkg_path <- file.path(data_dir(), "processed", params$input_gpkg)
  sf_socio <- st_read(gpkg_path, quiet = TRUE) %>% st_drop_geometry()

  # Ensure fid exists (consistent with other chunks)
  if (!"fid" %in% names(sf_socio)) {
    sf_socio$fid <- seq_len(nrow(sf_socio))
  }

  # Identify socio columns (heuristic)
  socio_cols <- grep("GDP|HDI|Pop|Gini", names(sf_socio), value = TRUE, ignore.case = TRUE)
  socio_cols <- socio_cols[!grepl("_chg", socio_cols)]

  if (length(socio_cols) > 0) {
    message("Found socio vars: ", paste(socio_cols, collapse=", "))

    # Join with LC drivers
    socio_lc <- sf_socio %>%
      select(fid, all_of(socio_cols)) %>%
      inner_join(lc_df, by = "fid")

    # Calculate correlations
    res_list <- list()
    for (d_name in names(drivers)) {
      d_col <- drivers[[d_name]]
      for (s_col in socio_cols) {

        valid_data <- socio_lc %>%
          select(d = all_of(d_col), s = all_of(s_col)) %>%
          filter(complete.cases(.))

        if (nrow(valid_data) > 100) {
           cor_val <- cor(valid_data$d, valid_data$s, method = "spearman")
           res_list[[length(res_list) + 1]] <- data.frame(
             Driver = d_name,
             SocioVar = s_col,
             SpearmanRho = cor_val
           )
        }
      }
    }

    if (length(res_list) > 0) {
      res_df <- bind_rows(res_list) %>% arrange(desc(abs(SpearmanRho)))
      print(knitr::kable(res_df, digits = 3, caption = "Correlation: LC Drivers vs Socioeconomic Status"))
    }
  }
}
```
