---
title: "Hotspot Spatial Synthesis"
subtitle: "Intensity, Enrichment, and Multi-service Overlap"
author: "Jerónimo Rodríguez Escobar"
date: last-modified
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
editor: source
params:
  input_gpkg: "10k_change_calc.gpkg"
  input_plt_long: "plt_long.rds"
  output_suffix: "" # e.g., "_v2"
---

## Overview

This notebook consolidates all spatial summary statistics for global hotspots into a single pipeline. It calculates:

1.  **Hotspot Intensity (Coverage)**: The percentage of land area within a spatial unit that is classified as a hotspot.
2.  **Global Share (Distribution)**: The proportion of all global hotspot gridcells that are located within a specific spatial unit.
3.  **Enrichment**: A ratio comparing the observed share of hotspots to the expected share based on land area (Observed / Expected).
4.  **Multi-service Hotspots & Hotness**: How many services is a single gridcell a hotspot for, and which regions have the highest average number of overlapping hotspots?

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
library(devtools)

# Initialize paths and load project functions
knitr::opts_knit$set(root.dir = here::here())
source(here::here("R","paths.R"))
devtools::load_all(quiet = TRUE)

options(dplyr.summarise.inform = FALSE)
```

## Cleanup Legacy Outputs
```{r}
#| label: cleanup-legacy

# 1. Clear summary CSVs
csvs_to_remove <- c(
file.path(data_dir(), "processed", "hotspot_area_stats.csv"),
file.path(data_dir(), "processed", "hotspot_multiservice_stats.csv")
)
for (f in csvs_to_remove) if (file.exists(f)) file.remove(f)

# 2. Clear intensity plots directory
intensity_dir <- file.path(out_plots(), paste0("intensity", params$output_suffix))
if (dir.exists(intensity_dir)) {
  unlink(list.files(intensity_dir, full.names = TRUE))
  message("Cleared legacy intensity plots in: ", intensity_dir)
}

# 3. Clear hotness plots from base plots directory
old_hotness <- list.files(out_plots(), pattern = "^hotness_.*\\.png$", full.names = TRUE)
if (length(old_hotness) > 0) {
  file.remove(old_hotness)
  message("Removed ", length(old_hotness), " old hotness plots.")
}
```

## Load Data

```{r}
#| label: load-data

# 1. Load Hotspot Data (plt_long)
rds_path <- file.path(data_dir(), "processed", params$input_plt_long)
if (file.exists(rds_path)) {
  message("Loading data from: ", rds_path)
  plt_long <- readRDS(rds_path)
} else {
  stop(params$input_plt_long, " not found. Please run analysis/hotspot_extraction.qmd first.")
}

# 2. Load Canonical Grid (for accurate denominators, including areas with no service)
gpkg_grid <- file.path(data_dir(), "processed", params$input_gpkg)
if (file.exists(gpkg_grid)) {
  message("Loading canonical grid from: ", gpkg_grid)
  grid_df <- sf::st_read(gpkg_grid, quiet = TRUE)
  grid_df <- sf::st_drop_geometry(grid_df)
  if (!"fid" %in% names(grid_df)) grid_df$fid <- seq_len(nrow(grid_df))
} else {
  stop("Canonical grid file not found: ", gpkg_grid)
}
```

## Configuration & Exclusions

```{r}
#| label: config

HOTS_CFG <- list(
  pct_cutoff      = 0.05,
  threshold_mode  = "percent",
  rule_mode       = "vectors",
  loss = c("Nature_Access","Pollination","N_Ret_Ratio","Sed_Ret_Ratio","C_Risk_Red_Ratio"),
  gain = c("Sed_export","N_export","C_Risk"),
  groupings = c("income_grp","region_wb","WWF_biome"),
  svc_order = c(
    "C_Risk","N_export","Sed_export",
    "C_Risk_Red_Ratio","N_Ret_Ratio","Sed_Ret_Ratio",
    "Pollination","Nature_Access"
  )
)

# Master Exclusion Lists (Applies to all downstream Intensity & Hotness calculations)
exclude_services <- c("C_Risk_NoHab", "USLE")
exclude_regions  <- c(
  "Seven seas (Open Ocean)", "Seven seas (open ocean)", "Antarctica",
  "Lakes", "Rock & Ice"
)

valid_groupings <- intersect(HOTS_CFG$groupings, names(grid_df))
```

## Extract Global Hotspots

```{r}
#| label: extract-hotspots

hs_global_obj <- extract_hotspots(
  df = plt_long,
  value_col = "pct_chg",
  pct_cutoff = HOTS_CFG$pct_cutoff,
  threshold_mode = HOTS_CFG$threshold_mode,
  rule_mode = HOTS_CFG$rule_mode,
  loss_services = HOTS_CFG$loss,
  gain_services = HOTS_CFG$gain,
  combos = NULL,
  id_cols = c("fid")
)
```

## Part 1: Intensity, Share, and Enrichment

```{r}
#| label: calc-intensity

hs_flags <- hs_global_obj$hotspots_df %>%
  dplyr::select(fid, service) %>%
  dplyr::mutate(is_hotspot = 1)

global_totals <- hs_flags %>%
  dplyr::count(service, name = "global_n_hot")

area_stats_list <- list()

for (g in valid_groupings) {
  stats_all <- lapply(HOTS_CFG$svc_order, function(svc) {
    grid_svc <- grid_df %>%
      dplyr::select(fid, !!g) %>%
      dplyr::mutate(service = svc) %>%
      dplyr::left_join(hs_flags %>% dplyr::filter(service == svc), by = c("fid", "service")) %>%
      dplyr::mutate(is_hotspot = tidyr::replace_na(is_hotspot, 0)) %>%
      # Apply master exclusions
      dplyr::filter(!service %in% exclude_services) %>%
      dplyr::filter(!.data[[g]] %in% exclude_regions) %>%
      dplyr::filter(!is.na(.data[[g]]))

    grid_svc %>%
      dplyr::group_by(service, group = .data[[g]]) %>%
      dplyr::summarise(
        n_total = dplyr::n(),
        n_hot   = sum(is_hotspot),
        pct_area = (n_hot / n_total) * 100,
        .groups = "drop"
      )
  })

  stats <- dplyr::bind_rows(stats_all) %>%
    dplyr::left_join(global_totals, by = "service") %>%
    dplyr::group_by(service) %>%
    dplyr::mutate(
      pct_share = (n_hot / global_n_hot) * 100,
      grouping_var = g,
      expected_share = (n_total / sum(n_total)) * 100,
      enrichment = ifelse(expected_share > 0, pct_share / expected_share, NA)
    ) %>%
    dplyr::ungroup() %>%
    dplyr::arrange(desc(pct_area))

  area_stats_list[[g]] <- stats
}

hotspot_area_stats <- dplyr::bind_rows(area_stats_list)

out_csv_path <- file.path(data_dir(), "processed", "hotspot_area_stats.csv")
readr::write_csv(hotspot_area_stats, out_csv_path)
message("Hotspot area statistics saved to: ", out_csv_path)
```

```{r}
#| label: plot-intensity
#| message: false

plot_area_stats <- function(df, g_var, metric = "pct_area") {
  metric_label <- switch(metric,
    "pct_area" = "% of Area that is a Hotspot (Intensity)",
    "pct_share" = "% Share of Global Hotspot Gridcells (Distribution)",
    "enrichment" = "Hotspot Enrichment (Observed/Expected)"
  )
  file_suffix <- switch(metric, "pct_area" = "coverage", "pct_share" = "share", "enrichment" = "enrichment")

  d_plot <- df %>%
    dplyr::filter(grouping_var == g_var) %>%
    dplyr::mutate(service = factor(service, levels = HOTS_CFG$svc_order)) %>%
    dplyr::filter(!is.na(service))

  if (nrow(d_plot) == 0) return(NULL)

  is_large_group <- dplyr::n_distinct(d_plot$group) > 20
  scales_val <- if (is_large_group) "free_y" else "free_x"

  if (!is_large_group) {
    d_plot <- d_plot %>%
      dplyr::arrange(dplyr::desc(group)) %>%
      dplyr::mutate(group_ordered = factor(group, levels = unique(group)))
  } else {
    d_plot <- d_plot %>% dplyr::mutate(group_ordered = group)
  }

  p <- ggplot2::ggplot(d_plot, ggplot2::aes(x = .data[[metric]], y = group_ordered, fill = service)) +
    ggplot2::geom_col(show.legend = FALSE) +
    ggplot2::facet_wrap(~ service, scales = scales_val, ncol = 3) +
    ggplot2::labs(
      title = paste0("Hotspot ", if (metric == "pct_area") "Intensity" else if (metric == "enrichment") "Enrichment" else "Distribution", " by ", g_var),
      subtitle = metric_label, x = metric_label, y = NULL
    ) +
    ggplot2::theme_minimal()

  if (metric == "enrichment") p <- p + ggplot2::geom_vline(xintercept = 1, linetype = "dashed", color = "black")
  if (is_large_group) p <- p + ggplot2::scale_y_discrete(labels = function(x) sub("___.*", "", x))

  out_dir <- file.path(out_plots(), paste0("intensity", params$output_suffix))
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
  ggsave(file.path(out_dir, paste0("hotspot_", file_suffix, "_", tolower(g_var), ".png")), p, width = 12, height = 8, bg = "white")
  return(p)
}

plots_intensity <- lapply(names(area_stats_list), function(g) plot_area_stats(hotspot_area_stats, g, "pct_area"))
plots_share     <- lapply(names(area_stats_list), function(g) plot_area_stats(hotspot_area_stats, g, "pct_share"))
plots_enrichment <- lapply(names(area_stats_list), function(g) plot_area_stats(hotspot_area_stats, g, "enrichment"))
```

## Part 2: Multi-service Overlap (Hotness)

```{r}
#| label: calc-overlap

hotspot_counts <- hs_global_obj$hotspots_df %>%
  dplyr::group_by(fid) %>%
  dplyr::summarise(hotspot_count = dplyr::n(), .groups = "drop")

df_overlap <- grid_df %>%
  dplyr::select(fid, dplyr::all_of(valid_groupings)) %>%
  dplyr::distinct(fid, .keep_all = TRUE) %>%
  dplyr::left_join(hotspot_counts, by = "fid") %>%
  dplyr::mutate(hotspot_count = tidyr::replace_na(hotspot_count, 0))

calc_hotness <- function(df, g_var) {
  df %>%
    dplyr::filter(!is.na(.data[[g_var]])) %>%
    # Apply master exclusions strictly here too
    dplyr::filter(!.data[[g_var]] %in% exclude_regions) %>%
    dplyr::group_by(group = .data[[g_var]]) %>%
    dplyr::summarise(
      mean_hotspots = mean(hotspot_count),
      pct_multi = mean(hotspot_count > 1) * 100,
      n_pixels = dplyr::n()
    ) %>%
    dplyr::mutate(grouping = g_var) %>%
    dplyr::arrange(desc(mean_hotspots))
}

hotness_stats <- dplyr::bind_rows(lapply(valid_groupings, function(g) calc_hotness(df_overlap, g)))
out_csv <- file.path(data_dir(), "processed", "hotspot_multiservice_stats.csv")
readr::write_csv(hotness_stats, out_csv)
```

```{r}
#| label: plot-overlap

plot_hotness <- function(df, g_var) {
  d_plot <- df %>% dplyr::filter(grouping == g_var) %>% dplyr::slice_head(n = 15) %>%
    dplyr::mutate(group = factor(group, levels = rev(sort(unique(as.character(group))))))

  p <- ggplot2::ggplot(d_plot, ggplot2::aes(x = group, y = mean_hotspots)) +
    ggplot2::geom_col(fill = "#e74c3c") +
    ggplot2::coord_flip() +
    ggplot2::labs(title = paste0("Average Hotspot Overlap by ", g_var), subtitle = "Hotspot Density (including zero-hotspot areas)", x = NULL, y = "Avg. Hotspot Count") +
    ggplot2::theme_minimal()

  ggsave(file.path(out_plots(), paste0("hotness_", tolower(g_var), ".png")), p, width = 10, height = 6, bg = "white")
  return(p)
}

# Plot B: Stacked Distribution (1, 2, 3+ services)
plot_stacked_overlap <- function(df, g_var) {
  d_summ <- df %>%
    dplyr::filter(!is.na(.data[[g_var]])) %>%
    dplyr::filter(!.data[[g_var]] %in% exclude_regions) %>%
    dplyr::filter(hotspot_count > 0) %>% # Conditional on being a hotspot
    dplyr::mutate(
      cat = dplyr::case_when(
        hotspot_count == 1 ~ "1",
        hotspot_count == 2 ~ "2",
        hotspot_count >= 3 ~ "3+",
        TRUE ~ as.character(hotspot_count)
      ),
      cat = factor(cat, levels = c("1", "2", "3+"))
    ) %>%
    dplyr::count(group = .data[[g_var]], cat) %>%
    dplyr::group_by(group) %>%
    dplyr::mutate(pct = n / sum(n)) %>%
    dplyr::ungroup()

  # Filter top 15 groups by "3+" share
  top_groups <- d_summ %>% dplyr::filter(cat == "3+") %>% dplyr::arrange(desc(pct)) %>% dplyr::slice_head(n = 15) %>% dplyr::pull(group)
  if (length(unique(d_summ$group)) > 20) d_summ <- d_summ %>% dplyr::filter(group %in% top_groups)

  # Order alphabetically (reversed for coord_flip to show A at top)
  d_summ <- d_summ %>% dplyr::mutate(group = factor(group, levels = rev(sort(unique(as.character(group))))))

  p <- ggplot2::ggplot(d_summ, ggplot2::aes(x = group, y = pct, fill = cat)) +
    ggplot2::geom_col(position = "fill") +
    ggplot2::scale_fill_viridis_d(option = "magma", direction = -1, name = "Services") +
    ggplot2::coord_flip() +
    ggplot2::labs(
      title = paste0("Multi-service Hotspot Distribution by ", g_var),
      subtitle = "Share of hotspot gridcells by number of overlapping services",
      x = NULL, y = "Proportion of hotspot gridcells"
    ) +
    ggplot2::theme_minimal()
  ggsave(file.path(out_plots(), paste0("hotness_dist_", tolower(g_var), ".png")), p, width = 10, height = 6, bg = "white")
  return(p)
}

plots_hotness <- lapply(valid_groupings, function(g) plot_hotness(hotness_stats, g))
plots_dist    <- lapply(valid_groupings, function(g) plot_stacked_overlap(df_overlap, g))
if (length(plots_hotness) > 0) plots_hotness[]
```