---
title: "DaTa Preparation NBS"    
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

I will use this project to pepare and extract insights from the raster
data for the polygons in the 5 prioritized areas.

# 1. Prepare Environemnt (Load necessary libraries)

```{r load libraries, include=FALSE}
packs <- c('terra', 'purrr', 'landscapemetrics', 'sf','dplyr',
           'here', 'gdalUtilities', 'jsonlite', 'devtools', 'stringr',
           'parallel', 'dplyr', 'tidyr', 'ggplot2', 'janitor', 'forcats', 'foreign')
sapply(packs, require, character.only = TRUE, quietly=TRUE)
rm(packs)
```

## 1.1 set class names and colors

```{r set symbols, include=FALSE}
#create vector with the country names
nam <- c("BRAZIL","MADAGASCAR","MEXICO","PERU","VIETNAM") 
lb <- data.frame(
  Class <- c("Cultivated", "Forest", "Shrubs and Grasses", "Sparse",
            "Mangroves", "Urban", "Bare", "Water"),
  color=c("#dcf064", "#00b809", "#d29000", "#ffebaf", "#009678", "#c31400", 
            "#fff5d7", "#0046c8"),
  value=c(1:8))
names(lb) <- c("Class", "color", "value")
```

## 1.3 Define align function
Because of course
```{r align list of rasters }


library(terra)

align_rasters <- function(raster_list, template, resample_method = "near") {
  lapply(raster_list, function(r) {
    if (!compareGeom(r, template, stopOnError = FALSE)) {
      message("Aligning raster: ", names(r))
      # Resample to align with the template
      resample(r, template, method = resample_method)
    } else {
      # Return raster as is if already aligned
      message("Raster already aligned: ", names(r))
      r
    }
  })
}


```


## 1.3 Define Data Preparation Function
```{r ES combination}
process_rasters <- function(nested_list, templates, resample_method = "bilinear") {
  lapply(seq_along(nested_list), function(i) {
    # Align all rasters in the sublist to the template
    aligned <- lapply(nested_list[[i]], function(r) {
      if (!compareGeom(r, templates[[i]], stopOnError = FALSE)) {
        message("Aligning raster: ", names(r))
        resample(r, templates[[i]], method = resample_method)
      } else {
        message("Raster already aligned: ", names(r))
        r
      }
    })
    
    # Stack all aligned rasters into a single multiband SpatRaster
    stacked <- do.call(c, aligned)
    
    # Merge the stacked raster with the corresponding template raster
    merged <- merge(stacked, templates[[i]])
    
    # Sum the layers of the merged raster
    app(merged, sum)
  })
}
```
## 1.4 Prepare Templates
```{r create tmp}
# add backgrounds/templates to align 
path_lc <- here('ESA_LC') 
# load reclassified land cover map
tf <- file.path(path_lc, list.files(path_lc, pattern= "rec"))
lc <- lapply(tf,rast)
# create rcl matrix
rcl <- matrix(c(
  0, Inf, 0   # Any value greater than 0 becomes 1
), ncol = 3, byrow = TRUE)

#create background pixels, subsitute all by 0
tmp <- lapply(lc, function(r){
  r <- classify(r[[1]], rcl)
})

rm(lc)
```
tfs <- 

## 1.4 Function to Normalize Rasters
```{r normalize}

normalize_raster <- function(r) {
  min_val <- min(values(r), na.rm = TRUE)
  max_val <- max(values(r), na.rm = TRUE)
  (r - min_val) / (max_val - min_val)
}

process_intervention_area <- function(raster_list) {
  # Normalize each raster in the list
  normalized_rasters <- lapply(raster_list, normalize_raster)
  # Combine the normalized rasters by summing them
  combined_raster <- do.call(sum, normalized_rasters)
  return(combined_raster)
}
```



# 2 Extract the data

## 2.1 List the Restoration Layers 
```{r select conservation data}
#Restoration:
tiffes <- file.path(here("cropped_raster_data" ), list.files(paste0(here("cropped_raster_data")),pattern= '.tif$'))
#Remove points, risk reduction baselin    
tiffes <- tiffes[c(1:5,51:55,71:75,81:90, 106:110)]
#tiffes <- tiffes[-c(6:10)]
#tiffes <- tiffes[c(6:10)]
```

## 2.2 Load and Sort the layers

Also run the normalizatio nand add them together
```{r apply mask AOI AP, eval=TRUE, include=FALSE}

# Step 1: Extract file names, product names, and country names
file_names <- basename(tiffes)  # Extract file names from paths

# Extract product names and country names
product_names <- sub("_[A-Z]+\\.tif$", "", file_names)  # Remove "_COUNTRY.tif" to get product name
country_names <- sub(".*_(.*?)\\.tif$", "\\1", file_names)  # Extract country name from file name

# Step 2: Create a dataframe to organize the information
file_info <- data.frame(
  FilePath = tiffes,
  Product = product_names,
  Country = country_names,
  stringsAsFactors = FALSE
)

# Step 3: Sort the dataframe by Country first, then by Product
file_info <- file_info %>%
  arrange(Country, Product)

# Step 4: Extract the sorted file paths
tiffes <- file_info$FilePath

baseES <- lapply(tiffes, rast)
baseES <- lapply(baseES, function(r){
  r <- normalize_raster(r)
})

# set the number of different products to split the list 
num <- length(unique(product_names))
# split again to have nested lists
baseES <- split(baseES, rep(1:length(nam), each = num))

# align, add background and sum all the values
baseES <- process_rasters(baseES,tmp, resample_method = "bilinear")
map(1:length(baseES), function(x) writeRaster(baseES[[x]], paste0(here("restoration"),'/', unique(country_names)[[x]], '_ES_sum.tif')))

```

## 2.3 Load Griscom Data and apply mask


```{r load Grsicoms data}
# Load Griscom Data
tiffes <- file.path(here("Restoration_Griscom"), list.files(paste0(here("Restoration_Griscom")),pattern= 'rec'))

pa_rc <- lapply(tiffes,rast)


baseES_m <- map2(
  baseES, 
  pa_rc, 
  function(rst, msk) {
    # Apply mask to each raster in the sublist using the corresponding country vector
    mask(rst, msk, maskvalue=0)
  }
)


map(1:length(baseES), function(x) writeRaster(baseES_m[[x]], paste0(here("restoration"),'/', unique(country_names)[[x]], '_ES_msk.tif'), overwrite=T))

```


# 3 Interventions

## 3.1 Brazil Espiritu Santo
```{r brazil intervention}
poly <- st_read(here("Interventions", "Brazil", "Espirito_Santo_Albers.shp"))
pol_wgs <- st_transform(poly, crs=crs(baseES_m[[1]]))
serv_bra <- baseES_m[[1]] %>% crop(pol_wgs) %>% mask(pol_wgs)
serv_bra <- project(serv_bra, crs(poly), method='cubic')
rest_sp <- crop(pa_rc[[1]], pol_wgs)
rest_sp <- mask(rest_sp,pol_wgs)
writeRaster(serv_bra, here("Interventions", "Brazil_intervention", "esp_santo_rest.tif"),overwrite=TRUE)
writeRaster(rest_sp, here("Interventions", "Brazil_intervention", "SS_griscom.tif"), overwrite=TRUE)

```

## 3.2 Madagascar (Not Necessary To Mask for te Interrventions)
```{r yucatan intervention}
poly <- st_read(here("Interventions", "Mex_intervention","mex_intervention2.geojson"))

pol_wgs <- st_transform(poly, crs=crs(baseES_m[[3]]))
serv_mdg <- baseES_m[[3]] 
serv_mdg <- project(serv_mdg, crs(pol_wgs), method='cubic')
rest_mdg <- crop(pa_rc[[3]], pol_wgs)
rest_sp <- mask(rest_sp,pol_wgs)
writeRaster(serv_mdg, here("Interventions", "Mdg_intervention", "mdg_int_rest.tif"))
writeRaster(rest_sp, here("Interventions", "Mdg_intervention", "mdg_griscom.tif"))

```


## 3.3  Yucatan
```{r yucatan intervention}
poly <- st_read(here("Interventions", "Mex_intervention","mex_intervention2.geojson"))

pol_wgs <- st_transform(poly, crs=crs(baseES_m[[3]]))
serv_mex <- baseES_m[[3]] %>% crop(pol_wgs) %>% mask(pol_wgs)
serv_mex <- project(serv_mex, crs(poly_wgs), method='cubic')
rest_mex <- crop(pa_rc[[3]], pol_wgs)
rest_mex <- mask(rest_mex,pol_wgs)
writeRaster(serv_mex, here("Interventions", "Mex_intervention", "yucatan_int_rest.tif"))
writeRaster(rest_mex, here("Interventions", "Mex_intervention", "Mex_griscom.tif"), overwrite=TRUE)

```

## 3.3  Peru
```{r yucatan intervention}
pol <- st_read(here("Interventions","Peru","Intervenciones", "commondata","ganaderia","GANADERIA.shp"))
pol <- st_make_valid(pol)
st_write(pol, here("Interventions", "ganaderia.shp"))
#get msk non zero values from the EC mask
msk_p <- classify(baseES[[4]],rcl)
pol_wgs <- st_transform(pol, crs=crs(baseES_m[[4]]))
serv_per_gan <- baseES[[4]] %>% crop(pol_wgs) %>% mask(pol_wgs)

rest_per <- project(pa_rc[[4]], serv_per)
rest_per <- classify(rest_per,rcl)
serv_per_gan <- project(serv_per_gan, serv_per, method='bilinear')
msk_g <- classify(serv_per_gan,rcl) 
msk_g <- subst(msk_g, from=0, to =1)
per1 <- project(per1,serv_per)

mask_all <- merge(msk_g,rest_per)
mask_all <- classify(mask_all,rcl)
maks <- list(rest_per, msk_g, mask_all)

mask_all_wgs <- project(mask_all, baseES[[4]])

mks_peru <- mask(baseES[[4]],mask_all_wgs, maskvalue=0)


areas <- lapply(maks,lsm_c_ca)
save(areas, file=here("Interventions", "Peru_intervention","Proportions.RData"))

writeRaster(mask_all, here("Interventions", "Peru_intervention","MDDmask_final.tif"), overwrite=TRUE)
writeRaster(mks_peru, here("Interventions", "Peru_intervention","MDDEES_final.tif"), overwrite=TRUE)


r <- project(rest_per, serv_per)

serv_per <- project(serv_per, crs(pol), method='cubic')
writeRaster(serv_mex, here("Interventions", "PER_intervention", "per_int_rest.tif"))
writeRaster(rest_mex, here("Interventions", "PER_intervention", "per_griscom.tif"), overwrite=TRUE)

```

```{r define vars}

bra <- rest_sp
mdg <- pa_rc[[2]]
mex <- rest_mex
per1 <- rest_per # Griscom Peru]
per2 <- msk_p #all modeled rest non zero values maybe merge
per3 <- max(per1,per2)
VN
```



I did not do Peru or Vietnam Because 1, in peru i will have to use the whole area or maybe add ganaderia. ?The resdt is included 
## 3.4 Peru

```{r Pero intervention}
# create rcl matrix
rcl <- matrix(c(
  0, Inf, 1   # Any value greater than 0 becomes 1
), ncol = 3, byrow = TRUE)

#check Area Grsicom
serv_tot <- classify(baseES[[4]], rcl)

plot(serv_toppoly <- st_read(here("Interventions", "Mex_intervention","mex_intervention2.geojson"))


pol_wgs <- st_transform(poly, crs=crs(baseES_m[[3]]))
serv_mex <- baseES_m[[3]] %>% crop(pol_wgs) %>% mask(pol_wgs)
serv_mex <- project(serv_mex, crs(poly_wgs), method='cubic')
rest_sp <- crop(pa_rc[[3]], pol_wgs)
rest_sp <- mask(rest_sp,pol_wgs)
writeRaster(serv_mex, here("Interventions", "Mex_intervention", "yucatan_int_rest.tif"))
writeRaster(rest_sp, here("Interventions", "Mex_intervention", "Mex_griscom.tif"))

## 3.3. Peru 
# 2 Crop the Data for the target areas

```{r load target area polygons}
poly <- st_read(here("vector", "areas_nbs2.geojson")) 
#temp: reproject if needed
#poly <- st_transform(poly, crs = crs(tmp))
#Split the vector into a list of individual polygons
#create vector with names
poly <- poly%>%split(.$COUNTR)
```

## 3.5 Peru
## 2.1. Restoration data
```{r select restroration data to process}

inpath <- here(("Downloaded_data_ES"))

tiffes <- file.path(here(inpath), list.files(paste0(here(inpath)),pattern= '.tif$'))
#Remove points, risk reduction baseline
tiffes <- tiffes[c(3,23,25,26,30)]
print(tiffes)

```


## 2.2 Acces data
```{r select accessibility data to process}

#Restoration:
tiffes <- file.path(here(inpath), list.files(paste0(here(inpath)),pattern= 'nature_access.*\\.tif$'))
tiffes <- tiffes[2]
# Access restoration
acces_rest <- rast(tiffes)
basename <- basename(tiffes)
clean_filename <- function(filename){
  sub("_md5.*", "", filename)
}
filename <- clean_filename(basename)

#acces_rest <- project(acces_rest, acc2, method='bilinear')
acces_rest <- map(1:length(poly), function(x) crop(acces_rest, poly[[x]]))
acces_rest <- map(1:length(acces_rest), function(x) mask(acces_rest[[x]], poly[[x]]))
map(1:length(acces_rest), function(x) writeRaster(acces_rest[[x]], paste0(here('cropped_raster_data'),"/",filename, '_', nam[x], '.tif'), overwrite=TRUE))


# pending access restoration. No need to export retrospected, just get everything done here. 

```

## 2.3 Acces data Restoration
```{r select accessibility restoration data to process, eval=FALSE, include=FALSE}

#Restoration:
tiffes <- file.path(here(inpath), list.files(paste0(here(inpath)),pattern= 'nature_access.*\\.tif$'))
tiffes <- tiffes[c(4)]
# Access restoration
basename <- basename(tiffes)
clean_filename <- function(filename){
  sub("_md5*", "", filename)
  sub(".tif.*", "", filename)
}
filename <- clean_filename(basename)

acces_rest <- rast(tiffes) 
acces_rest <- mapply(function(raster1, raster2) {
  project(raster1, raster2, method = "bilinear")  # Adjust method if needed
}, acces_rest, tmp, SIMPLIFY = FALSE)
acces_rest <- map(1:length(poly), function(x) crop(acces_rest[[1]], poly[[x]]))
acces_rest  <- map(1:length(acces_rest), function(x) mask(acces_rest[[x]], poly[[x]]))
map(1:length(acces_rest), function(x) writeRaster(acces_rest[[x]], paste0(here('cropped_raster_data'),"/",filename, '_', nam[x], '.tif'), overwrite=TRUE))


# pending access restoration. No need to export repporjected, just get everything done here. 

```


## 2.4 Coastal data Protection Conservation
```{r crop coastal protection data, eval=FALSE, include=FALSE}

#Restoration:
tiffes <- file.path(here("downloaded_data_ES"), list.files(paste0(here(inpath)),pattern= 'ESAmodVCFv2_cv_habitat_value_md5.*\\.tif$'))
# Access restoration
basename <- basename(tiffes)
acces_rest <- rast("~/Documents/WWF_nbs_op/Downloaded_data_ES/ESAmodVCFv2_cv_habitat_value_md5_c01e9b17aee323ead79573d66fa4020d.tif")

clean_filename <- function(filename){
  sub("_md5.*", "", filename)
  sub(".tif.*", "", filename)
}
filename <- clean_filename(basename)

acces_rest <- map(1:length(poly), function(x) crop(acces_rest, poly[[x]]))
acces_rest2 <- map(1:length(acces_rest), function(x) mask(acces_rest[[x]], poly[[x]]))
map(1:length(acces_rest), function(x) writeRaster(acces_rest[[x]], paste0(here('cropped_raster_data'),"/",filename, '_', nam[x], '.tif'), overwrite=TRUE))


# pending access restoration. No need to export repporjected, just get everything done here. 

```



This runs the same process in batch. Memory was optimized to not load all the data at thesame time, but process and write each raster individually before going to the next one.  


```{r iterate cropMask, eval=FALSE, include=FALSE}
clean_filename <- function(filename){
  sub("_md5.*", "", filename)
  sub(".tif.*", "", filename)
}
# Loop through each geotiff file
for (i in seq_along(tiffes)) {
  # Get the current tiff file
  tiff_file <- tiffes[i]
  
  # Load the current geotiff (don't load all at once)
  access <- rast(tiff_file)
  
  # Apply crop and mask operations to each polygon
  cropped <- map(1:length(poly), function(x) crop(access, poly[[x]]))
  masked <- map(1:length(cropped), function(x) mask(cropped[[x]], poly[[x]]))
  
  # Extract the base filename before 'md5'
  base_filename <- clean_filename(basename(tiff_file))
  
  # Write the cropped and masked rasters to files
  map(1:length(masked), function(x) {
    writeRaster(masked[[x]], paste0(here('cropped_raster_data'),"/", base_filename, "_", nam[x], ".tif"), overwrite = TRUE)
  })
  # Optionally, clear the variable to free up memory
  rm(access, cropped, masked)
  gc()  # Garbage collection to clear memory
}
```

## 2.3. Reproj and CropMask Accesibility Raster

The accessibility data was handled separately because it is  in another crs because of course, it is a
distance raster and an equal distance projection is required.

```{r reproj access, eval=FALSE, include=FALSE}

# Access 60m 2019
acces_rest <- rast(here(inpath, 'global_people_access_population_2019_60.0m_md5_d264d371bd0d0a750b002a673abbb383.tif'))
acces_rest <- project(acces_rest, acc2, method='bilinear')
acces_rest <- map(1:length(poly), function(x) crop(acces_rest, poly[[x]]))
acces_rest <- map(1:length(acces_rest), function(x) mask(acces_rest[[x]], poly[[x]]))


map(1:length(access_rest), function(x) writeRaster(acces_rest[[x]], paste0('/Volumes/Jero_HDD/WWF/NBS_OPS/raster/', 'global_people_access_population_2019_60.0m_', nam[x], '.tif'), overwrite=TRUE))
```


# 3 Explore Datasets

For summarizing geotiff raster files and comparing values across
protected areas or different land covers for an initial assessment,
especially for an audience without geospatial expertise, here are some
accessible metrics : 1. Basic Statistical Summaries:

## 3.1 Summarize Land covers

Extract the shares for the different land covers for the reclassified
ESA land cover maps for the 5 study areas for 1992 and 2020.

```{r basic summaries LC, eval=FALSE, include=FALSE}

path_lc <- here('ESA_LC') 
# load reclassified land cover map
tf <- file.path(path_lc, list.files(path_lc, pattern= "rec"))
lc <- lapply(tf,rast)


lsmet <- lapply(lc, function(r){
  mets <- lsm_c_ca(r)
})
lsmet <- lapply(seq_along(lsmet), function(i) {
  lsmet[[i]]$AREA <- nam[i]
  return(lsmet[[i]])
})
lsmet <- do.call(rbind, lsmet)
lsmet <- left_join(lsmet, lb, by= join_by(class == value))
lsmet <- lsmet %>% select(!c(level,id,metric))
lsmet <-  lsmet %>% group_by(AREA, layer) %>%
  mutate(percentage = value / sum(value) * 100) %>%  # Calculate percentage
  ungroup() %>% mutate(year=case_when(
    layer == 1 ~ 1992,
    layer == 2 ~ 2020
  ))

save(lsmet, file=here("output_data", 'shares_lc2.RData'))

```

## 3.2. ESA Land cover distribtion 1992-2020

-   Land Cover ESA
-   Area Proportion for both years (1992,2020)

## PlotLand Cover Shares
```{r land covers, echo=FALSE, fig.width=9, fig.height=7}
load(here("output_data", 'shares_lc2.RData'))
t <- ggplot(lsmet, aes(x = Class, y = percentage, fill = color)) +
  geom_bar(stat = "identity") +  # Use identity since we are plotting pre-calculated values
  scale_fill_identity() +  # Use the exact colors provided in the `color` column
  facet_grid(year ~ AREA) +  # Create a grid with rows for `AREA` and columns for `layer`
  labs(title = "Class Distribution by Area and Year",
       x = "Class",
       y = "Percentage") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +  # Rotate x-axis labels for readability
  geom_text(aes(label = scales::percent(percentage / 100, accuracy = 0.01)),  # Convert to percentage format
            vjust = -0.5,  # Position labels slightly above the bars
            size = 2.5)  # Adjust label size as needed

print(t)
```


# 4. Extract Summaries Inside/Outside AOIs

## 4.1 Conservation 

Assses ecosystem service output inside and outside protected areas. UICN cat National park Ia , Ib, II, III,
IV , V, VI and non-UICN

Barplots/distributions outside inside protected areas/land cover classes
as a way to see if protected areas are abnormally represented

## 4.2 Calculate Proportions

Here, we take the rasters with the target data (provided ecosystem service on time t (baseline) , recover (restoration potential)  and wthe Areas of Interest (AOI) to run comparisons between inside/outside to check the relative productivity of these areas. This is key for decision making/show donors/officials for example)

It is possible to do this systematically,  gain on replicabillity and to be able to make adjustments easily . We load the polygons/rasters, cropMask to get what's inside and the inverse.



### 4.2.1 Load the Polygons with the Areas of interest

```{r load protected  areas}
#load protected areas
pa = st_read(here("vector", "wdpa_nbs.shp"))
pa <- pa%>%split(.$COUNTR)

```

### 4.2.2 Extract the shares shares
```{r extract shares}
# add backgrounds/templates to align 
path_lc <- here('ESA_LC') 
# load reclassified land cover map
tf <- file.path(path_lc, list.files(path_lc, pattern= "rec"))
lc <- lapply(tf,rast)
# create rcl matrix
rcl <- matrix(c(
  0, Inf, 0   # Any value greater than 0 becomes 1
), ncol = 3, byrow = TRUE)

#create background pixels, subsitute all by 0
bkg <- lapply(lc, function(r){
  r <- classify(r[[1]], rcl)
})# Load the rasters with the target data.
# 
rm(lc) #This is extracts thebinary masks inside outside. Needed whrn input data is vector 
bkg_1 <- map2(bkg, pa, mask)
bkg_1 <- lapply(bkg_1, function(r){
  r <- subst(r, from = 0, to =1)
})
msk_ap <- map2(bkg_1,bkg, merge)
#Extrct feequencies (0 outside, 1 inside)
freq_ap <- lapply(msk_ap, function(r){
  f <- freq(r)
})
# get vector with the country names
country_names <- unique(country_names)
# Add the country names to tjhe extracted dataframes
freq_ap <- lapply(seq_along(freq_ap), function(i) {
  freq_ap[[i]]$country <- country_names[i]  # Add country column
  freq_ap[[i]]  # Return the modified dataframe
})
freq_ap <- as_tibble(do.call(rbind, freq_ap))

freq_ap <- freq_ap %>%
  group_by(country) %>%
  mutate(share = (count / sum(count)) * 100) %>%
  ungroup()

freq_ap <- freq_ap %>% filter(layer == 1)  %>%  filter(value == 1)
freq_ap <- freq_ap %>% select(!c(1,2,3))


```



### 4.1.2 Mask the Service Rasters by the AOI (1)
```{r apply mask AOI AP, eval=TRUE, include=FALSE}

# Step 1: Extract file names, product names, and country names
file_names <- basename(tiffes)  # Extract file names from paths

# Extract product names and country names
product_names <- sub("_[A-Z]+\\.tif$", "", file_names)  # Remove "_COUNTRY.tif" to get product name
country_names <- sub(".*_(.*?)\\.tif$", "\\1", file_names)  # Extract country name from file name

# Step 2: Create a dataframe to organize the information
file_info <- data.frame(
  FilePath = tiffes,
  Product = product_names,
  Country = country_names,
  stringsAsFactors = FALSE
)

# Step 3: Sort the dataframe by Country first, then by Product
file_info <- file_info %>%
  arrange(Country, Product)

# Step 4: Extract the sorted file paths
tiffes <- file_info$FilePath

baseES <- lapply(tiffes, rast)
# set the number of different products to split the list 
num <- length(unique(product_names))
# Group the list of ES as nested list, one for eah country
# split again to have nested lists
baseES <- split(baseES, rep(1:length(nam), each = num))


# Step 2: Apply mask for each group of rasters using the corresponding country vector
baseES_ap <- map2(
  baseES,   # The list of raster sublists
  pa,     # The list of country polygons
  function(rasters, vector) {
    # Apply mask to each raster in the sublist using the corresponding country vector
    lapply(rasters, function(raster) mask(raster, vector))
  }
)
# "api" stands for "outside". I don´t know why I gave tho name, it makes no sense 
baseES_api <- map2(
  baseES,   # The list of raster sublists
  pa,     # The list of country polygons
  function(rasters, vector) {
    # Apply mask to each raster in the sublist using the corresponding country vector
    lapply(rasters, function(raster) mask(raster, vector, inverse=TRUE))
  }
)


# Step 3: Flatten back the list to get a single list of masked rasters
baseES_ap <- unlist(baseES_ap, recursive = FALSE)
baseES_api <- unlist(baseES_api, recursive = FALSE)

```



################################
## 4.2 Calculate Shares (1)


```{r extract total  ES provision, eval=FALSE, include=FALSE}

#calculate ES output
# Define a function to compute the total sum for each SpatRaster
compute_sum <- function(raster) {
  total_sum <- global(raster, fun = "sum", na.rm = TRUE)[,1]  # Use global() to get the sum
  return(total_sum)
}
totSE_ap <-  lapply(baseES_ap, compute_sum)
totSE_api <-  lapply(baseES_api, compute_sum)

totSE_ap <- as_tibble(unlist(totSE_ap))
totSE_api <- as_tibble(unlist(totSE_api))
totSE <- cbind(file_info[c(2,3)], totSE_api, totSE_ap)


# dropoutside 

#rearrange columns 
names(totSE) <- c("Service", "country","Outside", "Inside")
totSE <- left_join(totSE,freq_ap)

# I need to add and then get the total!!!!!!!! 
totSE <- totSE %>% mutate(WholeArea= Outside+Inside)
# The freaking error is here!!!!!
totSE <- totSE %>% mutate(share_service= (Inside/WholeArea)*100)


names(totSE) <- c("Service", "Country","Outside", "Inside", "Share_AOI", "WholeArea", "Share_service")

#
totSE <- totSE %>%
  # Create a new column with simplified service names
  mutate(Service = case_when(
    Service == "global_n_retention_esamod2_compressed" ~ "Nitrogen Retention",
    Service == "global_sed_retention_esamod2_compressed" ~ "Sediment Retention",
    Service == "pollination_ppl_fed_on_ag_10s_esa2020" ~ "Pollination",
    Service == "ESAmodVCFv2_cv_habitat_value_md5_c01e9b17aee323ead79573d66fa4020d" ~ "Coastal Protection",
    TRUE ~ Service  # Keep other names unchanged, if any
  )) %>%
  # Select relevant columns and reshape the data into long format
  select(Service, Share_service, Share_AOI, Country) %>%
  pivot_longer(cols = c(Share_service, Share_AOI),
               names_to = "Metric",
               values_to = "Value")


totSE_l <- totSE %>%
  filter(!(Metric == "Share_AOI" & duplicated(paste(Country, Metric))))

#write.csv(totSE, file=here('output_data', "service_prov_PA.csv"))

totSE_l <- totSE_l %>% mutate(Service=case_when(
  Metric== "Share_AOI" ~ "Area",
  TRUE ~ Service 
))

```


## 4.3 Calculate Shares (2) 
```{r calculate Access Conservation}
tiffes <- file.path(here("cropped_raster_data" ), list.files(paste0(here("cropped_raster_data")),pattern= '.tif$'))
#Remove points, risk reduction baseline
tiffes <- tiffes[c(56:60)]
```

### 4.4 Mask the Service Rasters by the AOI (2)
```{r apply mask AOI AP2, eval=TRUE, include=FALSE}

# Step 1: Extract file names, product names, and country names
file_names <- basename(tiffes)  # Extract file names from paths

# Extract product names and country names
product_names <- sub("_[A-Z]+\\.tif$", "", file_names)  # Remove "_COUNTRY.tif" to get product name
country_names <- sub(".*_(.*?)\\.tif$", "\\1", file_names)  # Extract country name from file name

# Step 2: Create a dataframe to organize the information
file_info <- data.frame(
  FilePath = tiffes,
  Product = product_names,
  Country = country_names,
  stringsAsFactors = FALSE
)

# Step 3: Sort the dataframe by Country first, then by Product
file_info <- file_info %>%
  arrange(Country, Product)

# Step 4: Extract the sorted file paths
tiffes <- file_info$FilePath

baseES <- lapply(tiffes, rast)
# set the number of different products to split the list 
num <- length(unique(product_names))
# Group the list of ES as nested list, one for eah country
# split again to have nested lists
baseES <- split(baseES, rep(1:length(nam), each = num))


# Step 2: Apply mask for each group of rasters using the corresponding country vector
baseES_ap <- map2(
  baseES,   # The list of raster sublists
  pa,     # The list of country polygons
  function(rasters, vector) {
    # Apply mask to each raster in the sublist using the corresponding country vector
    lapply(rasters, function(raster) mask(raster, vector))
  }
)
# "api" stands for "outside". I don´t know why I gave tho name, it makes no sense 
baseES_api <- map2(
  baseES,   # The list of raster sublists
  pa,     # The list of country polygons
  function(rasters, vector) {
    # Apply mask to each raster in the sublist using the corresponding country vector
    lapply(rasters, function(raster) mask(raster, vector, inverse=TRUE))
  }
)


# Step 3: Flatten back the list to get a single list of masked rasters
baseES_ap <- unlist(baseES_ap, recursive = FALSE)
baseES_api <- unlist(baseES_api, recursive = FALSE)

```


### 4.4 Calculate Shares (2)


```{r extract total  ES provision}

#calculate ES output
# Define a function to compute the total sum for each SpatRaster
compute_sum <- function(raster) {
  total_sum <- global(raster, fun = "sum", na.rm = TRUE)[,1]  # Use global() to get the sum
  return(total_sum)
}
totSE_ap <-  lapply(baseES_ap, compute_sum)
totSE_api <-  lapply(baseES_api, compute_sum)

totSE_ap <- as_tibble(unlist(totSE_ap))
totSE_api <- as_tibble(unlist(totSE_api))
totSE <- cbind(file_info[c(2,3)], totSE_api, totSE_ap)


# dropoutside 

#rearrange columns 
names(totSE) <- c("Service", "country","Outside", "Inside")
totSE <- left_join(totSE,freq_ap)

# I need to add and then get the total!!!!!!!! 
totSE <- totSE %>% mutate(WholeArea= Outside+Inside)
# The freaking error is here!!!!!
totSE <- totSE %>% mutate(share_service= (Inside/WholeArea)*100)


names(totSE) <- c("Service", "Country","Outside", "Inside", "Share_AOI", "WholeArea", "Share_service")

#
totSE <- totSE %>%
  # Create a new column with simplified service names
  mutate(Service = case_when(
    Service == "nature_access_lspop2019_esa2020modVCFhab" ~ "Access",
    TRUE ~ Service  # Keep other names unchanged, if any
  )) %>%
  # Select relevant columns and reshape the data into long format
  select(Service, Share_service, Share_AOI, Country) %>%
  pivot_longer(cols = c(Share_service, Share_AOI),
               names_to = "Metric",
               values_to = "Value")


totSE_l2 <- totSE %>%
  filter(!(Metric == "Share_AOI" & duplicated(paste(Country, Metric))))
totSE_l <- rbind(totSE_l, totSE_l2)

totSE_l <- totSE_l %>% mutate(Service=case_when(
  Metric== "Share_AOI" ~ "Area",
  TRUE ~ Service 
))
write.csv(totSE_l, file=here('output_data', "service_prov_PA.csv"))
```
## 4.4  Plot the shares

```{r plot shares AP, fig.width=9, fig.height=6.5}

totSE_l <- as_tibble(read.csv(here('output_data', 'service_prov_PA.csv')))
totSE_l$Service <- factor(totSE_l$Service, levels = c('Area', 'Nitrogen Retention', 'Sediment Retention', "Pollination", "Access", "Coastal Protection"))

# Step 2: Create the bar plot
p <- ggplot(totSE_l, aes(x = Metric, y = Value, fill = Service)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) + 
  aes(width = ifelse(Metric == "Share_AOI", 0.5, 0.9)) + 
  scale_y_continuous(limits = c(0, 100)) +
  facet_wrap(~ Country, ncol = 5) +  # Create 5 columns (one for each country)
  scale_fill_manual(values = c(
    "Area" = "gray50",         # Medium gray for "Share Area"
    "Nitrogen Retention" = "#2c944c",  # Shade of green
    "Sediment Retention" = "#08306b",  # Shade of blue
    "Pollination" = "#dd1c77",          # Shade of purple
    "Access" = "#A57C00",
    "Coastal Protection" = "#7a0177"
  )) +
  labs(
    title = "Share of the protected areas and of the total ES provided inside them",
    x = "Metric",
    y = "%",
    fill = "Service"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    legend.position = "bottom",  # Place the legend at the bottom
    legend.box = "horizontal",  # Arrange legend items horizontally
    legend.spacing.x = unit(0.1, 'cm'),  # Add horizontal spacing between legend items
    legend.title = element_text(size = 12, face = "bold"),  # Adjust legend title size
    legend.text = element_text(size = 10),  # Adjust legend text size
    legend.box.just = "center"  # Center the legend items
  )


print(p)
```

########################################################################## 
Pendingthis same thing with the beneficiaries and other data wit other crs/spatial resolution

########################################################################## 

-   Coastal Risk Reduction
-   Restoration
    -   Nitrogen Export
    -   nature access restoration
    -   pollination restoration
    -   sediment export
    -   Conservation?

## 4.5 Produce Density Plots (Protected Areas) 


```{r set list of names, include=FALSE}

# Step 1: Extract file names, product names, and country names
file_names <- basename(tiffes)  # Extract file names from paths

# Extract product names and country names
product_names <- sub("_[A-Z]+\\.tif$", "", file_names)  # Remove "_COUNTRY.tif" to get product name
country_names <- sub(".*_(.*?)\\.tif$", "\\1", file_names)  # Extract country name from file name

# Step 2: Create a dataframe to organize the information
file_info <- data.frame(
  Product = product_names,
  Country = country_names,
  stringsAsFactors = FALSE
)

# Step 3: Sort the dataframe by Country first, then by Product
file_info <- file_info %>%
  arrange(Country, Product)


prod_nam <- unique(product_names)

num <- length(unique(product_names))
serv <- c("Coastal_Protection", "Nitrogen_Retention", "Sediment_Retention", "Pollination", "Access") 
#serv <- c("Access") 

new_nam <- cbind(prod_nam, serv)
```


```{r extract values conservation, eval=FALSE, include=FALSE}

baseES_v <- lapply(baseES_api, function(r){
  t <- as_tibble(values(r, na.rm=TRUE))
})

baseESap_v <- lapply(baseES_ap, function(r){
  t <- as_tibble(values(r, na.rm=TRUE))
})

# Add two new columns to each data frame in the list
baseES_v  <- Map(function(df, row) {
  names(df)[1] <- "Value"
  df$Service <- row[[1]]  
  df$Country <- row[[2]]  
  return(df)
}, baseES_v , split(file_info, seq(nrow(file_info))))

baseESap_v  <- Map(function(df, row) {
  names(df)[1] <- "Value"
  df$Service <- row[[1]]  
  df$Country <- row[[2]]  
  return(df)
}, baseESap_v , split(file_info, seq(nrow(file_info))))

           
##################### do not run inside the Rmd!!!!!!!!
baseES_v <- do.call(rbind, baseES_v)
baseESap_v <- do.call(rbind, baseESap_v)
##################### do not run inside the Rmd!!!!!!!!
baseES_v <- baseES_v%>% mutate(AOI = "0")
baseESap_v <- baseESap_v%>% mutate(AOI = "1")

baseES_v <- rbind(baseES_v, baseESap_v)
rm(baseESap_v)

baseES_v <- as_tibble(baseES_v)
baseES_v <- baseES_v %>%
  # Create a new column with simplified service names
  mutate(Service = case_when(
    Service == new_nam[1,][1] ~ new_nam[1,][2],
    Service == new_nam[2,][1] ~ new_nam[2,][2],
    Service == new_nam[3,][1] ~ new_nam[3,][2],
    Service == new_nam[4,][1] ~ new_nam[4,][2],
    Service == new_nam[5,][1] ~ new_nam[5,][2],
    TRUE ~ Service  # Keep other names unchanged, if any
  ))
# 
 write.csv(baseES_v, here('output_data', "distributionServ_pa.csv"))

```


```{r calculate percentiles 2 and 98, eval=FALSE, include=FALSE}

#baseES_vf <- read.csv(here('data', "distributionServ_pa.csv"))
baseES_v <- as_tibble(baseES_v)
# Calculate the 98th percentile for each combination of Service and Country
  baseES_v$Service <- factor(baseES_v$Service, levels = c(
    "Nitrogen_Retention", "Sediment_Retention", "Pollination", "Access","Coastal_Protection"))

# Calculate both the 2nd and 98th percentiles
percentiles <- baseES_v %>%
  group_by(Service, Country) %>%
  summarize(
    lower_threshold = quantile(Value, 0.02),
    upper_threshold = quantile(Value, 0.98)
  ) %>%
  ungroup()

# Join the thresholds back to the original data and filter values within the 2nd and 98th percentiles
baseES_filtered <- baseES_v %>%
  left_join(percentiles, by = c("Service", "Country")) %>%
  filter(Value >= lower_threshold & Value <= upper_threshold)
```

# Density Plots


### 4.3.1 BRazil

This is calcualted for the percentile 95 to remove extreme values
```{r density plots pa 1, echo=FALSE, fig.height=6, fig.width=9}

C <- 1

df <- baseES_filtered %>% filter(Country==nam[C])
t <- ggplot(df, aes(x = Value, fill = AOI)) +
  geom_density(alpha = 0.4) +  # Use transparency with alpha to see overlapping densities
  facet_wrap(~Service, scales = "free") +
  # Set both x and y axes to be free for each facet
  scale_fill_manual(values = c("blue", "orange")) +  # Set colors for the `restoration` variable
  labs(
    title = paste("Density Plot of Services Inside and Outside Protected Areas -", nam[C]),
    x = "Value (Log Scale)",
    y = "Density",
    fill = "Protected Area"
  ) +
  scale_x_continuous(trans = 'log10') +  # Apply log10 transformation to x-axis
  theme(
    strip.text = element_text(size = 12),  # Increase facet labels' font size
    legend.position = "bottom",  # Place the legend at the bottom
    legend.box = "horizontal"  # Arrange legend items horizontally
  )
print(t)
```

### 4.3.2 Madagascar
```{r density plots pa 2, echo=FALSE, fig.height=6, fig.width=9}

C <- 2

df <- baseES_filtered %>% filter(Country==nam[C])
t <- ggplot(df, aes(x = Value, fill = AOI)) +
  geom_density(alpha = 0.4) +  # Use transparency with alpha to see overlapping densities
  facet_wrap(~Service, scales = "free") +
  # Set both x and y axes to be free for each facet
  scale_fill_manual(values = c("blue", "orange")) +  # Set colors for the `restoration` variable
  labs(
    title = paste("Density Plot of Services Inside and Outside Protected Areas -", nam[C]),
    x = "Value (Log Scale)",
    y = "Density",
    fill = "Protected Area"
  ) +
  scale_x_continuous(trans = 'log10') +  # Apply log10 transformation to x-axis
  theme(
    strip.text = element_text(size = 12),  # Increase facet labels' font size
    legend.position = "bottom",  # Place the legend at the bottom
    legend.box = "horizontal"  # Arrange legend items horizontally
  )
print(t)
```
### 4.3.2 Mexico
```{r density plots pa 3, echo=FALSE, fig.height=6, fig.width=9}

C <- 3

df <- baseES_filtered %>% filter(Country==nam[C])
t <- ggplot(df, aes(x = Value, fill = AOI)) +
  geom_density(alpha = 0.4) +  # Use transparency with alpha to see overlapping densities
  facet_wrap(~Service, scales = "free") +
  # Set both x and y axes to be free for each facet
  scale_fill_manual(values = c("blue", "orange")) +  # Set colors for the `restoration` variable
  labs(
    title = paste("Density Plot of Services Inside and Outside Protected Areas -", nam[C]),
    x = "Value (Log Scale)",
    y = "Density",
    fill = "Protected Area"
  ) +
  scale_x_continuous(trans = 'log10') +  # Apply log10 transformation to x-axis
  theme(
    strip.text = element_text(size = 12),  # Increase facet labels' font size
    legend.position = "bottom",  # Place the legend at the bottom
    legend.box = "horizontal"  # Arrange legend items horizontally
  )
print(t)
```
### 4.3.4 Peru
```{r density plots pa 4, echo=FALSE, fig.height=6, fig.width=9}

C <- 4
df <- baseES_filtered %>% filter(Country==nam[C])
t <- ggplot(df, aes(x = Value, fill = AOI)) +
  geom_density(alpha = 0.4) +  # Use transparency with alpha to see overlapping densities
  facet_wrap(~Service, scales = "free") +
  # Set both x and y axes to be free for each facet
  scale_fill_manual(values = c("blue", "orange")) +  # Set colors for the `restoration` variable
  labs(
    title = paste("Density Plot of Services Inside and Outside Protected Areas -", nam[C]),
    x = "Value (Log Scale)",
    y = "Density",
    fill = "Protected Area"
  ) +
  scale_x_continuous(trans = 'log10') +  # Apply log10 transformation to x-axis
  theme(
    strip.text = element_text(size = 12),  # Increase facet labels' font size
    legend.position = "bottom",  # Place the legend at the bottom
    legend.box = "horizontal"  # Arrange legend items horizontally
  )
print(t)
```
### 4.3.5 Vietnam
```{r density plots pa 1, echo=FALSE, fig.height=6, fig.width=9}

C <- 5

df <- baseES_filtered %>% filter(Country==nam[C])
t <- ggplot(df, aes(x = Value, fill = AOI)) +
  geom_density(alpha = 0.4) +  # Use transparency with alpha to see overlapping densities
  facet_wrap(~Service, scales = "free") +
  # Set both x and y axes to be free for each facet
  scale_fill_manual(values = c("blue", "orange")) +  # Set colors for the `restoration` variable
  labs(
    title = paste("Density Plot of Services Inside and Outside Protected Areas -", nam[C]),
    x = "Value (Log Scale)",
    y = "Density",
    fill = "Protected Area"
  ) +
  scale_x_continuous(trans = 'log10') +  # Apply log10 transformation to x-axis
  theme(
    strip.text = element_text(size = 12),  # Increase facet labels' font size
    legend.position = "bottom",  # Place the legend at the bottom
    legend.box = "horizontal"  # Arrange legend items horizontally
  )
print(t)
```

# 5 Restoration Griscom

### 5.1. Fix Griscom's Data

```{r fix Griscom data, eval=FALSE, include=FALSE}
# Load Griscome Rest data
tiffes <- file.path(here("Restoration_Griscom"), list.files(paste0(here("Restoration_Griscom")),pattern= '.tif$'))
tiffes <- tiffes[-5]

pa <- lapply(tiffes,rast)
#Read the metadata because of course me
meta <- lapply(pa, cats)

rcl_matrices <- lapply(meta, function(x) x[[1]]) 

# rEclassify the rasters. 
pa_rc <- mapply(function(raster, rcl) {
  classify(raster, as.matrix(rcl[, c("Value", "Restore")]))
}, pa, rcl_matrices, SIMPLIFY = FALSE)
```




# Extracxt shares Restoration areas/total


## Fix file info

```{r clear fileinfo, eval=FALSE, include=FALSE}

file_info <-file_info %>% filter(Product!="nature_access_diff_Sc3v1_PNVnoag-esa2020")
```

```{r extract total  ES provision}

#calculate ES output
# Define a function to compute the total sum for each SpatRaster
compute_sum <- function(raster) {
  total_sum <- global(raster, fun = "sum", na.rm = TRUE)[,1]  # Use global() to get the sum
  return(total_sum)
}
totSE_ap <-  lapply(baseES_ap, compute_sum)
totSE_api <-  lapply(baseES_api, compute_sum)

totSE_ap <- as_tibble(unlist(totSE_ap))
totSE_api <- as_tibble(unlist(totSE_api))
totSE <- cbind(file_info[c(2,3)], totSE_api, totSE_ap)


# dropoutside 

#rearrange columns 
names(totSE) <- c("Service", "country","Outside", "Inside")
totSE <- left_join(totSE,freq_ap)

# I need to add and then get the total!!!!!!!! 
totSE <- totSE %>% mutate(WholeArea= Outside+Inside)
# The freaking error is here!!!!!
totSE <- totSE %>% mutate(share_service= (Inside/WholeArea)*100)


names(totSE) <- c("Service", "Country","Outside", "Inside", "Share_AOI", "WholeArea", "Share_service")

#
totSE <- totSE %>%
  # Create a new column with simplified service names
  mutate(Service = case_when(
    Service == "nature_access_diff_Sc3v1_PNVnoag-esa2020" ~ "Access",
    Service == "cv_habitat_value_Sc3v1-ESAmod2_v2_md5_64082b" ~ "Coastal Protection",
    Service == "nitrogen_ESAmod2-Sc3v1_md5_024a36" ~ "Nitrogen",
    Service == "pollination_ppl_fed_on_ag_10s_Sc3v1_PNVnoag-esa2020_md5_405c88" ~ "Pollination (People fed)",
    Service == "pollination_ppl_fed_on_hab_Sc3v1_PNV_no_ag-ESA_md5_576790" ~ "Pollination - (Habitat)",
    Service == "sediment_ESAmod2-Sc3v1_md5_149078" ~ "Sediment",
    TRUE ~ Service  # Keep other names unchanged, if any
  )) %>%
  # Select relevant columns and reshape the data into long format
  select(Service, Share_service, Share_AOI, Country) %>%
  pivot_longer(cols = c(Share_service, Share_AOI),
               names_to = "Metric",
               values_to = "Value")


totSE <- totSE %>%
  filter(!(Metric == "Share_AOI" & duplicated(paste(Country, Metric))))
#totSE_l. <- rbind(totSE_l, totSE_l2) # why did i do this?b i know, to add access.

totSE <- totSE %>% mutate(Metric=case_when(
  Metric== "Share_AOI" ~ "Area",
  TRUE ~ Service 
))

totSE <- totSE %>% mutate(Service=case_when(
  Metric== "Area" ~ "Area",
  TRUE ~ Service 
))


totSE <- totSE %>%
  filter(!(Metric == "Area" & duplicated(paste(Country, Metric))))

 write.csv(totSE, file=here('output_data', "service_prov_Rest.csv"))
```


```{r plot shares AP, fig.width=9, fig.height=6}

#totSE<- as_tibble(read.csv(here('output_data', "service_prov_Rest.csv")))
totSE$Service <- factor(totSE$Service, levels = c('Area', 'Nitrogen', 'Sediment', "Pollination (People fed)", "Pollination - (Habitat)","Access", "Coastal Protection"))

totSE <- totSE %>% mutate(Metric = case_when(
  Metric != 'Area' ~ "Share Service",
  TRUE ~ Metric
))

# Step 2: Create the bar plot
p <- ggplot(totSE, aes(x = Metric, y = Value, fill = Service)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) + 
  aes(width = ifelse(Metric == "Area", 0.5, 0.9)) + 
  scale_y_continuous(limits = c(0, 100)) +
  facet_wrap(~ Country, ncol = 5) +  # Create 5 columns (one for each country)
  scale_fill_manual(values = c(
    "Area" = "gray50",         # Medium gray for "Share Area"
    "Nitrogen" = "#2c944c",  # Shade of green
    "Sediment" = "#08306b",  # Shade of blue
    "Pollination (People fed)" = "#d1b66",          # Shade of purple
    "Pollination - (Habitat)" = "#dd1c99",
    "Access" = "#A57C00",
    "Coastal Protection" = "#7a0177"
  )) +
  labs(
    title = "Share of Restoration Amount occurring Inside Restoration Priority Areas",
    x = "Service",
    y = "%",
    fill = "Service"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    legend.position = "bottom",  # Place the legend at the bottom
    legend.box = "horizontal",  # Arrange legend items horizontally
    legend.spacing.x = unit(0.1, 'cm'),  # Add horizontal spacing between legend items
    legend.title = element_text(size = 12, face = "bold"),  # Adjust legend title size
    legend.text = element_text(size = 10),  # Adjust legend text size
    legend.box.just = "center"  # Center the legend items
  )


print(p)
```
```{r set list of names, include=FALSE}

# Step 1: Extract file names, product names, and country names
file_names <- basename(tiffes)  # Extract file names from paths

# Extract product names and country names
product_names <- sub("_[A-Z]+\\.tif$", "", file_names)  # Remove "_COUNTRY.tif" to get product name
country_names <- sub(".*_(.*?)\\.tif$", "\\1", file_names)  # Extract country name from file name

# Step 2: Create a dataframe to organize the information
file_info <- data.frame(
  Product = product_names,
  Country = country_names,
  stringsAsFactors = FALSE
)

# Step 3: Sort the dataframe by Country first, then by Product
file_info <- file_info %>%
  arrange(Country, Product)


prod_nam <- unique(product_names)

num <- length(unique(product_names))
serv <- c("Coastal_Protection", "Access", "Nitrogen_Retention", "Sediment_Retention", "Pollination (People fed)", "Pollination (Habitat)" ) 
#serv <- c("Access") 

new_nam <- cbind(prod_nam, serv)
```


```{r extract values conservation, eval=FALSE, include=FALSE}

baseES_v <- lapply(baseES_api, function(r){
  t <- as_tibble(values(r, na.rm=TRUE))
})

baseESap_v <- lapply(baseES_ap, function(r){
  t <- as_tibble(values(r, na.rm=TRUE))
})

# Add two new columns to each data frame in the list
baseES_v  <- Map(function(df, row) {
  names(df)[1] <- "Value"
  df$Service <- row[[1]]  
  df$Country <- row[[2]]  
  return(df)
}, baseES_v , split(file_info, seq(nrow(file_info))))

baseESap_v  <- Map(function(df, row) {
  names(df)[1] <- "Value"
  df$Service <- row[[1]]  
  df$Country <- row[[2]]  
  return(df)
}, baseESap_v , split(file_info, seq(nrow(file_info))))

           
##################### do not run inside the Rmd!!!!!!!!
baseES_v <- do.call(rbind, baseES_v)
baseESap_v <- do.call(rbind, baseESap_v)
##################### do not run inside the Rmd!!!!!!!!
baseES_v <- baseES_v%>% mutate(AOI = "0")
baseESap_v <- baseESap_v%>% mutate(AOI = "1")

baseES_v <- rbind(baseES_v, baseESap_v)
rm(baseESap_v)

baseES_v <- as_tibble(baseES_v)
baseES_v <- baseES_v %>%
  # Create a new column with simplified service names
  mutate(Service = case_when(
    Service == new_nam[1,][1] ~ new_nam[1,][2],
    Service == new_nam[2,][1] ~ new_nam[2,][2],
    Service == new_nam[3,][1] ~ new_nam[3,][2],
    Service == new_nam[4,][1] ~ new_nam[4,][2],
    Service == new_nam[5,][1] ~ new_nam[5,][2],
    Service == new_nam[6,][1] ~ new_nam[6,][2],
    TRUE ~ Service  # Keep other names unchanged, if any
  ))
# 
 write.csv(baseES_v, here('output_data', "distributionServ_rest.csv"))

```


```{r calculate percentiles 2 and 98, eval=FALSE, include=FALSE}

#baseES_vf <- read.csv(here('data', "distributionServ_pa.csv"))
baseES_v <- as_tibble(baseES_v)
# Calculate the 98th percentile for each combination of Service and Country
  baseES_v$Service <- factor(baseES_v$Service, levels = c(
    "Nitrogen_Retention", "Sediment_Retention", "Pollination (People fed)","Pollination (Habitat)", "Access","Coastal_Protection"))

# Calculate both the 2nd and 98th percentiles
percentiles <- baseES_v %>%
  group_by(Service, Country) %>%
  summarize(
    lower_threshold = quantile(Value, 0.02),
    upper_threshold = quantile(Value, 0.98)
  ) %>%
  ungroup()

# Join the thresholds back to the original data and filter values within the 2nd and 98th percentiles
baseES_filtered <- baseES_v %>%
  left_join(percentiles, by = c("Service", "Country")) %>%
  filter(Value >= lower_threshold & Value <= upper_threshold)
```

# Density Plots


### 4.3.1 Brazil

This is calcualted for the percentile 95 to remove extreme values
```{r density plots pa 1, echo=FALSE, fig.height=6, fig.width=9}

C <- 1

df <- baseES_filtered %>% filter(Country==nam[C])
t <- ggplot(df, aes(x = Value, fill = AOI)) +
  geom_density(alpha = 0.4) +  # Use transparency with alpha to see overlapping densities
  facet_wrap(~Service, scales = "free") +
  # Set both x and y axes to be free for each facet
  scale_fill_manual(values = c("blue", "orange")) +  # Set colors for the `restoration` variable
  labs(
    #title = paste(" Inside and Outside Rest. Potential Areas -", nam[C]),
    x = "Value (Log Scale)",
    y = "Density",
    fill = "restoration Potential"
  ) +
  scale_x_continuous(trans = 'log10') +  # Apply log10 transformation to x-axis
  theme(
    strip.text = element_text(size = 12),  # Increase facet labels' font size
    legend.position = "bottom",  # Place the legend at the bottom
    legend.box = "horizontal"  # Arrange legend items horizontally
  )
print(t)
```

### 4.3.2 Madagascar
```{r density plots pa 2, echo=FALSE, fig.height=6, fig.width=9}

C <- 2

df <- baseES_filtered %>% filter(Country==nam[C])
t <- ggplot(df, aes(x = Value, fill = AOI)) +
  geom_density(alpha = 0.4) +  # Use transparency with alpha to see overlapping densities
  facet_wrap(~Service, scales = "free") +
  # Set both x and y axes to be free for each facet
  scale_fill_manual(values = c("blue", "orange")) +  # Set colors for the `restoration` variable
  labs(
    #title = paste("Density Plot of Services Inside and Outside Protected Areas -", nam[C]),
    x = "Value (Log Scale)",
    y = "Density",
    fill = "Restoration Potential"
  ) +
  scale_x_continuous(trans = 'log10') +  # Apply log10 transformation to x-axis
  theme(
    strip.text = element_text(size = 12),  # Increase facet labels' font size
    legend.position = "bottom",  # Place the legend at the bottom
    legend.box = "horizontal"  # Arrange legend items horizontally
  )
print(t)
```
### 4.3.2 Mexico
```{r density plots pa 3, echo=FALSE, fig.height=6, fig.width=9}

C <- 3

df <- baseES_filtered %>% filter(Country==nam[C])
t <- ggplot(df, aes(x = Value, fill = AOI)) +
  geom_density(alpha = 0.4) +  # Use transparency with alpha to see overlapping densities
  facet_wrap(~Service, scales = "free") +
  # Set both x and y axes to be free for each facet
  scale_fill_manual(values = c("blue", "orange")) +  # Set colors for the `restoration` variable
  labs(
    #title = paste("Density Plot of Services Inside and Outside Protected Areas -", nam[C]),
    x = "Value (Log Scale)",
    y = "Density",
    fill = "Restoration Potential"
  ) +
  scale_x_continuous(trans = 'log10') +  # Apply log10 transformation to x-axis
  theme(
    strip.text = element_text(size = 12),  # Increase facet labels' font size
    legend.position = "bottom",  # Place the legend at the bottom
    legend.box = "horizontal"  # Arrange legend items horizontally
  )
print(t)
```
### 4.3.4 Peru
```{r density plots pa 4, echo=FALSE, fig.height=6, fig.width=9}

C <- 4
df <- baseES_v %>% filter(Country==nam[C])
t <- ggplot(df, aes(x = Value, fill = AOI)) +
  geom_density(alpha = 0.4) +  # Use transparency with alpha to see overlapping densities
  facet_wrap(~Service, scales = "free") +
  # Set both x and y axes to be free for each facet
  scale_fill_manual(values = c("blue", "orange")) +  # Set colors for the `restoration` variable
  labs(
    #title = paste("Density Plot of Services Inside and Outside Protected Areas -", nam[C]),
    x = "Value (Log Scale)",
    y = "Density",
    fill = "Restoration Potential"
  ) +
  scale_x_continuous(trans = 'log10') +  # Apply log10 transformation to x-axis
  theme(
    strip.text = element_text(size = 12),  # Increase facet labels' font size
    legend.position = "bottom",  # Place the legend at the bottom
    legend.box = "horizontal"  # Arrange legend items horizontally
  )
print(t)
```
### 4.3.5 Vietnam
```{r density plots pa 1, echo=FALSE, fig.height=6, fig.width=9}

C <- 5

df <- baseES_filtered %>% filter(Country==nam[C])
t <- ggplot(df, aes(x = Value, fill = AOI)) +
  geom_density(alpha = 0.4) +  # Use transparency with alpha to see overlapping densities
  facet_wrap(~Service, scales = "free") +
  # Set both x and y axes to be free for each facet
  scale_fill_manual(values = c("blue", "orange")) +  # Set colors for the `restoration` variable
  labs(
    title = paste("Density Plot of Services Inside and Outside Protected Areas -", nam[C]),
    x = "Value (Log Scale)",
    y = "Density",
    fill = "Protected Area"
  ) +
  scale_x_continuous(trans = 'log10') +  # Apply log10 transformation to x-axis
  theme(
    strip.text = element_text(size = 12),  # Increase facet labels' font size
    legend.position = "bottom",  # Place the legend at the bottom
    legend.box = "horizontal"  # Arrange legend items horizontally
  )
print(t)
```



####################################################################### 

Second part with justin.

# Explore Data.

The ecorregiosn Raster is not suited, 830 different values. Could be
done but does not make a lot of sense

```{r explore data}

ecoregions <- rast('/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/DataOBS_op/global_products/Ecoregions2017_compressed_md5_316061.tif')

ec_uniqe <- unique(ecoregions)
```

Use vector file, just the main biomes.

# PA

```{r filter out PAs}

pa <- st_read('/Users/sputnik/Documents/Natural_capital/NBS_OP/data/vector/global-2024-05-08.gpkg')

att <- unique(pa$IUCN_CAT)

filt <- c("Not Reported", "Not Assigned", "Not Applicable")
pa <- pa %>% filter(!(IUCN_CAT %in% filt))

st_write(pa, '/Users/sputnik/Documents/Natural_capital/NBS_OP/data/vector/PA_filt.gpkg')



countries <- st_read('/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/DataOBS_op/global_products/cartographic_ee_ee_r264_correspondence.gpkg')

pa <- st_transform(pa, crs=st_crs(countries))

```

split the rasters by biomes

```{r split countries}
wd. <- '/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/DataOBS_op/global_products'

tf <- list.files(wd., pattern='1992')
tf <- clean_filename(tf)

tf <- tf[-c(2,3,5)]


tiffes <- file.path(wd., list.files(wd., pattern='1992'))
rs <- lapply(tiffes, rast) 

rs <- rs[-3]

biomes <- st_read('/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/global_products/biomes_WWF.gpkg')


nam <- sort(biomes$BIOME)

biomes <- biomes %>% split(.$BIOME)

# Apply cropping and masking recursively using lapply and map
rs_msk <- lapply(rs, function(raster) {
  map(biomes, function(polygon) {
    # Crop the raster to the extent of the polygon
    cropped_raster <- crop(raster, polygon)
    masked_raster <- mask(cropped_raster, polygon)
    return(masked_raster)
  })
})

outpath <- '~/Documents/Natural_capital/Global_mapping/rest_masked_biomes'
# Use lapply to iterate over each list in nested_list
lapply(seq_along(rs_msk), function(i) {
  # For each outer list, iterate through the inner list (rasters)
  lapply(seq_along(rs_msk[[i]]), function(j) {
    
    # Construct the filename using vector_1 and vector_2
    file_name <- paste0(outpath, tf[[i]], '_', nam[[j]], '.tif')
    
    # Write the raster to the file
    writeRaster(rs_msk[[i]][[j]], file_name, overwrite = TRUE)
  })
})

summary_df <- map2_dfr(seq_along(rs_msk), rs_msk, function(i, inner_list) {
  
  # i corresponds to the index of the outer list (e.g., 1, 2, 3...)
  outer_label <- paste0("Outer_", i)  # Label for the outer list
  
  # Loop through each SpatRaster in the inner list
  map_dfr(inner_list, function(raster) {
    extract_raster_summary(raster, outer_label)
  })
})

```

# Split by Countries/continents

# mask by protected areas (all togeter, chategories I-VI)

# Reclassify Land Cover maps.

# Change Maps

```{r change_map_calc}
esat <- c('/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/DataOBS_op/global_products/ESACCI-LC-L4-LCCS-Map-300m-P1Y-2020-v2.1.1_md5_2ed6285e6f8ec1e7e0b75309cc6d6f9f.tif','/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/DataOBS_op/global_products/ESACCI-LC-L4-LCCS-Map-300m-P1Y-1992-v2.0.7cds_compressed_md5_60cf30.tif')


esas <- lapply(esat,rast)


nam <- c(2020,1992)

map(1:2, function(x) writeRaster(esas[[x]], paste0('ESA_LC_', nam[x],'.tif')))
cls <- lapply(esas,freq)


wd <- '~/Documents/Natural_capital/NBS_OP/data/ESA_LC'

tf <- file.path(wd, list.files(wd, pattern='tif'))

tf <- tf[c(2,3)]

esas <- lapply(tf,rast)



countries <- st_read('/Users/sputnik/Library/CloudStorage/OneDrive-TempleUniversity/personal files/PosDoc/DataOBS_op/global_products/cartographic_ee_ee_r264_correspondence.gpkg')

cont <- countries %>% filter(subregion=='Central America')

# Define the reclassification rules in one operation
reclass_table <- data.frame(
  from = c(10, 11, 12, 20, 30, 40, 50, 60, 61, 62, 70, 71, 72, 80, 81, 82, 90, 100,
           110, 120, 121, 122, 130, 140, 150, 151, 152, 153, 160, 170, 180,
           190, 200, 201, 202, 210, 220),
  to = c(1, 1, 1, 1, 1, 1,   # Cultivated
         2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2,  # Forests
         3, 3, 3, 3, 3,   # Grasses and Shrubs
         4, 4, 4, 4, 4,   # Sparse Vegetation
         5, 5, 5,   # Mangroves
         6,   # Urban
         7, 7, 7,   # Bare
         8,   # Water
         9)   # Ice
)

esas <- lapply(esas, function(r) {
  subst(r, from = reclass_table$from, to = reclass_table$to)
})


map(1:2, function(x) writeRaster(esas[[x]], paste0('~/Documents/Natural_capital/NBS_OP/data/ESA_LC/', "Forests_esa", year[x], '.tif')))

# Extract forests
esas <- lapply(esas, function(r) {
  subst(r, from = 2, to =1, others=NA)
})


chg <- ch_mapR(esas[[1]],esas[[2]])
freq(chg)

mat_1 <- crosstabm(esas[[1]],esas[[2]])

diffs_p <-  diffTablej(mat, digits=3, analysis = "change")

```

# Get Sankey Diagrams

```{r Sankey Diagram}
classes <- c("Cultivated", "Forests", "Grasses/shrubs", "Sparse", "Mangroves", "Urban", "Bare", "Water")
groupColor <- c("#dcf064","#00b809","#8ca000","#ffebaf",'#009678', "#c31400", "#fff5d7", "#0046c8")

years <- c(1992,2020)


nodeInfo <- nodeInfoR(years, classes, groupColor)

NodeCols<- sort(unique(nodeInfo$nodeCol))

#create paths to the rasters/bands
num_bands <- length(NodeCols)
tf <- list.files(wd, pattern= "Rec")
tf <- file.path(wd,tf)
fileInfo <- tibble(
  nodeCol = seq_along(tf),
  rast = tf,
  rasterBand = 1
)

# join path strings to nodeInfo
nodeInfo <- dplyr::left_join(nodeInfo, fileInfo, by='nodeCol') 
linkInfo <- linkInfoR(NodeCols, nodeInfo)


fontSize <- 0.5
nodeWidth <-30
fontFamily <- "sans-serif"
colorScaleJS <- sprintf("d3.scaleOrdinal().domain(%s).range(%s)",
                        jsonlite::toJSON(unique(nodeInfo$nodeGroup)),
                        jsonlite::toJSON(unique(nodeInfo$groupColor)))


sankeyNetwork(Links = linkInfo, Nodes = nodeInfo,
              Source = "source",
              Target = "target",
              Value = "value",
              NodeID = "nodeName",
              NodeGroup = "nodeGroup",
              LinkGroup = "LinkGroup",
              fontSize = fontSize,
              fontFamily = fontFamily,
              nodePadding = 10,
              margin=1,
              nodeWidth = nodeWidth,
              colourScale = JS(colorScaleJS))
```

# Next step

1.  Zonal Countries. one per service

mask for the forest

not sum, average.

Historic services.

name(test)

Map 4 maps for each servie for the average change fior that time peiod

map(1:2, function(x) writeRaster(esas[[x]],
paste0('\~/Documents/Natural_capital/NBS_OP/data/ESA_LC/',
"ESA_LC_Rec\_", year[x], '.tif')))

cls [[1]] layer value count 1 1 10 105009519 2 1 11 102087508 3 1 12
2415511 4 1 20 29796402 5 1 30 48088904 6 1 40 44613941 7 1 50 137316150
8 1 60 84649974 9 1 61 11061590 10 1 62 39391781 11 1 70 113472860 12 1
71 45563958 13 1 72 19882 14 1 80 109865531 15 1 81 55745 16 1 82 21 17
1 90 37417117 18 1 100 58048747 19 1 110 15730987 20 1 120 140408175 21
1 121 3007375 22 1 122 32663372 23 1 130 177774130 24 1 140 42447574 25
1 150 155885143 26 1 151 61 27 1 152 1072086 28 1 153 3664319 29 1 160
11677666 30 1 170 2394625 31 1 180 34499895 32 1 190 10933052 33 1 200
247641877 34 1 201 1572727 35 1 202 1244423 36 1 210 5675137546 37 1 220
871449826

So, now what does this mean?

Add column with the class names with multiple aggregation levels (ESA)

```         
Mean: The average value within each zone (e.g., inside/outside protected areas or across land cover types). This provides a simple understanding of the central tendency.
Median: The middle value when data is ordered, useful when your data has outliers that could skew the mean.
Standard Deviation: Indicates the variation or dispersion from the average, helping show if values are clustered or spread out.
Min/Max: The range of values, showing the extremes within each zone.
```

2.  Zonal Statistics (Per Category):

    Sum: The total value across a zone, particularly useful for
    variables like ecosystem services (e.g., total nitrogen retention
    within protected areas). Count of Pixels: The number of pixels in
    each zone, which could give an idea of the area covered by specific
    land covers. Percentage Area Contribution: For each zone or land
    cover class, express the variable as a percentage of the total
    landscape (e.g., “30% of nitrogen retention occurs in protected
    areas”).

3.  Comparisons by Category:

    Inside vs. Outside Protected Areas: You can calculate mean, sum, and
    standard deviation inside and outside protected areas to compare the
    effectiveness of protection measures in preserving ecosystem
    services. Land Cover Comparison: Use zonal means to compare how
    different land cover types contribute to ecosystem services like
    sediment retention or distance to beneficiaries. Express results as
    percentages, e.g., “Forests account for 60% of nitrogen retention.”

4.  Histogram or Distribution Analysis:

    A frequency distribution showing how the values of each variable are
    distributed across the entire study area or within protected vs.
    unprotected zones. This can help show if most of the service is
    concentrated in a few high-value areas or spread out.

5.  Change Detection (If Time Series Exists):

    If your rasters cover different time periods, you could show net
    changes in ecosystem service values (e.g., changes in nitrogen
    retention) inside and outside protected areas or across different
    land covers.

6.  Normalized or Standardized Scores:

    Convert variables into percentile ranks or Z-scores to facilitate
    comparisons across variables that might have different units (e.g.,
    how “high” or “low” the ecosystem service is in relation to the
    mean).

7.  Threshold Analysis:

    For variables where a threshold is meaningful (e.g., distance to
    beneficiaries \< 10 km), you can count the number of pixels above or
    below the threshold in each category (protected areas, land covers).
    For example, "80% of the people dependent on agriculture live within
    5 km of high-sediment retention areas."

8.  Spatial Autocorrelation (Advanced):

    If you're dealing with spatially explicit relationships, calculating
    spatial autocorrelation (Moran’s I) might help quantify clustering
    patterns of ecosystem services inside and outside protected areas.

9.  Cumulative Impact Maps:

    Create a composite map where multiple variables are aggregated to
    show high- vs. low-value areas for ecosystem services. This could
    simplify complex data into an easily interpretable map for
    non-specialists.

These summaries will provide enough detail to highlight key insights,
while also being interpretable by those without technical expertise. Let
me know if you'd like to dive deeper into any specific metric or
approach!

Produyces forest loss maps/gain permanences

Country/continent

Biomes. (Köppen)
