---
title: "Beneficiary-Mask Socioeconomic Disproportionality (HDI / Gini / GDP)"
subtitle: "v1.3.4 — Phase 4, 5-service redesign"
analysis_version: "v1.3.4"
date: now
date-format: "YYYY-MM-DD HH:mm zzz"
format:
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
    df-print: kable
editor: source
params:
  input_gpkg: "10k_change_calc.gpkg"
  coverage_csv: "tables/beneficiary_mask_coverage_10km.csv"
  output_suffix: ""
---

## Overview

Becky's ask (Slack, 2026-08-05/06): apply Rich's beneficiary union-coverage masks —
`combined_cross_category` and the `3+`/`4+` nested hotspot-count tiers — to the HDI,
Gini, and GDP rasters, to test whether beneficiary areas are socioeconomically
disproportionate compared to everywhere else.

This notebook reuses the exact two-sample KS + Cliff's Delta machinery built for the
WHO chapter (`analysis/KS_tests_hotspots.qmd`, `run_ks_hot_vs_non()` in
`R/ks_hotspots.R`), with one substitution: the grouping variable is not a
service's extreme-change hotspot definition, it is a spatial "inside the beneficiary
buffer mask" flag, produced by zonal-extracting Rich's fine-resolution
(~30 arcsec) `full_raster_extent_union_coverage.tif` masks onto the 10km analysis
grid (`Python_scripts/zonal_extract_beneficiary_masks.py`, run inside Rich's Docker
environment).

**Comparison is direct in-mask vs. out-of-mask**, not the WHO chapter's "median
background" tail comparison — there is no analogous "typical change" concept for a
spatial buffer-membership flag; Becky's question is simply mask vs. everywhere else.

**Threshold**: `full_raster_extent_union_coverage.tif` gives a fine-resolution
0/1 mask; zonal-extracting it onto a coarser 10km cell gives a *fraction* of that
cell's area covered by the mask. A 10km cell is flagged "inside" the beneficiary
area if that fraction is ≥ 0.5 (majority rule) — see `KS_CFG$coverage_threshold`
below to adjust.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false

library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(readr)
library(scales)
library(sf)
library(here)
library(devtools)

knitr::opts_knit$set(root.dir = here::here())
source(here::here("R","paths.R"))
devtools::load_all(quiet = TRUE)

options(dplyr.summarise.inform = FALSE)
Sys.setenv(GDAL_NUM_THREADS = "ALL_CPUS", PROJ_NETWORK = "ON")
set.seed(1)
```

## Config

```{r}
#| label: config
#| echo: true

KS_CFG <- list(
  categories = c(
    combined_cross = "combined_cross",
    tier_3plus     = "tier_3plus",
    tier_4plus     = "tier_4plus"
  ),
  coverage_threshold = 0.5,
  vars_to_test = c(
    "rast_gdpTot_1990_2020_30arcsec_2020_sum",
    "GHS_POP_E2020_GLOBE_sum",
    "hdi_raster_predictions_2020_mean",
    "rast_adm1_gini_disp_2020_mean"
  ),
  sampling          = "none",
  nonhot_cap        = 100000L,
  permute_n         = 0,
  compute_one_sided = TRUE,
  adjust_method     = "BH",
  transform_default = "auto",
  transform_by_var = c(
    "rast_gdpTot_1990_2020_30arcsec_2020_sum" = "log1p",
    "GHS_POP_E2020_GLOBE_sum"                 = "log1p"
  ),
  ecdf_top_k             = 3,
  ecdf_service_selection = "all",
  out_table = file.path(data_dir(), "processed", "tables",
                         paste0("ks_results_beneficiary_masks", params$output_suffix, ".csv")),
  out_dir   = out_plots(paste0("ks_beneficiary_masks", params$output_suffix))
)
```

## Load grid + coverage fractions, apply standard land-base filters

Same exclusions used throughout this project's grid-based analyses (Antarctica /
Seven seas / Lakes / Rock & Ice) so the land base here matches the already-verified
1,372,621-cell denominator used elsewhere in the 5-service redesign.

```{r}
#| label: load-grid
#| message: false
#| warning: false

gpkg <- file.path(data_dir(), "processed", params$input_gpkg)
stopifnot(file.exists(gpkg))

lyrs <- sf::st_layers(gpkg)$name
grid_sf <- sf::st_read(gpkg, layer = lyrs[1], quiet = TRUE)
grid <- sf::st_drop_geometry(grid_sf)
stopifnot("grid_fid" %in% names(grid), grid$grid_fid |> duplicated() |> any() == FALSE)

if ("continent" %in% names(grid)) {
  grid <- dplyr::filter(grid, !continent %in% c("Antarctica", "Seven seas (Open Ocean)"))
}
if ("WWF_biome" %in% names(grid)) {
  grid <- dplyr::filter(grid, !WWF_biome %in% c("Lakes", "Rock & Ice"))
}

n_land_cells <- nrow(grid)
message("Land base after standard exclusions: ", n_land_cells, " cells")
```

## Load beneficiary-mask coverage fractions

```{r}
#| label: load-coverage
#| message: false
#| warning: false

coverage <- readr::read_csv(file.path(data_dir(), "processed", params$coverage_csv),
                             show_col_types = FALSE)
stopifnot(all(c("grid_fid","category","coverage_frac") %in% names(coverage)))

# sanity: every requested category actually present
missing_cats <- setdiff(unname(KS_CFG$categories), unique(coverage$category))
if (length(missing_cats)) stop("Missing categories in coverage CSV: ", paste(missing_cats, collapse = ", "))
```

## Validation: cross-check against the independently-computed fine-resolution area %

Before trusting this join, cross-check it against `outputs/tables/
hotspot_5service_beneficiary_area_pct.csv` (2026-08-05), computed via a completely
independent path — `terra::expanse()` directly on the fine-resolution rasters, no
10km grid, no exact_extract, no Python zonal-stats pipeline at all. Because the
analysis grid is equal-area, `% of land cells flagged` (post-threshold) and the raw
area-weighted mean coverage fraction (pre-threshold) should both land close to that
reference number. A large mismatch would mean the grid_fid join is wrong — this
project's history has 4 prior incidents of exactly that failure mode, so this check
is load-bearing, not decorative.

```{r}
#| label: validate-coverage
#| message: false
#| warning: false

ref_pct <- readr::read_csv(project_dir("outputs","tables","hotspot_5service_beneficiary_area_pct.csv"),
                            show_col_types = FALSE) |>
  dplyr::filter(buffer_type == "union") |>
  dplyr::mutate(category_short = dplyr::case_match(
    category,
    "combined_cross_category_beneficiaries" ~ "combined_cross",
    "hotspot_count_3plus_beneficiaries"     ~ "tier_3plus",
    "hotspot_count_4plus_beneficiaries"     ~ "tier_4plus",
    "hotspot_count_1plus_beneficiaries"     ~ "tier_1plus",
    "hotspot_count_2plus_beneficiaries"     ~ "tier_2plus",
    "hotspot_count_5plus_beneficiaries"     ~ "tier_5plus",
    "water_overlap_downstream_beneficiaries"     ~ "water",
    "access_overlap_travel_time_beneficiaries"   ~ "access",
    .default = NA_character_
  )) |>
  dplyr::select(category_short, ref_pct_of_land = pct_of_total_land)

check <- coverage |>
  dplyr::semi_join(grid, by = "grid_fid") |>
  dplyr::group_by(category) |>
  dplyr::summarise(
    n_land_cells      = dplyr::n(),
    mean_coverage_pct = 100 * mean(coverage_frac, na.rm = TRUE),
    pct_flagged_ge50  = 100 * mean(coverage_frac >= KS_CFG$coverage_threshold, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::left_join(ref_pct, by = c("category" = "category_short")) |>
  dplyr::mutate(
    diff_meancov_vs_ref = mean_coverage_pct - ref_pct_of_land,
    diff_flagged_vs_ref = pct_flagged_ge50  - ref_pct_of_land
  )

knitr::kable(check, digits = 2)

# hard stop if any target category is wildly off (>5 pct-point) from the
# independent reference -- would indicate a grid-identity bug, not just
# expected 10km-vs-30arcsec discretization noise
target_cats <- unname(KS_CFG$categories)
bad <- check |> dplyr::filter(category %in% target_cats, abs(diff_meancov_vs_ref) > 5)
if (nrow(bad) > 0) {
  print(bad)
  stop("Coverage validation failed: >5pp mismatch vs independent area-pct reference for a target category.")
}
message("Validation passed: all target categories within 5pp of the independent area-pct reference.")
```

## Build hotspot / non-hotspot tables (in-mask vs. out-of-mask)

```{r}
#| label: build-groups
#| message: false
#| warning: false

grid_vars <- grid |>
  dplyr::select(grid_fid, dplyr::any_of(KS_CFG$vars_to_test))

stack <- purrr::map_dfr(names(KS_CFG$categories), function(cat_label) {
  cat_id <- KS_CFG$categories[[cat_label]]
  cov_cat <- coverage |>
    dplyr::filter(category == cat_id) |>
    dplyr::select(grid_fid, coverage_frac)

  grid_vars |>
    dplyr::inner_join(cov_cat, by = "grid_fid") |>
    dplyr::mutate(
      service = cat_id,
      flag = coverage_frac >= KS_CFG$coverage_threshold
    )
})

hotspots_df     <- dplyr::filter(stack, flag)
non_hotspots_df <- dplyr::filter(stack, !flag)

stopifnot(nrow(hotspots_df) > 0, nrow(non_hotspots_df) > 0)

dplyr::count(stack, service, flag)
```

## KS tests: beneficiary-mask vs. non-mask

```{r}
#| label: run-ks
#| message: false
#| warning: false

dir.create(dirname(KS_CFG$out_table), recursive = TRUE, showWarnings = FALSE)
ks_res <- run_ks_hot_vs_non(
  hotspots_df      = hotspots_df,
  inverse_df       = non_hotspots_df,
  vars             = KS_CFG$vars_to_test,
  sampling         = KS_CFG$sampling,
  nonhot_cap       = KS_CFG$nonhot_cap,
  permute_n        = KS_CFG$permute_n,
  compute_one_sided= KS_CFG$compute_one_sided,
  adjust_method    = KS_CFG$adjust_method,
  out_csv          = KS_CFG$out_table
)

dplyr::glimpse(ks_res, width = 120)
```

## Visualizations and exports

```{r}
#| label: ks-plots
#| message: false
#| warning: false
#| echo: false

ks_plot_dir <- KS_CFG$out_dir
if (dir.exists(ks_plot_dir)) unlink(list.files(ks_plot_dir, full.names = TRUE), recursive = TRUE)
dir.create(ks_plot_dir, recursive = TRUE, showWarnings = FALSE)

heat_path <- file.path(ks_plot_dir, "ks_heatmap.png")
ggplot2::ggsave(
  heat_path,
  plot_ks_heatmap(ks_res, fill_limits = c(0, NA), reorder = FALSE) +
    ggplot2::scale_fill_viridis_c(name = "KS Statistic (D)", option = "magma"),
  width = 8, height = 5, dpi = 300, bg = "white"
)

for (v in KS_CFG$vars_to_test) {
  safe_v <- gsub("[^A-Za-z0-9_]+", "_", v)
  p_bar <- plot_ks_bars(ks_res, v, top_k = length(KS_CFG$categories))
  ggplot2::ggsave(file.path(ks_plot_dir, paste0("ks_bars_", safe_v, ".png")),
                   p_bar, width = 7, height = 5, dpi = 300, bg = "white")
}

ecdf_dir <- file.path(ks_plot_dir, "ecdf")
dir.create(ecdf_dir, recursive = TRUE, showWarnings = FALSE)
ecdf_files <- save_ecdf_grids(
  hotspots_df, non_hotspots_df,
  vars               = KS_CFG$vars_to_test,
  ks_res             = ks_res,
  service_selection  = KS_CFG$ecdf_service_selection,
  transform_default  = KS_CFG$transform_default,
  transform_by_var   = KS_CFG$transform_by_var,
  out_dir            = ecdf_dir,
  prefix             = "ecdf",
  width = 9, height = 6, dpi = 300, use_ragg = FALSE, verbose = FALSE
)

dir_path <- file.path(ks_plot_dir, "directionality_cliffs_delta.png")
ks_res_plot <- ks_res |>
  dplyr::mutate(var_label = gsub("_", " ", var)) |>
  dplyr::mutate(var_label = stringr::str_wrap(var_label, 40))

signed_pow4 <- scales::trans_new(
  "signed_pow4",
  transform = function(x) sign(x) * abs(x)^0.25,
  inverse = function(x) sign(x) * abs(x)^4
)

p_direction <- ggplot(ks_res_plot, aes(x = service, y = cliffs_delta, fill = cliffs_delta)) +
  geom_bar(stat = "identity", position = "identity") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0,
                        name = "Cliff's Delta", trans = signed_pow4) +
  facet_wrap(~ var_label, ncol = 2, scales = "free_y") +
  coord_flip() +
  labs(
    title = "Directionality of Beneficiary-Mask Effects (Cliff's Delta)",
    subtitle = "Cliff's Delta > 0 (red): beneficiary-mask cells have higher values.\nCliff's Delta < 0 (blue): beneficiary-mask cells have lower values.",
    x = "Beneficiary mask category", y = "Cliff's Delta (Effect Size)"
  ) +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(size = 8))

ggsave(dir_path, p_direction, width = 10, height = 7, dpi = 300, bg = "white")
```

## Results table

```{r}
#| label: ks-table
#| echo: false

ks_clean <- ks_res |>
  dplyr::mutate(
    Variable = dplyr::case_when(
      grepl("gdpTot", var) ~ "GDP",
      grepl("POP", var)    ~ "Population",
      grepl("hdi", var)    ~ "HDI",
      grepl("gini", var)   ~ "Gini Index",
      TRUE ~ var
    )
  ) |>
  dplyr::select(Category = service, Variable, KS_Stat_D = D, Cliffs_Delta = cliffs_delta,
                p_adj, mean_hot, mean_non, median_hot, median_non, n_hot, n_non) |>
  dplyr::arrange(Variable, dplyr::desc(abs(Cliffs_Delta)))

knitr::kable(ks_clean, digits = 3)
```

## Preview

```{r}
#| label: ks-preview
#| echo: false
#| results: 'asis'

pngs <- list.files(ks_plot_dir, pattern = "\\.png$", recursive = TRUE, full.names = FALSE)
if (length(pngs)) {
  rel_dir <- file.path("..", "outputs", "plots", paste0("ks_beneficiary_masks", params$output_suffix))
  rel_paths <- file.path(rel_dir, pngs)
  md <- paste(sprintf("![](%s)", rel_paths), collapse = "\n\n")
  knitr::asis_output(md)
} else {
  knitr::asis_output("> No KS plots were generated; check upstream chunks.")
}
```
