---
title: "Prepare Granular Land Cover Data"
author: "Jeronimo Rodriguez-Escobar"
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
editor: source
---

## Overview

This script prepares the land cover data for the granular Land Cover Change analysis.
It extracts raw ESA CCI (1992) and C3S (2020) data from the source zip file and reclassifies them into a simplified 9-class scheme.

Crucially, this reclassification ensures that **Flooded Trees (classes 160, 170)** are mapped to **Forest**, which differs from some standard aggregations that might map them to wetlands or generic vegetation. This is essential for the "Forest Loss" model.

```{r}
#| label: setup
#| message: false
#| warning: false

library(terra)
library(glue)
library(here)
library(devtools)

# Initialize paths
knitr::opts_knit$set(root.dir = here::here())
source(here::here("R", "paths.R"))
devtools::load_all(quiet = TRUE)
```

## Configuration

```{r}
#| label: config

years_to_process <- c(1992, 2020)

# Output directory for granular analysis inputs
lc_dir_granular <- data_raw("LandCovers_Granular")
if (!dir.exists(lc_dir_granular)) dir.create(lc_dir_granular, recursive = TRUE)

# Raw zip file location
raw_zip_path <- file.path(data_raw(), "6e774a4e4a552fd229bcb88b4d3c02e6.zip")

# Reclassification matrix: Original ESA (37 classes) -> 9 simplified classes
# 1:Cropland, 2:Forest, 3:Grassland/Shrubland, 4:Sparse, 5:Wetland, 6:Urban, 7:Bare, 8:Water, 9:Snow
# Note: 160/170 (Flooded Trees) mapped to 2 (Forest) to capture mangroves/swamp forests.
reclass_37_to_9 <- matrix(c(
  10, 1, 11, 1, 12, 1, 20, 1, 30, 1, 40, 1,
  50, 2, 60, 2, 61, 2, 62, 2, 70, 2, 71, 2, 72, 2, 80, 2, 81, 2, 82, 2, 90, 2, 100, 2,
  110, 3, 120, 3, 121, 3, 122, 3, 130, 3,
  140, 4, 150, 4, 151, 4, 152, 4, 153, 4,
  160, 2, 170, 2, 180, 5,
  190, 6,
  200, 7, 201, 7, 202, 7,
  210, 8,
  220, 9
), ncol = 2, byrow = TRUE)
```

## Processing

```{r}
#| label: process-rasters

for (year in years_to_process) {
  out_file <- file.path(lc_dir_granular, paste0("landcover_gl_", year, ".tif"))

  # Force regeneration to ensure the correct classification (e.g. flooded trees) is applied.
  # if (file.exists(out_file)) {
  #   message(glue::glue("File for {year} already exists at {out_file}. Skipping."))
  #   next
  # }

  message(glue::glue("Processing {year} from raw zip..."))

  if (!file.exists(raw_zip_path)) {
    stop("Raw zip file not found at: ", raw_zip_path)
  }
  if (file.size(raw_zip_path) < 10000) {
    stop("Raw zip file appears corrupted or too small (<10KB): ", raw_zip_path)
  }

  # Unzip specific year file to temp
  ex_dir <- tempdir()
  file_list <- unzip(raw_zip_path, list = TRUE)

  # Find file matching year (NetCDF or TIF)
  # Pattern matches C3S/ESA naming conventions
  raw_file_name <- grep(paste0(year, ".*(nc|tif)$"), file_list$Name, value = TRUE)[1]

  if (is.na(raw_file_name)) {
    stop("Could not find raw file for year ", year, " in zip.")
  }

  message("Unzipping ", raw_file_name, "...")
  unzip(raw_zip_path, files = raw_file_name, exdir = ex_dir)
  raw_file_path <- file.path(ex_dir, raw_file_name)

  # Load and reclassify
  message("Reclassifying to 9 classes...")
  r_raw <- terra::rast(raw_file_path)

  # If NetCDF, it might have multiple layers, usually 'lccs_class' is what we want
  if (terra::nlyr(r_raw) > 1) {
    if ("lccs_class" %in% names(r_raw)) {
      r_raw <- r_raw[["lccs_class"]]
    } else {
      r_raw <- r_raw[[1]]
    }
  }

  r_9 <- terra::classify(r_raw, reclass_37_to_9, filename = out_file,
                         gdal = c("COMPRESS=LZW", "TILED=YES"), overwrite = TRUE)

  message("Generated: ", out_file)
  unlink(raw_file_path) # Clean up temp
}

message("Data preparation complete.")
```
