---
title: "Land Cover Change Metrics Workflow"
author: "Jeronimo Rodriguez-Escobar"
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
editor: source
---

------------------------------------------------------------------------

This Quarto document shows how to:

1.  Load land cover classification rasters and polygon units
2.  Apply `extract_zonal_lcc_metrics()` and `iterate_zonal_lcc_metrics()`
3.  Append the extracted metrics to the input polygon data.
4.  Summarize and plot key land cover transition metrics \[pending\]

```{r}
#| label: setup
#| message: false
#| warning: false

library(terra)
library(sf)
library(dplyr)
library(ggplot2)
library(glue)
library(tidyr)
library(purrr)
library(diffeR)
library(here)
library(stringr)
library(parallel)
library(devtools)

# Initialize paths
knitr::opts_knit$set(root.dir = here::here())
source(here::here("R", "paths.R"))
devtools::load_all(quiet = TRUE)

# source the helper functions
# source(here::here("R", "utils_lcc_metrics.R")) # Should be loaded by load_all() if in R/
# source(here("R","pct_change.R")) # Should be loaded by load_all()
```

# Prepare data

**Data Sources:**

-   **1992-2015:** ESA Climate Change Initiative (CCI) Land Cover v2.0.7.
    -   *Citation:* Defourny, P., Lamarche, C., Brockmann, C., Boettcher, M., Bontemps, S., De Maet, T., Duveiller, G. L. Harper, K., Hartley A., Kirches, G., Moreau, I., Peylin, P., Ottlé, C., Radoux J., Van Bogaert, E., Ramoino, F., Albergel, C., and Arino, O.: Observed annual global land-use change from 1992 to 2020 three times more dynamic than reported by inventory-based statistics, in preparation, 2023.
    -   *User Guide:* Defourny, P., Lamarche, C., Bontemps, S., De Maet, T., Van Bogaert, E., Moreau, I., Brockmann, C., Boettcher, M., Kirches, G., Wevers, J., Santoro, M., Ramoino, F., & Arino, O. (2017). Land Cover Climate Change Initiative - Product User Guide v2. Issue 2.0. [Download PDF](http://maps.elie.ucl.ac.be/CCI/viewer/download/ESACCI-LC-Ph2-PUGv2_2.0.pdf).
    -   *Download:* [ESA CCI Viewer](https://maps.elie.ucl.ac.be/CCI/viewer/download.php)
-   **2016-2020:** Copernicus Climate Change Service (C3S) Global Land Cover v2.1.1.
    -   Consistent with the ESA CCI series.
    -   *Download:* Copernicus Climate Change Service (C3S) Climate Data Store (CDS).

**Processing:** This workflow reclassifies the original LC maps into: a) A simplified 9-class map (`landcover_gl_[year].tif`) b) A binary Natural/Transformed map (`rec_landcover_gl_[year].tif`)

```{r}
#| label: reclassify-lc-maps
#| eval: false
#| include: false

# This chunk processes all available raw land cover data into reclassified versions.

# 1. Configuration
raw_lc_dir <- data_raw("LandCovers")
mask_poly <- st_read(data_vectors("hydrosheds_lv6_synth.gpkg"), quiet = TRUE)

# Reclassification matrix: Original ESA classes to 9 simplified classes
reclass_9_class_tbl <- data.frame(
  from = c(10, 11, 12, 20, 30, 40, 50, 60, 61, 62, 70, 71, 72, 80, 81, 82, 90, 100,
           110, 120, 121, 122, 130, 140, 150, 151, 152, 153, 160, 170, 180,
           190, 200, 201, 202, 210, 220),
  to = c(1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 3, 3, 3, 3, 3, 4, 4, 4, 4, 4, 5, 5, 5, 6, 7, 7, 7, 8, 9)
)

# Reclassification matrix: 9 simplified classes to 2 binary classes
reclass_2_class_tbl <- data.frame(
  from = c(3,4,5,6,7,8,9),
  to = c(2,2,2,1,2,2,2)
)

# 2. Helper function to process one raster layer
process_lc_raster <- function(r, year, out_dir, mask_poly, reclass_9, reclass_2) {
  message("Processing year: ", year)

  # --- 9-Class Reclassification ---
  out_9_path <- file.path(out_dir, paste0("landcover_gl_", year, ".tif"))
  if (!file.exists(out_9_path)) {
    r_9class <- subst(r, from = reclass_9$from, to = reclass_9$to)
    writeRaster(r_9class, out_9_path, overwrite = TRUE, gdal=c("COMPRESS=LZW", "TILED=YES"))
    message("  Saved 9-class: ", basename(out_9_path))
  } else {
    message("  9-class file already exists, skipping.")
    r_9class <- rast(out_9_path) # Load existing for next step
  }

  # --- 2-Class Reclassification ---
  out_2_path <- file.path(out_dir, paste0("rec_landcover_gl_", year, ".tif"))
  if (!file.exists(out_2_path)) {
    r_2class <- subst(r_9class, from = reclass_2$from, to = reclass_2$to)
    r_2class_masked <- mask(r_2class, mask_poly)
    writeRaster(r_2class_masked, out_2_path, overwrite = TRUE, gdal=c("COMPRESS=LZW", "TILED=YES"))
    message("  Saved 2-class: ", basename(out_2_path))
  } else {
    message("  2-class file already exists, skipping.")
  }
}

# 3. Process the 1992-2015 multi-band stack
stack_v207_path <- file.path(raw_lc_dir, "ESACCI-LC-L4-LCCS-Map-300m-P1Y-1992_2015-v2.0.7.tif")
if (file.exists(stack_v207_path)) {
  message("Processing 1992-2015 stack...")
  r_stack <- rast(stack_v207_path)
  # Extract years from band names, e.g., "ESACCI-LC-L4-LCCS-Map-300m-P1Y-1992-v2.0.7"
  years <- as.integer(stringr::str_extract(names(r_stack), "\\d{4}"))

  for (i in 1:nlyr(r_stack)) {
    process_lc_raster(r_stack[[i]], years[i], raw_lc_dir, mask_poly, reclass_9_class_tbl, reclass_2_class_tbl)
  }
}

# 4. Process the 2016-2020 individual files
message("Processing 2016-2020 individual files...")
years_v211 <- 2016:2020
for (year in years_v211) {
  # First, convert NetCDF to GeoTIFF if needed
  nc_file <- file.path(raw_lc_dir, sprintf("C3S-LC-L4-LCCS-Map-300m-P1Y-%d-v2.1.1.nc", year))
  tif_file <- file.path(raw_lc_dir, sprintf("C3S-LC-L4-LCCS-Map-300m-P1Y-%d-v2.1.1.tif", year))
  if (!file.exists(tif_file) && file.exists(nc_file)) {
    message("Converting ", basename(nc_file), " to GeoTIFF...")
    cmd <- sprintf(
      "gdalwarp -of Gtiff -co COMPRESS=LZW -co TILED=YES -ot Byte -te -180 -90 180 90 -tr 0.002777777777778 0.002777777777778 -t_srs EPSG:4326 NETCDF:\"%s\":lccs_class \"%s\"",
      nc_file, tif_file
    )
    system(cmd)
  }

  # Now process the GeoTIFF
  if (file.exists(tif_file)) {
    r <- rast(tif_file)
    process_lc_raster(r, year, raw_lc_dir, mask_poly, reclass_9_class_tbl, reclass_2_class_tbl)
  } else {
    warning("Could not find source file for year ", year)
  }
}
message("Reclassification process complete.")
```

## Load Inputs

```{r}
#| label: load-data

# Example file paths (replace with actual paths)
inpath <- data_raw("LandCovers")
landcover_files <- list.files(
  inpath,
  pattern = "^rec_landcover_gl_\\d{4}\\.tif$",
  full.names = TRUE
)
# Define the specific years for the analysis. This is more robust than index-based selection.
years_to_process <- c(1992, 2020)

# Create a regex pattern to match any of the desired years in the filename.
year_pattern <- paste(years_to_process, collapse = "|")

# Filter the file list to include only files for the desired years.
landcover_files <- landcover_files[stringr::str_detect(basename(landcover_files), year_pattern)]
# Name rasters by year
names(landcover_files) <- stringr::str_extract(basename(landcover_files), "\\d{4}")
# Load rasters as a list
landcover_rasters <- lapply(landcover_files, terra::rast)

# Check for duplicates to prevent list-columns in pivot_wider
if (any(duplicated(names(landcover_files)))) {
  stop("Duplicate land cover files found for the same year. Please check the input directory for multiple files matching the pattern.")
}

# Load canonical spatial units
# We use the 10km grid as the primary unit. Other available vectors in vector_basedata include:
# - Biome.gpkg (WWF Biomes)
# - cartographic_ee_r264_correpondence.gpkg (Regions/Countries)

# Load the raw canonical 10km grid (geometry only) to keep it lightweight
grid_path <- data_vectors("AOOGrid_10x10km_land_4326_clean.gpkg")
spatial_units <- st_read(grid_path, quiet = TRUE)

# FOR TESTING: Use a small subset of spatial units to speed up processing.
# Remove or comment out this line for the full analysis.
# spatial_units <- spatial_units %>% dplyr::slice_head(n = 500)

# Ensure we have a valid ID column (usually 'fid' in the canonical grid)
if (!"fid" %in% names(spatial_units)) {
  spatial_units$fid <- seq_len(nrow(spatial_units))
}
spatial_units <- spatial_units %>%
  dplyr::select(fid) # Keep only the ID for the extraction step
id_col_name <- "fid"
```

## Run Land Cover Change Metric Extraction

```{r}
#| label: run-lcc-extraction
#| message: true
#| warning: false

# This chunk runs the core land cover change metric extraction using the `diffeR`
# package logic wrapped in our helper functions.

# 1. Iterate over all time steps defined by the input raster list.
#    This calculates Gain, Persistence, Loss, etc. for each polygon in `spatial_units`.
#    We use chunking to handle the large dataset (2.5M cells) robustly.

chunk_size <- 50000  # Process 50k cells at a time
n_total <- nrow(spatial_units)
n_chunks <- ceiling(n_total / chunk_size)
chunk_dir <- file.path(data_dir(), "processed", "lcc_chunks")

if (!dir.exists(chunk_dir)) dir.create(chunk_dir, recursive = TRUE)

message(glue::glue("Processing {n_total} spatial units in {n_chunks} chunks of {chunk_size}..."))

lcc_m_list <- list()

for (i in 1:n_chunks) {
  chunk_file <- file.path(chunk_dir, paste0("lcc_chunk_", i, ".rds"))

  if (file.exists(chunk_file)) {
    message(glue::glue("Chunk {i}/{n_chunks} already exists. Loading..."))
    lcc_m_list[[i]] <- readRDS(chunk_file)
  } else {
    message(glue::glue("Processing chunk {i}/{n_chunks}..."))

    # Subset spatial units
    idx_start <- (i - 1) * chunk_size + 1
    idx_end <- min(i * chunk_size, n_total)
    chunk_units <- spatial_units[idx_start:idx_end, ]

    # Run extraction for this chunk
    chunk_res <- iterate_lcc_metrics(
      landcover_rasters,
      chunk_units,
      id_col = id_col_name,
      percent = TRUE,
      verbose = FALSE,
      mc = TRUE,
      ncores = 8,
      digits = 4
    )

    saveRDS(chunk_res, chunk_file)
    lcc_m_list[[i]] <- chunk_res

    rm(chunk_units, chunk_res)
    gc()
  }
}

message("Assembling chunks...")
lcc_m <- dplyr::bind_rows(lcc_m_list)

# 2. Calculate net change direction (Gain - Loss) for easier interpretation.
lcc_m <- lcc_m %>%
  mutate(
    Gain = as.numeric(Gain),
    Loss = as.numeric(Loss),
    dir_ch = round(Gain - Loss, digits = 4)
  )

# 3. Pivot the long-format metrics table to a wide format, where each
#    metric for each class and time step becomes a separate column.
message("Pivoting results to wide format...")
lcc_metrics_wide <- lcc_m %>%
  tidyr::pivot_wider(
    id_cols = all_of(id_col_name),
    names_from = c(Category, year_step),
    values_from = c(Gain, Persistence, Loss, Quantity, Exchange, dir_ch),
    names_glue = "{.value}_{Category}_{year_step}"
  ) %>%
  mutate(
    # Ensure the ID column type is correct for joining
    !!id_col_name := as.double(.data[[id_col_name]])
  )

# 4. Attribute the grid with Biomes and Regions (Spatial Join)
#    We do this NOW using the raw vectors, instead of relying on the pre-processed ES file.
message("Attributing grid with Biomes and Regions...")

biome_path <- data_vectors("Biome.gpkg")
# Use pattern matching to find the correspondence file (handles spelling variations)
ee_files <- list.files(data_vectors(), pattern = "cartographic_ee.*\\.gpkg$", full.names = TRUE)
regions_path <- if(length(ee_files) > 0) ee_files[1] else ""

if (file.exists(biome_path) && file.exists(regions_path)) {
  # Load vectors
  biomes_sf <- st_read(biome_path, quiet = TRUE) %>% st_transform(st_crs(spatial_units))
  regions_sf <- st_read(regions_path, quiet = TRUE) %>% st_transform(st_crs(spatial_units))

  # Use point-on-surface for robust spatial join (avoids 1 cell matching 2 polygons)
  spatial_units_pts <- st_point_on_surface(spatial_units)

  # Join attributes to the points
  spatial_units_attr <- spatial_units_pts %>%
    st_join(biomes_sf %>% select(any_of(c("WWF_biome", "BIOME")))) %>%
    st_join(regions_sf %>% select(any_of(c("region_wb", "income_grp", "nev_name")))) %>%
    st_drop_geometry() %>%
    select(fid, any_of(c("WWF_biome", "region_wb", "income_grp", "nev_name"))) %>%
    distinct(fid, .keep_all = TRUE)

  # Join attributes back to the main grid geometry
  spatial_units <- left_join(spatial_units, spatial_units_attr, by = "fid")
  message("Attributes joined successfully.")
} else {
  warning("Raw attribute vectors not found. Skipping spatial join.")
}

# 5. Join the LCC metrics to the attributed grid
message("Joining LCC metrics to attributed grid...")
spatial_units_with_lcc <- left_join(spatial_units, lcc_metrics_wide, by = id_col_name)

# Check for list columns which cause st_write to fail (excluding geometry)
list_cols <- sapply(spatial_units_with_lcc, is.list)
geom_col <- attr(spatial_units_with_lcc, "sf_column")
bad_cols <- list_cols & names(spatial_units_with_lcc) != geom_col
if (any(bad_cols)) {
  stop("List columns detected in output (likely due to duplicate keys in pivot): ", paste(names(spatial_units_with_lcc)[bad_cols], collapse = ", "))
}

# NEW: Check for and repair invalid geometries before saving
if (any(!sf::st_is_valid(spatial_units_with_lcc))) {
  message("Warning: Found invalid geometries. Attempting to repair with st_make_valid().")
  spatial_units_with_lcc <- sf::st_make_valid(spatial_units_with_lcc)
}

# NEW: Check for and remove empty geometries
if (any(st_is_empty(spatial_units_with_lcc))) {
  message("Warning: Found ", sum(st_is_empty(spatial_units_with_lcc)), " empty geometries. Removing them before saving.")
  spatial_units_with_lcc <- spatial_units_with_lcc[!st_is_empty(spatial_units_with_lcc), ]
}

# 6. Save the final enriched GeoPackage
out_gpkg <- file.path(data_dir(), "processed", "10k_lcc_metrics.gpkg")

# Ensure output directory exists
if (!dir.exists(dirname(out_gpkg))) dir.create(dirname(out_gpkg), recursive = TRUE)

# Rename 'fid' to 'grid_fid' to avoid conflict with GeoPackage reserved primary key
if ("fid" %in% names(spatial_units_with_lcc)) {
  message("Renaming 'fid' column to 'grid_fid' to avoid GPKG driver conflict.")
  names(spatial_units_with_lcc)[names(spatial_units_with_lcc) == "fid"] <- "grid_fid"
}

message("Saving output to: ", out_gpkg)
sf::st_write(spatial_units_with_lcc, out_gpkg, append = FALSE, delete_dsn = TRUE)

message("LCC metric extraction complete.")

```

## Inspect Results

```{r summarize-results}

# Summarize by year and metric
summary_df <- lcc_m %>%
  group_by(year_step, Category) %>%
  summarize(across(c(Gain, Persistence, Loss, Quantity, Exchange, Shift),
                   mean, na.rm = TRUE), .groups = "drop")

# Preview
head(summary_df)
```

```{r summarize-by-groups}
#| message: false

# Define the grouping columns of interest
grouping_cols <- c("region_wb", "income_grp", "nev_name", "WWF_biome")
existing_groups <- grouping_cols[grouping_cols %in% names(spatial_units_with_lcc)]

if (length(existing_groups) > 0) {
  message("Summarizing LCC metrics by groups: ", paste(existing_groups, collapse = ", "))

  group_summaries <- list()

  for (g in existing_groups) {
    # Calculate mean metrics per group (assuming equal area grid, mean % is appropriate)
    # Use lcc_m (long format) joined with attributes from spatial_units to preserve structure
    attrs <- spatial_units %>%
      sf::st_drop_geometry() %>%
      dplyr::select(all_of(id_col_name), all_of(g))

    summary_g <- lcc_m %>%
      dplyr::left_join(attrs, by = id_col_name) %>%
      group_by(across(all_of(g)), year_step, Category) %>%
      summarize(across(any_of(c("Gain", "Persistence", "Loss", "Quantity", "Exchange", "Shift", "dir_ch")),
                       mean, na.rm = TRUE), .groups = "drop") %>%
      mutate(grouping_var = g) %>%
      rename(group_val = !!g)

    group_summaries[[g]] <- summary_g
  }

  all_group_summaries <- bind_rows(group_summaries)

  # Save the grouped summary
  out_csv <- file.path(data_dir(), "processed", "lcc_summary_by_group.csv")
  readr::write_csv(all_group_summaries, out_csv)
  message("Grouped summaries saved to: ", out_csv)

  print(head(all_group_summaries))
} else {
  message("No grouping columns found in spatial units.")
}
```

```{r}
#| label: summarize-results-old
#| eval: false
#| include: false

# Summarize by year and metric
summary_df <- lcc_m %>%
  group_by(year_step, Category) %>%
  summarize(across(c(Gain, Persistence, Loss, Quantity, Exchange, Shift),
                   mean, na.rm = TRUE), .groups = "drop")

# Preview
head(summary_df)
```

The previous pivoting step produces many columns with complex, machine-generated names (e.g., `Gain_2_1995_2000`). Instead of manually decoding them, the following chunk acts as a **dynamic data dictionary generator**. It programmatically parses those names to create a clear, human-readable reference table (`ref_df`). This ensures that our documentation of what each column represents stays automatically in sync with the data

```{r}
#| label: produce-reference-for-attributes

col_names <- names(lcc_metrics_wide)[!names(lcc_metrics_wide) %in% id_col_name]

# Parse components from column names using regex
ref_df <- tibble(
  column_name = col_names
) %>%
  mutate(
    component = str_extract(column_name, "^(Gain|Persistence|Loss|Quantity|Exchange|Shift)"),
    class = case_when(
      str_detect(column_name, "_1_") ~ "Transformed (class 1)",
      str_detect(column_name, "_2_") ~ "Natural (class 2)",
      str_detect(column_name, "_Overall_") ~ "Overall change",
      TRUE ~ "Unknown"
    ),
    year_step = str_extract(column_name, "[0-9]{4}_[0-9]{4}"),
    description = paste(component, "for", class, "during", year_step)
  )

# Optional: arrange for readability
ref_df <- ref_df %>% select(column_name, description)

```

## Plot Change Metrics Over Time

```{r}
#| label: plot-changes
ggplot(summary_df, aes(x = year_step, y = Gain, group = as.factor(Category), color = as.factor(Category))) +
  geom_line(size = 1) +
  geom_point() +
  labs(title = "Average Gain in Transformed vs Natural Land Over Time",
       x = "Year Step",
       y = "% Area Changed") +
  theme_minimal()
```

```{r}
#| label: plot-regional-summary
#| message: false
#| warning: false

# Visualization: Gross Loss of Natural Land by Region
if (exists("all_group_summaries") && "region_wb" %in% all_group_summaries$grouping_var) {
  plot_data <- all_group_summaries %>%
    dplyr::filter(grouping_var == "region_wb", Category == 2) # Category 2 = Natural

  p <- ggplot(plot_data, aes(x = reorder(group_val, Loss), y = Loss)) +
    geom_col(fill = "#d95f02", alpha = 0.8) +
    coord_flip() +
    labs(title = "Gross Loss of Natural Land (1992-2020)",
         subtitle = "Average % Loss per 10km Grid Cell by Region",
         x = NULL, y = "Gross Loss (%)") +
    theme_minimal() +
    theme(panel.grid.major.y = element_blank())

  print(p)
}
```

```{r}
#| label: map-net-change
#| message: false
#| warning: false

# Visualization: Map of Net Change in Natural Land (dir_ch_2)
# Negative = Net Loss, Positive = Net Gain
target_col <- "dir_ch_2_2020_1992"
lcc_gpkg <- file.path(data_dir(), "processed", "10k_lcc_metrics.gpkg")

if (file.exists(lcc_gpkg)) {
  # Read data
  lcc_sf <- st_read(lcc_gpkg, quiet = TRUE)

  if (target_col %in% names(lcc_sf)) {
    # Filter Antarctica for better global view if continent column exists
    if ("continent" %in% names(lcc_sf)) {
      lcc_sf <- lcc_sf %>%
        dplyr::filter(!continent %in% c("Antarctica", "Seven seas (Open Ocean)"))
    }

    # Plot
    p_map <- ggplot(lcc_sf) +
      geom_sf(aes(fill = .data[[target_col]]), color = NA) +
      scale_fill_gradient2(
        low = "#d73027",   # Red for Loss
        mid = "#f7f7f7",   # White for Stable
        high = "#1a9850",  # Green for Gain
        midpoint = 0,
        limits = c(-100, 100),
        oob = scales::squish,
        name = "Net Change (%)"
      ) +
      labs(
        title = "Net Change of Natural Cover (1992-2020)",
        subtitle = "Net Change = Gain - Loss (Class 2)"
      ) +
      theme_minimal() +
      theme(legend.position = "bottom")

    # Save
    out_png <- file.path(out_plots(), "map_net_change_natural.png")
    ggsave(out_png, p_map, width = 12, height = 6, bg = "white")

    # Display saved image to avoid rendering heavy vector graphics in notebook output
    knitr::include_graphics(out_png)
  } else {
    message("Column ", target_col, " not found. Check column names.")
  }
}
```

\`\`\`