---
title: "Land Cover Change Raster Generation"
author: "Jeronimo Rodriguez-Escobar"
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
editor: source
---

## Overview

This notebook generates specific land cover change rasters for visualization and downstream analysis.
Based on the 9-class simplified land cover maps (ESA CCI / C3S) defined in `LC_change.qmd`, we derive:

1.  **Natural Land Change**: Single raster (1=Loss, 2=Gain).
2.  **Forest Change**: Single raster (1=Loss, 2=Gain).
3.  **Cropland Change**: Single raster (1=Expansion, 2=Abandonment).
4.  **Urban Change**: Single raster (1=Expansion, 2=Decline).

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(terra)
library(dplyr)
library(here)
library(stringr)
library(devtools)

# Initialize paths
knitr::opts_knit$set(root.dir = here::here())
source(here::here("R", "paths.R"))
devtools::load_all(quiet = TRUE)

# Output directory for change rasters
out_dir <- file.path(data_dir(), "processed", "lc_change_rasters")
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
```

## Class Definitions

We use the simplified 9-class schema derived from the original 37 ESA CCI classes (see `LC_change.qmd`).

| ID | Class Name | Original ESA Classes |
|----|------------|----------------------|
| 1  | Agriculture| 10-40                |
| 2  | Forest     | 50-100               |
| 3  | Grassland  | 110-130              |
| 4  | Sparse     | 140-150              |
| 5  | Wetland    | 160-180              |
| 6  | Urban      | 190                  |
| 7  | Bare       | 200-202              |
| 8  | Water      | 210                  |
| 9  | Snow/Ice   | 220                  |

**Binary Classification (Natural vs Transformed):**
*   **Transformed**: Agriculture (1), Urban (6).
*   **Natural**: Forest (2), Grassland (3), Sparse (4), Wetland (5), Bare (7), Water (8), Snow/Ice (9).

## Load Data

We load the simplified 9-class rasters for 1992 and 2020.

```{r}
#| label: load-data

raw_lc_dir <- data_raw("LandCovers_Granular")
lc_1992_path <- file.path(raw_lc_dir, "landcover_gl_1992.tif")
lc_2020_path <- file.path(raw_lc_dir, "landcover_gl_2020.tif")

if (!file.exists(lc_1992_path) || !file.exists(lc_2020_path)) {
  stop("Simplified 9-class land cover rasters not found. Please run the reclassification step in 'analysis/LC_change.qmd' first.")
}

r92 <- terra::rast(lc_1992_path)
r20 <- terra::rast(lc_2020_path)

# Ensure alignment
if (!terra::compareGeom(r92, r20, stopOnError = FALSE)) {
  warning("Geometries differ. Resampling 2020 to 1992...")
  r20 <- terra::resample(r20, r92, method = "near")
}
```

## Generate Change Rasters

We generate binary rasters where **1** indicates the specific transition occurred, and **NA** indicates no change or other transitions.

```{r}
#| label: generate-rasters

# Define classes
natural_classes <- c(2, 3, 4, 5, 7, 8, 9)
transformed_classes <- c(1, 6)

# Helper function to calculate change based on reclassification
# Logic: Class 1 -> Class 2 = Change Type 1 (e.g., Loss/Expansion)
#        Class 2 -> Class 1 = Change Type 2 (e.g., Gain/Abandonment)
# We map Start Class to 10, End Class to 20.
# r2 (20) - r1 (10) = 10 -> Type 1
# r2 (10) - r1 (20) = -10 -> Type 2
calc_change_raster <- function(r1, r2, rcl_matrix, out_path, others_val = NA) {
  # Reclassify both years
  r1_c <- terra::classify(r1, rcl_matrix, others = others_val)
  r2_c <- terra::classify(r2, rcl_matrix, others = others_val)

  # Calculate difference
  diff <- r2_c - r1_c

  # Remap to final values: 10 -> 1, -10 -> 2, others -> NA
  rcl_final <- matrix(c(10, 1, -10, 2), ncol = 2, byrow = TRUE)
  out <- terra::classify(diff, rcl_final, others = NA)

  terra::writeRaster(out, out_path, overwrite = TRUE, gdal = c("COMPRESS=LZW", "TILED=YES"))
  return(invisible(NULL))
}

# 1. Natural Land Change (Natural=10, Transformed=20)
message("Generating Natural Land Change...")
rcl_nat <- matrix(c(2, 10, 3, 10, 4, 10, 5, 10, 7, 10, 8, 10, 9, 10, 1, 20, 6, 20), ncol = 2, byrow = TRUE)
calc_change_raster(r92, r20, rcl_nat, file.path(out_dir, "change_natural_land_1992_2020.tif"))

# 2. Forest Change (Forest=10, Non-Forest=20)
message("Generating Forest Change...")
calc_change_raster(r92, r20, matrix(c(2, 10), ncol=2), file.path(out_dir, "change_forest_1992_2020.tif"), others_val = 20)

# 3. Cropland Change (Non-Crop=10, Crop=20)
message("Generating Cropland Change...")
calc_change_raster(r92, r20, matrix(c(1, 20), ncol=2), file.path(out_dir, "change_cropland_1992_2020.tif"), others_val = 10)

# 4. Urban Change (Non-Urban=10, Urban=20)
message("Generating Urban Change...")
calc_change_raster(r92, r20, matrix(c(6, 20), ncol=2), file.path(out_dir, "change_urban_1992_2020.tif"), others_val = 10)

message("All rasters generated in: ", out_dir)
```

## Reference

*   **Natural Land Change**: 1 = Loss (Natural -> Transformed), 2 = Gain (Transformed -> Natural).
*   **Forest Change**: 1 = Loss (Forest -> Non-Forest), 2 = Gain (Non-Forest -> Forest).
*   **Cropland Change**: 1 = Expansion (Non-Crop -> Crop), 2 = Abandonment (Crop -> Non-Crop).
*   **Urban Change**: 1 = Expansion (Non-Urban -> Urban), 2 = Decline (Urban -> Non-Urban).