---
title: "Land Cover Change Metrics Workflow"
author: "Jeronimo Rodriguez-Escobar"
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
editor: source
---

---

This Quarto document shows how to:

1. Load land cover classification rasters and polygon units
2. Apply `extract_zonal_lcc_metrics()` and `iterate_zonal_lcc_metrics()`
3. Append the extracted metrics to the input polygon data.
4. Summarize and plot key land cover transition metrics [pending]


```{r}
#| label: setup
#| message: false
#| warning: false

library(terra)
library(sf)
library(dplyr)
library(ggplot2)
library(glue)
library(tidyr)
library(purrr)
library(diffeR)
library(here)
library(stringr)
library(parallel)
library(devtools)

# Initialize paths
knitr::opts_knit$set(root.dir = here::here())
source(here::here("R", "paths.R"))
devtools::load_all(quiet = TRUE)

# source the helper functions
# source(here::here("R", "utils_lcc_metrics.R")) # Should be loaded by load_all() if in R/
# source(here("R","pct_change.R")) # Should be loaded by load_all()
```

# Prepare data

**Data Sources:**

*   **1992-2015:** ESA Climate Change Initiative (CCI) Land Cover v2.0.7.
    *   *Citation:* Defourny, P., Lamarche, C., Brockmann, C., Boettcher, M., Bontemps, S., De Maet, T., Duveiller, G. L. Harper, K., Hartley A., Kirches, G., Moreau, I., Peylin, P., Ottlé, C., Radoux J., Van Bogaert, E., Ramoino, F., Albergel, C., and Arino, O.: Observed annual global land-use change from 1992 to 2020 three times more dynamic than reported by inventory-based statistics, in preparation, 2023.
    *   *User Guide:* Defourny, P., Lamarche, C., Bontemps, S., De Maet, T., Van Bogaert, E., Moreau, I., Brockmann, C., Boettcher, M., Kirches, G., Wevers, J., Santoro, M., Ramoino, F., & Arino, O. (2017). Land Cover Climate Change Initiative - Product User Guide v2. Issue 2.0. [Download PDF](http://maps.elie.ucl.ac.be/CCI/viewer/download/ESACCI-LC-Ph2-PUGv2_2.0.pdf).
    *   *Download:* [ESA CCI Viewer](https://maps.elie.ucl.ac.be/CCI/viewer/download.php)
*   **2016-2020:** Copernicus Climate Change Service (C3S) Global Land Cover v2.1.1.
    *   Consistent with the ESA CCI series.
    *   *Download:* Copernicus Climate Change Service (C3S) Climate Data Store (CDS).

**Processing:**
This workflow reclassifies the original LC maps into:
a) A simplified 9-class map (`landcover_gl_[year].tif`)
b) A binary Natural/Transformed map (`rec_landcover_gl_[year].tif`)

```{r}
#| label: reclassify-lc-maps
#| eval: false
#| include: false

# This chunk processes all available raw land cover data into reclassified versions.

# 1. Configuration
raw_lc_dir <- data_raw("LandCovers")
mask_poly <- st_read(data_vectors("hydrosheds_lv6_synth.gpkg"), quiet = TRUE)

# Reclassification matrix: Original ESA classes to 9 simplified classes
reclass_9_class_tbl <- data.frame(
  from = c(10, 11, 12, 20, 30, 40, 50, 60, 61, 62, 70, 71, 72, 80, 81, 82, 90, 100,
           110, 120, 121, 122, 130, 140, 150, 151, 152, 153, 160, 170, 180,
           190, 200, 201, 202, 210, 220),
  to = c(1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 3, 3, 3, 3, 3, 4, 4, 4, 4, 4, 5, 5, 5, 6, 7, 7, 7, 8, 9)
)

# Reclassification matrix: 9 simplified classes to 2 binary classes
reclass_2_class_tbl <- data.frame(
  from = c(3,4,5,6,7,8,9),
  to = c(2,2,2,1,2,2,2)
)

# 2. Helper function to process one raster layer
process_lc_raster <- function(r, year, out_dir, mask_poly, reclass_9, reclass_2) {
  message("Processing year: ", year)

  # --- 9-Class Reclassification ---
  out_9_path <- file.path(out_dir, paste0("landcover_gl_", year, ".tif"))
  if (!file.exists(out_9_path)) {
    r_9class <- subst(r, from = reclass_9$from, to = reclass_9$to)
    writeRaster(r_9class, out_9_path, overwrite = TRUE, gdal=c("COMPRESS=LZW", "TILED=YES"))
    message("  Saved 9-class: ", basename(out_9_path))
  } else {
    message("  9-class file already exists, skipping.")
    r_9class <- rast(out_9_path) # Load existing for next step
  }

  # --- 2-Class Reclassification ---
  out_2_path <- file.path(out_dir, paste0("rec_landcover_gl_", year, ".tif"))
  if (!file.exists(out_2_path)) {
    r_2class <- subst(r_9class, from = reclass_2$from, to = reclass_2$to)
    r_2class_masked <- mask(r_2class, mask_poly)
    writeRaster(r_2class_masked, out_2_path, overwrite = TRUE, gdal=c("COMPRESS=LZW", "TILED=YES"))
    message("  Saved 2-class: ", basename(out_2_path))
  } else {
    message("  2-class file already exists, skipping.")
  }
}

# 3. Process the 1992-2015 multi-band stack
stack_v207_path <- file.path(raw_lc_dir, "ESACCI-LC-L4-LCCS-Map-300m-P1Y-1992_2015-v2.0.7.tif")
if (file.exists(stack_v207_path)) {
  message("Processing 1992-2015 stack...")
  r_stack <- rast(stack_v207_path)
  # Extract years from band names, e.g., "ESACCI-LC-L4-LCCS-Map-300m-P1Y-1992-v2.0.7"
  years <- as.integer(stringr::str_extract(names(r_stack), "\\d{4}"))

  for (i in 1:nlyr(r_stack)) {
    process_lc_raster(r_stack[[i]], years[i], raw_lc_dir, mask_poly, reclass_9_class_tbl, reclass_2_class_tbl)
  }
}

# 4. Process the 2016-2020 individual files
message("Processing 2016-2020 individual files...")
years_v211 <- 2016:2020
for (year in years_v211) {
  # First, convert NetCDF to GeoTIFF if needed
  nc_file <- file.path(raw_lc_dir, sprintf("C3S-LC-L4-LCCS-Map-300m-P1Y-%d-v2.1.1.nc", year))
  tif_file <- file.path(raw_lc_dir, sprintf("C3S-LC-L4-LCCS-Map-300m-P1Y-%d-v2.1.1.tif", year))
  if (!file.exists(tif_file) && file.exists(nc_file)) {
    message("Converting ", basename(nc_file), " to GeoTIFF...")
    cmd <- sprintf(
      "gdalwarp -of Gtiff -co COMPRESS=LZW -co TILED=YES -ot Byte -te -180 -90 180 90 -tr 0.002777777777778 0.002777777777778 -t_srs EPSG:4326 NETCDF:\"%s\":lccs_class \"%s\"",
      nc_file, tif_file
    )
    system(cmd)
  }

  # Now process the GeoTIFF
  if (file.exists(tif_file)) {
    r <- rast(tif_file)
    process_lc_raster(r, year, raw_lc_dir, mask_poly, reclass_9_class_tbl, reclass_2_class_tbl)
  } else {
    warning("Could not find source file for year ", year)
  }
}
message("Reclassification process complete.")
```

## Load Inputs

```{r}
#| label: load-data

# Example file paths (replace with actual paths)
inpath <- data_raw("LandCovers")
landcover_files <- list.files(
  inpath,
  pattern = "rec.*\\.tif$",
  full.names = TRUE
)
# Define the specific years for the analysis. This is more robust than index-based selection.
years_to_process <- c(1992, 2020)

# Create a regex pattern to match any of the desired years in the filename.
year_pattern <- paste(years_to_process, collapse = "|")

# Filter the file list to include only files for the desired years.
landcover_files <- landcover_files[stringr::str_detect(basename(landcover_files), year_pattern)]
# Name rasters by year
names(landcover_files) <- stringr::str_extract(basename(landcover_files), "\\d{4}")
# Load rasters as a list
landcover_rasters <- lapply(landcover_files, terra::rast)

# Load canonical spatial units
# We use the 10km grid as the primary unit. Other available vectors in vector_basedata include:
# - Biome.gpkg (WWF Biomes)
# - cartographic_ee_r264_correpondence.gpkg (Regions/Countries)

# Load the raw canonical 10km grid (geometry only) to keep it lightweight
grid_path <- data_vectors("AOOGrid_10x10km_land_4326_clean.gpkg")
spatial_units <- st_read(grid_path, quiet = TRUE)

# Ensure we have a valid ID column (usually 'fid' in the canonical grid)
if (!"fid" %in% names(spatial_units)) {
  spatial_units$fid <- seq_len(nrow(spatial_units))
}
spatial_units <- spatial_units %>%
  dplyr::select(fid) # Keep only the ID for the extraction step
id_col_name <- "fid"
```

## Run Land Cover Change Metric Extraction

```{r}
#| label: run-lcc-extraction
#| message: true
#| warning: false

# This chunk runs the core land cover change metric extraction using the `diffeR`
# package logic wrapped in our helper functions.

# 1. Iterate over all time steps defined by the input raster list.
#    This calculates Gain, Loss, Persistence, etc. for each polygon in `spatial_units`.
message("Running LCC metric extraction for all time steps...")
lcc_m <- iterate_lcc_metrics(
  landcover_rasters,
  spatial_units,
  id_col = id_col_name,
  percent = TRUE,
  verbose = TRUE,
  mc = TRUE,
  ncores = 8,
  digits = 4
)

# 2. Calculate net change direction (Gain - Loss) for easier interpretation.
lcc_m <- lcc_m %>%
  mutate(
    Gain = as.numeric(Gain),
    Loss = as.numeric(Loss),
    dir_ch = round(Gain - Loss, digits = 4)
  )

# 3. Pivot the long-format metrics table to a wide format, where each
#    metric for each class and time step becomes a separate column.
message("Pivoting results to wide format...")
lcc_metrics_wide <- lcc_m %>%
  tidyr::pivot_wider(
    id_cols = all_of(id_col_name),
    names_from = c(Category, year_step),
    values_from = c(Gain, Persistence, Loss, Quantity, Exchange, dir_ch),
    names_glue = "{.value}_{Category}_{year_step}"
  ) %>%
  mutate(
    # Ensure the ID column type is correct for joining
    !!id_col_name := as.double(.data[[id_col_name]])
  )

# 4. Attribute the grid with Biomes and Regions (Spatial Join)
#    We do this NOW using the raw vectors, instead of relying on the pre-processed ES file.
message("Attributing grid with Biomes and Regions...")

biome_path <- data_vectors("Biome.gpkg")
# Use pattern matching to find the correspondence file (handles spelling variations)
ee_files <- list.files(data_vectors(), pattern = "cartographic_ee.*\\.gpkg$", full.names = TRUE)
regions_path <- if(length(ee_files) > 0) ee_files[1] else ""

if (file.exists(biome_path) && file.exists(regions_path)) {
  # Load vectors
  biomes_sf <- st_read(biome_path, quiet = TRUE) %>% st_transform(st_crs(spatial_units))
  regions_sf <- st_read(regions_path, quiet = TRUE) %>% st_transform(st_crs(spatial_units))

  # Use point-on-surface for robust spatial join (avoids 1 cell matching 2 polygons)
  spatial_units_pts <- st_point_on_surface(spatial_units)

  # Join attributes to the points
  spatial_units_attr <- spatial_units_pts %>%
    st_join(biomes_sf %>% select(any_of(c("WWF_biome", "BIOME")))) %>%
    st_join(regions_sf %>% select(any_of(c("iso3", "region_wb", "income_grp", "nev_name")))) %>%
    st_drop_geometry() %>%
    select(fid, any_of(c("WWF_biome", "iso3", "region_wb", "income_grp", "nev_name")))

  # Join attributes back to the main grid geometry
  spatial_units <- left_join(spatial_units, spatial_units_attr, by = "fid")
  message("Attributes joined successfully.")
} else {
  warning("Raw attribute vectors not found. Skipping spatial join.")
}

# 5. Join the LCC metrics to the attributed grid
message("Joining LCC metrics to attributed grid...")
spatial_units_with_lcc <- left_join(spatial_units, lcc_metrics_wide, by = id_col_name)

# 6. Save the final enriched GeoPackage
out_gpkg <- file.path(data_dir(), "processed", "10k_lcc_metrics.gpkg")
message("Saving output to: ", out_gpkg)
sf::st_write(spatial_units_with_lcc, out_gpkg, append = FALSE, delete_dsn = TRUE)

message("LCC metric extraction complete.")

```

## Inspect Results

```{r summarize-results}
#| label: summarize-results

# Summarize by year and metric
summary_df <- lcc_m %>%
  group_by(year_step, Category) %>%
  summarize(across(c(Gain, Persistence, Loss, Quantity, Exchange, Shift),
                   mean, na.rm = TRUE), .groups = "drop")

# Preview
head(summary_df)
```

```{r summarize-by-groups}
#| label: summarize-by-groups
#| message: false

# Define the grouping columns of interest
grouping_cols <- c("region_wb", "income_grp", "nev_name", "WWF_biome")
existing_groups <- grouping_cols[grouping_cols %in% names(spatial_units_with_lcc)]

if (length(existing_groups) > 0) {
  message("Summarizing LCC metrics by groups: ", paste(existing_groups, collapse = ", "))

  group_summaries <- list()

  for (g in existing_groups) {
    # Calculate mean metrics per group (assuming equal area grid, mean % is appropriate)
    summary_g <- spatial_units_with_lcc %>%
      sf::st_drop_geometry() %>%
      group_by(across(all_of(g)), year_step, Category) %>%
      summarize(across(c(Gain, Persistence, Loss, Quantity, Exchange, Shift, dir_ch),
                       mean, na.rm = TRUE), .groups = "drop") %>%
      mutate(grouping_var = g) %>%
      rename(group_val = !!g)

    group_summaries[[g]] <- summary_g
  }

  all_group_summaries <- bind_rows(group_summaries)

  # Save the grouped summary
  out_csv <- file.path(data_dir(), "processed", "lcc_summary_by_group.csv")
  readr::write_csv(all_group_summaries, out_csv)
  message("Grouped summaries saved to: ", out_csv)

  print(head(all_group_summaries))
} else {
  message("No grouping columns found in spatial units.")
}
```

```{r}
#| label: summarize-results-old
#| eval: false
#| include: false

# Summarize by year and metric
summary_df <- lcc_m %>%
  group_by(year_step, Category) %>%
  summarize(across(c(Gain, Persistence, Loss, Quantity, Exchange, Shift),
                   mean, na.rm = TRUE), .groups = "drop")

# Preview
head(summary_df)
```

The previous pivoting step produces many columns with complex, machine-generated names (e.g., `Gain_2_1995_2000`). Instead of manually decoding them, the following chunk acts as a **dynamic data dictionary generator**. It programmatically parses those names to create a clear, human-readable reference table (`ref_df`). This ensures that our documentation of what each column represents stays automatically in sync with the data

```{r}
#| label: produce-reference-for-attributes

col_names <- names(lcc_metrics_wide)[!names(lcc_metrics_wide) %in% id_col_name]

# Parse components from column names using regex
ref_df <- tibble(
  column_name = col_names
) %>%
  mutate(
    component = str_extract(column_name, "^(Gain|Persistence|Loss|Quantity|Exchange|Shift)"),
    class = case_when(
      str_detect(column_name, "_1_") ~ "Transformed (class 1)",
      str_detect(column_name, "_2_") ~ "Natural (class 2)",
      str_detect(column_name, "_Overall_") ~ "Overall change",
      TRUE ~ "Unknown"
    ),
    year_step = str_extract(column_name, "[0-9]{4}_[0-9]{4}"),
    description = paste(component, "for", class, "during", year_step)
  )

# Optional: arrange for readability
ref_df <- ref_df %>% select(column_name, description)

```
## Plot Change Metrics Over Time

```{r}
#| label: plot-changes
ggplot(summary_df, aes(x = year_step, y = Gain, group = as.factor(Category), color = as.factor(Category))) +
  geom_line(size = 1) +
  geom_point() +
  labs(title = "Average Gain in Transformed vs Natural Land Over Time",
       x = "Year Step",
       y = "% Area Changed") +
  theme_minimal()
```