---
title: "Granular Land Cover Change Analysis"
subtitle: "Forest Loss & Urban/Cropland Expansion Models"
author: "Jeronimo Rodriguez-Escobar"
date: last-modified
format:
  html:
    toc: true
    theme: cosmo
editor: source
---

## Overview

This document implements a more detailed Land Cover Change (LCC) analysis, moving beyond the simple "Natural vs. Transformed" binary model. It runs two separate, policy-relevant analyses based on the `diffeR` methodology (Pontius et al.), which calculates detailed transition metrics like Gain, Loss, Persistence, Quantity, Exchange, and Shift.

1.  **Forest Loss Model:** Reclassifies land cover into a binary `Forest` vs. `Non-Forest` map to specifically track and quantify forest dynamics.
2.  **Urban & Cropland Expansion Model:** Reclassifies land cover into a three-class map (`Cropland`, `Urban`, `Other`) to quantify the expansion of anthropogenic land uses.

The metrics are extracted for the canonical 10km grid and saved to a new GeoPackage for integration with the main hotspot analysis.

```{r}
#| label: setup
#| message: false
#| warning: false

library(terra)
library(sf)
library(dplyr)
library(ggplot2)
library(glue)
library(tidyr)
library(purrr)
library(diffeR)
library(here)
library(stringr)
library(devtools)

# Initialize paths and load helper functions
knitr::opts_knit$set(root.dir = here::here())
source(here::here("R", "paths.R"))
devtools::load_all(quiet = TRUE)
```

## 1. Load Input Data

This analysis relies on simplified 9-class land cover maps (`landcover_gl_[year].tif`) prepared by `analysis/LC_change_preparation.qmd`.

```{r}
#| label: load-data

# Define years for analysis
years_to_process <- c(1992, 2020)

# Path to pre-processed 9-class rasters
lc_dir_granular <- data_raw("LandCovers_Granular")
if (!dir.exists(lc_dir_granular)) {
  stop("Granular LC directory not found. Run analysis/LC_change_preparation.qmd first.")
}

# Find the files for the specified years
landcover_files <- list.files(
  lc_dir_granular,
  pattern = paste0("landcover_gl_(", paste(years_to_process, collapse = "|"), ")\\.tif$"),
  full.names = TRUE
)

# Check if files were found
if (length(landcover_files) != length(years_to_process)) {
  stop("Could not find all required 9-class land cover rasters for years: ", paste(years_to_process, collapse = ", "))
}

# Load rasters into a named list
names(landcover_files) <- stringr::str_extract(basename(landcover_files), "\\d{4}")
landcover_rasters_9class <- lapply(landcover_files, terra::rast)

message("Loaded 9-class rasters for years: ", paste(names(landcover_rasters_9class), collapse = ", "))

# Load canonical 10km grid
grid_path <- data_vectors("AOOGrid_10x10km_land_4326_clean.gpkg")
spatial_units <- st_read(grid_path, quiet = TRUE)

if (!"fid" %in% names(spatial_units)) {
  spatial_units$fid <- seq_len(nrow(spatial_units))
}
spatial_units <- spatial_units %>% dplyr::select(fid)
id_col_name <- "fid"

# --- TESTING MODE ---
# Set to TRUE to run on a small subset for verification.
TEST_MODE <- FALSE
if (TEST_MODE) {
  message("!!! RUNNING IN TEST MODE (subset of 1000 contiguous cells) !!!")
  # Select a contiguous block of cells to optimize raster I/O during testing
  set.seed(123)
  start_idx <- sample(seq_len(nrow(spatial_units) - 1000), 1)
  spatial_units <- spatial_units %>% dplyr::slice(start_idx:(start_idx + 999))
}

message("Loaded ", nrow(spatial_units), " spatial units from the 10km grid.")
```

## 2. Define and Run LCC Models

We define a helper function to run the analysis for a given reclassification scheme. This avoids code duplication and makes the process modular. The function reclassifies the input rasters, runs the chunked `iterate_lcc_metrics` process, and returns the results in a tidy, long format.

```{r}
#| label: define-analysis-runner

run_granular_lcc_analysis <- function(model_name, rasters_9class, reclass_matrix, class_labels, spatial_units, id_col) {
  message(glue::glue("\n--- Starting Granular LCC Analysis: {model_name} ---"))

  # 1. Reclassify rasters based on the model's matrix
  message("Reclassifying rasters...")
  reclassified_rasters <- list()
  for (year in names(rasters_9class)) {
    reclassified_rasters[[year]] <- terra::classify(rasters_9class[[year]], reclass_matrix)
  }

  # 2. Run LCC metric extraction in chunks
  chunk_size <- 50000
  n_total <- nrow(spatial_units)
  n_chunks <- ceiling(n_total / chunk_size)
  chunk_dir <- file.path(data_dir(), "processed", "lcc_chunks", model_name)
  # Isolate test chunks to prevent them from blocking a full run later
  if (exists("TEST_MODE") && TEST_MODE) {
    chunk_dir <- paste0(chunk_dir, "_test")
  }
  if (!dir.exists(chunk_dir)) dir.create(chunk_dir, recursive = TRUE)

  message(glue::glue("Processing {n_total} units in {n_chunks} chunks..."))

  lcc_m_list <- list()
  for (i in 1:n_chunks) {
    chunk_file <- file.path(chunk_dir, paste0("lcc_chunk_", i, ".rds"))

    if (file.exists(chunk_file)) {
      message(glue::glue("Chunk {i}/{n_chunks} exists. Loading..."))
      lcc_m_list[[i]] <- readRDS(chunk_file)
    } else {
      message(glue::glue("Processing chunk {i}/{n_chunks}..."))
      idx_start <- (i - 1) * chunk_size + 1
      idx_end <- min(i * chunk_size, n_total)
      chunk_units <- spatial_units[idx_start:idx_end, ]

      chunk_res <- iterate_lcc_metrics(
        reclassified_rasters,
        chunk_units,
        id_col = id_col,
        percent = TRUE,
        mc = TRUE,
        ncores = 8,
        digits = 4
      )
      saveRDS(chunk_res, chunk_file)
      lcc_m_list[[i]] <- chunk_res
      rm(chunk_units, chunk_res); gc()
    }
  }

  # 3. Assemble and label results
  message("Assembling and labeling results...")
  lcc_m <- dplyr::bind_rows(lcc_m_list) %>%
    mutate(
      Category = recode(Category, !!!class_labels),
      model = model_name
    )

  message(glue::glue("--- Finished: {model_name} ---"))
  return(lcc_m)
}
```

### Model 1: Forest Loss

This model uses a binary classification to isolate changes related to forest cover.

```{r}
#| label: run-forest-model

# Reclassification: 9-class -> Forest (1) vs. Non-Forest (2)
forest_reclass_matrix <- matrix(c(
  1, 2, # Cropland -> Non-Forest
  2, 1, # Forest -> Forest
  3, 2, # Grassland -> Non-Forest
  4, 2, # Sparse -> Non-Forest
  5, 2, # Wetland -> Non-Forest
  6, 2, # Urban -> Non-Forest
  7, 2, # Bare -> Non-Forest
  8, 2, # Water -> Non-Forest
  9, 2  # Snow/Ice -> Non-Forest
), ncol = 2, byrow = TRUE)

forest_class_labels <- c("1" = "Forest", "2" = "NonForest")

forest_results <- run_granular_lcc_analysis(
  model_name = "ForestLoss",
  rasters_9class = landcover_rasters_9class,
  reclass_matrix = forest_reclass_matrix,
  class_labels = forest_class_labels,
  spatial_units = spatial_units,
  id_col = id_col_name
)
```

### Model 2: Urban & Cropland Expansion

This model uses a three-way classification to track expansion of key anthropogenic land uses.

```{r}
#| label: run-expansion-model

# Reclassification: 9-class -> Cropland (1), Urban (2), Other (3)
expansion_reclass_matrix <- matrix(c(
  1, 1, # Cropland -> Cropland
  2, 3, # Forest -> Other
  3, 3, # Grassland -> Other
  4, 3, # Sparse -> Other
  5, 3, # Wetland -> Other
  6, 2, # Urban -> Urban
  7, 3, # Bare -> Other
  8, 3, # Water -> Other
  9, 3  # Snow/Ice -> Other
), ncol = 2, byrow = TRUE)

expansion_class_labels <- c("1" = "Cropland", "2" = "Urban", "3" = "Other")

expansion_results <- run_granular_lcc_analysis(
  model_name = "Expansion",
  rasters_9class = landcover_rasters_9class,
  reclass_matrix = expansion_reclass_matrix,
  class_labels = expansion_class_labels,
  spatial_units = spatial_units,
  id_col = id_col_name
)
```

## 3. Consolidate and Export Results

The results from both models are combined and pivoted into a wide format, creating one column for each metric, class, and model. This final dataset is then joined to the grid geometry and saved.

```{r}
#| label: consolidate-and-save

message("Combining results from all models...")
all_results_long <- bind_rows(forest_results, expansion_results)

# Pivot to wide format
message("Pivoting to wide format...")
lcc_granular_wide <- all_results_long %>%
  # Calculate net change for easier interpretation
  mutate(dir_ch = round(as.numeric(Gain) - as.numeric(Loss), digits = 4)) %>%
  tidyr::pivot_wider(
    id_cols = all_of(id_col_name),
    names_from = c(model, Category, year_step),
    values_from = c(Gain, Persistence, Loss, Quantity, Exchange, Shift, dir_ch),
    names_glue = "{model}_{.value}_{Category}_{year_step}"
  ) %>%
  mutate(
    !!id_col_name := as.double(.data[[id_col_name]])
  )

# Join metrics to the spatial grid
message("Joining metrics to spatial grid...")
spatial_units_with_granular_lcc <- left_join(spatial_units, lcc_granular_wide, by = id_col_name)

# Check for and repair invalid geometries
if (any(!sf::st_is_valid(spatial_units_with_granular_lcc))) {
  message("Warning: Found invalid geometries. Repairing with st_make_valid().")
  spatial_units_with_granular_lcc <- sf::st_make_valid(spatial_units_with_granular_lcc)
}

# Save the final enriched GeoPackage
out_gpkg <- file.path(data_dir(), "processed", "10k_lcc_granular_metrics.gpkg")
if (exists("TEST_MODE") && TEST_MODE) {
  out_gpkg <- file.path(data_dir(), "processed", "10k_lcc_granular_metrics_test.gpkg")
}
if (!dir.exists(dirname(out_gpkg))) dir.create(dirname(out_gpkg), recursive = TRUE)

# Rename 'fid' to 'grid_fid' to avoid conflict with GeoPackage reserved primary key
if ("fid" %in% names(spatial_units_with_granular_lcc)) {
  message("Renaming 'fid' column to 'grid_fid' to avoid GPKG driver conflict.")
  names(spatial_units_with_granular_lcc)[names(spatial_units_with_granular_lcc) == "fid"] <- "grid_fid"
}

message("Saving output to: ", out_gpkg)
sf::st_write(spatial_units_with_granular_lcc, out_gpkg, append = FALSE, delete_dsn = TRUE)

message("Granular LCC metric extraction complete.")

# Optional: Save a summary CSV for quick inspection
summary_by_class <- all_results_long %>%
  group_by(model, year_step, Category) %>%
  summarize(across(c(Gain, Persistence, Loss, Quantity, Exchange, Shift),
                   mean, na.rm = TRUE), .groups = "drop")

out_csv <- file.path(data_dir(), "processed", "lcc_granular_summary.csv")
readr::write_csv(summary_by_class, out_csv)
message("Saved summary table to: ", out_csv)

print(head(summary_by_class))
```