---
title: "Legacy: ID Grid Assignment"
author: "Jerónimo Rodríguez-Escobar"
status: "Archived / Superseded"
format: html
editor: source
---

::: callout-important
## DEPRECATION NOTICE

This workflow has been superseded by the logic in `Consolidation.qmd` and the standard `AOOGrid_10x10km_land.gpkg` preparation.

**Why was this replaced?**
- This script used a rasterization-based approach to assign country IDs to grid cells.
- The current pipeline uses spatial joins (`st_join`) with the canonical vector grid.
- This file is preserved for archival purposes.
:::

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE) # Disable execution globally
```

# 1. Join Country Data (improve Analysis)

## 1.1. Rasterize Country Data to the ESA LC

```{python add country data}
#| echo: true
import geopandas as gpd
import rasterio
from rasterio import features
from rasterio.windows import Window
import numpy as np
from tqdm import tqdm  # <-- add this

# -------------------------------------
# CONFIG
# -------------------------------------
countries_fp = "/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_ES_TimeSeries/vector/cartographic_ee_ee_r264_correspondence.gpkg"
esa_raster_fp = "/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_NCP/data/input_rasters/LandCovers/ESACCI-LC-L4-LCCS-Map-300m-P1Y-2020-v2.1.1_md5_2ed6285e6f8ec1e7e0b75309cc6d6f9f.tif"
output_raster_fp = "/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/Ref_Data/Raster/countries_fid_300m_COMPRESSED.tif"
updated_countries_fp = "/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/Ref_Data/vector/ee_r264_correspondence_fid.gpkg"

# -------------------------------------
# 1. Load Countries & Add fid Column
# -------------------------------------
countries = gpd.read_file(countries_fp).reset_index(drop=True)
countries["fid"] = countries.index  # Explicit integer ID

# Save updated vector file for future use
countries[
    ["fid", *[c for c in countries.columns if c not in ["fid", "geometry"]], "geometry"]
].to_file(updated_countries_fp, driver="GPKG")

print(f"✓ Updated countries saved with fid column to {updated_countries_fp}")

# Keep only fid + geometry for rasterization
countries = countries[["fid", "geometry"]]

# -------------------------------------
# 2. Load reference raster
# -------------------------------------
with rasterio.open(esa_raster_fp) as ref_raster:
    meta = ref_raster.meta.copy()
    transform = ref_raster.transform
    crs = ref_raster.crs
    height, width = ref_raster.height, ref_raster.width

# -------------------------------------
# 3. Reproject countries to raster CRS
# -------------------------------------
countries = countries.to_crs(crs)

# -------------------------------------
# 4. Prepare shapes for rasterization
# -------------------------------------
shapes = [(geom, int(fid)) for fid, geom in zip(countries["fid"], countries.geometry)]

# Update metadata to use compression
meta.update(
    dtype=rasterio.int32,
    count=1,
    nodata=-1,
    compress="DEFLATE",  # lossless compression
    predictor=2,         # improves compression for continuous data
    zlevel=6             # compression level (1–9, trade-off size/speed)
)

with rasterio.open(output_raster_fp, "w", **meta) as dst:
    tile_height, tile_width = 2000, 2000  # Adjust chunk size

    for row_off in tqdm(range(0, height, tile_height)):
        for col_off in range(0, width, tile_width):
            win = Window(
                col_off,
                row_off,
                min(tile_width, width - col_off),
                min(tile_height, height - row_off),
            )

            win_transform = rasterio.windows.transform(win, transform)

            raster_chunk = features.rasterize(
                shapes=shapes,
                out_shape=(win.height, win.width),
                transform=win_transform,
                fill=-1,
                dtype=np.int32,
            )

            dst.write(raster_chunk, 1, window=win)

print(f"✓ Compressed raster saved to {output_raster_fp}")

```

## 1.2. Assign Country Fid to the Grid

```{python assign ID}

import geopandas as gpd
import rasterio
from rasterio import features
from rasterio.transform import from_origin
import numpy as np
from pathlib import Path

# -----------------------------
# CONFIG
# -----------------------------
countries_fp = "/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_ES_TimeSeries/vector/cartographic_ee_ee_r264_correspondence.gpkg"
output_raster_fp = "/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/Ref_Data/Raster/countries_fid_5km_COMPRESSED.tif"
updated_countries_fp = "/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/Ref_Data/vector/ee_r264_correspondence_fid.gpkg"

# -----------------------------
# 1. Load countries & add fid
# -----------------------------
countries = gpd.read_file(countries_fp).reset_index(drop=True)
countries["fid"] = countries.index  # Explicit integer ID

# Save updated version (so fid is always available in future runs)
# countries.to_file(updated_countries_fp, driver="GPKG")
# print(f"✓ Updated countries saved with fid column to {updated_countries_fp}")

# Keep only necessary columns
countries = countries[["fid", "geometry"]]

# -----------------------------
# 2. Create 10 km reference raster (WGS84)
# -----------------------------
xmin, ymin, xmax, ymax = -180, -90, 180, 90
resolution = 0.05  # ~10 km at equator
width = int((xmax - xmin) / resolution)
height = int((ymax - ymin) / resolution)

transform = from_origin(xmin, ymax, resolution, resolution)

meta = {
    "driver": "GTiff",
    "dtype": rasterio.int32,
    "count": 1,
    "width": width,
    "height": height,
    "crs": "EPSG:4326",
    "transform": transform,
    "nodata": -1,
    "compress": "DEFLATE",  # lossless compression
    "predictor": 2,
    "zlevel": 6
}

# -----------------------------
# 3. Rasterize
# -----------------------------
shapes = [(geom, int(fid)) for fid, geom in zip(countries["fid"], countries.geometry)]

with rasterio.open(output_raster_fp, "w", **meta) as dst:
    rasterized = features.rasterize(
        shapes=shapes,
        out_shape=(height, width),
        transform=transform,
        fill=-1,
        dtype=np.int32
    )
    dst.write(rasterized, 1)

print(f"✓ Compressed 10 km raster saved to {output_raster_fp}")


```

## 1.3 Assign Id to 10 km grid

```{python alicate ids to grid}

import geopandas as gpd
import rasterio
import pandas as pd

# Paths
grid_fp = "/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/Ref_Data/vector/grid10_km_c.gpkg"
raster_fp = "/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/Ref_Data/Raster/countries_fid_5km_COMPRESSED.tif"
output_fp = "/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/Ref_Data/vector/grid10_km_c2.gpkg"
# Load grid
grid = gpd.read_file(grid_fp)

# Compute centroids
grid_centroids = grid.copy()
grid_centroids["geometry"] = grid_centroids.centroid

# Open raster
with rasterio.open(raster_fp) as src:
    coords = [(geom.x, geom.y) for geom in grid_centroids.geometry]
    values = list(src.sample(coords))
    values = [int(v[0]) if v[0] != src.nodata else None for v in values]

# Add country_fid column
grid["country_fid"] = values

# Save
grid.to_file(output_fp, driver="GPKG")
print(f"✓ Grid with country_fid saved to {output_fp}")
```

# Final Edits/cleaning

Some fixzes to column nam,es because of course

```{r wrf}

grid10 <- st_read("/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/Ref_Data/vector/grid10_km_c.gpkg")


sf_f <- st_read(paste0(inpath,'/', '10k_grid_ES_change_benef.gpkg'))
sf_f <- sf_f %>% mutate(fid = dplyr::row_number())
grid10 <- grid10 %>% mutate(fid= dplyr::row_number())

grid10 <- grid10 %>% filter(!is.na(country_fid))
grid10 <- st_drop_geometry(grid10)

sf_f <- left_join(sf_f, grid10)
sf_f <- rename(sf_f, c_fid = country_fid)

st_write(sf_f, paste0(inpath, '/', '10k_grid_ES_change_benef.gpkg'), append=FALSE)
st_write(grid10, "/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/Ref_Data/vector/grid10_km_c.gpkg", append=FALSE)


```

Not RUN. This is is an earlier approach that Becky told me ( i don't remember the exact reason) was not great). Rasterizing to the ESA resoluton and the using exactextractr. That's why i did not use that at the end.

```{r add c protection}
#| eval: false
#| include: false
# Update: include the coastal protection data:
sf_cp <- st_read('/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_NCP/summary_pipeline_workspace/grid_10k_synth_zonal_2025_07_20_23_45_27.gpkg')

# Define name replacements per object. Had to do it manually, there is just too much variability to automatize, did it by hand, but this is the kind of thing an LLM helps to write easily. Still, there must be a smarter, more flexible way to do it (maybe the best approach is to keep a consistent naming convention for the modeled rasters, this has been a PIA since i started this work)
sf_cp <- sf_cp %>% mutate(fid = dplyr::row_number())
# Calculate bitemporal change in % and absolute terms. Drop original columns
sf_cp <- compute_variable_change(sf_cp, suffix = c("_mean"), change_type= "both", drop_columns = TRUE)

sf_cp <- st_drop_geometry(sf_cp)
# load previous version
sf_f <- st_read(paste0(inpath, '/', '10k_grid_ES_change_benef.gpkg'))
sf_f <-  sf_f %>% mutate(fid = dplyr::row_number()) # not sure why fid is missing. Nevermind, added here
sf_f <- left_join(sf_f, sf_cp)
# reorganize columns:

sf_f <- sf_f %>%
  relocate(
    coastal_protection_Rt_abs_chg,
    coastal_protection_Rt_pct_chg,
    .before = GHS_BUILT_S_E2020_mean
  )
```

### 1.1.2 Include Coastal Protection (QGIS)

This approach was not great, because I did the calculation on QGIS Here are the instructions:

Option 2: QGIS (no coding)

Use the “Join attributes by location (summary)” tool:

```
Input layer: your cp2

Join layer: your points

Geometric predicate: intersects (or within, if points are entirely inside)

Choose the column to summarize

Statistics to calculate: check Mean and Max

Run and save the result
```

Quite easy, but doing this leaves no record, abd I just wasted two days trying to remember where the hell was I, so it may be better to run it here so we have ALL the steps on. single location