---
title: "Data Preparation"
subtitle: "v1.3.4"
author: "Jerónimo Rodríguez Escobar"
date: last-modified
date-format: "YYYY-MM-DD HH:mm zzz"
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
 # pdf:
 #   toc: true
 #   number-sections: true
 #   geometry: margin=1in
editor: source
---

```{r setup, message=FALSE, warning=FALSE}
library(terra)
library(sf)
library(dplyr)
library(ggplot2)
library(glue)
library(tidyr)
library(purrr)
library(diffeR)
library(here)
library(stringr)
library(tidytext)
library(rlang)
library(tidyr)
library(forcats)
library(scales)
library(RColorBrewer)
library(htmltools)
library(leaflet)
library(devtools)
library(reticulate)
library(exactextractr)
library(httpgd)
load_all()

# set venv for Python
use_virtualenv("/home/jeronimo/venvs/coastal_snap_env", required = TRUE)

# Centralize data roots (set GLOBAL_NCP_DATA in ~/.Renviron)
data_root <- data_dir()
interim_dir <- data_interim()
processed_dir <- data_processed()

processed_tables_dir <- file.path(processed_dir, "tables")
processed_hotspots_dir <- file.path(processed_dir, "hotspots")
processed_coastal_dir <- file.path(processed_dir, "coastal")

# Source unified hotspot/change helpers
source(here::here("R", "utils_hotspot.R"))

# Legacy locations (kept for tracking):
# inpath <- "/Users/rodriguez/Library/CloudStorage/OneDrive-WorldWildlifeFund,Inc/global_NCP/output_data"
# inpath <- "/home/jeronimo/OneDrive/PROJECTS/Global_NCP/Final_outputs"
inpath <- processed_dir
# This maps raw column names from synth outputs to canonical names.
rename_list <- list(
  c(
    usle_1992_mean = "global_usle_marine_mod_ESA_1992_mean",
    usle_2020_mean = "global_usle_marine_mod_ESA_2020_mean",
    nature_access_1992_mean = "nature_access_lspop2019_ESA1992_mean",
    nature_access_2020_mean = "nature_access_lspop2019_ESA2020_mean",
    n_ret_ratio_1992_mean = "N_ret_ratio_1992_mean",
    n_ret_ratio_2020_mean = "N_ret_ratio_2020_mean",
    sed_ret_ratio_1992_mean = "Sed_ret_ratio_1992_mean",
    sed_ret_ratio_2020_mean = "Sed_ret_ratio_2020_mean",
    coastal_protection_Rt_1992_mean = "coastal_protection_Rt_1992_mean",
    coastal_protection_Rt_2020_mean = "coastal_protection_Rt_2020_mean",
    coastal_protection_Rt_nohab_all_1992_mean = "coastal_protection_Rt_nohab_all_1992_mean",
    coastal_protection_Rt_nohab_all_2020_mean = "coastal_protection_Rt_nohab_all_2020_mean",
    coastal_protection_Rt_ratio_1992_mean = "coastal_protection_Rt_ratio_1992_mean",
    coastal_protection_Rt_ratio_2020_mean = "coastal_protection_Rt_ratio_2020_mean"
  ),
  c(
    n_export_1992_sum = "global_n_export_tnc_esa1992_sum",
    n_export_2020_sum = "global_n_export_tnc_esa2020_sum",
    n_retention_1992_sum = "global_n_retention_ESAmar_1992_fertilizer_sum",
    n_retention_2020_sum = "global_n_retention_ESAmar_2020_fertilizer_sum",
    sed_export_1992_sum = "global_sed_export_marine_mod_ESA_1992_sum",
    sed_export_2020_sum = "global_sed_export_marine_mod_ESA_2020_sum",
    pollination_1992_sum = "realized_polllination_on_ag_ESA1992_sum",
    pollination_2020_sum = "realized_polllination_on_ag_ESA2020_sum"
  )
)

```

::: callout-warning
**TODO / cleanup (run heavy chunks only when inputs are ready)**

- Paths are centralized via `data_dir()`; keep `GLOBAL_NCP_DATA` set.
- Keep expensive consolidation chunks `eval: false` until the latest `10k_grid_synth_{serv,benef,coastal}_*.gpkg` files are in `interim/`.
- Use this file as the living pipeline record; promote stable text to README/methods.
- Maintain the AOO equal-area grid as canonical; use the 4326 clean copy for raster extraction, then wrap dateline for display/output hygiene.
:::

::: callout-note
## Legacy Status

This document (`prepare_data.qmd`) details the one-time setup process for creating the project's canonical 10km grid from raw source files. The code chunks here are generally set to `eval: false` as they do not need to be re-run unless the base grid itself needs to be recreated.

The main data processing and change calculation workflow is handled in `process_data.qmd`.
:::

# Pipeline Overview

For a detailed description of the full pipeline methodology, see README_pipeline.md.

This notebook focuses on the **consolidation** step: synthesizing zonal stats, calculating change, and preparing the canonical dataset.

::: {.callout-tip icon="true"}
## Future Tasks & Ideas

Here are some ideas and future tasks for this analysis:

1.  **Adapt analysis for multi-temporal data:** Adapt analysis to handle updated modeled ES layers and multiple points in time (beyond bi-temporal T0, T1). Strategize for incorporating multi-temporal data.
2.  **Quantify hotspot vs. non-hotspot change:** Develop a method to quantify and visualize the share of total change (from bar plots) that occurs within hotspots versus outside of them, possibly using stacked bar plots.

3.  **V2 Pipeline Simplification:** Ensure the canonical grid's `fid` is strictly preserved through the Python zonal stats pipeline, allowing us to drop the complex spatial coordinate joins in the downstream R scripts and return to simple tabular joins.
:::

## Data Inputs

Global 10 km grid covering terrestrial areas worldwide (polygon-based template).

### Canonical 10 km grid (IUCN AOO, land-only)

We do **not** generate the grid from scratch. We use the published IUCN
Area of Occupancy (AOO) 10×10 km equal-area grid and subset it to land
by intersecting with the correspondence polygons. We keep the full grid
cells that touch land (no clipping) and append attributes used for
subregional aggregation (country/continent/region/biome).


```{r}
#| label: prep-grid-aoo-land
#| eval: true
#| echo: true
# This chunk creates the canonical, analysis-ready 10km grid. It performs
# several critical steps:
# 1. Filters the global AOO grid to only land-intersecting cells.
# 2. Spatially joins all necessary attributes (country, region, biome) from their source files.
# 3. Fixes the dateline wraparound artifact to ensure global maps render correctly.
# 4. Reprojects the final grid to EPSG:4326 for compatibility with raster extraction tools.
#
# Run this chunk only if you need to regenerate the base grid from scratch.

library(sf)
library(dplyr)

message("Loading raw data sources...")
grid_raw <- st_read(data_vectors("AOOGrid_10x10kmShp/AOOGrid_10x10km.shp"), quiet = TRUE)

# Add a check to standardize the ID column name. The raw shapefile might use
# 'id' (lowercase) instead of 'ID'. This makes the script more robust.
id_candidates <- c("ID", "id", "fid", "FID") # Common names for unique identifiers
found_id_col <- NULL
for (col in id_candidates) {
    if (col %in% names(grid_raw)) {
        found_id_col <- col
        break
    }
}

if (is.null(found_id_col)) {
    stop("Could not find a suitable ID column in the raw grid. Looked for: ", paste(id_candidates, collapse = ", "))
} else if (found_id_col != "ID") {
    message(glue::glue("Standardizing ID column: renaming '{found_id_col}' to 'ID'."))
    grid_raw <- dplyr::rename(grid_raw, ID = all_of(found_id_col))
}

corr_path <- file.path(data_dir(), "vector_basedata", "cartographic_ee_r264_correspondence.gpkg")
land_polys <- st_read(corr_path, quiet = TRUE)

biome_path <- file.path(data_dir(), "vector_basedata", "Biome.gpkg")
biome_sf <- st_read(biome_path, quiet = TRUE)

message("Filtering grid to land areas...")
grid_land <- st_filter(grid_raw, st_transform(land_polys, st_crs(grid_raw)), .predicate = st_intersects)

message("Attaching attributes via spatial join...")
# Use point-on-surface for a robust join that avoids duplicating grid cells that touch multiple polygons
grid_pts <- st_point_on_surface(grid_land)

grid_attrs <- grid_pts %>%
  st_join(st_transform(land_polys, st_crs(grid_pts)) %>% select(iso3, nev_name, region_wb, income_grp, continent, region_un, subregion)) %>%
  st_join(st_transform(biome_sf, st_crs(grid_pts)) %>% select(WWF_biome)) %>%
  st_drop_geometry() %>%
  select(fid = ID, iso3, nev_name, region_wb, income_grp, continent, region_un, subregion, WWF_biome) %>%
  distinct(fid, .keep_all = TRUE) # Ensure one row per grid cell

grid_land_attrs <- grid_land %>%
  select(fid = ID) %>%
  left_join(grid_attrs, by = "fid")

message("Fixing dateline wraparound artifact...")
grid_clean <- st_wrap_dateline(grid_land_attrs, options = c("WRAPDATELINE=YES", "DATELINEOFFSET=180"))

message("Reprojecting to EPSG:4326 for final output...")
grid_4326 <- st_transform(grid_clean, "EPSG:4326")

out_path <- file.path(data_dir(), "vector_basedata", "AOOGrid_10x10km_land_4326_clean.gpkg")
message("Writing final clean grid to: ", out_path)
st_write(grid_4326, out_path, delete_dsn = TRUE)

message("Base grid preparation complete.")
```

```{python}
#| label: make-grid-valid-shapely
#| eval: false
#| message: false
#| warning: false

# Optional: create a Shapely-validated grid copy to avoid GEOS errors in Python.
# Requires geopandas/shapely in the Quarto Python environment.
import os
import geopandas as gpd

data_root = os.environ.get("GLOBAL_NCP_DATA", "")
grid_in = os.path.join(data_root, "vector_basedata", "AOOGrid_10x10km_land.gpkg")
grid_out = os.path.join(data_root, "vector_basedata", "AOOGrid_10x10km_land_shpvalid.gpkg")

gdf = gpd.read_file(grid_in)
gdf["geometry"] = gdf.geometry.make_valid()
gdf = gdf[gdf.geometry.notna() & gdf.is_valid]

gdf.to_file(grid_out, driver="GPKG")
print(f"wrote {len(gdf):,} features -> {grid_out}")
```

```{python}
#| label: grid-4326-clean
#| eval: false
#| message: false
#| warning: false

# Create a 4326 version for raster extraction (avoids to_crs errors at runtime).
# This keeps the same cells, only expressed in EPSG:4326.
import geopandas as gpd
from shapely.ops import transform
from pyproj import Transformer

grid_in = os.path.join(data_root, "vector_basedata", "AOOGrid_10x10km_land_shpvalid.gpkg")
grid_out = os.path.join(data_root, "vector_basedata", "AOOGrid_10x10km_land_4326_clean.gpkg")

gdf = gpd.read_file(grid_in)
transformer = Transformer.from_crs(gdf.crs, "EPSG:4326", always_xy=True)

bad = 0
new_geoms = []
for geom in gdf.geometry:
    try:
        g = transform(transformer.transform, geom)
        if g.is_empty or not g.is_valid:
            bad += 1
            new_geoms.append(None)
        else:
            new_geoms.append(g)
    except Exception:
        bad += 1
        new_geoms.append(None)

gdf["geometry"] = new_geoms
gdf = gdf[gdf.geometry.notna()]
gdf = gdf.set_crs("EPSG:4326", allow_override=True)
gdf.to_file(grid_out, driver="GPKG")
print(f"dropped {bad:,} invalid geometries; wrote -> {grid_out}")
```

**Canonical grids used downstream**

- **Equal-area canonical grid**: `AOOGrid_10x10km_land.gpkg`
- **Extraction grid (EPSG:4326)**: `AOOGrid_10x10km_land_4326_clean.gpkg`

The 4326 file is only for raster extraction; the cell geometry is the same
10×10 km equal-area grid expressed in a different CRS.

Coastal protection model outputs (1992 and 2020) originally as points spaced \~500 m along shorelines.

Global raster datasets for nitrogen retention/export/ratio, sediment retention/export/USLE, pollination, and nature access (1992 and 2020).

Country polygon dataset for spatial aggregation and attribution.
