---
title: "Data Processing"
subtitle: "v1.2.1"
author: "Jerónimo Rodríguez Escobar"
date: last-modified
date-format: "YYYY-MM-DD HH:mm zzz"
format:
#  html:
#    toc: true
#    number-sections: true
#   theme: cosmo
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
editor: source
---

```{r setup, message=FALSE, warning=FALSE}
library(terra)
library(sf)
library(dplyr)
library(ggplot2)
library(glue)
library(tidyr)
library(purrr)
library(diffeR)
library(here)
library(stringr)
library(tidytext)
library(rlang)
library(tidyr)
library(forcats)
library(scales)
library(RColorBrewer)
library(htmltools)
library(leaflet)
library(devtools)
library(reticulate)
library(exactextractr)
library(httpgd)
load_all()

# set venv for Python
use_virtualenv("/home/jeronimo/venvs/coastal_snap_env", required = TRUE)

# Centralize data roots (set GLOBAL_NCP_DATA in ~/.Renviron)
data_root <- data_dir()
interim_dir <- data_interim()
processed_dir <- data_processed()

processed_tables_dir <- file.path(processed_dir, "tables")
processed_hotspots_dir <- file.path(processed_dir, "hotspots")

# Source unified hotspot/change helpers
source(here::here("R", "utils_hotspot.R"))

# This maps raw column names from synth outputs to canonical names.
rename_list <- list(
  c(
    usle_1992_mean = "global_usle_marine_mod_ESA_1992_mean",
    usle_2020_mean = "global_usle_marine_mod_ESA_2020_mean",
    nature_access_1992_mean = "nature_access_lspop2019_ESA1992_mean",
    nature_access_2020_mean = "nature_access_lspop2019_ESA2020_mean",
    n_ret_ratio_1992_mean = "N_ret_ratio_1992_mean",
    n_ret_ratio_2020_mean = "N_ret_ratio_2020_mean",
    sed_ret_ratio_1992_mean = "Sed_ret_ratio_1992_mean",
    sed_ret_ratio_2020_mean = "Sed_ret_ratio_2020_mean",
    c_risk_1992_mean = "C_Risk_1992_mean",
    c_risk_2020_mean = "C_Risk_2020_mean",
    c_risk_red_ratio_1992_mean = "C_Risk_Red_Ratio_1992_mean",
    c_risk_red_ratio_2020_mean = "C_Risk_Red_Ratio_2020_mean"
  ),
  c(
    n_export_1992_sum = "global_n_export_tnc_esa1992_sum",
    n_export_2020_sum = "global_n_export_tnc_esa2020_sum",
    n_retention_1992_sum = "global_n_retention_ESAmar_1992_fertilizer_sum",
    n_retention_2020_sum = "global_n_retention_ESAmar_2020_fertilizer_sum",
    sed_export_1992_sum = "global_sed_export_marine_mod_ESA_1992_sum",
    sed_export_2020_sum = "global_sed_export_marine_mod_ESA_2020_sum",
    pollination_1992_sum = "realized_polllination_on_ag_ESA1992_sum",
    pollination_2020_sum = "realized_polllination_on_ag_ESA2020_sum"
  )
)
```

# Data processing

## Consolidate Synthesis Data

Assemble the final gridded Change/Beneficiary dataset by joining the interim synthesis files (Services, Beneficiaries).

```{r}
#| label: assemble-final-grid
#| eval: true
#| include: false
source(here::here("R", "paths.R"))
interim_dir <- data_interim()
processed_dir <- data_processed()

latest_interim <- function(pattern) {
  files <- list.files(interim_dir, pattern = pattern, full.names = TRUE)
  if (!length(files)) stop("No interim files match pattern: ", pattern)
  files[which.max(file.mtime(files))]
}

inpath <- processed_dir
# Optional: export name indexes to CSV
write_index <- FALSE
write_base_synth <- TRUE

# New workflow: means + sums are in the same synth output.
serv_path  <- latest_interim("^10k_grid_services_base\\.gpkg$")
ben_path   <- latest_interim("^10k_grid_benef(.*)?\\.gpkg$")

S <- st_read(serv_path)
B <- st_read(ben_path)

if (!"fid" %in% names(S)) S$fid <- seq_len(nrow(S))
if (!"fid" %in% names(B)) B$fid <- seq_len(nrow(B))

# Normalize grouping column names to canonical forms (handle casing issues)
expected_attrs <- c(
  "iso3", "continent", "region_un", "subregion", "nev_name",
  "region_wb", "income_grp", "WWF_biome"
)
curr_names <- names(S)
for (target in expected_attrs) {
  match_idx <- which(tolower(curr_names) == tolower(target))
  if (length(match_idx) == 1) names(S)[match_idx] <- target
}

# Immediately normalize variable names after joining
S_tbl <- st_drop_geometry(S)
B_tbl <- st_drop_geometry(B)

sf_f <- S %>%
  left_join(B_tbl, by = "fid")

# Canonical renaming: apply once, early, and only here
rename_map <- c(rename_list[[1]], rename_list[[2]])
sf_f <- dplyr::rename(sf_f, !!!rename_map)


# Ensure c_fid is present (critical for downstream hotspot extraction)
if (!"c_fid" %in% names(sf_f)) {
  sf_f$c_fid <- sf_f$fid
}

# Optional: save the base synthesis (raw 1992/2020 + beneficiaries + coastal)
if (isTRUE(write_base_synth)) {
  st_write(sf_f, file.path(processed_dir, "10k_grid_synth_all.gpkg"), delete_dsn = TRUE)
}

 # Services-only synth lives in interim (10k_grid_services.gpkg) if needed for QA.

```

### next
```{r compute-bitemporal-change}
#| eval: true
#| include: false
# Assumes `sf_f` is already created in the consolidation chunk above.
# If not, load the latest base synthesis from processed.
# This keeps the change calculation resumable without rerunning consolidation.
if (!exists("processed_dir")) {
  if (!exists("data_processed")) source(here::here("R", "paths.R"))
  processed_dir <- data_processed()
}
if (!exists("compute_change")) {
  devtools::load_all(here::here(), quiet = TRUE)
}

if (!exists("sf_f")) {
  synth_path <- file.path(processed_dir, "10k_grid_synth_all.gpkg")
  sf_f <- st_read(synth_path, quiet = TRUE)
}
# Calculate bi-temporal change in % and absolute terms.
# pct_chg uses the earliest year as baseline: ((t1 - t0) / t0) * 100.
# Division by zero yields Inf/NaN in standard pct change; symmetric change handles this by bounding at +/-200%.
# However, if any Inf values persist (e.g. appearing from 0), we clamp them to the max symmetric range (+/- 200%).
# Use pct_mode = "symm" to avoid counter-intuitive sign flips and handle zero baselines.
sf_f <- compute_change(
  sf_f,
  suffix = c("_sum", "_mean"),
  change_type = "both",
  drop_columns = FALSE,
  pct_mode = "symm"
)

# Safety: Replace any accidental Inf with +/-200% (appearing/disappearing)
sf_f <- sf_f %>%
  mutate(across(ends_with("_pct_chg"), ~ case_when(
    is.infinite(.) & . > 0 ~ 200,
    is.infinite(.) & . < 0 ~ -200,
    TRUE ~ .
  )))

# NaN are divisions by 0, NA is when values are NA
write_full_change <- TRUE

# Expanded change synthesis (raw + change) for QA/validation
if (isTRUE(write_full_change)) {
  st_write(
    sf_f,
    file.path(processed_dir, "10k_grid_ES_change_benef.gpkg"),
    layer = "10k_grid_ES_change_benef",
    delete_dsn = TRUE
  )
}


```
# ============================================================================
# AUTHORITATIVE GRID + GROUPING + CHANGE DATASET
# ============================================================================
# This chunk produces 10k_change_calc.gpkg, the canonical input for:
#   - analysis/hotspot_extraction.qmd
#   - analysis/KS_tests_hotspots.qmd
#   - Any downstream aggregation/plotting
#
# Grid attributes (fid, c_fid, iso3, continent, region_wb, income_grp,
# region_un, subregion, WWF_biome, BIOME) are attached via Consolidation.qmd
# and NOT re-derived elsewhere.
#
# If this file is missing or stale, regenerate by running the full
# Consolidation.qmd workflow.
# ============================================================================
```{r filter-canonical-change-prep}
#| eval: true
#| include: false
# STEP 1: Prepare all column lists and perform an initial slim-down of sf_f.
# This avoids doing too many complex operations in a single chunk.

if (!"fid" %in% names(sf_f)) {
  sf_f$fid <- seq_len(nrow(sf_f))
}

# Define service rename mapping (same as in consolidation chunk).

keep_fixed <- c(
  "fid", "c_fid", "iso3", "continent", "region_un", "subregion",
  "nev_name", "region_wb", "income_grp", "WWF_biome"
)

benef_pattern <- "^(GHS_|GlobPOP_|hdi_|rast_gdpTot_|rast_adm1_gini_|fields_mehrabi_)"

# Drop raw/base-year service columns; we only keep change metrics here.
message("Dropping raw year columns to reduce memory footprint...")
year_cols <- grep("1992|2020", names(sf_f), value = TRUE)
benef_cols <- grep(benef_pattern, names(sf_f), value = TRUE)
drop_year_cols <- setdiff(year_cols, benef_cols)

if (length(drop_year_cols) > 0) {
  sf_f <- sf_f[, !names(sf_f) %in% drop_year_cols]
  # free memory after the drop
  rm(year_cols, benef_cols, drop_year_cols)
  gc()
}
message("Initial column drop complete.")

# Now, define the final set of columns to keep.
chg_cols <- grep("(_(abs|pct)_chg$)", names(sf_f), value = TRUE)
if (!length(chg_cols)) {
  message("No change columns found in sf_f; probing full dataset file I/O (n_max=1)...")
  # quick probe to ensure the file is accessible and not stuck on IO
  try({
    sf::st_read(file.path(processed_dir, "10k_grid_ES_change_benef.gpkg"), n_max = 1, quiet = TRUE)
    message('Probe read OK; performing full read...')
  }, silent = TRUE)
  sf_f <- st_read(file.path(processed_dir, "10k_grid_ES_change_benef.gpkg"), quiet = FALSE)
  if (!"fid" %in% names(sf_f)) {
    sf_f$fid <- seq_len(nrow(sf_f))
  }
  chg_cols <- grep("(_(abs|pct)_chg$)", names(sf_f), value = TRUE)
}

# Ensure c_fid is preserved/normalized before final selection
# TODO: Clarify legacy usage of c_fid. Originally used for non-spatial country joins.
# Currently serves as a stable unique identifier for grid cells (often == fid).
# In future refactors, consider unifying strictly on 'fid' if c_fid is redundant.
if (!"c_fid" %in% names(sf_f)) {
  if ("c_fid.x" %in% names(sf_f)) sf_f <- dplyr::rename(sf_f, c_fid = c_fid.x)
  else if ("c_fid.y" %in% names(sf_f)) sf_f <- dplyr::rename(sf_f, c_fid = c_fid.y)
  else if ("id" %in% names(sf_f))      sf_f <- dplyr::rename(sf_f, c_fid = id)
}
if (!"c_fid" %in% names(sf_f)) {
  sf_f$c_fid <- sf_f$fid
}

# Define canonical services and beneficiaries to keep
svc_cols <- names(c(rename_list[[1]], rename_list[[2]]))
svc_base <- unique(stringr::str_replace(svc_cols, "_(1992|2020)_", "_"))

canonical_change <- unique(c(
  paste0(svc_base, "_abs_chg"),
  paste0(svc_base, "_pct_chg")
))
chg_cols <- intersect(chg_cols, canonical_change)

benef_keep <- intersect(
  c(
    paste0(c("fields_mehrabi_2017", "GHS_BUILT_S_E2020", "hdi_raster_predictions_2020", "rast_adm1_gini_disp_2020"), "_mean"),
    paste0(c("GHS_POP_E2020_GLOBE", "GlobPOP_Count_30arc_2020", "rast_gdpTot_1990_2020_30arcsec_2020"), "_sum")
  ),
  names(sf_f)
)

message("Beneficiary columns to keep (n = ", length(benef_keep), "): ")
print(benef_keep)

message("Change columns to keep (n = ", length(chg_cols), "): ")
print(head(chg_cols, 25))
message('Freeing temporary objects before heavy selection...')

# This is the final list of columns for our canonical dataset
keep_cols <- unique(c(
  keep_fixed,
  benef_keep,
  chg_cols
))
gc()

```

```{r filter-canonical-change-select}
#| eval: true
#| include: false
# STEP 2: Perform the main, heavy selection operation.
# This is isolated to prevent memory hangs.
message("Creating canonical subset by selecting final columns...")
# attempt a small test of selection to detect schema issues quickly
message('Testing select on a small subset (first 10 rows)')
try({ tmp_test <- sf_f[seq_len(min(10, nrow(sf_f))), ] |> dplyr::select(any_of(keep_cols)); rm(tmp_test) }, silent = TRUE)
gc()
t0 <- Sys.time();
cols_to_keep <- intersect(keep_cols, names(sf_f))
sf_canon <- sf_f[, cols_to_keep]
rm(sf_f); gc()
message('Canonical subset created. Elapsed (mins): ', round(as.numeric(difftime(Sys.time(), t0, units = 'mins')), 2))
```

```{r filter-canonical-change-join}
#| eval: true
#| include: false
# STEP 3: Join missing grid attributes if necessary and finalize schema.
# If fixed grouping columns are missing, we proceed without them (per user request),
# assuming they might not be critical or will be handled downstream.
# The complex join logic here has been removed to avoid errors when AOOGrid is unavailable
# or mismatches fids, relying on the input services GPKG to provide these attributes.
missing_cols <- setdiff(keep_fixed, names(sf_canon))
if (length(missing_cols) > 0) {
  message("WARNING: The following grouping columns are missing from the final dataset: ",
          paste(missing_cols, collapse=", "),
          ". Downstream analysis (hotspots) may require them.")
}

# Re-apply selection to ensure only intended columns remain after the join
sf_canon <- sf_canon %>% dplyr::select(any_of(keep_cols))
message("Final column schema enforced.")
```

```{r write-canonical-change}
#| eval: true
#| include: false
# Clean service change names only (drop _mean/_sum for canonical services).
rename_change <- function(nm) {
  stringr::str_replace(nm, "_(mean|sum)_(abs|pct)_chg$", "_\\2_chg")
}
svc_change <- c(
  paste0(svc_base, "_abs_chg"),
  paste0(svc_base, "_pct_chg")
)
svc_change <- intersect(svc_change, names(sf_canon))
if (length(svc_change)) {
  sf_canon <- dplyr::rename_with(sf_canon, rename_change, .cols = all_of(svc_change))
}

st_write(
  sf_canon,
  file.path(processed_dir, "10k_change_calc.gpkg"),
  layer = "10k_change_calc",
  delete_dsn = TRUE
)
```

Note: coastal protection is prepared upstream (join → ratios → rasterize → summary)
via `Python_scripts/coastal_protection_join.py`, `Python_scripts/rasterize_coastal.py`,
and `analysis_configs/c_protection_synth.yaml`. This document consumes the resulting
coastal synthesis GPKG from `data_interim()`.

## Add Updated Access Data

```
