---
title: "Process Zonal Stats V2"
subtitle: "Calculate Bi-temporal Change from New Zonal Stats"
author: "Jeronimo Rodriguez-Escobar"
date: last-modified
format: html
editor: source
---

## Overview

This script processes the new zonal statistics generated by the `zonal_stats_toolkit`.
It performs the following steps:
1.  Loads the new raw statistics (separate columns for 1992 and 2020).
2.  Loads the original `10k_change_calc.gpkg` to retrieve spatial attributes (Region, Biome, etc.) and ensure consistency.
3.  Calculates **Absolute Change** (`2020 - 1992`).
4.  Calculates **Symmetric Percentage Change** (`(2020 - 1992) / ((|2020| + |1992|) / 2)`).
5.  Exports `processed/10k_change_calc_v2.gpkg`.

```{r setup, message=FALSE, warning=FALSE}
library(sf)
library(dplyr)
library(stringr)
library(here)
library(devtools)

# Initialize paths
knitr::opts_knit$set(root.dir = here::here())
source(here::here("R", "paths.R"))
devtools::load_all(quiet = TRUE)
```

## 1. Load Data

```{r load-data}
# Path to new raw stats
new_stats_path <- "/home/jeronimo/projects/zonal_stats_toolkit/output_raw_fid/grid_10km_with_stats.gpkg"
if (!file.exists(new_stats_path)) stop("New stats file not found.")

message("Loading new zonal stats...")
df_new <- st_read(new_stats_path, quiet = TRUE) %>% st_drop_geometry()

# User reported grid_fid is NULL in new stats. Using row number (implicit fid) as join key.
# Force creation of fid from row number to ensure alignment by position
df_new$fid <- seq_len(nrow(df_new))

# Path to original data (for attributes)
old_gpkg_path <- file.path(data_dir(), "processed", "10k_change_calc.gpkg")
if (!file.exists(old_gpkg_path)) stop("Original 10k_change_calc.gpkg not found.")

message("Loading original spatial attributes...")
sf_old <- st_read(old_gpkg_path, quiet = TRUE)

# Ensure 'fid' exists, as it can be a special column not read by default
# Force creation of fid from row number to ensure alignment by position
sf_old$fid <- seq_len(nrow(sf_old))

# Select only ID and grouping columns from old data
grouping_cols <- c("income_grp", "region_wb", "continent", "region_un", "WWF_biome", "nev_name", "iso3", "subregion")
grouping_cols <- intersect(grouping_cols, names(sf_old))

# Socioeconomic variables for KS tests
socio_cols <- c(
  "rast_gdpTot_1990_2020_30arcsec_2020_sum",
  "GHS_POP_E2020_GLOBE_sum",
  "GlobPOP_Count_30arc_2020_sum",
  "GHS_BUILT_S_E2020_mean",
  "fields_mehrabi_2017_mean",
  "hdi_raster_predictions_2020_mean",
  "rast_adm1_gini_disp_2020_mean"
)
socio_cols <- intersect(socio_cols, names(sf_old))

df_attribs <- sf_old %>%
  dplyr::select(fid, c_fid, dplyr::all_of(grouping_cols), dplyr::all_of(socio_cols))

# Ensure FID consistency and validate alignment
if (!"fid" %in% names(df_new)) {
  # If new data lacks FID, assume row order matches if nrow is same (RISKY, check first)
  if (nrow(df_new) == nrow(df_attribs)) {
    message("WARNING: New data lacks 'fid'. Assuming row order matches old data.")
    df_new$fid <- df_attribs$fid
  } else {
    stop("New data lacks 'fid' and row count differs from old data.")
  }
} else {
  # If fid exists, check overlap
  common_fids <- intersect(df_new$fid, df_attribs$fid)
  message("FID overlap: ", length(common_fids), " out of ", nrow(df_attribs), " target rows.")
  if (length(common_fids) == 0) {
    stop("CRITICAL ERROR: No FID overlap between new stats and original attributes. Check FID columns (types/values).")
  }
}

# Validate alignment using common columns (e.g. iso3, region_wb) to ensure we aren't mixing rows
common_cols <- setdiff(intersect(names(df_new), names(df_attribs)), "fid")
if (length(common_cols) > 0) {
  message("Validating alignment using common columns: ", paste(common_cols, collapse = ", "))
  for (col in common_cols) {
    mismatches <- sum(df_new[[col]] != df_attribs[[col]], na.rm = TRUE)
    if (mismatches > 0) {
      warning("Mismatch in column ", col, ": ", mismatches, " rows differ. Check data alignment!")
    }
  }
  # Drop redundant columns from df_new to avoid .x/.y suffixes in join
  df_new <- df_new %>% dplyr::select(-dplyr::all_of(common_cols))
}
```

## 2. Calculate Change Metrics

```{r calc-change}
# Identify service columns (assuming format: service_1992_mean, service_2020_mean)
cols <- names(df_new)
cols_1992 <- cols[grepl("(_)?1992", cols)]
services <- unique(gsub("(_)?1992.*", "", cols_1992))

message("Found services: ", paste(services, collapse = ", "))

# Canonical lookup to ensure V2 names match V1/Hotspot config
canonical_lookup <- c(
  sed_export       = "Sed_export",
  n_export         = "N_export",
  n_retention      = "N_retention",
  nature_access    = "Nature_Access",
  pollination      = "Pollination",
  usle             = "USLE",
  n_ret_ratio      = "N_Ret_Ratio",
  sed_ret_ratio    = "Sed_Ret_Ratio",
  coastal_protection_rt            = "C_Risk",
  coastal_protection_rt_nohab_all  = "C_Risk_NoHab",
  coastal_protection_rt_ratio      = "C_Risk_Red_Ratio",
  rt_ratio         = "C_Risk_Red_Ratio",
  rt               = "C_Risk",
  c_risk           = "C_Risk",
  c_risk_red_ratio = "C_Risk_Red_Ratio",
  realized_pollination_on_ag_esa = "Pollination"
)

df_calc <- df_new
change_cols <- c()

for (svc in services) {
  # Normalize svc name for output column
  svc_clean <- gsub("^(mean|sum)_", "", tolower(svc))
  svc_norm <- dplyr::recode(svc_clean, !!!canonical_lookup, .default = svc_clean)

  # Regex to find matching 1992 and 2020 columns (handling _mean or _sum suffixes)
  col_92 <- cols[grepl(paste0("^", svc, "(_)?1992"), cols)]
  col_20 <- cols[grepl(paste0("^", svc, "(_)?2020"), cols)]

  if (length(col_92) == 1 && length(col_20) == 1) {
    val_92 <- df_calc[[col_92]]
    val_20 <- df_calc[[col_20]]

    # Absolute Change
    abs_col <- paste0(svc_norm, "_abs_chg")
    df_calc[[abs_col]] <- val_20 - val_92

    # Symmetric Percentage Change
    # Formula: (y2 - y1) / ((|y2| + |y1|) / 2)
    pct_col <- paste0(svc_norm, "_pct_chg")
    denom <- (abs(val_20) + abs(val_92)) / 2

    # Handle division by zero (if both are 0, change is 0)
    pct_val <- (val_20 - val_92) / denom
    pct_val[denom == 0] <- 0

    df_calc[[pct_col]] <- pct_val
    change_cols <- c(change_cols, abs_col, pct_col)
  }
}

# Diagnostic: Check if df_calc has valid data before join
na_check <- sapply(df_calc[change_cols], function(x) sum(is.na(x)))
if (all(na_check == nrow(df_calc))) {
  stop("CRITICAL: All calculated change columns in df_calc are NA *before* the join. Check input stats (val_92/val_20).")
}
```

## 3. Merge and Export

```{r export}
# Join calculated metrics with original spatial attributes and geometry
# Only keep the calculated change columns (drop base years, stdev, valid_count)
sf_final_v2 <- df_attribs %>%
  left_join(df_calc %>% dplyr::select(fid, dplyr::all_of(change_cols)), by = "fid")

# CRITICAL CHECK: Verify we didn't generate a dataset of NAs
na_counts <- sapply(sf_final_v2[change_cols], function(x) sum(is.na(x)))
if (any(na_counts == nrow(sf_final_v2))) {
  stop("CRITICAL ERROR: Some change columns are 100% NA. The join failed to match rows.")
}

# Rename 'fid' to 'grid_fid' to avoid GPKG driver conflict (reserved primary key)
if ("fid" %in% names(sf_final_v2)) {
  sf_final_v2 <- dplyr::rename(sf_final_v2, grid_fid = fid)
}

out_path <- file.path(data_dir(), "processed", "10k_change_calc_v2.gpkg")
message("Saving V2 dataset to: ", out_path)
st_write(sf_final_v2, out_path, delete_dsn = TRUE, quiet = TRUE)
```