---
title: "Results Interpretation and Narrative Synthesis"
subtitle: "Where are the hotspots, who is affected, and what are the drivers?"
author: "Jerónimo Rodríguez Escobar"
date: last-modified
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
editor: source
---

## Overview

This document synthesizes the primary outputs of the Global NCP hotspot analysis. It loads the final summary tables to build a narrative around the three core questions of the project:

1.  **Where are the hotspots?** (Geographic Distribution & Intensity)
2.  **Who is affected?** (Socioeconomic Profile of Exposed Populations)
3.  **What are the drivers?** (Land Cover Change Attribution)

The goal is to move from quantitative tables to qualitative insights and key takeaways for the final manuscript.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
library(knitr)
library(scales)

# Initialize paths
knitr::opts_knit$set(root.dir = here::here())
source(here::here("R","paths.R"))

options(scipen = 999) # Disable scientific notation for easier reading
```

## Load Synthesized Data

This section loads the three key data products generated by the upstream analysis notebooks.

```{r}
#| label: load-data

# 1. Hotspot Area & Intensity Statistics
area_stats_path <- file.path(data_dir(), "processed", "tables", "hotspot_area_stats.csv")
if (file.exists(area_stats_path)) {
  area_stats <- readr::read_csv(area_stats_path, show_col_types = FALSE)
  message("Loaded area stats.")
} else {
  stop("hotspot_area_stats.csv not found. Run hotspot_synthesis.qmd first.")
}

# 2. Socioeconomic Exposure Statistics
pop_exposure_path <- file.path(data_dir(), "processed", "tables", "hotspot_pop_exposure.csv")
if (file.exists(pop_exposure_path)) {
  pop_exposure <- readr::read_csv(pop_exposure_path, show_col_types = FALSE)
  message("Loaded population exposure stats.")
} else {
  warning("hotspot_pop_exposure.csv not found. Run hotspot_synthesis.qmd to generate it.")
}

# 3. LCC Attribution Statistics (Update with the correct filename)
# lcc_stats_path <- file.path(data_dir(), "processed", "tables", "lcc_es_hotspot_overlap_summary.csv")
# if (file.exists(lcc_stats_path)) {
#   lcc_stats <- readr::read_csv(lcc_stats_path, show_col_types = FALSE)
#   message("Loaded LCC attribution stats.")
# } else {
#   warning("LCC attribution stats not found. Check filename and upstream scripts.")
# }

# 4. Multi-service overlap stats
multiservice_stats_path <- file.path(data_dir(), "processed", "tables", "hotspot_multiservice_stats.csv")
if (file.exists(multiservice_stats_path)) {
  multiservice_stats <- readr::read_csv(multiservice_stats_path, show_col_types = FALSE)
  message("Loaded multi-service stats.")
} else {
  warning("hotspot_multiservice_stats.csv not found. Run hotspot_synthesis.qmd first.")
}
```

## Part 1: Where are the hotspots? (Geographic Concentration)

This section uses `hotspot_area_stats.csv` and `hotspot_multiservice_stats.csv` to identify the regions and biomes bearing a disproportionate burden of ecosystem service decline.

### Key Questions:
*   Which regions/biomes have the highest **Hotspot Intensity** (i.e., largest percentage of their own area in crisis)?
*   Which regions/biomes contain the largest **Global Share** of hotspots for each service?
*   Which regions/biomes show the highest **Relative Intensity**, indicating a disproportionate burden?
*   Where are multi-service hotspots concentrated (high **"Hotness"**)?

### Analysis: Regional Hotspot Concentration (World Bank Regions)

```{r}
#| label: analyze-regional-intensity

# Example: Find the top 3 regions with the highest relative intensity for Nitrogen Export
area_stats %>%
  filter(grouping_var == "region_wb", service == "N_export") %>%
  arrange(desc(relative_intensity)) %>%
  select(group, pct_area, pct_share, relative_intensity) %>%
  head(3) %>%
  kable(caption = "Top 3 WB Regions by Relative Intensity for N_export Hotspots")

# Narrative Point: "The analysis of relative intensity reveals that [Region X] bears a burden of Nitrogen Export hotspots that is Y times greater than would be expected given its land area."
```

## Part 2: Who is affected? (Socioeconomic Exposure)

This section uses `hotspot_pop_exposure.csv` to quantify the human dimension of ES degradation.

### Key Questions:
*   How many people are living in hotspot areas for each service?
*   How does this exposure break down across World Bank Income Groups?
*   Are populations with lower Human Development Index (HDI) disproportionately exposed?

### Analysis: Population Exposure by Income Group

```{r}
#| label: analyze-pop-exposure-income

# Example: Total population exposed to Nature Access decline, by income group
if (exists("pop_exposure")) {
  pop_exposure %>%
    filter(service == "Nature_Access") %>%
    group_by(income_grp) %>%
    summarise(total_exposed_pop = sum(exposed_population, na.rm = TRUE)) %>%
    mutate(total_exposed_pop = round(total_exposed_pop)) %>%
    arrange(desc(total_exposed_pop)) %>%
    kable(caption = "Population in Nature Access Hotspots by Income Group",
          format.args = list(big.mark = ","))
}

# Narrative Point: "A total of [X million] people live in areas experiencing critical declines in Nature Access. Of these, [Y%] reside in low- and lower-middle-income countries, highlighting a significant equity concern."
```

## Part 3: What are the drivers? (LCC Attribution)

This section will use the `lcc_es_hotspot_overlap_... .csv` tables to connect ES hotspots to specific land cover change dynamics (e.g., Forest Loss, Cropland Expansion).

*This section is a placeholder until the LCC attribution tables are finalized and loaded.*

## Part 4: Synthesis & Key Takeaways

This section is for drafting the high-level summary points for the manuscript's discussion and conclusion.

*   **Takeaway 1: Geographic Mismatch.** *e.g., "Our results show a clear geographic mismatch between the areas providing the bulk of a service and the areas experiencing the most intense relative degradation..."*
*   **Takeaway 2: Equity and Vulnerability.** *e.g., "The burden of ES degradation is not borne equally. We find that low-income countries and populations with low HDI are disproportionately exposed to declines in critical services like..."*
*   **Takeaway 3: Drivers of Change.** *e.g., "While land conversion is a major driver for services like [Service A], a significant portion of hotspots for [Service B] occur in stable landscapes, pointing towards degradation-in-place..."*