---
title: "Results Interpretation and Narrative Synthesis"
subtitle: "Where are the hotspots, who is affected, and what are the drivers?"
author: "Jerónimo Rodríguez Escobar"
date: last-modified
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
editor: source
---

## Overview

This document synthesizes the primary outputs of the Global NCP hotspot analysis. It loads the final summary tables to build a narrative around the three core questions of the project:

1.  **Where are the hotspots?** (Geographic Distribution & Intensity)
2.  **Who is affected?** (Socioeconomic Profile of Exposed Populations)
3.  **What are the drivers?** (Land Cover Change Attribution)

The goal is to move from quantitative tables to qualitative insights and key takeaways for the final manuscript.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
library(knitr)
library(scales)

# Initialize paths
knitr::opts_knit$set(root.dir = here::here())
source(here::here("R","paths.R"))

# Diagnostic: Print the resolved data_dir() path to help with local setup
local_data_root <- data_dir()
message(paste("Resolved data root for local execution:", local_data_root))
if (!dir.exists(file.path(local_data_root, "processed", "tables"))) {
  warning(paste("Warning: The 'processed/tables' directory does not exist at", file.path(local_data_root, "processed", "tables"), ". Please ensure data is downloaded and GLOBAL_NCP_DATA is set correctly."))
}

options(scipen = 999) # Disable scientific notation for easier reading
```

## Load Synthesized Data

This section loads the three key data products generated by the upstream analysis notebooks.

```{r}
#| label: load-data
#| echo: false

# 1. Hotspot Area & Intensity Statistics
area_stats_path <- file.path(data_dir(), "processed", "tables", "hotspot_area_stats.csv")
if (file.exists(area_stats_path)) {
  area_stats <- readr::read_csv(area_stats_path, show_col_types = FALSE)
  message("Loaded area stats.")
} else {
  stop("hotspot_area_stats.csv not found. Run hotspot_synthesis.qmd first.")
}

# 2. Socioeconomic Exposure Statistics
pop_exposure_path <- file.path(data_dir(), "processed", "tables", "hotspot_pop_exposure.csv")
if (file.exists(pop_exposure_path)) {
  pop_exposure <- readr::read_csv(pop_exposure_path, show_col_types = FALSE)
  message("Loaded population exposure stats.")
} else {
  warning("hotspot_pop_exposure.csv not found. Run hotspot_synthesis.qmd to generate it.")
}

# 3. LCC Attribution Statistics (Update with the correct filename)
lcc_stats_path <- file.path(data_dir(), "processed", "tables", "lcc_es_hotspot_overlap_pct.csv")
if (file.exists(lcc_stats_path)) {
  lcc_stats <- readr::read_csv(lcc_stats_path, show_col_types = FALSE)
  message("Loaded LCC attribution stats.")
} else {
  warning("LCC attribution stats not found. Check filename and upstream scripts.")
}

# 4. KS Test Socioeconomic Results
ks_results_path <- file.path(data_dir(), "processed", "tables", "ks_results_hot_vs_non.csv")
if (file.exists(ks_results_path)) {
  ks_res <- readr::read_csv(ks_results_path, show_col_types = FALSE)
  message("Loaded KS test results.")
} else {
  warning("ks_results_hot_vs_non.csv not found. Run KS_tests_hotspots.qmd to generate it.")
}

# 4. Multi-service overlap stats
multiservice_stats_path <- file.path(data_dir(), "processed", "tables", "hotspot_multiservice_stats.csv")
if (file.exists(multiservice_stats_path)) {
  multiservice_stats <- readr::read_csv(multiservice_stats_path, show_col_types = FALSE)
  message("Loaded multi-service stats.")
} else {
  warning("hotspot_multiservice_stats.csv not found. Run hotspot_synthesis.qmd first.")
}
```

## Part 1: Where are the hotspots? (Geographic Concentration)

This section uses `hotspot_area_stats.csv` and `hotspot_multiservice_stats.csv` to identify the regions and biomes bearing a disproportionate burden of ecosystem service decline. Hotspots of ecosystem service decline are not randomly distributed; they cluster heavily in the Global South, creating epicenters of **Compound Risk** where communities face multiple simultaneous ecological failures.

### Key Questions:
*   Which regions/biomes have the highest **Hotspot Intensity** (i.e., largest percentage of their own area in crisis)?
*   Which regions/biomes contain the largest **Global Share** of hotspots for each service?
*   Which regions/biomes show the highest **Relative Intensity**, indicating a disproportionate burden?
*   Where are multi-service hotspots concentrated (high **"Hotness"**)?

### Analysis: Regional Hotspot Concentration (World Bank Regions)

```{r}
#| label: analyze-regional-intensity
#| echo: false
#| results: 'asis'

if (exists("area_stats")) {
  # By World Bank Region
  cat("#### Top World Bank Region by Relative Hotspot Intensity\n\n")
  area_stats %>%
    filter(grouping_var == "region_wb") %>%
    group_by(service) %>%
    slice_max(order_by = relative_intensity, n = 1) %>%
    ungroup() %>%
    select(Service = service, `Top Region` = group, `Relative Intensity` = relative_intensity) %>%
    arrange(Service) %>%
    kable(digits = 2)

  # By WWF Biome
  cat("\n\n#### Top WWF Biome by Relative Hotspot Intensity\n\n")
  area_stats %>%
    filter(grouping_var == "WWF_biome") %>%
    group_by(service) %>%
    slice_max(order_by = relative_intensity, n = 1) %>%
    ungroup() %>%
    select(Service = service, `Top Biome` = group, `Relative Intensity` = relative_intensity) %>%
    arrange(Service) %>%
    kable(digits = 2)
}
```

**Narrative Point:** Latin America and Sub-Saharan Africa carry a vastly disproportionate share of the global damage compared to their landmass. For example, Latin America contains roughly 15.6% of global land, but hosts 22.2% of all global Sediment Retention hotspots (a relative intensity of 1.42).

```{r}
#| label: plot-regional-intensity
#| fig-cap: "Relative intensity of hotspots by World Bank Region. A value greater than 1 indicates a region has a disproportionately high share of global hotspots compared to its land area."
#| fig-width: 10
#| fig-height: 8
#| echo: false

if (exists("area_stats")) {
  svc_order <- c("C_Risk", "N_export", "Sed_export", "C_Risk_Red_Ratio", "N_Ret_Ratio", "Sed_Ret_Ratio", "Pollination", "Nature_Access")
  
  plot_data <- area_stats %>%
    filter(grouping_var == "region_wb") %>%
    mutate(service = factor(service, levels = svc_order))

  ggplot(plot_data, aes(x = relative_intensity, y = reorder(group, relative_intensity), fill = relative_intensity > 1)) +
    geom_col() +
    geom_vline(xintercept = 1, linetype = "dashed", color = "gray30") +
    scale_fill_manual(values = c("TRUE" = "darkred", "FALSE" = "gray"), name = "Disproportionate Burden") +
    facet_wrap(~service, ncol = 2, scales = "free_x") +
    labs(
      title = "Disproportionate Burden: Regional Concentration of Hotspots",
      subtitle = "Relative intensity shows which regions have a higher share of hotspots than their share of global land area.",
      x = "Relative Intensity (Share of Hotspots / Share of Area)",
      y = "World Bank Region"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
}
```

## Part 2: Who is affected? (Socioeconomic Exposure)

This section uses `hotspot_pop_exposure.csv` to quantify the human dimension of ES degradation.

### Key Questions:
*   How many people are living in hotspot areas for each service?
*   What is the **absolute** exposure across different income groups? (Total number of people affected).
*   What is the **relative** socioeconomic profile of hotspot areas? (Are hotspots statistically richer/poorer, more/less populated than the background landscape?).

### Analysis 1: Relative Socioeconomic Profile (KS Tests)

```{r}
#| label: analyze-socioeconomic-profile
#| echo: false

if (exists("ks_res")) {
  ks_res %>%
    filter(grepl("gdp|POP", var)) %>%
    mutate(
      Variable = if_else(grepl("gdp", var), "GDP", "Population")
    ) %>%
    select(Service = service, Variable, `Cliff's Delta` = cliffs_delta, p_adj) %>%
    pivot_wider(names_from = Variable, values_from = `Cliff's Delta`) %>%
    kable(caption = "Socioeconomic Profile of Hotspots (Cliff's Delta)", digits = 3)
}
```

```{r}
#| label: plot-ks-directionality
#| fig-cap: "Directionality of socioeconomic effects in hotspots. Positive values (red) indicate hotspots have higher values of the variable (e.g., higher GDP, more population) than the background. Negative values (blue) indicate the opposite."
#| echo: false

ks_plot_path <- here("outputs", "plots", "ks", "directionality_cliffs_delta.png")
if(file.exists(ks_plot_path)) {
  knitr::include_graphics(ks_plot_path)
} else {
  "Directionality plot not found. Run KS_tests_hotspots.qmd to generate it."
}
```

**Narrative Point:** The socioeconomic profile of hotspots reveals a critical distinction. For services like **Nature Access** and **Coastal Risk Reduction**, hotspots are consistently found in areas with significantly higher GDP and Population than the background landscape. This points to degradation driven by urbanization and development. Conversely, hotspots for **Pollination** loss and increased **Sediment Export** are found in areas with lower GDP and population, highlighting a different set of rural and agricultural pressures.

### Analysis 2: Absolute Population Exposure

While the KS tests show the *relative* profile of hotspot areas, this analysis shows the *absolute* number of people living within them. This helps us understand the total scale of human impact.

```{r}
#| label: plot-pop-exposure-income
#| fig-cap: "Total population living within ecosystem service hotspots, broken down by World Bank income group."
#| fig-width: 10
#| fig-height: 8
#| echo: false

if (exists("pop_exposure")) {
  # Defensively find the income group column, same as in the table chunk.
  income_col_name <- names(pop_exposure)[grepl("income_grp", names(pop_exposure), ignore.case = TRUE)]
  
  svc_order <- c("C_Risk", "N_export", "Sed_export", "C_Risk_Red_Ratio", "N_Ret_Ratio", "Sed_Ret_Ratio", "Pollination", "Nature_Access")
  income_order <- c("High income", "Upper middle income", "Lower middle income", "Low income")
  
  if (!is.na(income_col_name)) {
    plot_data <- pop_exposure %>%
      # Use the dynamically found column name and clean it
      rename(income_group = all_of(income_col_name)) %>%
      group_by(service, income_group) %>%
      summarise(total_exposed_pop = sum(exposed_population, na.rm = TRUE), .groups = "drop") %>%
      mutate(service = factor(service, levels = svc_order),
             income_group = factor(income_group, levels = income_order))

    ggplot(plot_data, aes(x = income_group, y = total_exposed_pop, fill = income_group)) +
      geom_col() + # The legend is now shown by default
      scale_y_continuous(labels = scales::label_number(suffix = " M", scale = 1e-6)) +
      # Add a clear name for the legend title
      scale_fill_brewer(palette = "RdYlBu", direction = -1, name = "Income Group") +
      facet_wrap(~service, ncol = 2, scales = "free_y") +
      labs(
        title = "Population Exposure to Hotspots by Income Group",
        subtitle = "Total population living in areas with critical ecosystem service decline.",
        x = "World Bank Income Group",
        y = "Total Exposed Population"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1), 
        strip.text = element_text(face = "bold"),
        legend.position = "bottom" # Move legend to bottom for clarity
      )
  }
}
```

## Part 3: What are the drivers? (LCC Attribution)

This section uses the `lcc_es_hotspot_overlap_pct.csv` table to connect ES hotspots to specific land cover change dynamics (e.g., Forest Loss, Cropland Expansion). The plot below shows the percentage of each service's hotspot area that can be attributed to a major land conversion driver occurring in the same 10km grid cell.

```{r}
#| label: plot-lcc-attribution
#| fig-cap: "Percentage of ecosystem service hotspots attributed to different land cover change drivers."
#| fig-width: 12
#| fig-height: 9
#| echo: false

if (exists("lcc_stats")) {
  svc_order <- c("C_Risk", "N_export", "Sed_export", "C_Risk_Red_Ratio", "N_Ret_Ratio", "Sed_Ret_Ratio", "Pollination", "Nature_Access")
  
  plot_data <- lcc_stats %>%
    rename_with(~"overlap_pct", any_of(c("pct_overlap", "percent", "value", "pct_hotspot_in_lcc"))) %>%
    rename_with(~"Attribution", any_of(c("attribution", "Driver", "driver", "Category"))) %>%
    mutate(service = factor(service, levels = svc_order)) %>%
    filter(Attribution != "Change Occuring Off-Pixel")

  ggplot(plot_data, aes(x = overlap_pct, y = reorder(Attribution, overlap_pct), fill = Attribution)) +
    geom_col(show.legend = FALSE) +
    scale_x_continuous(labels = scales::percent_format(scale = 1)) +
    facet_wrap(~service, ncol = 2) +
    labs(
      title = "Drivers of Decline: Attributing ES Hotspots to Land Cover Change",
      subtitle = "Percentage of hotspot area for each service that overlaps with a major land conversion driver.",
      x = "% of Hotspot Area Overlapping with Driver",
      y = "Land Cover Change Driver"
    ) +
    theme_minimal() +
    theme(strip.text = element_text(face = "bold"))
}
```

## Part 4: Synthesis & Key Takeaways

This section is for drafting the high-level summary points for the manuscript's discussion and conclusion.

*   **Takeaway 1: Geographic Mismatch.** *e.g., "Our results show a clear geographic mismatch between the areas providing the bulk of a service and the areas experiencing the most intense relative degradation..."*
*   **Takeaway 2: Equity and Vulnerability.** *e.g., "The burden of ES degradation is not borne equally. We find that low-income countries and populations with low HDI are disproportionately exposed to declines in critical services like..."*
*   **Takeaway 3: Drivers of Change.** *e.g., "While land conversion is a major driver for services like [Service A], a significant portion of hotspots for [Service B] occur in stable landscapes, pointing towards degradation-in-place..."*